### 2021-07-22 notes/notes_FB/MaxEnt_une_ligne_v3.tex
Add files via upload

% !TEX encoding = UTF-8 Unicode
\documentclass{svmult}

% choose options for [] as required from the list
% in the Reference Guide, Sect. 2.2

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{latexsym}
%\usepackage{psfig}
\usepackage{graphicx}
\usepackage{multicol}
%\usepackage[frenchb]{babel}

\usepackage[utf8]{inputenc}
 \usepackage{url}
 \usepackage{bm}

\newcommand{\ii}{{[i\cup j]}}

\usepackage{natbib}
%\bibliographystyle{ksfh_nat}
\bibliographystyle{apalike}

% etc.
% see the list of further useful packages
% in the Reference Guide, Sects. 2.3, 3.1-3.3

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[toctitles]{titlesec}% http://ctan.org/pkg/titlesec
\titleformat{\section}%
  [hang]% <shape>
  {\normalfont\bfseries\Large}% <format>
  {}% <label>
  {0pt}% <sep>
  {}% <before code>
\renewcommand{\thesection}{}% Remove section references...
\renewcommand{\thesubsection}{}% Remove subsection references...
%\renewcommand{\thesubsection}{\arabic{subsection}}%... from subsections

\titlerunning{Maximum d'entropie, une ligne de bus (juillet 2021)}

\begin{document}

\title*{Maximum d'entropie, une seule ligne de bus}

% Use \titlerunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\author{FB-RL-GG}
% Use \authorrunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\institute{UNIL}
%
% Use the package "url.sty" to avoid
% problems with special characters
% used in your e-mail or web address
%

\maketitle
%%%%%%%%%
%\usepackage{fancyhdr}
%\pagestyle{fancy}
%\pagestyle{fancyplain}
%\rfoot{}
% \cfoot{\fancyplain{}{\small JADT 2012 : $\mbox{11}^{\mbox{\tiny es}}$ Journ\'ees internationales d'Analyse statistique des  Donn\'ees Textuelles}} 
%\renewcommand{\headrulewidth}{0pt}
%\renewcommand{\footrulewidth}{0.5pt}
%%%%%%%%%

%\textwidth = 6.5 in
%\textheight = 8 in
%\oddsidemargin = 0.0 in
%\evensidemargin = 0.0 in
%\topmargin = 0.2 in
%\headheight = 0.1 in
%\parindent = 0.2 in

\section*{Notations et formalisation du problème}
Une ligne orientée allant de l'arrêt  $i=1$ jusqu'à $i=l$. Soit $x_i$ le nombre de passagers montant à l'arrêt $i$, et $y_i$ le nombre de passagers descendant à l'arrêt $i$. On a 
$y_1=0$ et $x_l=0$.

\

Soit $N_{ij}$ (avec $i<j$; sinon $N_{ij}=0$) le nombre de personnes montant en $i$ et descendant en $j$.

\

Soit $n_{i,i+1}$ le nombre de personnes transportées dans le tronçon $i$, $i+1$. Par construction: 
\begin{equation}
\label{ }
n_{i,i+1}=n_{i-1,i}+x_i-y_i\qquad\qquad n_{01}=0
\end{equation}

\

On veut estimer les données $N_{ij}$. Par construction, 
\begin{equation}
\label{constraints_margins}
N_{i\bullet}=x_i\qquad\qquad N_{\bullet j}=y_j\qquad\qquad N_{\bullet\bullet}=x_\bullet \stackrel{!}{=}y_\bullet
\end{equation}
Soit $f_{ij}^D:=\frac{N_{ij}}{N_{\bullet\bullet}}$ la distribution empirique à estimer. Soit $g_i:=\frac{x_i}{x_\bullet}$ et  $h_j:=\frac{y_j}{y_\bullet}$ les distributions marginales correspondantes. En fait $f_{ij}^D$ est une matrice $(l-1)\times (l-1)$, où $i=1,\ldots, l-1$ et $j=2,\ldots, l$

\

Pour la distribution théorique $f^M$, donnée par une matrice $(l-1)\times (l-1)$, on peut imaginer un prior
\begin{equation}
\label{nrem}
f_{ij}^M=a_ib_j 1(i<j)\qquad\qquad \sum_{i=1}^{l-1} a_i B_i\stackrel{!}{=}1\qquad\qquad B_i:=\sum_{j=i+1}^l b_j
\end{equation}
dépendant des $2l-3$ paramètres libres $a_1,\ldots, a_{l-1}$ et  $b_2,\ldots, b_{l}$ (contraints par la normalisation).

\

Les contraintes se réécrivent 
\begin{equation}
\label{ }
f_{i\bullet}^D=g_i\quad i=1,\ldots, l-1
\qquad\qquad \qquad\qquad 
f_{\bullet j}^D=h_j\quad j=2,\ldots, l
\end{equation}
Il y en a aussi $2l-3$ (car $\sum_if_{i\bullet}^D=\sum_j f_{\bullet j}^D$). Cela permet d'espérer de déterminer $a$ et $b$ de telle sorte que $f^D=f^M\equiv f$, qui donnerait un minimum absolu de $K(f^D||f^M)$.

\

Les premiers termes non nuls sont $f_{12}=a_1b_2$, $f_{13}=a_1b_3$.... $f_{1,l}=a_1b_l$, dont la somme $f_{1\bullet}=a_1B_1$ doit être $g_1$.

Puis $f_{23}=a_2b_3$, $f_{24}=a_2b_4$.... $f_{2,l}=a_2b_l$, dont la somme $f_{2\bullet}=a_2B_2$ doit être $g_2$.

\

En général, on a $f_{i\bullet}=a_iB_i\stackrel{!}{=}g_i$ pour  $i=1,\ldots, l-1$. De même, la normalisation (\ref{nrem}) peut aussi s'écrire 
\begin{displaymath}
 \sum_{j=2}^{l} A_j b_j\stackrel{!}{=}1\qquad\qquad A_j=\sum_{a=1}^{j-1} a_i
\end{displaymath}
d'où l'on tire que $f_{\bullet j}=A_jb_j\stackrel{!}{=}h_j$ pour  $j=2,\ldots, l$.

\

Pas sûr que $f^D=f^M$ puisse être réalisé, peut-être faut il changer de prior $f^M$ : piste à ne pas abandonner. Mais...

\section{Approche: ``estimer une table de contingence $N$ dont les marges sont fixées"}
... le problème d'estimer une table de contingence $N$ dont les marges sont fixées  (équation (\ref{constraints_margins})) a fait l'objet d'une énorme littérature... A étudier et poursuivre.

\section{Modèle de Guillaume}
(avec quelques notations utilisées ici). 
\begin{enumerate}
  \item[$\bullet$] Probabilité de monter en $i$:  $p_i^{\mbox{\tiny in}}=x_i/x_{\bullet}$. 
    \item[$\bullet$] Probabilité de descendre en $i$: $p_i^{\mbox{\tiny out}}=y_i/n_{i-1,i}$. 
      \item[$\bullet$] Probabilité de continuer de $i$ à $i+1$ : $c_i=1-p_i^{\mbox{\tiny out}}$. 
          \item[$\bullet$] Probabilité $P_{ij}$ de trajet de $i$ à $j>i$: 
\begin{equation}
\label{ }
P_{ij}=p_i^{\mbox{\tiny in}}c_{i+1}\ldots c_{j-1}p_j^{\mbox{\tiny out}}\quad \mbox{ pour } j\ge i+2
\qquad\qquad P_{i,i+1}=p_i^{\mbox{\tiny in}}p_{i+1}^{\mbox{\tiny out}}
\end{equation}
Le produit commence par $c_{i+1}$, car, si l'on est monté en $i$, la probabilité d'effectuer le tronçon $i\to i+1$ vaut 1.
\end{enumerate}

Soit $X_i:=\sum_{k=1}^i x_k$ le nombre cumulé de montées, et $Y_i:=\sum_{k=1}^i y_k$ le nombre cumulé de descentes. On a 
\begin{equation}
\label{ }
X_i\ge Y_i\quad i=1,\ldots,l\qquad\qquad X_l=Y_l\qquad\qquad n_{i,i+1}=X_i-Y_i
\end{equation}
Il est pratique de définir le ``transit d'avant"  $t_i:=X_{i-1}-Y_{i-1}=n_{i-1,i}$, en posant $t_1=1$ (au lieu de $t_0=0$) afin que  
$p_1^{\mbox{\tiny out}}=y_1/t_1=0/1=0$. On a alors $p_i^{\mbox{\tiny out}}=y_i/t_i$ pour tout $i=1,\ldots, l$, avec 
\begin{displaymath}
p_l^{\mbox{\tiny out}}=\frac{y_l}{t_l}=\frac{y_l}{X_{l-1}-Y_{l-1}}=\frac{y_l}{y_l}=1
\end{displaymath}
comme il se doit, où on a utilisé $X_{l-1}-Y_{l-1}=X_{l-1}+0-Y_{l-1}-y_l+y_l=X_l-Y_l+y_l=y_l$.

\

On observe que $P_{\bullet\bullet}=1$. On va redéfinir comme avant $f_{ij}:=P_{ij}$. 
Le nombre attendu de trajets $N_{ij}$ est alors $N_{ij}=x_{\bullet} f_{ij}$. On observe que $N_{i\bullet}=x_i$ et $N_{j\bullet}=y_j$.

\

On observe aussi que les trajets attendus sont (pour les cas étudiés) de la forme (cf. (\ref{nrem}))
\begin{equation}
\label{nrem2}
N_{ij}=N_{\bullet\bullet}a_ib_jI(j>i)
\end{equation}
et qu'ainsi la forme des histogrammes du nombre de sorties $j$ depuis un départ $i$ variable reste la même: 
\begin{equation}
\label{nrem3}
N_{j|i}:=\frac{N_{ij}}{N_{i\bullet}}=\frac{N_{\bullet\bullet}a_ib_jI(j>i)}{N_{\bullet\bullet}a_i\sum_{k>i}b_k}=\frac{b_jI(j>i)}{B_i}
\end{equation}
avec $B_i:=\sum_{k>i}b_k$.

 Par construction, $a_l$ et $b_1$ sont indéfinis dans  (\ref{nrem2}); on peut les poser égaux à zéro. On peut noter que 
 $N_{ij}=0$ si $x_i=0$ ou si $y_j=0$.  On peut alors poser 
 \begin{displaymath}
a_i=:x_i\alpha_i\qquad\qquad\mbox{et}\qquad\qquad b_j=:y_j\beta_j
\end{displaymath}
et déterminer $\alpha$ et $\beta$ par itération. Les conditions $N_{i\bullet}=x_i$ et $N_{j\bullet}=y_j$ donnent 
\begin{equation}
\label{iterFit}
\alpha_i=\frac{1}{\sum_{j>i}\beta_j y_j}\qquad i<l \qquad\qquad \beta_j =\frac{1}{\sum_{i<j}\alpha_i x_i}\qquad j>1
\end{equation}
qu'on peut itérer (iterative fitting) à partir (par exemple) des conditions initiales $\beta^{(0)}=(0,\frac{1}{l-1},\frac{1}{l-1},\cdots,\frac{1}{l-1})$, itérées par exemple $500$ fois.

\subsection{Simulations numériques}
 Voir test\_markov\_Guillaume\_Francois.R : tout semble jouer avec l'exemple 1, pour lequel le transit $X_i-Y_i$ n'est jamais nul (sauf en $i=l$). Mais difficultés avec l'exemple 2 ($l=10$), pour lequel le bus est vide entre les arrêts 8 et 9 $(X_8-Y_8=0$), et donc  $p_9^{\mbox{\tiny out}}$ {\em n'est pas défini}.

 Clairement, l'absence de voyageurs entre les arrêts 8 et 9 ``simplifie" le problème, qui doit être résolu comme deux problèmes ``disjoints": de la station 1 à la station 8 d'une part, et de la station 9 à la station 10 d'autre part: il faut commencer par déterminer les tronçons vides, puis résoudre les sous-problèmes délimités par les tronçons vides.

 \section{Estimation des param. $a$, $b$ dans $n_{ij}=a_ib_j I(i<j)$}
On a $n_{i\bullet}=x_i$ et $n_{i\bullet }=y_j$. Soit 
\begin{equation}
\label{ }
s:=\min_i \{i | x_i>0\}
\qquad\qquad
t:=\max_j \{j | y_j>0\}
\end{equation}
Par construction, $1\le s<t\le l$. On suppose que $n_{ij}$ est irréductible (pas de tronçon à vide); sinon il faut considérer chaque tronçon plein séparément.  
De fait, l'expression $n_{ij}=a_ib_j I(i<j)$ présuppose explicitement l'irréductibilité.

\

En particulier, 
$a_i=0$ pour $i<s$ et $i=l$; $b_j=0$ pour $j>t$ et $j=1$; 
$a_s>0$ et $b_t>$. Alors, pour $ s\le i <j\le t$, 
\begin{equation}
\label{ }
n_{sj}=a_sb_j\qquad\qquad n_{it}=a_ib_t\qquad\qquad n_{st}=a_sb_t
\end{equation}
et donc
\begin{equation}
\label{ }
\mbox{\fbox{$a_ib_j=\frac{n_{it}n_{sj}}{n_{st}}$}}
\end{equation}

\end{document}

### 2022-03-28 old_code/tableau_vitesse.R
Test RStudio

@@ -1,0 +1,120 @@
+############ Création d'un tableau comportant le temps de déplacement à pied d'un arrêt i à un arrêt j
+# Possible aussi de d'avoir d'autres moyens de transport
+
+
+# Database connexion
+library("RPostgreSQL")
+# Connexion à la base de données
+conn <- dbConnect("PostgreSQL", dbname = 'tl_small', host = 'localhost', user = 'rloup', password = '')
+
+library("sf")
+# Réseau à cartographier
+reseau <- st_read(conn, query="
+select distinct code_arret_theo, libelle_arret_theo, lat_arret, lon_arret
+from freq_theo;")
+
+# To avoid to connect to the database
+# write.csv(reseau,"reseau.csv", row.names = FALSE)
+# setwd("/Users/rloup/Desktop")
+# reseau <- read.csv("reseau.csv")
+
+# OTP connexion
+# Set the working directory
+setwd("/Users/rloup/Documents/otpServ") # choose the location of your server
+# install.packages("curl")
+# install.packages("remotes")
+# remotes::install_github("ITSleeds/opentripplanner")
+library("opentripplanner")
+# https://github.com/ropensci/opentripplanner/blob/master/R/otp-plan.R
+
+# Set the server
+otp = paste0(getwd(), "/otp-1.4.0-shaded.jar")
+graphdir <- paste0(getwd(), "/graphs/default")
+# Allocate the RAM (in Mo) and set the name of the router, possible to allocate less memory but it will be very slow
+# Be sure to have Java insalled
+log <- otp_setup(otp, getwd(), memory = 20000, router = "default") # possible to change the port adding 'port = 8081' before the last bracket
+
+# Connexion to OTP
+otpcon <- otp_connect(hostname =  "localhost",
+                      router = "default",
+                      port = 8080)
+
+# Function that give the walking time [s] from a point i to a point j
+ok <- function(){
+  origin <- reseau$code_arret_theo[i]
+  destination <- reseau$code_arret_theo[j]
+  time <- otp_plan(otpcon, 
+                   fromPlace = c(reseau$lon_arret[i], reseau$lat_arret[i]), 
+                   toPlace = c(reseau$lon_arret[j], reseau$lat_arret[j]),
+                   mode = "WALK"
+  )$walkTime
+  
[…]

### 2022-03-28 compute_distrib.R
Old code moved

@@ -1,47 +1,0 @@
-setwd("~/Documents/recherche/projet_tl")
-data = read.csv("data/Ptr_marges.csv")
-
-nom_arrets = names(data)
-nij = as.matrix(data)
-n_arrets = dim(nij)[1]
-l = n_arrets - 1
-
-n_trajets = c()
-red_nij = nij
-for(i in 1:l){
-  red_nij = red_nij[-dim(red_nij)[2],-1]
-  if(!is.null(dim(red_nij))) n_trajets = c(n_trajets, sum(diag(red_nij)))
-  else n_trajets = c(n_trajets, red_nij)
-}
-
-names(n_trajets) = 1:l
-barplot(n_trajets)
-
-for(id_arret in 1:(length(nom_arrets)-1)){
-  rep_ligne = nij[id_arret,]
-  names(rep_ligne) = nom_arrets
-  rep_ligne[-(1:id_arret)]
-  barplot(rep_ligne, main=nom_arrets[id_arret][])
-}
-
-n_montees = rowSums(nij)
-n_descentes = colSums(nij)
-
-# Calcul du transit (doit correspondre dans les données) 
-transit_ij = n_montees[1] 
-for (i in 2:(n_arrets-1)){
-  transit_ij = c(transit_ij, transit_ij[i-1] + n_montees[i] - n_descentes[i])
-}
-
-p_in = n_montees / sum(n_montees) # Probabilité de montées 
-p_out = n_descentes / c(1, transit_ij) # Probabilité de descente
-p_ij = 1 - p_out # Probabilité de continuer
-
-Pij = nij / sum(nij)
-j = 3
-n_descentes[j] / sum(n_descentes[-(1:(j-1))])
-Pij[j-1, j] / sum(Pij[j-1, ])
-
-
-
-

### 2022-03-28 graphs_freq_tl_lignes.R

@@ -1,52 +1,0 @@
-############ Graphique de fréquentation annuelle des lignes TL
-
-setwd("/Users/rloup/Desktop")
-
-### Connexion à la base de données
-library('RPostgreSQL')
-library('config')
-
-# Connexion à la base de données
-conn <- dbConnect("PostgreSQL", dbname = 'tl_small', host = 'localhost', user = 'rloup', password = '')
-
-# Fréquentation annuelle : somme des montées/descentes par an sur les ligne, direction et arrêt
-stops_frequentation <- st_read(conn, query="select stop_order.code_arret_theo, freq_theo.libelle_arret_theo2, stop_order.code_ligne_theo, freq_theo.direction_voy_theo, stop_order.sequence_theo, sum(freq_theo.monte) AS montees, sum(freq_theo.descendu) AS descentes
-from freq_theo
-inner join stop_order
-on freq_theo.code_ligne_theo = stop_order.code_ligne_theo
-and freq_theo.code_arret_theo = stop_order.code_arret_theo
-and freq_theo.direction_voy_theo = stop_order.direction_voy_theo
-group by stop_order.code_arret_theo, freq_theo.libelle_arret_theo2, stop_order.code_ligne_theo, freq_theo.direction_voy_theo, stop_order.sequence_theo
-order by stop_order.code_ligne_theo, freq_theo.direction_voy_theo, stop_order.sequence_theo;")
-
-
-### Initialisation of the table
-table <- data.frame(matrix(ncol = 7, nrow = 0))
-namesTable <- c("code_arret", "libelle_arret", "code_ligne", "sequence", "direction", "montees", "descentes")
-colnames(table) <- namesTable
-
-# Itérations sur i, row 1 et j row 2
-i <- 1 # row number
-j <- 1
-
-while (i <= length(stops_frequentation$code_ligne_theo)) {
-  #row1 <- i
-  while (stops_frequentation$sequence_theo[j] < stops_frequentation$sequence_theo[j+1]) {
-    if (stops_frequentation$sequence_theo[j]<stops_frequentation$sequence_theo[j+1] &&
-        stops_frequentation$direction_voy_theo[j]<=stops_frequentation$direction_voy_theo[j+1] &&
-        stops_frequentation$code_ligne_theo[j]<=stops_frequentation$code_ligne_theo[j+1]
-    ) {
-      j<-j+1
-    }
-    else{}
-  }
[…]

### 2022-03-28 markov_tl_lignes.R

@@ -1,107 +1,0 @@
-setwd("/Users/rloup/Desktop/markov")
-
-############# Connexion à la base de données
-
-library('RPostgreSQL')
-library('config')
-
-# Connexion à la base de données
-conn <- dbConnect("PostgreSQL", dbname = 'tl_small', host = 'localhost', user = 'rloup', password = '')
-
-# Fréquentation annuelle : somme des montées/descentes par an sur les ligne, direction et arrêt
-stops_frequentation <- st_read(conn, query="select stop_order.code_arret_theo, freq_theo.libelle_arret_theo2, stop_order.code_ligne_theo, freq_theo.direction_voy_theo, stop_order.sequence_theo, sum(freq_theo.monte) AS montees, sum(freq_theo.descendu) AS descentes
-from freq_theo
-
-inner join stop_order
-on freq_theo.code_ligne_theo = stop_order.code_ligne_theo
-and freq_theo.code_arret_theo = stop_order.code_arret_theo
-and freq_theo.direction_voy_theo = stop_order.direction_voy_theo
-
-group by stop_order.code_arret_theo, freq_theo.libelle_arret_theo2, stop_order.code_ligne_theo, freq_theo.direction_voy_theo, stop_order.sequence_theo
-order by stop_order.code_ligne_theo, freq_theo.direction_voy_theo, stop_order.sequence_theo;")
-
-
-############# Petit script de calcul des trajets via une chaîne de Markov
-
-# Fonction chaîne de Markov, à itérer sur chaque ligne TL
-markov_Ptr <- function(n_montees, n_descentes, l){
-  # Calcul du transit (doit correspondre dans les données) 
-  transit_ij = n_montees[1] 
-  for (i in 2:(l-1)){
-    transit_ij = c(transit_ij, transit_ij[i-1] + n_montees[i] - n_descentes[i])
-  }
-  
-  p_in = n_montees / sum(n_montees) # Probabilité de montées 
-  p_out = n_descentes / c(1, transit_ij) # Probabilité de descente
-  p_ij = 1 - p_out # Probabilité de continuer
-  
-  ########## Calcul avec chaîne de Markov 
-  
-  # On construit la matrice W
-  W_tr = rbind(cbind(rep(0,l-1), diag(p_ij[-l])), rep(0, l))
[…]

### 2022-03-28 tableau_vitesse_v2.R

@@ -1,117 +1,0 @@
-############ Création d'un tableau comportant le temps de déplacement à pied d'un arrêt i à un arrêt j
-# Possible aussi de d'avoir d'autres moyens de transport
-
-
-# Database connexion
-library("RPostgreSQL")
-# Connexion à la base de données
-conn <- dbConnect("PostgreSQL", dbname = 'tl_small', host = 'localhost', user = 'rloup', password = '')
-
-library("sf")
-# Réseau à cartographier
-reseau <- st_read(conn, query="
-select distinct code_arret_theo, libelle_arret_theo, lat_arret, lon_arret
-from freq_theo
-order by code_arret_theo;")
-
-# To avoid to connect to the database
-# write.csv(reseau,"reseau.csv", row.names = FALSE)
-setwd("/Users/rloup/Desktop")
-# reseau <- read.csv("reseau.csv")
-
-# OTP connexion
-# Set the working directory
-setwd("/Users/rloup/Documents/otpServ") # choose the location of your server
-# install.packages("curl")
-# install.packages("remotes")
-# remotes::install_github("ITSleeds/opentripplanner")
-library("opentripplanner")
-# https://github.com/ropensci/opentripplanner/blob/master/R/otp-plan.R
-
-# Set the server
-otp = paste0(getwd(), "/otp-1.4.0-shaded.jar")
-graphdir <- paste0(getwd(), "/graphs/default")
-# Allocate the RAM (in Mo) and set the name of the router, possible to allocate less memory but it will be very slow
-# Be sure to have Java insalled
-log <- otp_setup(otp, getwd(), memory = 20000, router = "default") # possible to change the port adding 'port = 8081' before the last bracket
-
-# Connexion to OTP
-otpcon <- otp_connect(hostname =  "localhost",
-                      router = "default",
-                      port = 8080)
-
-# Function that give the walking time [s] from a point i to a point j
-ok <- function(i, j){
-  origin <- reseau$code_arret_theo[i]
-  destination <- reseau$code_arret_theo[j]
-  time <- otp_plan(otpcon, 
-                   fromPlace = c(reseau$lon_arret[i], reseau$lat_arret[i]), 
-                   toPlace = c(reseau$lon_arret[j], reseau$lat_arret[j]),
-                   mode = "WALK",
[…]

### 2022-03-28 test_markov_Guillaume_Francois_v3.R

@@ -1,81 +1,0 @@
-# Petit script de calcul des trajets via une chaîne de Markov (+/-)
-# Orginal: Guillaume (+ qqes notations de François)
-
-
-# génération de montées x et descentes y d'une ligne de longueur (= nombre arrêts) l
-
-
-#  Exemple 1 : l=8 
-x = c(30, 15, 50, 15, 0, 20, 5, 0) # Nombre de gens qui montent
-y = c(0, 10, 30, 40, 25, 0, 10, 20) # Nombre de gens qui descendent
-
-#  Exemple 2 : l=10 
-#   x = c(30, 15, 50, 15, 0, 20, 5, 30,20,0) # Nombre de gens qui montent
-#   y = c(0, 10, 30, 40, 25, 0, 10, 50,0,20) # Nombre de gens qui descendent
-# ici, le bus est vide entre les arrêts 8 et 9, et la chaine est réductible
-
-
-l = length(x) # Longueur du trajet
-
-X=cumsum(x) # Nombre cumulé de gens qui montent
-Y=cumsum(y) # Nombre cumulé de gens qui descendent. Il faut que $X_j\ge Y_j$ pour tout $j$, et que $X_l=Y_l$
-
-transit=X-Y  #  ici un vecteur de longueur l, équivalent à transit_ij qui était de longueur l-1
-t=c(1,transit)[-(l+1)] # "transit d'avant". La valeur t[1]=1 permet de définir p_out[1]=0/1=0
-
-p_in = x / sum(x) # Probabilité de montée 
-p_out = y / t # Probabilité de descente
-c = 1 - p_out # Probabilité de continuer
-
-
-Ptr = matrix(0, l, l)
-for(i in 1:(l-1)){
-  for(j in (i+1):l){
-  	c_trunc_right=c[-(j:l)]
-  	c_trunc=c(1,c_trunc_right)[-(1:(i+1))]
-  	Ptr[i, j]=p_in[i]*prod(c_trunc)*p_out[j]   
-  }
-}
-
-# Note: le code marche parce que, pour j=i+1, on a prod(numeric(0))=1, ce qui est assez chanceux... 
-
-
-# Affichage en probs 
-Ptr
-# Affichage en effectifs (les marges donnent x et y)
-N=Ptr*sum(x); N 
-
-# les marges donnent x et y:
[…]

### 2022-03-28 test_markov_new.R

@@ -1,28 +1,0 @@
-############# Petit script de calcul des trajets via une chaîne de Markov
-
-n_montees = c(10, 20, 50, 20, 30, 20, 10, 0) # Nombre de gens qui montent
-n_descentes = c(0, 5, 10, 25, 50, 20, 20, 30) # Nombre de gens qui descendent
-l = length(n_montees) # Longueur du trajet1
-
-# Calcul du transit (doit correspondre dans les données) 
-transit_ij = n_montees[1] 
-for (i in 2:(l-1)){
-  transit_ij = c(transit_ij, transit_ij[i-1] + n_montees[i] - n_descentes[i])
-}
-
-p_in = n_montees / sum(n_montees) # Probabilité de montées 
-p_out = n_descentes / c(1, transit_ij) # Probabilité de descente
-p_ij = 1 - p_out # Probabilité de continuer
-
-########## Calcul avec chaîne de Markov 
-
-# On construit la matrice W
-W_tr = rbind(cbind(rep(0,l-1), diag(p_ij[-l])), rep(0, l))
-W = rbind(cbind(rep(0, l), rbind(p_in, cbind(W_tr[-1,-1], p_out[-1]))), rep(0, l+1))
-
-# On trouve la matrice fondamentale
-Z = solve(diag(dim(W)[1]) - W)
-
-# On construit la matrice de prob
-Ptr = outer(p_in[-l], rep(1,l-1)) * Z[-c(1, l+1), -c(1, l+1)] * outer(rep(1,l-1), p_out[-1])
-Ptr = rbind(cbind(rep(0, l-1), Ptr), rep(0, l))

### 2022-03-28 old_code/compute_distrib.R
File moved

@@ -1,0 +1,47 @@
+setwd("~/Documents/recherche/projet_tl")
+data = read.csv("data/Ptr_marges.csv")
+
+nom_arrets = names(data)
+nij = as.matrix(data)
+n_arrets = dim(nij)[1]
+l = n_arrets - 1
+
+n_trajets = c()
+red_nij = nij
+for(i in 1:l){
+  red_nij = red_nij[-dim(red_nij)[2],-1]
+  if(!is.null(dim(red_nij))) n_trajets = c(n_trajets, sum(diag(red_nij)))
+  else n_trajets = c(n_trajets, red_nij)
+}
+
+names(n_trajets) = 1:l
+barplot(n_trajets)
+
+for(id_arret in 1:(length(nom_arrets)-1)){
+  rep_ligne = nij[id_arret,]
+  names(rep_ligne) = nom_arrets
+  rep_ligne[-(1:id_arret)]
+  barplot(rep_ligne, main=nom_arrets[id_arret][])
+}
+
+n_montees = rowSums(nij)
+n_descentes = colSums(nij)
+
+# Calcul du transit (doit correspondre dans les données) 
+transit_ij = n_montees[1] 
+for (i in 2:(n_arrets-1)){
+  transit_ij = c(transit_ij, transit_ij[i-1] + n_montees[i] - n_descentes[i])
+}
+
+p_in = n_montees / sum(n_montees) # Probabilité de montées 
+p_out = n_descentes / c(1, transit_ij) # Probabilité de descente
+p_ij = 1 - p_out # Probabilité de continuer
+
+Pij = nij / sum(nij)
+j = 3
+n_descentes[j] / sum(n_descentes[-(1:(j-1))])
+Pij[j-1, j] / sum(Pij[j-1, ])
+
+
+
+

### 2022-03-28 old_code/graphs_freq_tl_lignes.R

@@ -1,0 +1,52 @@
+############ Graphique de fréquentation annuelle des lignes TL
+
+setwd("/Users/rloup/Desktop")
+
+### Connexion à la base de données
+library('RPostgreSQL')
+library('config')
+
+# Connexion à la base de données
+conn <- dbConnect("PostgreSQL", dbname = 'tl_small', host = 'localhost', user = 'rloup', password = '')
+
+# Fréquentation annuelle : somme des montées/descentes par an sur les ligne, direction et arrêt
+stops_frequentation <- st_read(conn, query="select stop_order.code_arret_theo, freq_theo.libelle_arret_theo2, stop_order.code_ligne_theo, freq_theo.direction_voy_theo, stop_order.sequence_theo, sum(freq_theo.monte) AS montees, sum(freq_theo.descendu) AS descentes
+from freq_theo
+inner join stop_order
+on freq_theo.code_ligne_theo = stop_order.code_ligne_theo
+and freq_theo.code_arret_theo = stop_order.code_arret_theo
+and freq_theo.direction_voy_theo = stop_order.direction_voy_theo
+group by stop_order.code_arret_theo, freq_theo.libelle_arret_theo2, stop_order.code_ligne_theo, freq_theo.direction_voy_theo, stop_order.sequence_theo
+order by stop_order.code_ligne_theo, freq_theo.direction_voy_theo, stop_order.sequence_theo;")
+
+
+### Initialisation of the table
+table <- data.frame(matrix(ncol = 7, nrow = 0))
+namesTable <- c("code_arret", "libelle_arret", "code_ligne", "sequence", "direction", "montees", "descentes")
+colnames(table) <- namesTable
+
+# Itérations sur i, row 1 et j row 2
+i <- 1 # row number
+j <- 1
+
+while (i <= length(stops_frequentation$code_ligne_theo)) {
+  #row1 <- i
+  while (stops_frequentation$sequence_theo[j] < stops_frequentation$sequence_theo[j+1]) {
+    if (stops_frequentation$sequence_theo[j]<stops_frequentation$sequence_theo[j+1] &&
+        stops_frequentation$direction_voy_theo[j]<=stops_frequentation$direction_voy_theo[j+1] &&
+        stops_frequentation$code_ligne_theo[j]<=stops_frequentation$code_ligne_theo[j+1]
+    ) {
+      j<-j+1
+    }
+    else{}
+  }
[…]

### 2022-03-28 old_code/markov_tl_lignes.R

@@ -1,0 +1,107 @@
+setwd("/Users/rloup/Desktop/markov")
+
+############# Connexion à la base de données
+
+library('RPostgreSQL')
+library('config')
+
+# Connexion à la base de données
+conn <- dbConnect("PostgreSQL", dbname = 'tl_small', host = 'localhost', user = 'rloup', password = '')
+
+# Fréquentation annuelle : somme des montées/descentes par an sur les ligne, direction et arrêt
+stops_frequentation <- st_read(conn, query="select stop_order.code_arret_theo, freq_theo.libelle_arret_theo2, stop_order.code_ligne_theo, freq_theo.direction_voy_theo, stop_order.sequence_theo, sum(freq_theo.monte) AS montees, sum(freq_theo.descendu) AS descentes
+from freq_theo
+
+inner join stop_order
+on freq_theo.code_ligne_theo = stop_order.code_ligne_theo
+and freq_theo.code_arret_theo = stop_order.code_arret_theo
+and freq_theo.direction_voy_theo = stop_order.direction_voy_theo
+
+group by stop_order.code_arret_theo, freq_theo.libelle_arret_theo2, stop_order.code_ligne_theo, freq_theo.direction_voy_theo, stop_order.sequence_theo
+order by stop_order.code_ligne_theo, freq_theo.direction_voy_theo, stop_order.sequence_theo;")
+
+
+############# Petit script de calcul des trajets via une chaîne de Markov
+
+# Fonction chaîne de Markov, à itérer sur chaque ligne TL
+markov_Ptr <- function(n_montees, n_descentes, l){
+  # Calcul du transit (doit correspondre dans les données) 
+  transit_ij = n_montees[1] 
+  for (i in 2:(l-1)){
+    transit_ij = c(transit_ij, transit_ij[i-1] + n_montees[i] - n_descentes[i])
+  }
+  
+  p_in = n_montees / sum(n_montees) # Probabilité de montées 
+  p_out = n_descentes / c(1, transit_ij) # Probabilité de descente
+  p_ij = 1 - p_out # Probabilité de continuer
+  
+  ########## Calcul avec chaîne de Markov 
+  
+  # On construit la matrice W
+  W_tr = rbind(cbind(rep(0,l-1), diag(p_ij[-l])), rep(0, l))
[…]

### 2022-03-28 old_code/tableau_vitesse_v2.R

@@ -1,0 +1,117 @@
+############ Création d'un tableau comportant le temps de déplacement à pied d'un arrêt i à un arrêt j
+# Possible aussi de d'avoir d'autres moyens de transport
+
+
+# Database connexion
+library("RPostgreSQL")
+# Connexion à la base de données
+conn <- dbConnect("PostgreSQL", dbname = 'tl_small', host = 'localhost', user = 'rloup', password = '')
+
+library("sf")
+# Réseau à cartographier
+reseau <- st_read(conn, query="
+select distinct code_arret_theo, libelle_arret_theo, lat_arret, lon_arret
+from freq_theo
+order by code_arret_theo;")
+
+# To avoid to connect to the database
+# write.csv(reseau,"reseau.csv", row.names = FALSE)
+setwd("/Users/rloup/Desktop")
+# reseau <- read.csv("reseau.csv")
+
+# OTP connexion
+# Set the working directory
+setwd("/Users/rloup/Documents/otpServ") # choose the location of your server
+# install.packages("curl")
+# install.packages("remotes")
+# remotes::install_github("ITSleeds/opentripplanner")
+library("opentripplanner")
+# https://github.com/ropensci/opentripplanner/blob/master/R/otp-plan.R
+
+# Set the server
+otp = paste0(getwd(), "/otp-1.4.0-shaded.jar")
+graphdir <- paste0(getwd(), "/graphs/default")
+# Allocate the RAM (in Mo) and set the name of the router, possible to allocate less memory but it will be very slow
+# Be sure to have Java insalled
+log <- otp_setup(otp, getwd(), memory = 20000, router = "default") # possible to change the port adding 'port = 8081' before the last bracket
+
+# Connexion to OTP
+otpcon <- otp_connect(hostname =  "localhost",
+                      router = "default",
+                      port = 8080)
+
+# Function that give the walking time [s] from a point i to a point j
+ok <- function(i, j){
+  origin <- reseau$code_arret_theo[i]
+  destination <- reseau$code_arret_theo[j]
+  time <- otp_plan(otpcon, 
+                   fromPlace = c(reseau$lon_arret[i], reseau$lat_arret[i]), 
+                   toPlace = c(reseau$lon_arret[j], reseau$lat_arret[j]),
+                   mode = "WALK",
[…]

### 2022-03-28 old_code/test_markov_new.R

@@ -1,0 +1,28 @@
+############# Petit script de calcul des trajets via une chaîne de Markov
+
+n_montees = c(10, 20, 50, 20, 30, 20, 10, 0) # Nombre de gens qui montent
+n_descentes = c(0, 5, 10, 25, 50, 20, 20, 30) # Nombre de gens qui descendent
+l = length(n_montees) # Longueur du trajet1
+
+# Calcul du transit (doit correspondre dans les données) 
+transit_ij = n_montees[1] 
+for (i in 2:(l-1)){
+  transit_ij = c(transit_ij, transit_ij[i-1] + n_montees[i] - n_descentes[i])
+}
+
+p_in = n_montees / sum(n_montees) # Probabilité de montées 
+p_out = n_descentes / c(1, transit_ij) # Probabilité de descente
+p_ij = 1 - p_out # Probabilité de continuer
+
+########## Calcul avec chaîne de Markov 
+
+# On construit la matrice W
+W_tr = rbind(cbind(rep(0,l-1), diag(p_ij[-l])), rep(0, l))
+W = rbind(cbind(rep(0, l), rbind(p_in, cbind(W_tr[-1,-1], p_out[-1]))), rep(0, l+1))
+
+# On trouve la matrice fondamentale
+Z = solve(diag(dim(W)[1]) - W)
+
+# On construit la matrice de prob
+Ptr = outer(p_in[-l], rep(1,l-1)) * Z[-c(1, l+1), -c(1, l+1)] * outer(rep(1,l-1), p_out[-1])
+Ptr = rbind(cbind(rep(0, l-1), Ptr), rep(0, l))

### 2022-03-28 single_line/notes/notes_GG/Notes_trajet.tex
Structure changed

\documentclass[11p]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{tikz}

\tikzset{main/.style = {draw, circle, minimum width={width("$l-1$")+10pt}}}
\tikzset{dashdot/.style = {draw, dashed, circle, minimum width={width("$l-1$")+10pt}}}
%opening
\title{Lignes de bus, approche Markov}
\author{notes GG}

\begin{document}

\maketitle
Soit $x_i$, resp. $y_i$, le nombre de passagers qui montent, resp. descendent, en $i$ ($1 \leq i \leq l$). On pose le $n_{i, i+1}$ le nombre de passagers qui sont transportées entre $i$
à $i+1$. On cherche $N_{ij}$ avec les bonnes marges. \\

Posons $p^\text{in}_i$ et $p^\text{out}_i$ comme :

\begin{align}
p^\text{in}_i &:= \frac{x_i}{x_\bullet} \\
p^\text{out}_i &:= \frac{y_i}{n_{i-1, i}} = \frac{y_i}{\sum_{1 \leq k \leq (i-1)} (x_k - y_k)}
\end{align}

On peut alors supposer une distribution $f_{ij} = \frac{N_{ij}}{N_{\bullet \bullet}}$ de la forme suivante :

\begin{equation}
f_{ij} = \begin{cases}
p^\text{in}_i (1 - p^\text{out}_{i+1}) \cdots (1 - p^\text{out}_{j-1}) p^\text{out}_j & \text{si } i < j - 1 \\
p^\text{in}_i p_j^\text{out} & \text{si } i = j - 1 \\
0 & \text{sinon}
\end{cases}
\end{equation}

\section{Chaîne de Markov}

La modélisation par une chaîne de Markov se fait de la façon suivante :

\begin{center}
\begin{tikzpicture}
\node[dashdot] (1) at (0,0) {$1$};
\node[main] (2) at (2,0) {$2$};
\node[main] (3) at (4,0) {$3$};
\node[] (dots) at (6,0) { \vspace{2cm} ...  \vspace{2cm} };
\node[main] (lp) at (8,0) {${l-1}$};
\node[main] (l) at (10,0) {$l$};

\node[main] (s) at (6, 3) {$s$};
\node[main] (t) at (6, -3) {$t$};

\path [->] (s) edge node[left] {\tiny$p^\text{in}_{1}$} (2);
\path [->] (s) edge node[left] {\tiny$p^\text{in}_{2}$} (3);
\path [->] (s) edge node[left] {\tiny$p^\text{in}_{l-2}$} (lp);
\path [->] (s) edge node[right] {\tiny$p^\text{in}_{1-1}$} (l);

\path [->, dashed] (1) edge node[above] {} (2);
\path [->] (2) edge node[above] {\tiny$1 - p^\text{out}_{2}$} (3);
\path [->] (3) edge node[above] {\tiny$1 - p^\text{out}_{3}$} (dots);
\path [->] (dots) edge node[above] {\tiny$1 - p^\text{out}_{l-2}$} (lp);
\path [->] (lp) edge node[above] {\tiny$1 - p^\text{out}_{l-1}$} (l);

\path [->] (2) edge node[left] {\tiny$p^\text{out}_{2}$} (t);
\path [->] (3) edge node[left] {\tiny$p^\text{out}_{3}$} (t);
\path [->] (lp) edge node[left] {\tiny$p^\text{out}_{l-1}$} (t);
\path [->] (l) edge node[right] {\tiny$p^\text{out}_{1}=1$} (t);
\end{tikzpicture} 
\end{center}

qui est donc représentée par la matrice de transition $\mathbf{W}$ suivante :

\begin{equation}
\mathbf{W} = \left(
\begin{array}{c|ccccccc|c}
	0 & p^\text{in}_1 & p^\text{in}_2 & p^\text{in}_3 & \ldots & p^\text{in}_{l-3} & p^\text{in}_{l-2} & p^\text{in}_{l-1} & 0 \\
	\hline
	0 & 0 & 1 - p^\text{out}_{2} & 0 & \cdots & 0 & 0 & 0 & p^\text{out}_{2} \\	
	0 & 0 & 0 & 1 - p^\text{out}_{3} & \ddots & 0 & 0 & 0 & p^\text{out}_{3} \\
	0 & 0 & 0 & 0 & \ddots & 0 & 0 & 0 & p^\text{out}_{4} \\
	\rule{0pt}{18pt} \vdots & \vdots & \vdots & \vdots & \ddots & \ddots & \ddots & \vdots & \vdots \\
	\rule{0pt}{18pt} 0 & 0 & 0 & 0 & \cdots & 0 & 1 - p^\text{out}_{l-2} & 0 & p^\text{out}_{l-2} \\
	\rule{0pt}{15pt} 0 & 0 & 0 & 0 & \cdots & 0 & 0 & 1 - p^\text{out}_{l-1} & p^\text{out}_{l-1} \\
	\rule{0pt}{15pt} 0 & 0 & 0 & 0 & \cdots & 0 & 0 & 0 & p^\text{out}_{1} \\
	\hline
	\rule{0pt}{15pt} 0 & 0 & 0 & 0 & \cdots & 0 & 0 & 0 & 0
\end{array}
\right)
\end{equation}

Pour représenter le fait qu'une personne qui monte en $i$ va nécessairement en $i + 1$ avant de descendre, la probabilité de monter à l'arrêt $i$, i.e $p^\text{in}_i$, est directement connecté au noeud $i+1$. Le noeud représentant l'arrêt $1$ n'est donc plus nécessaire. \\

A partir de cette matrice, on peut calculer la \emph{matrice fondamentale}, $\mathbf{Z}$, de la façon suivante :

\begin{equation}
\mathbf{Z} := \mathbf{I} + \mathbf{W} + \mathbf{W}^2 + \mathbf{W}^3 + \ldots =  (\mathbf{I} - \mathbf{W})^{-1}
\end{equation}

Comme cette chaîne ne possède pas de boucle, on sait que les composantes de cette matrice $z_{kl}$ contiennent la \emph{probabilité d'atteindre $l$ en partant de $k$} (REF, p.ex. Saerens).  On a donc :

\begin{equation}
z_{kl} = \begin{cases}
(1 - p^\text{out}_{k}) \cdots (1 - p^\text{out}_{l-1}) & \text{si } k < l \\
1 & \text{si } k = l \\
0 & \text{sinon}
\end{cases} \qquad \forall k, l \neq s, t
\end{equation}

la probabilité $f_{ij}$ s'obtient donc avec :

\begin{equation}
f_{ij} = p^\text{in}_i z_{i+1, j} p^\text{out}_j
\end{equation}

\subsection{Marges}

Les marges de $f_{ij}$ doivent vérifier $f_{i \bullet} = x_i / x_\bullet$ et $f_{\bullet j} = y_j / y_\bullet$. On a :

\begin{align}
f_{i \bullet} = \sum_{j=1}^l p^\text{in}_i z_{i+1, j} p^\text{out}_j = p^\text{in}_i \sum_{j=1}^l z_{i+1, j} p^\text{out}_j 
\end{align}
Comme $z_{i+1, j} p^\text{out}_j$ représente la probabilité d'avoir une trajectoire partant de $i+1$ et étant absorbé juste après avoir atteint $j$, on a :

\begin{align}
\sum_{j=1}^l z_{i+1, j} p^\text{out}_j &= 1 \\
\Rightarrow f_{i \bullet} = p^\text{in}_i &= \frac{x_i}{x_\bullet}
\end{align}

Pour l'autre marge, on fait par réccurence. Pour j = 2 :

\begin{align}
f_{\bullet 2} &= p^\text{in}_1 p^\text{out}_2 = \frac{x_1}{x_\bullet} \frac{y_2}{(x_1 - y_1)} \notag \\
			  &= \frac{x_1}{x_\bullet} \frac{y_2}{x_1} = \frac{y_2}{x_\bullet} = \frac{y_2}{y_\bullet}
\end{align}

En supposant que $f_{\bullet j} = y_j / y_\bullet$, calculons $f_{\bullet, j+1}$. Pour commencer, remarquons que :

\begin{align}
z_{kl} = \begin{cases}
z_{k, (l-1)} (1 - p^\text{out}_{l-1}) & \text{si } k < l \\
1 & \text{si } k = l \\
0 & \text{sinon}
\end{cases}
\end{align}

On a donc :

\begin{align}
f_{\bullet, (j+1)} &= \sum_{i=1}^l p^\text{in}_i z_{(i+1), (j+1)} p^\text{out}_{j+1} = \sum_{i=1}^{j-1} p^\text{in}_i z_{i+1, j} (1 - p^\text{out}_{j}) p^\text{out}_{j+1} + p^\text{in}_j p^\text{out}_{j+1} \notag \\
&= \frac{f_{\bullet, j}}{p^\text{out}_j}(1 - p^\text{out}_j)p^\text{out}_{j+1} + \frac{x_j}{x_\bullet} p^\text{out}_{j+1} = \left( \frac{y_j}{y_\bullet} \left( \frac{1}{p^\text{out}_j} - 1 \right)+ \frac{x_j}{x_\bullet} \right) p^\text{out}_{j+1} \notag \\
&= \left( \frac{n_{j-1,j} - y_j + x_j}{y_\bullet} \right) \frac{y_{j+1}}{n_{j,j+1}} = \frac{y_{j+1}}{y_\bullet}
\end{align}

\end{document}

### 2022-03-28 single_line/compute_distrib.R

@@ -1,4 +1,7 @@
-setwd("~/Documents/recherche/projet_tl")
+# Set working directory 
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Load data
 data = read.csv("data/Ptr_marges.csv")
 
 nom_arrets = names(data)

### 2022-03-28 single_line/graphs_freq_tl_lignes.R

@@ -1,6 +1,7 @@
 ############ Graphique de fréquentation annuelle des lignes TL
 
-setwd("/Users/rloup/Desktop")
+# Set working directory 
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
 ### Connexion à la base de données
 library('RPostgreSQL')

### 2022-03-28 single_line/markov_tl_lignes.R

@@ -1,4 +1,5 @@
-setwd("/Users/rloup/Desktop/markov")
+# Set working directory 
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
 ############# Connexion à la base de données
 

### 2022-03-28 multiple_lines/Multiligne_6789_bynode.R
Added multiple lines script

@@ -1,0 +1,544 @@
+###############################################################
+# CODE R : multi-line model
+###############################################################
+
+############################################################### 
+####### LIBRARIES
+
+library(igraph)
+
+############################################################### 
+####### Setting paths
+
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+############################################################### 
+####### PARAMETERS
+
+compute_sp_data = F
+epsilon = 1e-30
+conv_thres_if = 1e-15
+conv_thres_algo = 1e-15
+local_lambda = 1
+flow_error = 1e-20
+
+############################################################### 
+####### LOADING AND PREPROCESSING MATRICES
+
+
+# --- Loading adjacency matrices 
+
+# Loading adjacency matrix without loop
+adj_line = read.csv("data_multiligne/adj_next.csv", sep=",", row.names = 1)
+adj_line = as.matrix(adj_line)
+diag(adj_line) = 0
+
+# Loading adjacency matrix with loops on the end of the line
+adj_line_wloop = read.csv("data_multiligne/adj_next_bind.csv", sep=",", row.names = 1)
+adj_line_wloop = as.matrix(adj_line_wloop)
+diag(adj_line_wloop) = 0
+
+# Component for line
+g_adj_wloop = graph_from_adjacency_matrix(adj_line_wloop, mode = "directed")
+line_comp = components(g_adj_wloop)$membership
+
+# Loading proximity matrix
+adj_prox = read.csv("data_multiligne/adj_prox.csv", sep=",", row.names = 1)
+adj_prox = as.matrix(adj_prox)
+diag(adj_prox) = 0
+
+# Erease same line join
+for(id_line in 1:max(line_comp)){
+  adj_prox[line_comp == id_line, line_comp == id_line] = 0
+}
+
+# Saving stop names and renaming columns
+stop_names = trimws(paste0("S", rownames(adj_line)))
+colnames(adj_line) = stop_names
+rownames(adj_line) = stop_names
+colnames(adj_line_wloop) = stop_names
+rownames(adj_line_wloop) = stop_names
+colnames(adj_prox) = stop_names
+rownames(adj_prox) = stop_names
+
+# Number of stops
+n = length(stop_names)
+
+# Correction for the 7th line 
[…]

### 2022-03-28 github.R
Added github connection script

@@ -1,0 +1,55 @@
+#### 1. Sign up at GitHub.com ################################################
+
+## If you do not have a GitHub account, sign up here:
+## https://github.com/join
+
+# ----------------------------------------------------------------------------
+
+#### 2. Install git ##########################################################
+
+## If you do not have git installed, please do so: 
+## Windows ->  https://git-scm.com/download/win
+## Mac     ->  https://git-scm.com/download/mac
+## Linux   ->  https://git-scm.com/download/linux
+##             or: $ sudo dnf install git-all
+
+# ----------------------------------------------------------------------------
+
+### 3. Configure git with Rstudio ############################################
+
+## set your user name and email:
+usethis::use_git_config(user.name = "YourName", user.email = "your@mail.com")
+
+## create a personal access token for authentication:
+usethis::create_github_token() 
+## in case usethis version < 2.0.0: usethis::browse_github_token() (or even better: update usethis!)
+
+## set personal access token:
+credentials::set_github_pat("YourPAT")
+
+## or store it manually in '.Renviron':
+usethis::edit_r_environ()
+## store your personal access token with: GITHUB_PAT=xxxyyyzzz
+## and make sure '.Renviron' ends with a newline
+
+# ----------------------------------------------------------------------------
+
+#### 4. Restart R! ###########################################################
+
+# ----------------------------------------------------------------------------
+
+#### 5. Verify settings ######################################################
+
+usethis::git_sitrep()
+
+## Your username and email should be stated correctly in the output. 
+## Also, the report shoud cotain something like:
+## 'Personal access token: '<found in env var>''
+
+## If you are still having troubles, read the output carefully.
+## It might be that the PAT is still not updated in your `.Renviron` file.
+## Call `usethis::edit_r_environ()` to update that file manually.
+
+# ----------------------------------------------------------------------------
+
+## THAT'S IT!

### 2022-03-28 multiple_lines/Multiligne_6789_bynode.R
Corrected bug for the final display

@@ -507,20 +507,33 @@ df_free_edge = data.frame(from=in_node_nn, to=out_node_nn,
                           quant=total_edge_flow[(round(total_edge_flow, 10) > 0) & free_edge_vec] * to_num)
 
 # Computing in-network/def, out-network/def
-transfer_out = rep(0, n)
-transfer_out[non_null_edge[, 1]] = df_free_edge$quant
-transfer_in = rep(0, n)
-transfer_in[non_null_edge[, 2]] = df_free_edge$quant
+transfer_out = c()
+transfer_in = c()
+for(id_node in 1:n){
+  
+  # Free out/in-going flow from node in and out
+  id_free_edge_from = intersect(which(edge_mat[,1] == id_node), which(free_edge_vec))
+  id_free_edge_to = intersect(which(edge_mat[,2] == id_node), which(free_edge_vec))
+  
+  # Outgoing and ingoing flow
+  free_outgoing_flow = sum(total_edge_flow[id_free_edge_from])
+  free_ingoing_flow = sum(total_edge_flow[id_free_edge_to])
+  
+  # flow transfer 
+  transfer_out = c(transfer_out, free_outgoing_flow)
+  transfer_in = c(transfer_in, free_ingoing_flow)
+}
   
 full_df = inout_cor
 full_df["montees_initiales"] = round(sigma_in * to_num)
-full_df["transferts_in"] = transfer_in
+full_df["transferts_in"] = round(transfer_in * to_num)
 full_df["diff_in"] = full_df["montees_initiales"] + full_df["transferts_in"] - full_df["montees"]
 full_df["descentes_finales"] = round(sigma_out * to_num)
-full_df["transferts_out"] = transfer_out
+full_df["transferts_out"] = round(transfer_out * to_num)
 full_df["diff_out"] = full_df["descentes_finales"] + full_df["transferts_out"] - full_df["descentes"]
 
 View(full_df)
+View(df_free_edge)
 
 # View
 id_arret = 30

### 2022-03-28 multiple_lines/notes/latex/Notes_trajet.tex
Added note for multilines

\documentclass[11p]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}

%opening
\title{Bus lines, multilines approach}
\author{notes GG}

\begin{document}

\maketitle

\section{Formalism}

\subsection{Network definition}

Let $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ be a oriented graph and $\mathbf{A}$ its adjacency matrix, representing a multimodal transportation network between $|\mathcal{V}| = n$ stops and possessing $l$ lines. Each stop belong to only one line, i.e. $\mathcal{V} = \bigcup_{k=1}^l \mathcal{V}_k$ and $\bigcap_{k=1}^l \mathcal{V}_k = \emptyset$, where $\mathcal{V}_k$ represents the set of nodes in line $k$. The edge set $\mathcal{E}$, can also be decomposed with $\mathcal{E} = \left(\bigcup_{k=1}^l \mathcal{E}_k\right) \cup \mathcal{E}_\text{trsf}$, where $\mathcal{E}_k$ contains edges connecting node inside transportation line $k$, and $\mathcal{E}_\text{trsf}$ contains edges permitting transfer between different transportation lines. We suppose that edges belong to only a unique set, i.e. $\mathbf{A} = \sum_{k=1}^l \mathbf{A}_k + \mathbf{A}_\text{trsf}$, and there are an uniquely define route inside lines, i.e. $a^k_{i \bullet} \leq 1$ and $a^k_{\bullet i} \leq 1$ $\forall i,k$. Transfer edges also define a subset of nodes $\mathcal{V}_\text{f} \subset \mathcal{V}$, containing \emph{free nodes}, i.e. nodes connected to a transfer edge. We have $i \in \mathcal{V}_\text{f} \iff a^\text{trsf}_{i \bullet} + a^\text{trsf}_{\bullet i} > 0$.

\subsection{Problem definition}

Suppose that we possess two $n$-length vectors $\mathbf{x}_\text{in} = (x^\text{in}_i)$ and $\mathbf{y}_\text{out} = (y^\text{out}_i)$, representing respectively the flow entering the network at stop $i$ and going out of the network at stop $i$.

\end{document}

### 2022-03-29 multiple_lines/notes/latex/Notes_trajet.tex
Multiline latex change

@@
 
-Suppose that we possess two $n$-length vectors $\mathbf{x}_\text{in} = (x^\text{in}_i)$ and $\mathbf{y}_\text{out} = (y^\text{out}_i)$, representing respectively the flow entering the network at stop $i$ and going out of the network at stop $i$.
+A \emph{flow matrix} $\mathbf{N} = (n_{ij})$.
+Suppose that we possess two $n$-length vectors $\mathbf{x}_\text{in} = (x^\text{in}_i)$ and $\mathbf{y}_\text{out} = (y^\text{out}_i)$, representing respectively the flow entering a line at stop $i$ and going out of the line at stop $i$.
 

### 2022-03-29 multiple_lines/notes/latex/Notes_trajet.tex
Multiple lines notes updated

--- 2022-03-28 multiple_lines/notes/latex/Notes_trajet.tex
@@
 
-\subsection{Problem definition}
+\subsection{Flow definition}
 
-Suppose that we possess two $n$-length vectors $\mathbf{x}_\text{in} = (x^\text{in}_i)$ and $\mathbf{y}_\text{out} = (y^\text{out}_i)$, representing respectively the flow entering the network at stop $i$ and going out of the network at stop $i$.
+A matrix $\mathbf{N} = (n_{ij})$, verifying:
+\begin{enumerate}
+	\item $n_{ij} \geq 0$
+	\item $a_{ij} = 0 \Rightarrow n_{ij} = 0$
+\end{enumerate}
+is a \emph{flow matrix} defined on $\mathcal{G}$. Again, we can decompose this matrix with 
+\begin{equation}
+\mathbf{N} = \mathbf{N}_\text{W} + \mathbf{N}_\text{B} \qquad \mathbf{N}_\text{W} := \sum_{k=1}^l \mathbf{N}_k
+\end{equation}
+where the flow $\mathbf{N}_\text{W}$ is the flow inside the lines, $\mathbf{N}_\text{B}$ the flow between lines, and $\mathbf{N}_l$ the flow inside line $l$.
 
+Suppose that we possess two $n$-length vectors $\mathbf{l}_\text{in} = (l^\text{in}_i)$ and $\mathbf{l}_\text{out} = (l^\text{out}_i)$, representing ,respectively, the flow entering the line associated with stop $i$ and going out of this line at stop $i$.
+These vectors can in fact be expressed with:
+\begin{align}
+\mathbf{l}_\text{in} &= \bm{\sigma}_\text{in} + 
+\end{align}
 \end{document}

### 2022-03-30 data_processing/adjency_matrix.R
Dry code

@@ -1,0 +1,199 @@
+############################################################### 
+####### Setting paths
+
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+setwd("/Users/rloup/switchdrive/theseRomain/data")
+
+##### Distance Matrix #####
+
+
+# Database connexion
+library("RPostgreSQL")
+# Connexion à la base de données
+conn <- dbConnect("PostgreSQL", dbname = 'tl_small', host = 'localhost', user = 'rloup', password = '')
+
+library("sf")
+
+# Réseau de tous les arrêts
+df <- st_read(conn, query="
+select distinct stop_order.code_ligne_theo, stop_order.code_arret_theo, stop_order.libelle_arret_theo, stop_order.sequence_theo, stop_order.distance_arret_prec_theo, stop_order.direction_voy_theo,
+freq_theo.lat_arret, freq_theo.lon_arret, freq_theo.geom
+from stop_order
+inner join freq_theo
+on stop_order.code_arret_theo = freq_theo.code_arret_theo
+order by code_ligne_theo, direction_voy_theo, sequence_theo;")
+
+df2 <- df
+df <- df2
+df$new_seq <- 1
+i <- 1
+j <- 1
+
+while (i <= length(df$code_arret_theo)) {
+  df$new_seq[i] <- j
+  if(df$code_ligne_theo[i] == df$code_ligne_theo[i+1] && df$code_ligne_theo[i] == df$code_ligne_theo[i+1] && df$direction_voy_theo[i] == df$direction_voy_theo[i+1]) {
+    j <- j + 1
+  }
+  else{
+    j <- 1
+  }
+  i <- i + 1
+}
+
+# Lier dernier arrêt de l'aller avec le premier du retour
+while (i <= length(df$code_arret_theo)) {
+  df$new_seq[i] <- j
+  if(df$code_ligne_theo[i] == df$code_ligne_theo[i+1] && df$code_ligne_theo[i] == df$code_ligne_theo[i+1]) {
+    j <- j + 1
+  }
+  else{
+    j <- 1
+  }
+  i <- i + 1
+}
+  
+# Libraries
+library(geosphere)
+df2 <- df
+#Database
+df <- read.csv("arrets_geom_inner.csv", header = TRUE)
+
[…]

### 2022-03-30 multiple_lines/notes/latex/Notes_trajet.tex

--- 2022-03-28 multiple_lines/notes/latex/Notes_trajet.tex
@@
 
-\subsection{Network definition}
+\subsection{The multi-lines transportation network}
 
-Let $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ be a oriented graph and $\mathbf{A}$ its adjacency matrix, representing a multimodal transportation network between $|\mathcal{V}| = n$ stops and possessing $l$ lines.
-Each stop belong to only one line, i.e.
-$\mathcal{V} = \bigcup_{k=1}^l \mathcal{V}_k$ and $\bigcap_{k=1}^l \mathcal{V}_k = \emptyset$, where $\mathcal{V}_k$ represents the set of nodes in line $k$.
-The edge set $\mathcal{E}$, can also be decomposed with $\mathcal{E} = \left(\bigcup_{k=1}^l \mathcal{E}_k\right) \cup \mathcal{E}_\text{trsf}$, where $\mathcal{E}_k$ contains edges connecting node inside transportation line $k$, and $\mathcal{E}_\text{trsf}$ contains edges permitting transfer between different transportation lines.
-We suppose that edges belong to only a unique set, i.e.
-$\mathbf{A} = \sum_{k=1}^l \mathbf{A}_k + \mathbf{A}_\text{trsf}$, and there are an uniquely define route inside lines, i.e.
-$a^k_{i \bullet} \leq 1$ and $a^k_{\bullet i} \leq 1$ $\forall i,k$.
-Transfer edges also define a subset of nodes $\mathcal{V}_\text{f} \subset \mathcal{V}$, containing \emph{free nodes}, i.e.
-nodes connected to a transfer edge.
-We have $i \in \mathcal{V}_\text{f} \iff a^\text{trsf}_{i \bullet} + a^\text{trsf}_{\bullet i} > 0$.
+Let $\mathcal{G} = (\mathcal{V}, \mathcal{E})$ be a simple, oriented, and connected graph representing a transportation network between $|\mathcal{V}| = n$ nodes, having $|\mathcal{E}| = m$ edges, and possessing $p$ different transportation lines.
+Each node belongs to only one line, i.e.
+$\mathcal{V} = \bigcup_{k=1}^p \mathcal{V}_k$ and $\bigcap_{k=1}^p \mathcal{V}_k = \emptyset$, where $\mathcal{V}_k$ represents the set of nodes in line $k$.
+The edge set $\mathcal{E}$, can also be decomposed with
+\begin{equation}
+\mathcal{E} = \mathcal{E}_\text{W} \cup \mathcal{E}_\text{B}, \qquad \mathcal{E}_\text{W} := \bigcup_{k=1}^p \mathcal{E}_k, \\
+\mathcal{E}_\text{W} \cap \mathcal{E}_\text{B} = \emptyset, \quad \mathcal{E}_k \cap \mathcal{E}_l = \emptyset, \quad \forall k, l.
+\end{equation}
+where $\mathcal{E}_k$ is the set of edges composing line $k$, $\mathcal{E}_\text{W}$ the set containing all edges inside lines, and $\mathcal{E}_\text{B}$ the set of transfer edges, connecting the different lines.
+\\
+The graph $\mathcal{G}$ can be represented by its adjacency matrix $\mathbf{A} = (a_{ij})$, which can also be decomposed with 
+\begin{equation}
+\mathbf{A} = \mathbf{A}_\text{W} + \mathbf{A}_\text{B}, \qquad \mathbf{A}_\text{W} = \sum_{k=1}^p \mathbf{A}_k,
+\end{equation}
+with $\mathbf{A}_k = (a^k_{ij})$ are edges of line $k$, $\mathbf{A}_\text{W} = (a^\text{W}_{ij})$ edges of inside all lines, and $\mathbf{A}_\text{B} = (a^\text{B}_{ij})$ transfer edges.
+We suppose that there is an uniquely define route inside lines, i.e.
+\begin{equation}
+a^k_{i \bullet} \leq 1 \text{ and } a^k_{\bullet i} \leq 1, \qquad \forall i,k.
+\end{equation}
+where $\bullet$ designs a summation over the replaced index.
+
+\subsection{The origin-destination matrix}
+
+The $(n \times n)$ \emph{origin-destination matrix}, denoted by $\mathbf{N} = (n_{st})$, $n_{st} \geq 0, \; \forall s,t$, contains the flow (e.g.
+the number of passengers) entering the network in source node $s$ and leaving it in target node $t$.
+We can denote its margins with
+\begin{align}
+\bm{\sigma}_\text{in} := \mathbf{N} \mathbf{e}_n \\
+\bm{\sigma}_\text{out} := \mathbf{N}^\top \mathbf{e}_n \\
+\end{align}
+where $\mathbf{e}_n$ is the vector of ones of size $n$.
+The vector $\bm{\sigma}_\text{in} = (\sigma^\text{in}_i)$ is the \emph{vector of flow entering the network} and $\bm{\sigma}_\text{in} = (\sigma^\text{in}_i)$ is the \emph{vector of flow escaping the network}.
+We have
+\begin{equation}
+\sigma^\text{in}_{\bullet} = \sigma^\text{out}_{\bullet}.
+\end{equation}
+
+\subsection{The flow matrix}
+
+A flow on edges is represented by the $(n \times n)$ \emph{flow matrix} $\mathbf{X} = (x_{ij})$, verifying
+\begin{align}
+x_{ij} &\geq 0, \qquad \forall i,j, \\
+a_{ij} = 0 &\Rightarrow x_{ij} = 0, \qquad \forall i,j, \\ 
+x_{i\bullet} + \sigma^\text{in}_i &= x_{\bullet i} + \sigma^\text{out}_i, \qquad \forall i.
+\end{align}
+Again, we can decompose the flow matrix with 
+\begin{equation}
+\mathbf{X} = \mathbf{X}_\text{W} + \mathbf{X}_\text{B} \qquad \mathbf{X}_\text{W} := \sum_{k=1}^l \mathbf{X}_k
+\end{equation}
+where $\mathbf{X}_k$ represent the flow inside line $k$, $\mathbf{X}_\text{W}$ is the flow inside all lines, and $\mathbf{X}_\text{B}$ the flow between lines. This decomposition allows us to define 
+the \emph{vector of flow entering lines} $\bm{\rho}_\text{in} = (\rho^\text{in}_i)$ and the \emph{vector of flow escaping lines} $\bm{\rho}_\text{out} = (\rho^\text{out}_i)$, with
+\begin{align}
+\bm{\rho}_\text{in} &:= \bm{\sigma}_\text{in} + \mathbf{X}^\top_\text{B} \mathbf{e}_n, \\
+\bm{\rho}_\text{out} &:= \bm{\sigma}_\text{out} + \mathbf{X}_\text{B} \mathbf{e}_n,
+\end{align}
+where $\mathbf{e}_n$ is the vector of ones of size $n$. It is easy to see that we still have $\rho^\text{in}_\bullet = \rho^\text{out}_\bullet$. \\
+
+\subsection{Shortest-paths flow}
+
+Let $\mathcal{P}_{st}$ be the set of \emph{admissible} shortest-paths between $s$ and $t$ on $\mathcal{G}$.
+We can denote by $\text{P}_{st}(i, j)$ the probability of having edge $(i, j) \in \wp$ when drawing a path $\wp$ from $\mathcal{P}_{st}$.
+We have
+\begin{equation}
+\text{P}_{st}(i, j) := \frac{1}{|\mathcal{P}_{st}|}\sum_{\wp \in \mathcal{P}_{st}} \delta((i, j) \in \wp),
+\end{equation}
+where $\delta(.)$ designate the indicator function.
+If we are given an origin-destination matrix $\mathbf{N} = (n_{st})$, we can compute the \emph{shortest-path flow matrix}, noted $\mathbf{X}_\text{sp} = (x^\text{sp}_{ij})$, with:
+\begin{equation}
+x^\text{sp}_{ij} = \sum_{st} \text{P}_{st}(i, j) n_{st} \label{sp_comp}
+\end{equation}
+This matrix contains the flow on each edge if we suppose that the flow follows shortest-paths.
 
@@
 
-Suppose that we possess two $n$-length vectors $\mathbf{x}_\text{in} = (x^\text{in}_i)$ and $\mathbf{y}_\text{out} = (y^\text{out}_i)$, representing respectively the flow entering the network at stop $i$ and going out of the network at stop $i$.
+We suppose that we know the flow entering and leaving each line, i.e.
+$\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{out}$ and we are interested in finding trajectories $n_{st}$ of the flow in the network knowing the set $\mathcal{A}$ of admissible pair $(s, t)$, i.e.
+pair of nodes where there is a possible use of the network for traveling.
+This set can be given by the matrix $\mathbf{T} = (t_{st})$ defined by
+\begin{equation}
+t_{st} = \begin{cases}
+1 & \text{if } (s, t) \in \mathcal{A}, \\
+0 & \text{otherwise}.
+\end{cases}
+\end{equation}
+
+\subsection{Algorithm}
+Set $\bm{\sigma}^{(0)}_\text{in} = \bm{\rho}_\text{in} + \epsilon$ and $\bm{\sigma}^{(0)}_\text{out} = \bm{\rho}_\text{out} + \epsilon$, with $\epsilon$ a small positive scalar. Until convergence, do:
+\begin{enumerate}
+	\item Compute $\mathbf{N}^{(i)} = \textbf{Diag}(\mathbf{a}^{(i)}) (\mathbf{T} + \epsilon)\textbf{Diag}(\mathbf{b}^{(i)})$ with iterative fitting, such that $\mathbf{N}^{(i)} \mathbf{e}_n = \bm{\sigma}^{(i)}_\text{in}$ and $\left(\mathbf{N}^{(i)}\right)^\top \mathbf{e}_n = \bm{\sigma}^{(i)}_\text{out}$.
+	\item Compute the associated shortest-path flow matrix $\mathbf{X}^{(i)}$ with (\ref{sp_comp}).
+	\item Compute $\bm{\sigma}^{(i+1)}_\text{in} = \bm{\rho}_\text{in} -  (\mathbf{X}^{(i)})^\top_\text{B} \mathbf{e}_n$ and $\bm{\sigma}^{(i+1)}_\text{out} = \bm{\rho}_\text{out} - \mathbf{X}^{(i)}_\text{B} \mathbf{e}_n$.
+\end{enumerate}
 

### 2022-03-30 data_processing/adjency_matrix.R
Update

@@ -2,34 +2,35 @@
 ####### Setting paths
 
 setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
-setwd("/Users/rloup/switchdrive/theseRomain/data")
+# setwd("/Users/rloup/switchdrive/theseRomain/data")
+
+
+###### Libraries
+library(RPostgreSQL) # Database connection
+library(tidyverse) # Data filter
+library(geosphere) # Distances between points
+
+
 
 ##### Distance Matrix #####
 
-
 # Database connexion
-library("RPostgreSQL")
+
 # Connexion à la base de données
 conn <- dbConnect("PostgreSQL", dbname = 'tl_small', host = 'localhost', user = 'rloup', password = '')
 
-library("sf")
 
 # Réseau de tous les arrêts
-df <- st_read(conn, query="
-select distinct stop_order.code_ligne_theo, stop_order.code_arret_theo, stop_order.libelle_arret_theo, stop_order.sequence_theo, stop_order.distance_arret_prec_theo, stop_order.direction_voy_theo,
-freq_theo.lat_arret, freq_theo.lon_arret, freq_theo.geom
-from stop_order
-inner join freq_theo
-on stop_order.code_arret_theo = freq_theo.code_arret_theo
-order by code_ligne_theo, direction_voy_theo, sequence_theo;")
+df <- st_read(conn, query="SELECT * FROM stop_frequentation_tot;")
+
+# Save data
+write.csv2(df, "stop_frequentation.csv", row.names = FALSE)
 
 df2 <- df
-df <- df2
-df$new_seq <- 1
-i <- 1
-j <- 1
 
-while (i <= length(df$code_arret_theo)) {
+# Rectify stop order
+j <- 1
+for (i in 1:dim(df)[1]) {
   df$new_seq[i] <- j
   if(df$code_ligne_theo[i] == df$code_ligne_theo[i+1] && df$code_ligne_theo[i] == df$code_ligne_theo[i+1] && df$direction_voy_theo[i] == df$direction_voy_theo[i+1]) {
     j <- j + 1
@@ -37,10 +38,11 @@ while (i <= length(df$code_arret_theo)) {
   else{
     j <- 1
   }
-  i <- i + 1
 }
 
-# Lier dernier arrêt de l'aller avec le premier du retour
+i<-1
+j<-1
[…]

### 2022-03-31 multiple_lines/old_code/Multiligne_6789_bynode_old.R
New version of the algorithm

@@ -6,7 +6,6 @@
 ####### LIBRARIES
 
 library(igraph)
-library(Matrix)
 
 ############################################################### 
 ####### Setting paths
@@ -19,7 +18,7 @@ setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 compute_sp_data = F
 epsilon = 1e-30
 conv_thres_if = 1e-15
-conv_thres_algo = 1e-30
+conv_thres_algo = 1e-15
 local_lambda = 1
 flow_error = 1e-20
 
@@ -263,10 +262,10 @@ for(id_line in id_line_vec){
 }
 
 # Fixed distributions of in and out
-rho_in = inout_cor$montees + epsilon
-rho_in = rho_in / sum(rho_in)
-rho_out = inout_cor$descentes + epsilon
-rho_out = rho_out / sum(rho_out)
+pi_in = inout_cor$montees + epsilon
+pi_in = pi_in / sum(pi_in)
+pi_out = inout_cor$descentes + epsilon
+pi_out = pi_out / sum(pi_out)
 
 ############################################################### 
 ####### Algorithm 
@@ -277,7 +276,7 @@ sigma_out = pi_out
 p_sigma_in = sigma_in
 p_sigma_out = sigma_out
 converge_algo = F
-it_algo = 1
+epoch_algo = 1
 
 while(!converge_algo){
   
@@ -285,37 +284,135 @@ while(!converge_algo){
   p_sigma_in_old = p_sigma_in
   p_sigma_out_old = p_sigma_out
   
-  # --- --- Iterative fitting 
+  # Shuffle free node
+  free_node_shuffled = sample(which(free_node_vec))
+  # Reset id 
+  id_sample = 1
+  # edge_cycle_done
+  node_cycle_done = F
+  # Has mod
+  modification_occured = T
   
-  converge_if = F
-  results_mat = admissible_sp_mat + epsilon
-  while(!converge_if){
-    # Saving old results
-    results_mat_old = results_mat 
-    # Normalizing by row
-    results_mat = results_mat * sigma_in / rowSums(results_mat)
-    # Normalizing by columns 
-    results_mat = t(t(results_mat) * sigma_out / rowSums(t(results_mat)))
-    # Checking for convergence
-    if(sum(abs(results_mat_old - results_mat)) < conv_thres_if){
-      converge_if = T
-    }
-  }
-  
-  # Find flow on edges 
[…]

### 2022-03-31 multiple_lines/notes/latex/Notes_trajet.tex
New version of the algorithm
Algo is refined

@@
 %opening
-\title{Bus lines, multilines approach}
+\title{Transportation network with multiple lines}
 \author{notes GG}
@@
 
-\subsection{The multi-lines transportation network}
+\subsection{The transportation network with multiple lines}
 
@@
 \begin{equation}
-\mathcal{E} = \mathcal{E}_\text{W} \cup \mathcal{E}_\text{B}, \qquad \mathcal{E}_\text{W} := \bigcup_{k=1}^p \mathcal{E}_k, \\
-\mathcal{E}_\text{W} \cap \mathcal{E}_\text{B} = \emptyset, \quad \mathcal{E}_k \cap \mathcal{E}_l = \emptyset, \quad \forall k, l.
+\mathcal{E} = \mathcal{E}_\text{W} \cup \mathcal{E}_\text{B}, \qquad \mathcal{E}_\text{W} := \bigcup_{k=1}^p \mathcal{E}_k, \qquad \mathcal{E}_\text{W} \cap \mathcal{E}_\text{B} = \emptyset, \quad \mathcal{E}_k \cap \mathcal{E}_l = \emptyset, \quad \forall k, l.
 \end{equation}
@@
 \end{equation}
-where $\bullet$ designs a summation over the replaced index.
+where $\bullet$ designates a summation over the replaced index.
 
@@
 \bm{\sigma}_\text{in} := \mathbf{N} \mathbf{e}_n \\
-\bm{\sigma}_\text{out} := \mathbf{N}^\top \mathbf{e}_n \\
+\bm{\sigma}_\text{out} := \mathbf{N}^\top \mathbf{e}_n
 \end{align}
@@
 $\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{out}$ and we are interested in finding trajectories $n_{st}$ of the flow in the network knowing the set $\mathcal{A}$ of admissible pair $(s, t)$, i.e.
-pair of nodes where there is a possible use of the network for traveling.
+pairs of nodes where there is a possible use of the network for traveling.
 This set can be given by the matrix $\mathbf{T} = (t_{st})$ defined by
@@
 	\item Compute the associated shortest-path flow matrix $\mathbf{X}^{(i)}$ with (\ref{sp_comp}).
-	\item Compute $\bm{\sigma}^{(i+1)}_\text{in} = \bm{\rho}_\text{in} -  (\mathbf{X}^{(i)})^\top_\text{B} \mathbf{e}_n$ and $\bm{\sigma}^{(i+1)}_\text{out} = \bm{\rho}_\text{out} - \mathbf{X}^{(i)}_\text{B} \mathbf{e}_n$.
+	\item Compute $\bm{\sigma}^{(i+1)}_\text{in} = \bm{\rho}_\text{in} -  (\mathbf{X}^{(i)}_\text{B})^\top \mathbf{e}_n$ and $\bm{\sigma}^{(i+1)}_\text{out} = \bm{\rho}_\text{out} - \mathbf{X}^{(i)}_\text{B} \mathbf{e}_n$.
 \end{enumerate}

### 2022-03-31 multiple_lines/Multiligne_6789_bynode.R

@@ -272,18 +272,17 @@ rho_out = rho_out / sum(rho_out)
 ####### Algorithm 
 
 # Init
-sigma_in = pi_in
-sigma_out = pi_out
-p_sigma_in = sigma_in
-p_sigma_out = sigma_out
+sigma_in = rho_in
+sigma_out = rho_out
 converge_algo = F
 it_algo = 1
+X_b = matrix(1, n, n)
 
 while(!converge_algo){
   
-  # Save old distrib values
-  p_sigma_in_old = p_sigma_in
-  p_sigma_out_old = p_sigma_out
+  # --- --- Save old X_b
+  
+  X_b_old = X_b 
   
   # --- --- Iterative fitting 
   
@@ -312,62 +311,17 @@ while(!converge_algo){
   X_b = sparseMatrix(i=free_edge_loc[, 1],j=free_edge_loc[, 2],x=total_edge_flow[free_edge_vec], dims=c(n, n))
   
   # Update sigma_in and sigma_out
-  sigma_in = pi_in - colSums(X_b)
-  sigma_out = pi_out - rowSums(X_b)
+  sigma_in = rho_in - colSums(X_b)
+  sigma_out = rho_out - rowSums(X_b)
   sigma_in[sigma_in < 0] = 0
   sigma_out[sigma_out < 0] = 0
   
-  # --- --- Compute in-out balance 
-  
-  # IF
-  converge_if = F
-  results_mat = admissible_sp_mat + epsilon
-  while(!converge_if){
-    # Saving old results
-    results_mat_old = results_mat 
-    # Normalizing by row
-    results_mat = results_mat * sigma_in / rowSums(results_mat)
-    # Normalizing by columns 
-    results_mat = t(t(results_mat) * sigma_out / rowSums(t(results_mat)))
-    # Checking for convergence
-    if(sum(abs(results_mat_old - results_mat)) < conv_thres_if){
-      converge_if = T
-    }
-  }
-  
-  # Find flow on edges 
-  total_edge_flow = as.vector(t(sp_edge_mat) %*% c(results_mat))
-  
-  # Find flow on free edges
-  in_error_vec = c()
-  out_error_vec = c()
-  for(id_node in 1:n){
-    
-    # Free out/in-going flow from node in and out
-    id_free_edge_from = intersect(which(edge_mat[,1] == id_node), which(free_edge_vec))
[…]

### 2022-04-05 multiple_lines/notes/latex/Notes_trajet.tex
update notes and began prototyping functions

@@
 where $\mathbf{e}_n$ is the vector of ones of size $n$.
-The vector $\bm{\sigma}_\text{in} = (\sigma^\text{in}_i)$ is the \emph{vector of flow entering the network} and $\bm{\sigma}_\text{in} = (\sigma^\text{in}_i)$ is the \emph{vector of flow escaping the network}.
+The vector $\bm{\sigma}_\text{in} = (\sigma^\text{in}_i)$ is the \emph{vector of flow entering the network} and $\bm{\sigma}_\text{in} = (\sigma^\text{in}_i)$ is the \emph{vector of flow leaving the network}.
 We have
@@
 where $\mathbf{X}_k$ represent the flow inside line $k$, $\mathbf{X}_\text{W}$ is the flow inside all lines, and $\mathbf{X}_\text{B}$ the flow between lines. This decomposition allows us to define 
-the \emph{vector of flow entering lines} $\bm{\rho}_\text{in} = (\rho^\text{in}_i)$ and the \emph{vector of flow escaping lines} $\bm{\rho}_\text{out} = (\rho^\text{out}_i)$, with
+the \emph{vector of flow entering lines} $\bm{\rho}_\text{in} = (\rho^\text{in}_i)$ and the \emph{vector of flow leaving lines} $\bm{\rho}_\text{out} = (\rho^\text{out}_i)$, with
 \begin{align}

### 2022-04-05 multiple_lines/Multiligne_6789_bynode.R

@@ -47,7 +47,7 @@ adj_prox = read.csv("data_multiligne/adj_prox.csv", sep=",", row.names = 1)
 adj_prox = as.matrix(adj_prox)
 diag(adj_prox) = 0
 
-# Erease same line join
+# Erase same line join
 for(id_line in 1:max(line_comp)){
   adj_prox[line_comp == id_line, line_comp == id_line] = 0
 }
@@ -361,7 +361,7 @@ while(!converge_algo){
   to_red_out_edge = as.vector(t(nodefrom_edge_mat[,free_edge_vec]) %*% p_to_red_to)
   to_red_max_edge = pmax(to_red_in_edge, to_red_out_edge)
   to_red = apply(edge_to_sp * to_red_max_edge, 2, max)
-  to_red_mat = matrix(to_red, n, n, byrow=T) * admissible_sp_mat
+  to_red_mat = matrix(to_red, n, n, byrow=T)
   max_to_red = max(to_red)
   
   ## FOR DEBUG
@@ -378,7 +378,7 @@ while(!converge_algo){
   ## FOR DEBUG
   
   # Reduce the dep
-  relationship_ref_mat = relationship_ref_mat * (1 - to_red_mat)
+  relationship_ref_mat = (relationship_ref_mat * (1 - to_red_mat)) * admissible_sp_mat
   
   # --- --- Check for convergence and iterate
   

### 2022-04-05 multiple_lines/transnet_functions.R

@@ -1,0 +1,36 @@
+#-------------------------------------------------------------------------------
+#
+# Functions for transportation lines algorithms 
+#
+#-------------------------------------------------------------------------------
+
+
+#-------------------------------------------------------------------------------
+# Function: build_transport_structure
+#
+# Data_in:
+# Data out:
+# - A_W:  The (n x n) adjacency matrix within transportation lines.
+# - A_B:  The (n x n) adjacency matrix between transportation lines.
+#-------------------------------------------------------------------------------
+
+#-------------------------------------------------------------------------------
+# Function: build_sp_data
+#
+# Data_in:
+# Data out:
+# - T:  The (n x n) matrix of admissible shortest-paths in the network, 
+#       containing (n_sp) ones and (n^2 - n_sp) zeroes.
+# - S:  The (n_sp x m) shortest-path - edges matrix, with s_{ij} = 1 iff edge j
+#       is in shortest-path i.
+#-------------------------------------------------------------------------------
+
+#-------------------------------------------------------------------------------
+# Function: build_in_out_flow 
+#
+# Data_in:
+# Data out:
+# - sigma_in: The n-length vector of flow entering lines.
+# - sigma_out: The n-length vector of flow leaving lines.
+#-------------------------------------------------------------------------------
+

### 2022-04-05 multiple_lines/notes/latex/Notes_trajet.tex
Updated mutlilines notes

@@
 a_{ij} = 0 &\Rightarrow x_{ij} = 0, \qquad \forall i,j, \\ 
-x_{i\bullet} + \sigma^\text{in}_i &= x_{\bullet i} + \sigma^\text{out}_i, \qquad \forall i.
+x_{i\bullet} + \sigma^\text{out}_i &= x_{\bullet i} + \sigma^\text{in}_i, \qquad \forall i.
 \end{align}
@@
 \end{equation}
-This matrix contains the flow on each edge if we suppose that the flow follows shortest-paths.
+This matrix contains the flow on each edge if we suppose that the flow follows shortest-paths. \\
+
+We can rewrite equation (\ref{sp_comp}) by defining the $(n^2 \times n^2)$ \emph{shortest-path - edge matrix} $\mathbf{S} = (s_{\alpha \beta})$ with
+\begin{equation}
+s_{\alpha \beta} = \begin{cases}
+\text{P}_{st}(i, j) & \text{if } \alpha = s + n(t - 1) \text{ and }  \beta = i + n(j - 1), \\
+0 & \text{otherwise}.
+\end{cases}
+\end{equation}
+Then (\ref{sp_comp}) writes
+\begin{equation}
+\
+\end{equation}
 

### 2022-04-05 multiple_lines/transnet_functions.R

@@ -9,9 +9,12 @@
 # Function: build_transport_structure
 #
 # Data_in:
+# - stopnames_df:   a dataframe containing, stop names, line number 
+#                   and direction.
 # Data out:
 # - A_W:  The (n x n) adjacency matrix within transportation lines.
 # - A_B:  The (n x n) adjacency matrix between transportation lines.
+# - stop_names: The n-length vectors containing unique id for each stops.
 #-------------------------------------------------------------------------------
 
 #-------------------------------------------------------------------------------

### 2022-04-05 multiple_lines/Multiligne_6789_bynode_swing.R
Updated notes

@@ -1,0 +1,471 @@
+###############################################################
+# CODE R : multi-line model
+###############################################################
+
+############################################################### 
+####### LIBRARIES
+
+library(igraph)
+library(Matrix)
+
+############################################################### 
+####### Setting paths
+
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+############################################################### 
+####### PARAMETERS
+
+compute_sp_data = F
+epsilon = 1e-40
+conv_thres_if = 1e-5
+conv_thres_algo = 1e-5
+mu_n = 1
+sigma_n = 5
+
+############################################################### 
+####### LOADING AND PREPROCESSING MATRICES
+
+
+# --- Loading adjacency matrices 
+
+# Loading adjacency matrix without loop
+adj_line = read.csv("data_multiligne/adj_next.csv", sep=",", row.names = 1)
+adj_line = as.matrix(adj_line)
+diag(adj_line) = 0
+
+# Loading adjacency matrix with loops on the end of the line
+adj_line_wloop = read.csv("data_multiligne/adj_next_bind.csv", sep=",", row.names = 1)
+adj_line_wloop = as.matrix(adj_line_wloop)
+diag(adj_line_wloop) = 0
+
+# Component for line
+g_adj_wloop = graph_from_adjacency_matrix(adj_line_wloop, mode = "directed")
+line_comp = components(g_adj_wloop)$membership
+
+# Loading proximity matrix
+adj_prox = read.csv("data_multiligne/adj_prox.csv", sep=",", row.names = 1)
+adj_prox = as.matrix(adj_prox)
+diag(adj_prox) = 0
+
+# Erase same line join
+for(id_line in 1:max(line_comp)){
+  adj_prox[line_comp == id_line, line_comp == id_line] = 0
+}
+
+# Saving stop names and renaming columns
+stop_names = trimws(paste0("S", rownames(adj_line)))
+colnames(adj_line) = stop_names
+rownames(adj_line) = stop_names
+colnames(adj_line_wloop) = stop_names
+rownames(adj_line_wloop) = stop_names
+colnames(adj_prox) = stop_names
+rownames(adj_prox) = stop_names
+
+# Number of stops
+n = length(stop_names)
+
+# Correction for the 7th line 
[…]

### 2022-04-07 multiple_lines/notes/latex/Notes_trajet.tex
Updated notes
Notes updated

@@
 \begin{align}
-\bm{\sigma}_\text{in} := \mathbf{N} \mathbf{e}_n \\
-\bm{\sigma}_\text{out} := \mathbf{N}^\top \mathbf{e}_n
+\bm{\sigma}_\text{in} := \mathbf{N} \mathbf{e}_n \label{N_rowsum}\\
+\bm{\sigma}_\text{out} := \mathbf{N}^\top \mathbf{e}_n \label{N_colsum}
 \end{align}
@@
 \end{equation}
+
+Note that if only $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ are given, a flow matrix $\mathbf{N}$ can be computed relatively to an \emph{orgin-destination affinity matrix} $\mathbf{S} = (s_{st})$, $0 \leq s_{st} \leq 1$, where $s_{st} = 1$ denote a perfect affinity and $s_{st} = 0$ no affinity, through
+
+\begin{equation}
+\mathbf{N} = \textbf{Diag}(\mathbf{a}) (\mathbf{S} + \epsilon) \textbf{Diag}(\mathbf{b}),
+\end{equation}
+where $\textbf{Diag}(.)$ denote the diagonal matrix obtained from a vector, $\epsilon$ a very small quantity, and vectors $\mathbf{a}$ and $\mathbf{b}$ are found through \emph{proportional iterative fitting} algorithm in order to have margin constraints (\ref{N_rowsum}) and (\ref{N_colsum}) respected for $\mathbf{N}$ (a small $\epsilon$ has to be added to $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ if they possess null components).
 
@@
 where $\delta(.)$ designate the indicator function.
-If we are given an origin-destination matrix $\mathbf{N} = (n_{st})$, we can compute the \emph{shortest-path flow matrix}, noted $\mathbf{X}_\text{sp} = (x^\text{sp}_{ij})$, with:
+Note that if there is an unique shortest-path between node $s$ and $t$, noted $\wp_{st}$, we have $\text{P}_{st}(i, j) = \delta((i, j) \in \wp_{st})$.
+\\
+
+If we are given an origin-destination matrix $\mathbf{N} = (n_{st})$, we can compute the \emph{shortest-path flow matrix}, noted $\mathbf{X}_\text{sp} = (x^\text{sp}_{ij})$, with
 \begin{equation}
-x^\text{sp}_{ij} = \sum_{st} \text{P}_{st}(i, j) n_{st} \label{sp_comp}
+x^\text{sp}_{ij} = \sum_{st} \text{P}_{st}(i, j) n_{st} \label{sp_comp}.
 \end{equation}
-This matrix contains the flow on each edge if we suppose that the flow follows shortest-paths. \\
+This matrix contains the flow on each edge if we suppose that the flow follows shortest-paths from origin to destination. \\
 
-We can rewrite equation (\ref{sp_comp}) by defining the $(n^2 \times n^2)$ \emph{shortest-path - edge matrix} $\mathbf{S} = (s_{\alpha \beta})$ with
+We can rewrite equation (\ref{sp_comp}) by defining the $(n^2 \times n^2)$ \emph{shortest-path - edge matrix} $\mathbf{P} = (p_{\alpha \beta})$ with
 \begin{equation}
-s_{\alpha \beta} = \begin{cases}
-\text{P}_{st}(i, j) & \text{if } \alpha = s + n(t - 1) \text{ and }  \beta = i + n(j - 1), \\
+p_{\alpha \beta} = \begin{cases}
+\text{P}_{st}(i, j) & \text{if } \alpha = t + n(s - 1) \text{ and }  \beta = j + n(i - 1), \\
 0 & \text{otherwise}.
@@
 \begin{equation}
-\
+\textbf{vec}(\mathbf{X}_\text{sp}) = \mathbf{P}^\top \textbf{vec}(\mathbf{N}),
 \end{equation}
+where $\textbf{vec}(.)$ denotes the vectorization function of a matrix, obtained by stacking matrix columns on top of one another. \\
+
+From equation (\ref{sp_comp}), we see that 
+\begin{equation}
+\frac{\partial x^\text{sp}_{ij}}{\partial n_{st}} = \text{P}_{st}(i, j),
+\end{equation}
+which equal to $1$ if there is a unique shortest-path between $s$ and $t$ and $(i,j)$ belongs to this path.
 
@@
 
-We suppose that we know the flow entering and leaving each line, i.e.
-$\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{out}$ and we are interested in finding trajectories $n_{st}$ of the flow in the network knowing the set $\mathcal{A}$ of admissible pair $(s, t)$, i.e.
-pairs of nodes where there is a possible use of the network for traveling.
-This set can be given by the matrix $\mathbf{T} = (t_{st})$ defined by
-\begin{equation}
-t_{st} = \begin{cases}
-1 & \text{if } (s, t) \in \mathcal{A}, \\
-0 & \text{otherwise}.
-\end{cases}
-\end{equation}
+\paragraph{The problem:} We suppose that we know the flow entering and leaving each line, i.e.
+$\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{out}$ and we want to find origin-destination trajectories $n_{st}$.
+\\
+
+By setting the problem like that, we easily see that it is ill posed.
+Several solutions exists (e.g.
+units remain on the same line and follows a first-in/first-out scheme) and we need to add some hypotheses to restrain it.
+
+\paragraph{Hypothesis 1:} Trajectories follow shortest-paths from origin $s$ to destination $t$. \\
+
+\paragraph{Hypothesis 2:} The number of trajectories $n_{st}$ should be as close as possible to $\rho^\text{in}_s \rho^\text{out}_t s_{st}$, where $S = (s_{st})$ is a given affinity matrix between origin and destination nodes.
+\\
 
 \subsection{Algorithm}
-Set $\bm{\sigma}^{(0)}_\text{in} = \bm{\rho}_\text{in} + \epsilon$ and $\bm{\sigma}^{(0)}_\text{out} = \bm{\rho}_\text{out} + \epsilon$, with $\epsilon$ a small positive scalar. Until convergence, do:
+Set $\bm{\sigma}^{(1)}_\text{in} = \bm{\rho}_\text{in}$, $\bm{\sigma}^{(1)}_\text{out} = \bm{\rho}_\text{out}$, and $\mathbf{S}^{(1)} = S$. Until convergence, do:
 \begin{enumerate}
-	\item Compute $\mathbf{N}^{(i)} = \textbf{Diag}(\mathbf{a}^{(i)}) (\mathbf{T} + \epsilon)\textbf{Diag}(\mathbf{b}^{(i)})$ with iterative fitting, such that $\mathbf{N}^{(i)} \mathbf{e}_n = \bm{\sigma}^{(i)}_\text{in}$ and $\left(\mathbf{N}^{(i)}\right)^\top \mathbf{e}_n = \bm{\sigma}^{(i)}_\text{out}$.
+	\item Compute $\mathbf{N}^{(i)} = \textbf{Diag}(\mathbf{a}^{(i)}) (\mathbf{S}^{(i)} + \epsilon)\textbf{Diag}(\mathbf{b}^{(i)})$ with iterative fitting, such that $\mathbf{N}^{(i)} \mathbf{e}_n = \bm{\sigma}^{(i)}_\text{in} + \epsilon$ and $\left(\mathbf{N}^{(i)}\right)^\top \mathbf{e}_n = \bm{\sigma}^{(i)}_\text{out}  + \epsilon$.
 	\item Compute the associated shortest-path flow matrix $\mathbf{X}^{(i)}$ with (\ref{sp_comp}).
 	\item Compute $\bm{\sigma}^{(i+1)}_\text{in} = \bm{\rho}_\text{in} -  (\mathbf{X}^{(i)}_\text{B})^\top \mathbf{e}_n$ and $\bm{\sigma}^{(i+1)}_\text{out} = \bm{\rho}_\text{out} - \mathbf{X}^{(i)}_\text{B} \mathbf{e}_n$.
+	\item Set negative components of $\bm{\sigma}^{(i+1)}_\text{in}$ and $\bm{\sigma}^{(i+1)}_\text{out}$ to 0.
+	\item Update $\mathbf{S}^{(i + 1)}$ according to excess flow.
 \end{enumerate}

### 2022-04-07 multiple_lines/Multiligne_6789_bynode_swing.R

@@ -20,7 +20,8 @@ compute_sp_data = F
 epsilon = 1e-40
 conv_thres_if = 1e-5
 conv_thres_algo = 1e-5
-max_p_sigma = 0.99
+max_p_sigma = 0.9999
+lambda_sigma = 100
 
 ############################################################### 
 ####### LOADING AND PREPROCESSING MATRICES
@@ -339,14 +340,19 @@ while(!converge_algo){
   ## FOR DEBUG
   
   # Get the flow allowed to pass through each nodes 
-  # allowed_from_free_flow = tanh(from_free_flow / rho_in) * rho_in
-  # allowed_to_free_flow = tanh(to_free_flow / rho_out) * rho_out
+  
+  # allowed_from_free_flow = rho_in * (1 - exp(-lambda_sigma * from_free_flow / rho_in))
+  # allowed_to_free_flow = rho_out * (1 - exp(-lambda_sigma * to_free_flow / rho_out))
+  # allowed_from_free_flow[allowed_from_free_flow > from_free_flow] = from_free_flow[allowed_from_free_flow > from_free_flow]
+  # allowed_to_free_flow[allowed_to_free_flow > to_free_flow] = to_free_flow[allowed_to_free_flow > to_free_flow]
+  
   allowed_from_free_flow = from_free_flow
   allowed_to_free_flow = to_free_flow
-  allowed_from_free_flow[allowed_from_free_flow > rho_in*max_p_sigma] = 
+  allowed_from_free_flow[allowed_from_free_flow > rho_in*max_p_sigma] =
     rho_in[allowed_from_free_flow > rho_in*max_p_sigma]*max_p_sigma
-  allowed_to_free_flow[allowed_to_free_flow > rho_out*max_p_sigma] = 
+  allowed_to_free_flow[allowed_to_free_flow > rho_out*max_p_sigma] =
     rho_out[allowed_to_free_flow > rho_out*max_p_sigma]*max_p_sigma
+  
   p_allowed_from = allowed_from_free_flow / (from_free_flow + epsilon)
   p_allowed_to = allowed_to_free_flow / (to_free_flow + epsilon)
   p_allowed_from[p_allowed_from == 0] = 1
@@ -457,10 +463,12 @@ for(id_node in 1:n){
 full_df = inout_cor
 full_df["montees_initiales"] = round(sigma_in * to_num, 3)
 full_df["transferts_in"] = round(transfer_in * to_num, 3)
[…]

### 2022-04-07 data_processing/speed_matrix.R
Data/result too large -> switchdrive

@@ -1,0 +1,449 @@
+############################################################### 
+####### Setting paths
+
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+# setwd("/Users/rloup/Desktop")
+
+# Command lines to cut a pbf file -->
+#   osmconvert /Users/rloup/Desktop/switzerland-latest.osm.pbf -b=6.48008,46.48580,6.89584,46.72089 --complete-ways -o=lausanne.pbf
+
+####### Libraries
+library("opentripplanner")
+library(tidyverse) # Data filter
+
+
+# path_data <- file.path(tempdir(), "OTP")
+# dir.create(path_data)
+
+path_data <- file.path("/otp")
+# dir.create(path_data)
+
+path_otp <- file.path("/otp/otp-2.0.0-shaded.jar")
+
+# path_otp <- otp_dl_jar(path_data, cache = FALSE)
+
+
+# Demo
+otp_dl_demo(path_data)
+log1 <- otp_build_graph(otp = path_otp, dir = path_data)
+log2 <- otp_setup(otp = path_otp, dir = path_data)
+
+log1 <- otp_build_graph(otp = "/Users/rloup/Documents/r_projects/tl_study/data_processing/otp/otp-2.0.0-shaded.jar", dir = "/Users/rloup/Documents/r_projects/tl_study/data_processing/otp", memory = 30000)
+
+
+log2 <- otp_setup(otp = "/Users/rloup/Documents/r_projects/tl_study/data_processing/otp/otp-2.0.0-shaded.jar", dir = "/Users/rloup/Documents/r_projects/tl_study/data_processing/otp", memory = 16000, port = 8080)
+# log2 <- otp_setup(otp = "otp-2.0.0-shaded.jar", dir = "/Users/rloup/otp2", memory = 30000, port = 8080, router = "default")
+
+
+otpcon <- otp_connect(hostname =  "localhost",
+                      router = "default",
+                      port = 8080)
+
+otp_plan(otpcon,
+         fromPlace = c(6.62801,46.52025),
+         toPlace = c(6.633971156608949, 46.522829744515576),
+         mode = c("WALK"),
+         # maxWalkDistance = 1000,
+         # date_time = as.POSIXct(strptime("2019-11-14 13:02", "%Y-%m-%d %H:%M")),
+         numItineraries = 1,
+         # fromID = "hello",
+         # toID = "ciao",
+         # maxWalkDistance = 5
[…]

### 2022-04-11 multiple_lines/notes/latex/Notes_trajet.tex
Notes updated
Multi-lines notes updated

@@
 
-Note that if only $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ are given, a flow matrix $\mathbf{N}$ can be computed relatively to an \emph{orgin-destination affinity matrix} $\mathbf{S} = (s_{st})$, $0 \leq s_{st} \leq 1$, where $s_{st} = 1$ denote a perfect affinity and $s_{st} = 0$ no affinity, through
+Note that if only $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ are given, a flow matrix $\mathbf{N}$ can be computed relatively to an \emph{origin-destination affinity matrix} $\mathbf{S} = (s_{st})$, $0 \leq s_{st} \leq 1$, where $s_{st} = 1$ denote a perfect affinity and $s_{st} = 0$ no affinity, through
 
@@
 which equal to $1$ if there is a unique shortest-path between $s$ and $t$ and $(i,j)$ belongs to this path.
+This equation means that if we multiply the flow by a factor $\alpha \geq 0$ on each $s, t$ which contains $(i, j)$ on their shortest-paths, the resulting flow on $(i, j)$ will also be multiplied by $\alpha$.
 
@@
 Several solutions exists (e.g.
-units remain on the same line and follows a first-in/first-out scheme) and we need to add some hypotheses to restrain it.
+units remain on the same line and follow a first-in/first-out scheme) and we need to add some hypotheses to restrain it.
 
-\paragraph{Hypothesis 1:} Trajectories follow shortest-paths from origin $s$ to destination $t$. \\
+\paragraph{Hypothesis 1:} Trajectories in the network follow shortest-paths from origin $s$ to destination $t$. \\
 
-\paragraph{Hypothesis 2:} The number of trajectories $n_{st}$ should be as close as possible to $\rho^\text{in}_s \rho^\text{out}_t s_{st}$, where $S = (s_{st})$ is a given affinity matrix between origin and destination nodes.
+\paragraph{Hypothesis 2:} The number of trajectories $\mathbf{N} = (n_{st})$ should be as close as possible to $s_{st}$, where $S = (s_{st})$ is a given affinity matrix between origin and destination nodes, in the sens that
+\begin{equation}
+	K(\mathbf{N} | \mathbf{S}) := \sum_{st} \frac{n_{st}}{n_{\bullet \bullet}} \log \left( \frac{n_{st}/n_{\bullet \bullet}}{s_{st} / s_{\bullet \bullet}} \right), \label{kl_div}
+\end{equation}
+i.e.
+the \emph{Kullback-Leibler divergence} between the probability of selecting an origin-destination path according to $\mathbf{N}$ relatively to the probabilty of selecting an origin-destination path according to $\mathbf{S}$, is minimum.
+Note that the divergence (\ref{kl_div}) is well defined only if $s_{st} > 0, \; \forall s, t$.
 \\
 
+With these two additional hypotheses, we can find a solution with the following algorithm.
+
 \subsection{Algorithm}
-Set $\bm{\sigma}^{(1)}_\text{in} = \bm{\rho}_\text{in}$, $\bm{\sigma}^{(1)}_\text{out} = \bm{\rho}_\text{out}$, and $\mathbf{S}^{(1)} = S$. Until convergence, do:
+Set $\bm{\sigma}^{(1)}_\text{in} = \bm{\rho}_\text{in}$, $\bm{\sigma}^{(1)}_\text{out} = \bm{\rho}_\text{out}$, and $\mathbf{S}^{(1)} = \mathbf{S}$. Until convergence, do:
 \begin{enumerate}
-	\item Compute $\mathbf{N}^{(i)} = \textbf{Diag}(\mathbf{a}^{(i)}) (\mathbf{S}^{(i)} + \epsilon)\textbf{Diag}(\mathbf{b}^{(i)})$ with iterative fitting, such that $\mathbf{N}^{(i)} \mathbf{e}_n = \bm{\sigma}^{(i)}_\text{in} + \epsilon$ and $\left(\mathbf{N}^{(i)}\right)^\top \mathbf{e}_n = \bm{\sigma}^{(i)}_\text{out}  + \epsilon$.
-	\item Compute the associated shortest-path flow matrix $\mathbf{X}^{(i)}$ with (\ref{sp_comp}).
-	\item Compute $\bm{\sigma}^{(i+1)}_\text{in} = \bm{\rho}_\text{in} -  (\mathbf{X}^{(i)}_\text{B})^\top \mathbf{e}_n$ and $\bm{\sigma}^{(i+1)}_\text{out} = \bm{\rho}_\text{out} - \mathbf{X}^{(i)}_\text{B} \mathbf{e}_n$.
-	\item Set negative components of $\bm{\sigma}^{(i+1)}_\text{in}$ and $\bm{\sigma}^{(i+1)}_\text{out}$ to 0.
-	\item Update $\mathbf{S}^{(i + 1)}$ according to excess flow.
+	\item Compute 
+	\begin{equation}
+		\mathbf{N}^{(\tau)} = \textbf{Diag}(\mathbf{a}^{(\tau)}) (\mathbf{S}^{(\tau)} + \epsilon)\textbf{Diag}(\mathbf{b}^{(\tau)}),
+	\end{equation}
+	with iterative fitting, such that $\mathbf{N}^{(\tau)} \mathbf{e}_n = \bm{\sigma}^{(\tau)}_\text{in} + \epsilon$ and $\left(\mathbf{N}^{(\tau)}\right)^\top \mathbf{e}_n = \bm{\sigma}^{(\tau)}_\text{out}  + \epsilon$.
+
+	\item Compute the associated shortest-path flow matrix $\mathbf{X}^{(\tau)}$ with (\ref{sp_comp}).
+
+	\item Compute the vectors of \emph{between-lines flow entering and leaving each nodes}, i.e.
+	\begin{align}
+		\mathbf{x}^{(\tau)}_\text{B,in} &= (\mathbf{X}^{(\tau)}_\text{B})^\top \mathbf{e}_n \\
+		\mathbf{x}^{(\tau)}_\text{B,out} &= \mathbf{X}^{(\tau)}_\text{B} \mathbf{e}_n
+	\end{align}
+
+	\item Compute the vectors of \emph{between-line allowed flow entering and leaving each nodes}, written resp.
+$\tilde{\mathbf{x}}^{(\tau)}_\text{B,in} = (\tilde{x}^{\text{B,in},(\tau)}_i)$ and $\tilde{\mathbf{x}}^{(\tau)}_\text{B,out} = (\tilde{x}^{\text{B,out},(\tau)}_i)$, with
+	\begin{align}
+		\tilde{x}^{\text{B,in},(\tau)}_i &= \rho_i^\text{in} \phi \left( \frac{x^{\text{B,in},(\tau)}_i}{\rho_i^\text{in}} \right), \\
+		\tilde{x}^{\text{B,in},(\tau)}_i &= \rho_i^\text{out} \phi \left( \frac{x^{\text{B,out},(\tau)}_i}{\rho_i^\text{out}} \right),
+	\end{align}
+	where $\phi(x)$ is a positive increasing function which should be the identity when $x \to 0$ and with $\phi(x) \leq 1$, $\forall x$. For example
+	\begin{enumerate}
+		\item $\phi(x) = \min(x, 1)$,
+		\item $\phi(x) = \min(x, 1 - \exp(-\lambda x))$, with $\lambda > 0$ a parameter.
+	\end{enumerate}
+
+	\item Compute the \emph{between-lines allowed flow on edges}, noted $\widetilde{\mathbf{X}}_\text{B}^{(\tau)} = (\widetilde{x}_{ij}^{\text{B},(\tau)})$ with 
+	\begin{equation}
+		\widetilde{x}_{ij}^{\text{B},(\tau)} = \min \left( \frac{\tilde{x}^{\text{B,out},(\tau)}_i}{x^{\text{B,out},(\tau)}_i} , \frac{\tilde{x}^{\text{B,in},(\tau)}_j}{x^{\text{B,in},(\tau)}_j} \right)  x^{\text{B},(\tau)}_{ij}.
+	\end{equation}
+	\item Update the flow entering and leaving the network with
+		\begin{align}
+			\bm{\sigma}^{(\tau+1)}_\text{in} &= \bm{\rho}_\text{in} -  (\widetilde{\mathbf{X}}^{(\tau)}_\text{B})^\top \mathbf{e}_n, \\
+			\bm{\sigma}^{(\tau+1)}_\text{out} &= \bm{\rho}_\text{out} -  \widetilde{\mathbf{X}}^{(\tau)}_\text{B} \mathbf{e}_n.
+		\end{align}
+	\item Compute the \emph{reducing factor matrix} $\mathbf{R}^{(\tau)} = (r^{(\tau)}_{st})$ with 
+	\begin{equation}
+		r^{(\tau)}_{st} = 1 - \max_{ij} \left( \text{P}_{st}(i, j) \frac{x_{ij}^{\text{B},(\tau)}  - \widetilde{x}_{ij}^{\text{B},(\tau)}}{x_{ij}^{\text{B},(\tau)} + \epsilon} \right).
+	\end{equation}
+	\item Update the origin-destination affinity matrix with
+	\begin{equation}
+		\mathbf{S}^{(\tau + 1)} = \mathbf{S}^{(\tau)} \odot \mathbf{R}^{(\tau)},
+	\end{equation}
+	where $\odot$ designate the Hadamard (component-wise) product of matrices. 
 \end{enumerate}

### 2022-04-12 multiple_lines/notes/latex/Notes_trajet.tex
Notes updated, gitignore too

@@
 Each node belongs to only one line, i.e.
-$\mathcal{V} = \bigcup_{k=1}^p \mathcal{V}_k$ and $\bigcap_{k=1}^p \mathcal{V}_k = \emptyset$, where $\mathcal{V}_k$ represents the set of nodes in line $k$.
+$\mathcal{V} = \bigcup_{k=1}^p \mathcal{V}_k$ and $\mathcal{V}_k \cap \mathcal{V}_l = \emptyset, \, \forall k,l$, where $\mathcal{V}_k$ represents the set of nodes in line $k$.
 The edge set $\mathcal{E}$, can also be decomposed with
@@
 \begin{equation}
-\mathbf{X} = \mathbf{X}_\text{W} + \mathbf{X}_\text{B} \qquad \mathbf{X}_\text{W} := \sum_{k=1}^l \mathbf{X}_k
+\mathbf{X} = \mathbf{X}_\text{W} + \mathbf{X}_\text{B} \qquad \mathbf{X}_\text{W} := \sum_{k=1}^p \mathbf{X}_k
 \end{equation}
@@
 where $\delta(.)$ designate the indicator function.
-Note that if there is an unique shortest-path between node $s$ and $t$, noted $\wp_{st}$, we have $\text{P}_{st}(i, j) = \delta((i, j) \in \wp_{st})$.
+Note that if there is an unique shortest-path between node $s$ and $t$, noted $\wp_{st}$, we have $\text{P}_{st}(i, j) = 1$ if $(i, j) \in \wp_{st}$,  $\text{P}_{st}(i, j) = 0$ otherwise.
 \\
@@
 which equal to $1$ if there is a unique shortest-path between $s$ and $t$ and $(i,j)$ belongs to this path.
-This equation means that if we multiply the flow by a factor $\alpha \geq 0$ on each $s, t$ which contains $(i, j)$ on their shortest-paths, the resulting flow on $(i, j)$ will also be multiplied by $\alpha$.
+This equation means that if we multiply $n_{st}$ by a factor $\alpha \geq 0$ on each $s, t$ which contains $(i, j)$ on their shortest-paths, the resulting flow on $(i, j)$ will also be multiplied by $\alpha$.
 
@@
 \paragraph{The problem:} We suppose that we know the flow entering and leaving each line, i.e.
-$\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{out}$ and we want to find origin-destination trajectories $n_{st}$.
+$\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{out}$, and we want to find origin-destination trajectories $n_{st}$.
 \\
@@
 By setting the problem like that, we easily see that it is ill posed.
-Several solutions exists (e.g.
-units remain on the same line and follow a first-in/first-out scheme) and we need to add some hypotheses to restrain it.
+Several solutions exists, with some of them trival (e.g.
+units remain on the same line and follow a first-in/first-out scheme), and we need to add some hypotheses to restrain it.
 
@@
 
-\paragraph{Hypothesis 2:} The number of trajectories $\mathbf{N} = (n_{st})$ should be as close as possible to $s_{st}$, where $S = (s_{st})$ is a given affinity matrix between origin and destination nodes, in the sens that
+\paragraph{Hypothesis 2:} The number of trajectories $\mathbf{N} = (n_{st})$ should be as close as possible to $s_{st}$, where $\mathbf{S} = (s_{st})$ is a given affinity matrix between origin and destination nodes, in the sens that
 \begin{equation}
@@
 	\end{equation}
-	with iterative fitting, such that $\mathbf{N}^{(\tau)} \mathbf{e}_n = \bm{\sigma}^{(\tau)}_\text{in} + \epsilon$ and $\left(\mathbf{N}^{(\tau)}\right)^\top \mathbf{e}_n = \bm{\sigma}^{(\tau)}_\text{out}  + \epsilon$.
+	with proportional iterative fitting, such that $\mathbf{N}^{(\tau)} \mathbf{e}_n = \bm{\sigma}^{(\tau)}_\text{in} + \epsilon$ and $\left(\mathbf{N}^{(\tau)}\right)^\top \mathbf{e}_n = \bm{\sigma}^{(\tau)}_\text{out}  + \epsilon$.
 
-	\item Compute the associated shortest-path flow matrix $\mathbf{X}^{(\tau)}$ with (\ref{sp_comp}).
+	\item Compute the associated shortest-path flow matrix $\mathbf{X}^{(\tau)}$ with
+	\begin{equation}
+		\textbf{vec}(\mathbf{X}^{(\tau)}) = \mathbf{P}^\top \textbf{vec}(\mathbf{N}^{(\tau)}).
+	\end{equation}
 
@@
 	\begin{equation}
-		\widetilde{x}_{ij}^{\text{B},(\tau)} = \min \left( \frac{\tilde{x}^{\text{B,out},(\tau)}_i}{x^{\text{B,out},(\tau)}_i} , \frac{\tilde{x}^{\text{B,in},(\tau)}_j}{x^{\text{B,in},(\tau)}_j} \right)  x^{\text{B},(\tau)}_{ij}.
+		\widetilde{x}_{ij}^{\text{B},(\tau)} = \begin{cases}
+			\min \left(\frac{\tilde{x}^{\text{B,out},(\tau)}_i}{x^{\text{B,out},(\tau)}_i} , \frac{\tilde{x}^{\text{B,in},(\tau)}_j}{x^{\text{B,in},(\tau)}_j} \right)  x^{\text{B},(\tau)}_{ij} & \text{if } x^{\text{B},(\tau)}_{ij} > 0\\
+			0 & \text{otherwise}
+			\end{cases}
 	\end{equation}

### 2022-04-12 multiple_lines/Multiligne_6789_bynode_swing.R
Function build_network_structure is written

@@ -341,17 +341,17 @@ while(!converge_algo){
   
   # Get the flow allowed to pass through each nodes 
   
-  # allowed_from_free_flow = rho_in * (1 - exp(-lambda_sigma * from_free_flow / rho_in))
-  # allowed_to_free_flow = rho_out * (1 - exp(-lambda_sigma * to_free_flow / rho_out))
-  # allowed_from_free_flow[allowed_from_free_flow > from_free_flow] = from_free_flow[allowed_from_free_flow > from_free_flow]
-  # allowed_to_free_flow[allowed_to_free_flow > to_free_flow] = to_free_flow[allowed_to_free_flow > to_free_flow]
+  allowed_from_free_flow = rho_in * (1 - exp(-lambda_sigma * from_free_flow / rho_in))
+  allowed_to_free_flow = rho_out * (1 - exp(-lambda_sigma * to_free_flow / rho_out))
+  allowed_from_free_flow[allowed_from_free_flow > from_free_flow] = from_free_flow[allowed_from_free_flow > from_free_flow]
+  allowed_to_free_flow[allowed_to_free_flow > to_free_flow] = to_free_flow[allowed_to_free_flow > to_free_flow]
   
-  allowed_from_free_flow = from_free_flow
-  allowed_to_free_flow = to_free_flow
-  allowed_from_free_flow[allowed_from_free_flow > rho_in*max_p_sigma] =
-    rho_in[allowed_from_free_flow > rho_in*max_p_sigma]*max_p_sigma
-  allowed_to_free_flow[allowed_to_free_flow > rho_out*max_p_sigma] =
-    rho_out[allowed_to_free_flow > rho_out*max_p_sigma]*max_p_sigma
+  # allowed_from_free_flow = from_free_flow
+  # allowed_to_free_flow = to_free_flow
+  # allowed_from_free_flow[allowed_from_free_flow > rho_in*max_p_sigma] =
+  #   rho_in[allowed_from_free_flow > rho_in*max_p_sigma]*max_p_sigma
+  # allowed_to_free_flow[allowed_to_free_flow > rho_out*max_p_sigma] =
+  #   rho_out[allowed_to_free_flow > rho_out*max_p_sigma]*max_p_sigma
   
   p_allowed_from = allowed_from_free_flow / (from_free_flow + epsilon)
   p_allowed_to = allowed_to_free_flow / (to_free_flow + epsilon)

### 2022-04-12 multiple_lines/transnet_functions.R

@@ -21,6 +21,34 @@
 
 build_network_structure = function(line_mbr, tour_mbr, D, thres_D){
   
+  # Get the number of stops 
+  n = length(line_mbr)
+  # Get levels of lines
+  lines_lvl = as.factor(line_mbr)
+  # Get levels of tour
+  tour_lvl = as.factor(tour_mbr)
+  
+  # --- Build A_W
+  
+  # Make links to the next stop
+  A_W = matrix(0, n, n)
+  for(i in 1:(n-1)){
+    if(lines_lvl[i] == lines_lvl[i + 1]) A_W[i, i + 1] = 1
+  }
+  
+  # --- Build A_B
+  
+  # With distance threshold
+  A_B = 1*(D < thres_D)
+  # Removing diagonal
+  diag(A_B) = 0
+  # Removing same tour links
+  for(t_lvl in levels(tour_lvl)){
+    A_B[tour_lvl == t_lvl, tour_lvl == t_lvl] = 0
+  }
+  
+  # Return the results
+  return(list(A_W=A_W, A_B=A_B))
 }
 
 #-------------------------------------------------------------------------------

### 2022-04-12 multiple_lines/1_build_multilines_data.R
Restructured files

@@ -1,0 +1,30 @@
+#-------------------------------------------------------------------------------
+#
+# Preprocess data and save it
+#
+#-------------------------------------------------------------------------------
+
+#--------------------------------
+# Head
+#--------------------------------
+
+# Set working directory path
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Load functions
+source("local_functions.R")
+
+#--------------------------------
+# Parameters
+#--------------------------------
+
+# The file containing the dataset
+line_data_file = ""
+# The file containing the pedestrian distance between stops
+d_file = ""
+
+
+#--------------------------------
+# Process
+#--------------------------------
+

### 2022-04-12 multiple_lines/Multiligne_6789_bynode.R

@@ -20,7 +20,8 @@ compute_sp_data = F
 epsilon = 1e-40
 conv_thres_if = 1e-5
 conv_thres_algo = 1e-5
-min_sigma = 0.0001
+max_p_sigma = 0.9999
+lambda_sigma = 100
 
 ############################################################### 
 ####### LOADING AND PREPROCESSING MATRICES
@@ -274,8 +275,8 @@ rho_out = inout_cor$descentes + epsilon
 rho_out = rho_out / sum(rho_out)
 
 
-
-
+############################################################### 
+############################################################### 
 ############################################################### 
 ####### Algorithm 
 
@@ -321,7 +322,7 @@ while(!converge_algo){
   # Find flow on edges and free edges
   total_edge_flow = as.vector(t(sp_edge_mat) %*% c(t(results_mat)))
   free_edge_flow = total_edge_flow[free_edge_vec]
-    
+  
   # --- --- In and out-flow correction on nodes
   
   # The flow on free edges
@@ -338,29 +339,43 @@ while(!converge_algo){
   # print(df_edge)
   ## FOR DEBUG
   
+  # Get the flow allowed to pass through each nodes 
+  
+  allowed_from_free_flow = rho_in * (1 - exp(-lambda_sigma * from_free_flow / rho_in))
+  allowed_to_free_flow = rho_out * (1 - exp(-lambda_sigma * to_free_flow / rho_out))
+  allowed_from_free_flow[allowed_from_free_flow > from_free_flow] = from_free_flow[allowed_from_free_flow > from_free_flow]
+  allowed_to_free_flow[allowed_to_free_flow > to_free_flow] = to_free_flow[allowed_to_free_flow > to_free_flow]
+  
+  # allowed_from_free_flow = from_free_flow
+  # allowed_to_free_flow = to_free_flow
+  # allowed_from_free_flow[allowed_from_free_flow > rho_in*max_p_sigma] =
+  #   rho_in[allowed_from_free_flow > rho_in*max_p_sigma]*max_p_sigma
+  # allowed_to_free_flow[allowed_to_free_flow > rho_out*max_p_sigma] =
+  #   rho_out[allowed_to_free_flow > rho_out*max_p_sigma]*max_p_sigma
+  
[…]

### 2022-04-12 multiple_lines/old_code/Multiligne_6789_bynode_brute.R

@@ -20,8 +20,7 @@ compute_sp_data = F
 epsilon = 1e-40
 conv_thres_if = 1e-5
 conv_thres_algo = 1e-5
-max_p_sigma = 0.9999
-lambda_sigma = 100
+min_sigma = 0.0001
 
 ############################################################### 
 ####### LOADING AND PREPROCESSING MATRICES
@@ -275,8 +274,8 @@ rho_out = inout_cor$descentes + epsilon
 rho_out = rho_out / sum(rho_out)
 
 
-############################################################### 
-############################################################### 
+
+
 ############################################################### 
 ####### Algorithm 
 
@@ -322,7 +321,7 @@ while(!converge_algo){
   # Find flow on edges and free edges
   total_edge_flow = as.vector(t(sp_edge_mat) %*% c(t(results_mat)))
   free_edge_flow = total_edge_flow[free_edge_vec]
-  
+    
   # --- --- In and out-flow correction on nodes
   
   # The flow on free edges
@@ -339,43 +338,29 @@ while(!converge_algo){
   # print(df_edge)
   ## FOR DEBUG
   
-  # Get the flow allowed to pass through each nodes 
-  
-  allowed_from_free_flow = rho_in * (1 - exp(-lambda_sigma * from_free_flow / rho_in))
-  allowed_to_free_flow = rho_out * (1 - exp(-lambda_sigma * to_free_flow / rho_out))
-  allowed_from_free_flow[allowed_from_free_flow > from_free_flow] = from_free_flow[allowed_from_free_flow > from_free_flow]
-  allowed_to_free_flow[allowed_to_free_flow > to_free_flow] = to_free_flow[allowed_to_free_flow > to_free_flow]
-  
-  # allowed_from_free_flow = from_free_flow
-  # allowed_to_free_flow = to_free_flow
-  # allowed_from_free_flow[allowed_from_free_flow > rho_in*max_p_sigma] =
-  #   rho_in[allowed_from_free_flow > rho_in*max_p_sigma]*max_p_sigma
-  # allowed_to_free_flow[allowed_to_free_flow > rho_out*max_p_sigma] =
-  #   rho_out[allowed_to_free_flow > rho_out*max_p_sigma]*max_p_sigma
-  
[…]

### 2022-04-12 multiple_lines/data_multiligne/data_old_format/convert_data.R
Script to change data in new format written. Data converted for 6789

@@ -1,0 +1,80 @@
+#-------------------------------------------------------------------------------
+#
+# Convert old format into new format
+#
+#-------------------------------------------------------------------------------
+
+#---------------------------
+# Set working directory path
+#---------------------------
+
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+#---------------------------
+# Loading the datasets 
+#---------------------------
+
+# --- Passengers in and out df
+
+# Loading in/out data
+inout_data = read.csv("stops_frequentation_6789.csv", sep=",", row.names = 1)
+stop_names = trimws(paste(paste0("S", inout_data$code_ligne_theo), 
+                          inout_data$direction_voy_theo, 
+                          inout_data$code_arret_theo, sep="_"))
+
+# --- Mean stopping time df
+
+# Mean stop time at bus stops
+stop_time_df = read.csv("waiting_time_s_tp.csv", sep=",")
+# Get the name of stop
+stop_time_rownames = trimws(paste(paste0("S", stop_time_df$code_ligne_theo), 
+                                  stop_time_df$direction_voy_theo, 
+                                  stop_time_df$code_arret_theo, sep="_"))
+# Get the mean stopping time 
+stop_time_vec = c()
+for (i in 1:length(stop_names)){
+  stop_time = stop_time_df[which(stop_time_rownames == stop_names[i]),
+                           ]$mean_waiting_time
+  stop_time_vec = c(stop_time_vec, stop_time)
+}
+
+# --- Mean travelling time df
+
+# Bus time between stops
+travel_time_df = read.csv("mean_time_i-j_s_tp.csv", sep=",")
+# Get the name of starting stop
+travel_time_rownames = trimws(paste(paste0("S", travel_time_df$code_ligne_theo), 
+                                    travel_time_df$direction_voy_theo, 
+                                    travel_time_df$code_arret_prec_theo, sep="_"))
+# Get the mean travel time 
+travel_time_vec = c()
+for (i in 1:length(stop_names)){
+  travel_time = travel_time_df[which(travel_time_rownames == stop_names[i])[1],
+                               ]$mean_time
+  if(length(travel_time) == 0) travel_time = NA
+  travel_time_vec = c(travel_time_vec, travel_time)
+}
+
+# --- Pedestrian time matrix
+
+# Pedestrian time martix 
+ped_time_mat = read.csv("time_mat_s_6789_names.csv", sep=",")
+ped_time_mat = as.matrix(ped_time_mat[, -1])
[…]

### 2022-04-12 multiple_lines/data_multiligne/data_old_format/convert_data.R
Corrected a bug in data transformation

@@ -76,5 +76,5 @@ bus_df = data.frame(stop_names=stop_names,
                     travel_time=travel_time_vec,
                     stopping_time=stop_time_vec)
 
-write.table(bus_df, "bus_df.csv", sep=",")
+write.table(bus_df, "bus_df.csv", sep=",", row.names = F)
 write.table(ped_time_mat, "ped_time.csv", sep=",", row.names=F, col.names=F)

### 2022-04-20 multiple_lines/local_functions.R
Speed optimized, bug fixed, timer removed.

@@ -308,15 +308,6 @@ compute_origin_destination = function(rho_in, rho_out, edge_ref, sp_ref, p_mat,
                                       conv_thres_algo=1e-5, epsilon=1e-40,
                                       max_it=200){
   
-  # --- Timers
-  
-  time_if = c()
-  time_betw_flow = c()
-  time_allowed_flow_node = c()
-  time_allowed_flow_edge = c()
-  time_flow_update = c()
-  time_affinity_update = c()
-  
   # --- Get the network structure 
   
   # The number of stops 
@@ -327,21 +318,15 @@ compute_origin_destination = function(rho_in, rho_out, edge_ref, sp_ref, p_mat,
   edge_btw_ref = edge_ref[where_edge_btw, 1:2]
   # The shortest-path - edge matrix retrained on transfer edges
   p_btw_mat = p_mat[, where_edge_btw]
-  # The matrix of admissible sp 
+  # The shortest-path order reference
+  sp_order_ref = mapply(function(i, j) i+n*(j-1), sp_ref[, 1], sp_ref[, 2])
   
-  ######################
-  # POTENTIAL ERROR
-  ######################
-  where_adm_sp_mat = as.matrix(sparseMatrix(i=sp_ref[, 1], j=sp_ref[, 2], 
-                                        dims=c(n, n)))
-  where_adm_sp = c(where_adm_sp_mat)
   # If s_mat is not given, build it
   if(is.null(s_mat)){
-    s_mat = 1*where_adm_sp_mat
+    # The matrix of admissible sp 
+    s_mat = 1*as.matrix(sparseMatrix(i=sp_ref[, 1], j=sp_ref[, 2], 
+                                     dims=c(n, n)))
   }
-  ######################
-  # POTENTIAL ERROR
-  ######################
   
   # --- Algorithm 
   
@@ -365,8 +350,6 @@ compute_origin_destination = function(rho_in, rho_out, edge_ref, sp_ref, p_mat,
     
     # --- Iterative fitting 
     
-    t1 = Sys.time()
-    
     n_mat = s_it_mat + epsilon
     converge_if = F
     while(!converge_if){
@@ -382,21 +365,10 @@ compute_origin_destination = function(rho_in, rho_out, edge_ref, sp_ref, p_mat,
       }
     }
     
-    t2 = Sys.time()
-    time_if = c(time_if, as.numeric(t2-t1))
-    
     # --- Find between flow for edges and nodes
     
[…]

### 2022-04-20 multiple_lines/multilines_6789_bynode.R

@@ -295,13 +295,6 @@ relationship_ref_mat = admissible_sp_mat
 total_rel = sum(admissible_sp_mat)
 max_to_red = 1
 
-time_if = c()
-time_betw_flow = c()
-time_allowed_flow_node = c()
-time_allowed_flow_edge = c()
-time_flow_update = c()
-time_affinity_update = c()
-
 while(!converge_algo & it_algo <= 30){
   
   # --- --- Save old X_b
@@ -310,8 +303,6 @@ while(!converge_algo & it_algo <= 30){
   max_to_red_old = max_to_red
   
   # --- --- Iterative fitting 
-  
-  t1 = Sys.time()
   
   converge_if = F
   results_mat = relationship_ref_mat + epsilon
@@ -328,11 +319,6 @@ while(!converge_algo & it_algo <= 30){
     }
   }
   
-  t2 = Sys.time()
-  time_if = c(time_if, as.numeric(t2-t1))
-  
-  t1 = Sys.time()
-
   # Find flow on edges and free edges
   total_edge_flow = as.vector(t(sp_edge_mat) %*% c(t(results_mat)))
   free_edge_flow = total_edge_flow[free_edge_vec]
@@ -347,9 +333,6 @@ while(!converge_algo & it_algo <= 30){
   from_free_flow = colSums(X_b)
   to_free_flow = rowSums(X_b)
   
-  t2 = Sys.time()
-  time_betw_flow = c(time_betw_flow, as.numeric(t2-t1))
-  
   ## FOR DEBUG
   # df_edge = data.frame(free_edge_loc[, 1],free_edge_loc[, 2],free_edge_flow)
   # df_edge = df_edge[df_edge$free_edge_flow > 0, ]
@@ -357,8 +340,6 @@ while(!converge_algo & it_algo <= 30){
   ## FOR DEBUG
   
   # Get the flow allowed to pass through each nodes 
-  
-  t1 = Sys.time()
   
   allowed_from_free_flow = rho_in * (1 - exp(-lambda_sigma * from_free_flow / rho_in))
   allowed_to_free_flow = rho_out * (1 - exp(-lambda_sigma * to_free_flow / rho_out))
@@ -372,11 +353,6 @@ while(!converge_algo & it_algo <= 30){
   # allowed_to_free_flow[allowed_to_free_flow > rho_out*max_p_sigma] =
   #   rho_out[allowed_to_free_flow > rho_out*max_p_sigma]*max_p_sigma
   
-  t2 = Sys.time()
[…]

### 2022-04-20 multiple_lines/2_compute_OD_matrix.R
Added a tolerance parameter for the reduction, to make values close to zero null

@@ -22,19 +22,19 @@ source("local_functions.R")
 data_folder = "multilines_data/preprocessed_data/test_6789"
 
 # Conv threshold for iterative fitting
-conv_thres_if = 100
+conv_thres_if = 1
 # Conv threshold
-conv_thres_algo = 1e-5
+conv_thres_algo = 10
 # Smooth or strict limit
 smooth_limit = F
 # lambda for exponential law
-exp_lambda = 100
+exp_lambda = 1.5
 # prop limit 
-prop_limit = 0.2
+prop_limit = 1 - 0.582812
 # epsilon
 epsilon = 1e-40
 # max iteration
-max_it = 30
+max_it = 5000
 
 #--------------------------------
 # Process

### 2022-04-20 multiple_lines/local_functions.R

@@ -69,7 +69,6 @@ build_network_structure = function(line_mbr, tour_mbr, dist_mat, dist_thres){
 }
 
 
-
 #-------------------------------------------------------------------------------
 # Function: build_sp_data()
 #
@@ -190,7 +189,6 @@ build_sp_data = function(line_mbr, tour_mbr, travel_t, wait_t, dist_mat,
 }
 
 
-
 #-------------------------------------------------------------------------------
 # Function: build_in_out_flow 
 #
@@ -306,7 +304,7 @@ compute_origin_destination = function(rho_in, rho_out, edge_ref, sp_ref, p_mat,
                                       s_mat=NULL, smooth_limit=T, exp_lambda=10,
                                       prop_limit=0.1, conv_thres_if=1e-5,
                                       conv_thres_algo=1e-5, epsilon=1e-40,
-                                      max_it=200){
+                                      max_it=200, red_tol=1e-4){
   
   # --- Get the network structure 
   
@@ -432,18 +430,20 @@ compute_origin_destination = function(rho_in, rho_out, edge_ref, sp_ref, p_mat,
     p_to_red = (btw_edge_flow - allowed_edge_flow) / (btw_edge_flow + epsilon)
     # Compute the maximum reduction needed on the shortest-path list
     max_p_to_red = apply(t(p_btw_mat) * p_to_red, 2, max)
+    max_p_to_red[max_p_to_red < red_tol] = 0
     # Convert it to a reduction factor and transform it in a s,t matrix
     red_mat = sparseMatrix(i=sp_ref[, 1], j=sp_ref[, 2], 
                            x=(1-max_p_to_red), dims=c(n, n))
     # Compute the updated version of origin-destination affinity matrix
-    s_it_mat = s_it_mat * red_mat
+    s_it_mat = as.matrix(s_it_mat * red_mat)
     
     # --- Check for convergence and iterate
     
     diff = sum(abs(n_mat_old - n_mat))
+    couple = which(red_mat == (1 - max(max_p_to_red)), arr.ind=T)[1,]
     
-    cat("It", it_algo, ": diff =", diff, ", to_red_max =", max(max_p_to_red),
-        "\n")
+    cat("It", it_algo, ": diff =", diff, ", to_red_max =", 
+        max(max_p_to_red), ", which =", couple,"\n")
     if(diff < conv_thres_algo){
       converge_algo = T
     }

### 2022-04-20 multiple_lines/local_functions.R
Added more stats to iterations

@@ -304,7 +304,7 @@ compute_origin_destination = function(rho_in, rho_out, edge_ref, sp_ref, p_mat,
                                       s_mat=NULL, smooth_limit=T, exp_lambda=10,
                                       prop_limit=0.1, conv_thres_if=1e-5,
                                       conv_thres_algo=1e-5, epsilon=1e-40,
-                                      max_it=200, red_tol=1e-4){
+                                      max_it=200){
   
   # --- Get the network structure 
   
@@ -345,6 +345,9 @@ compute_origin_destination = function(rho_in, rho_out, edge_ref, sp_ref, p_mat,
     # --- Save old values
     
     n_mat_old = n_mat
+    s_it_mat_old = s_it_mat
+    sigma_in_old = sigma_in
+    sigma_out_old = sigma_out
     
     # --- Iterative fitting 
     
@@ -430,7 +433,6 @@ compute_origin_destination = function(rho_in, rho_out, edge_ref, sp_ref, p_mat,
     p_to_red = (btw_edge_flow - allowed_edge_flow) / (btw_edge_flow + epsilon)
     # Compute the maximum reduction needed on the shortest-path list
     max_p_to_red = apply(t(p_btw_mat) * p_to_red, 2, max)
-    max_p_to_red[max_p_to_red < red_tol] = 0
     # Convert it to a reduction factor and transform it in a s,t matrix
     red_mat = sparseMatrix(i=sp_ref[, 1], j=sp_ref[, 2], 
                            x=(1-max_p_to_red), dims=c(n, n))
@@ -439,11 +441,19 @@ compute_origin_destination = function(rho_in, rho_out, edge_ref, sp_ref, p_mat,
     
     # --- Check for convergence and iterate
     
+    # Compute iteration statistics
     diff = sum(abs(n_mat_old - n_mat))
+    diff_s_it = sum(abs(s_it_mat_old - s_it_mat))
+    diff_sigma_in = sum(abs(sigma_in_old - sigma_in))
+    diff_sigma_out = sum(abs(sigma_out_old - sigma_out))
     couple = which(red_mat == (1 - max(max_p_to_red)), arr.ind=T)[1,]
     
-    cat("It", it_algo, ": diff =", diff, ", to_red_max =", 
-        max(max_p_to_red), ", which =", couple,"\n")
+    # Print iteration statistics
+    cat("It", it_algo, ": diff =", diff, ", diff_si =", diff_s_it, 
[…]

### 2022-04-21 presentation_bb/presentation_bb.tex
Presentation started

\documentclass[10pt]{beamer}
\usetheme[progressbar=frametitle]{metropolis}

\usepackage [autostyle, english = american]{csquotes}
\usepackage{bm}

\MakeOuterQuote{"}
\newcommand{\imp}[1]{\textbf{\color{cyan}#1}}

%---------------------------------------------------
% Title

\title{Estimation of flow trajectories in a multiple lines network}
\subtitle{Case studies with \textit{transports publics de la région lausannoise} (tl) data}
\date{}
\author{Guillaume Guex \\ Romain Loup \\ François Bavaud}
\institute{University of Lausanne}

\begin{document}

	%------------------------------------------------------------------

	\maketitle

	%------------------------------------------------------------------

	\begin{frame}{Table of contents}
		\setbeamertemplate{section in toc}[sections numbered]
		\tableofcontents%[hideallsubsections]
	\end{frame}

	%------------------------------------------------------------------

	\section[Introduction]{Introduction}

	%------------------------------------------------------------------

	\begin{frame}{Context}

		The \imp{tl dataset}, used by Romain Loup for his PhD:

		\begin{itemize}
			\item 1 year of data (2019).
			\item 115 millions of passengers.
			\item 42 bus and subway lines.
			\item 1361 stops and 497 "superstops".
			\item Every journey data: traveling time, waiting time, embarking and disembarking passengers at each stops, etc.
		\end{itemize}

	\end{frame}

	%------------------------------------------------------------------

	\begin{frame}[fragile]{Context}
	\scriptsize
\begin{center}
\begin{verbatim}
     ##   stop_id  stop_name line_id direction order embarkment disembarkment
     ## 1 MALAD_N  Maladière       1         A     1     164558             0
     ## 2 MTOIE_E    Montoie       1         A     2     136236         12705
     ## 3 BATEL_E  Batelière       1         A     3     203045         13409
     ## 4 RTCOU_E Riant-Cour       1         A     4     156015         24909

     ##    stop_id  stop_name line_id direction order embarkment disembarkment
     ## 42 RTCOU_O Riant-Cour       1         R    19      23634        132201
     ## 43 BATEL_O  Batelière       1         R    20      13707        168884
     ## 44 MTOIE_O    Montoie       1         R    21       4259        128255
     ## 45 MALAD_N  Maladière       1         R    22          0        146798
\end{verbatim}

\includegraphics[width=0.7\textwidth]{img/tl_network.png}
\end{center}
	\end{frame}

	%------------------------------------------------------------------

	\begin{frame}{The multiple lines network}
		\begin{columns}
			\begin{column}{0.5\textwidth}
				Having only lines data, we have a \imp{disconnected oriented graph}. \linebreak

				In addition to \imp{line edges}, it is possible to construct \imp{transfer edges} to make the graph connected, by using, e.g.,
				\begin{itemize}
					\item Superstops names,
					\item Pedestrian time,
					\item Distance.
				\end{itemize} 
				\vspace{0.4cm}
				This is what we call a \imp{multiple lines network}.
			\end{column}
			\begin{column}{0.5\textwidth} 
				\begin{center}
					\includegraphics[width=0.8\textwidth]{img/stop_area_c.png} \\
					\vspace{0.5cm}
					\includegraphics[width=0.7\textwidth]{img/edge_type2.png}
				\end{center}
			\end{column}
		\end{columns}
	\end{frame}

	%------------------------------------------------------------------

	\begin{frame}{The problematic}

		This dataset offers multiple axes of research. In this presentation, we will focus on one question: \vspace{0.4cm}

		\begin{quotation}
			\large
			Knowing \imp{(1) the network structure} and \imp{(2) the number of passengers embarking and disembarking at each stop}, can we deduce \imp{trajectories of the passengers} in the network ?
		\end{quotation}

	\end{frame}

	%------------------------------------------------------------------

	\begin{frame}{The problematic}

		Short answer: {\large \imp{No}}.

		\begin{center}
			\huge Thank you for your attention ! \\
			Questions ?
		\end{center}

	\end{frame}

	%------------------------------------------------------------------

	\begin{frame}{The problematic}

		Exact trajectories are impossible to know, but with additional hypotheses, we can \imp{estimate} them. \\
		\vspace{0.4cm}
		We will divide this problematic into two parts:
		\begin{itemize}
			\item The estimation of trajectories on a \imp{single line}.
			\item The estimation of trajectories on the \imp{multiple lines network}.
		\end{itemize}

	\end{frame}

	%------------------------------------------------------------------

	\section{The single line problem}

	%------------------------------------------------------------------

	\begin{frame}{Formal problem definition}
		Let a line (in one direction), which have $n$ stops. Let $\bm{\rho}_\text{in} = (\rho^\text{in}_i)$ and $\bm{\rho}_\text{out} = (\rho^\text{out}_i)$ be two vectors representing, respectively, the passengers entering and leaving the line at each stops.
	\end{frame}

	%------------------------------------------------------------------

	\begin{frame}[allowframebreaks]{References}

		\bibliography{demo}
		\bibliographystyle{abbrv}

	\end{frame}

\end{document}

### 2022-04-25 presentation_bb/presentation_bb.tex
Presentation updated

@@
 \usepackage [autostyle, english = american]{csquotes}
+\usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{bm}
@@
 			\begin{column}{0.5\textwidth}
-				Having only lines data, we have a \imp{disconnected oriented graph}. \linebreak
+				Having only lines data, the structure is a \imp{disconnected oriented graph}. \linebreak
 
@@
 				\vspace{0.4cm}
-				This is what we call a \imp{multiple lines network}.
+				With transfer edges, we have a \imp{unilaterally connected graph}.
 			\end{column}
@@
 		Let a line (in one direction), which have $n$ stops.
-Let $\bm{\rho}_\text{in} = (\rho^\text{in}_i)$ and $\bm{\rho}_\text{out} = (\rho^\text{out}_i)$ be two vectors representing, respectively, the passengers entering and leaving the line at each stops.
+Let $\bm{\rho}_\text{in} = (\rho^\text{in}_s)$ and $\bm{\rho}_\text{out} = (\rho^\text{out}_t)$ be two vectors representing, respectively, the \imp{passengers entering and leaving lines at each stop}.
+\\
+		\vspace{0.4cm}
+		We search a $(n \times n)$ \imp{origin-destination matrix} $\mathbf{N} = (n_{st})$ where components represents
+		$$
+			n_{st} = \text{"\imp{the number of passengers entering line at $s$ and leaving at $t$}"}.
+		$$
+		These components must verify
+		\begin{enumerate}
+			\item $n_{st} \geq 0$,
+			\item $n_{s\bullet} = \rho^\text{in}_s$,
+			\item $n_{\bullet t} = \rho^\text{out}_t$.
+		\end{enumerate}
+		\small ($\bullet$ indicates a sum on the replaced index)
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Formal problem definition}
+		$$
+		 \mathbf{N} = \bordermatrix{
+			   & \sigma^\text{out}_1 & \sigma^\text{out}_2  & \cdots & \sigma^\text{out}_{n-1} & \sigma^\text{out}_n   \cr
+			\sigma^\text{in}_1 & n_{11} & n_{12} & \cdots & n_{1,n-1} & n_{1n} \cr
+			\sigma^\text{in}_2 & n_{21} & n_{22} & \cdots & n_{2,n-1} & n_{2n} \cr
+			\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
+			\sigma^\text{in}_{n-1} & n_{n-1,1} & n_{n-1,2} & \cdots & n_{n-1,n-1} & n_{n-1,n} \cr
+			\sigma^\text{in}_{n} & n_{n,1} & n_{n,2} & \cdots & n_{n,n-1} & n_{n,n}}
+		$$
 	\end{frame}

### 2022-04-25 multiple_lines/3_results_mapping.R
Mapping

@@ -84,7 +84,7 @@ stops_sel_shp = stops_shp %>%
 
 # --- Creating big shapefiles from results  
 
-id_stop = 50
+id_stop = 160
 
 # Getting the ingoing and outgoing flow for each stops
 init_stop = stop_names[id_stop]

### 2022-04-25 presentation_bb/presentation_bb.tex
Presentation updated

@@
 	\begin{frame}{Formal problem definition}
+		\small
+		It reads:
 		$$
@@
 		$$
+
+		In fact, we already know that some components are null:
+		$$
+		\mathbf{N} = \bordermatrix{
+			& \imp{0} & \sigma^\text{out}_2  & \cdots & \sigma^\text{out}_{n-1} & \sigma^\text{out}_n   \cr
+			\sigma^\text{in}_1 & \imp{0} & n_{12} & \cdots & n_{1,n-1} & n_{1n} \cr
+			\sigma^\text{in}_2 & \imp{0} & \imp{0} & n_{23} & \cdots & n_{2n} \cr
+			\vdots & \vdots & \vdots & \ddots & \ddots & \vdots \cr
+			\sigma^\text{in}_{n-1} & \imp{0} & \imp{0} & \cdots & \imp{0} & n_{n-1,n} \cr
+			\imp{0} & \imp{0} & \imp{0} & \cdots & \imp{0} & \imp{0}}
+		$$
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Formal problem definition}
+
+		In this form, the problem is \imp{ill posed}, because it has multiple solutions. An exemple of solution is to make passengers follow a \imp{first in, first out (FIFO)} scheme. \\
+		\vspace{0.4cm}
+		In such a case, a principle of mathematical modeling is to find the solution which makes the \imp{least assumptions about passenger behavior}.
+
 	\end{frame}

### 2022-04-25 presentation_bb/presentation_bb.tex
Presentation updated

--- 2022-04-25 presentation_bb/presentation_bb.tex
@@
 \usepackage [autostyle, english = american]{csquotes}
+\DeclareFontShape{OT1}{cmss}{b}{n}{<->ssub * cmss/bx/n}{} 
 \usepackage{amsmath}
@@
 
-		Exact trajectories are impossible to know, but with additional hypotheses, we can \imp{estimate} them. \\
+		Exact trajectories are impossible to know, but we can \imp{estimate} them. \\
 		\vspace{0.4cm}
@@
 	\begin{frame}{Formal problem definition}
-		Let a line (in one direction), which have $n$ stops.
+		Let a line (in one direction), which have $n$ stops, indexed by line order.
 Let $\bm{\rho}_\text{in} = (\rho^\text{in}_s)$ and $\bm{\rho}_\text{out} = (\rho^\text{out}_t)$ be two vectors representing, respectively, the \imp{passengers entering and leaving lines at each stop}.
@@
 	\begin{frame}{Formal problem definition}
+		\small
+		It reads:
 		$$
 		\mathbf{N} = \bordermatrix{
-			   & \sigma^\text{out}_1 & \sigma^\text{out}_2  & \cdots & \sigma^\text{out}_{n-1} & \sigma^\text{out}_n   \cr
-			\sigma^\text{in}_1 & n_{11} & n_{12} & \cdots & n_{1,n-1} & n_{1n} \cr
-			\sigma^\text{in}_2 & n_{21} & n_{22} & \cdots & n_{2,n-1} & n_{2n} \cr
+			& \rho^\text{out}_1 & \rho^\text{out}_2  & \cdots & \rho^\text{out}_{n-1} & \rho^\text{out}_n   \cr
+			\rho^\text{in}_1 & n_{11} & n_{12} & \cdots & n_{1,n-1} & n_{1n} \cr
+			\rho^\text{in}_2 & n_{21} & n_{22} & \cdots & n_{2,n-1} & n_{2n} \cr
 			\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
-			\sigma^\text{in}_{n-1} & n_{n-1,1} & n_{n-1,2} & \cdots & n_{n-1,n-1} & n_{n-1,n} \cr
-			\sigma^\text{in}_{n} & n_{n,1} & n_{n,2} & \cdots & n_{n,n-1} & n_{n,n}}
+			\rho^\text{in}_{n-1} & n_{n-1,1} & n_{n-1,2} & \cdots & n_{n-1,n-1} & n_{n-1,n} \cr
+			\rho^\text{in}_{n} & n_{n,1} & n_{n,2} & \cdots & n_{n,n-1} & n_{n,n}}
+		$$
+
+		In fact, we already know that some components are null:
+		$$
+		\mathbf{N} = \bordermatrix{
+			& \imp{0} & \rho^\text{out}_2  & \cdots & \rho^\text{out}_{n-1} & \rho^\text{out}_n \cr
+			\rho^\text{in}_1 & \imp{0} & n_{12} & \cdots & n_{1,n-1} & n_{1n} \cr
+			\rho^\text{in}_2 & \imp{0} & \imp{0} & n_{23} & \cdots & n_{2n} \cr
+			\vdots & \vdots & \vdots & \ddots & \ddots & \vdots \cr
+			\rho^\text{in}_{n-1} & \imp{0} & \imp{0} & \cdots & \imp{0} & n_{n-1,n} \cr
+			\imp{0} & \imp{0} & \imp{0} & \cdots & \imp{0} & \imp{0}}
 		$$
 	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Formal problem definition}
+
+		In this form, the problem is \imp{ill posed}, because it has multiple solutions. An exemple of solution is to make passengers follow a \imp{first in, first out (FIFO)} scheme. \\
+		\vspace{0.4cm}
+		A principle of mathematical modeling is to find the solution which makes the \imp{least assumptions about passenger behavior}, in other words the \imp{maximum entropy solution}. \\
+		\vspace{0.4cm}
+		In this case, it translates by supposing that there is the \imp{same probability of leaving the line for every passenger which have traveled at least one stop}.
+
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Solution with Markov chain modeling}
+
+		We can then model passenger flow with a \imp{Markov chain}:
+
+		\begin{center}
+			\includegraphics[width=0.8\linewidth]{img/line_markov.png} 
+		\end{center}
+
+		with $p^\text{in}_i = \frac{\rho^\text{in}_i}{\rho^\text{in}_\bullet}$ and $p^\text{out}_i = \frac{\rho^\text{out}_i}{\sum_{1 \leq k \leq (i-1)} (\rho^\text{in}_k - \rho^\text{out}_k)}$.
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Solution with Markov chain modeling}
+
+		\begin{center}
+			\includegraphics[width=\textwidth]{img/stop_distributions.png} 
+		\end{center}
+
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Solution with Markov chain modeling}
+
+		\begin{center}
+			\includegraphics[width=\textwidth]{img/length_exp.png} 
+		\end{center}
+
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\section{Iterative proportial fitting}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Iterative proportional fitting}
+
+	The same solution can be obtained with the \imp{iterative proportional fitting (IPF)} algorithm. Let 
+	\begin{enumerate}
+		\item $\mathbf{P} = (p_{ij})$ a $(n \times m)$ matrix, 
+		\item $\mathbf{u} = (u_i)$ a n-length vector, and 
+		\item $\mathbf{v} = (v_i)$ a m-length vector,
+	\end{enumerate}
+	all of them with strictly positive components. We can find two vectors $\mathbf{a} = (a_i)$ and $\mathbf{b} = (b_i)$ such that the matrix $\mathbf{Q} = (q_{ij})$, defined with
+	$$
+		q_{ij} = a_i b_j p_{ij},
+	$$
+	verifies
+	\begin{itemize}
+		\item $q_{i\bullet} = u_i$,
+		\item $q_{\bullet j} = v_j$,
+		\item $K(\mathbf{Q}| \mathbf{P}) := \sum_{ij} \frac{q_{ij}}{q_{\bullet \bullet}} \log \left( \frac{q_{ij} / q_{\bullet \bullet}}{p_{ij} / p_{\bullet \bullet}} \right)$ is minimum.
+	\end{itemize}
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Solution with iterative proportional fitting}
+
+		In our context, it means that if we define an \imp{origin-destination affinity matrix} $\mathbf{S} = (s_{st})$ with 
+			$$
+				\mathbf{S} = \left( \begin{array}{ccccc}
+				0 & 1 & \cdots & 1 & 1 \\
+				0 & 0 & 1 & \cdots & 1 \\
+				\vdots & \vdots & \ddots & \ddots & \vdots \\
+				0 & 0 & \cdots & 0 & 1 \\
+				0 & 0 & \cdots & 0 & 0 
+				\end{array} \right)
+			$$
+		we can find the \imp{maximum entropy solution with iterative proportional fitting}, i.e we find $\mathbf{a} = (a_s)$ and $\mathbf{b} = (b_t)$, such that $n_{ij} = a_s b_t s_{st}$ verify:
+		\begin{enumerate}
+			\item $n_{s\bullet} = \rho^\text{in}_s$,
+			\item $n_{\bullet t} = \rho^\text{out}_t$,
+			\item $K(\mathbf{N}| \mathbf{S})$ is minimum.
+		\end{enumerate}
+		\small (a small number $\epsilon$ has to be added on null components).
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Solution with iterative proportional fitting}
+		By \imp{decreasing (resp. increasing) $s_{st}$}, we \imp{reduce (resp. expand)} the resulting number of passengers going from $s$ to $t$ obtained with IPF. \\
+		\vspace{0.4cm}
+		Thus, this approach is more \imp{flexible}, because we could give a \imp{specific affinity matrix $\mathbf{S} = (s_{st})$}, based on other data (additional assumptions).\\
+		\vspace{0.4cm}
+		The relationship $n_{ij} = a_s b_t s_{st}$ can be seen as a \imp{gravity model} (still to investigate).
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\section{The multiple lines problem}
 

### 2022-04-25 multiple_lines/3_results_mapping.R

@@ -84,7 +84,7 @@ stops_sel_shp = stops_shp %>%
 
 # --- Creating big shapefiles from results  
 
-id_stop = 160
+id_stop = 50
 
 # Getting the ingoing and outgoing flow for each stops
 init_stop = stop_names[id_stop]

### 2022-04-26 presentation_bb/presentation_bb.tex
Presentation updated

@@
 
+		Short answer: {\large \imp{No}}.
 	\end{frame}
@@
 
-		Short answer: {\large \imp{No}}.
-
 		\begin{center}
 			\huge Thank you for your attention ! \\
-			Questions ?
+			Questions ? \\
+			\vspace{1cm}
+			\small
+			(just kidding)
 		\end{center}
@@
 
-		Exact trajectories are impossible to know, but we can \imp{estimate} them. \\
+		Exact trajectories are impossible to know, but, with additional assumptions, we can \imp{estimate} them. \\
 		\vspace{0.4cm}
@@
 			\rho^\text{in}_1 & \imp{0} & n_{12} & \cdots & n_{1,n-1} & n_{1n} \cr
-			\rho^\text{in}_2 & \imp{0} & \imp{0} & n_{23} & \cdots & n_{2n} \cr
+			\rho^\text{in}_2 & \imp{0} & \imp{0} & \ddots & \cdot & n_{2n} \cr
 			\vdots & \vdots & \vdots & \ddots & \ddots & \vdots \cr
@@
 				\mathbf{S} = \left( \begin{array}{ccccc}
-				0 & 1 & \cdots & 1 & 1 \\
+				0 & 1 & 1 & \cdots & 1 \\
 				0 & 0 & 1 & \cdots & 1 \\
@@
 
+	\begin{frame}{The multiple lines problem}
+
+		In this problem, we will have to use the whole multiple lines network, which is composed of \imp{line edges} and \imp{transfer edges}.
+
+		\begin{center}
+			\includegraphics[width=0.5\textwidth]{img/edge_type2.png}
+		\end{center}
+
+		We cannot use a Markov chain modeling in this case, but the \imp{iterative proportional fitting} approach is still promising. However, there are \imp{additional difficulties}.
+
+	\end{frame}
+
+	%-----------------------------------------------^-------------------
+
+	\begin{frame}{The flow in/out lines/network}
+
+		To begin with, we have to distinguish between
+
+		\begin{itemize}
+			\item The passengers \imp{entering and leaving lines} at each stops, represented by vectors \imp{$\bm{\rho}_\text{in}$} and \imp{$\bm{\rho}_\text{out}$} and 
+			\item The passengers \imp{entering and leaving the network} at each stops, represented by vectors \imp{$\bm{\sigma}_\text{in}$} and \imp{$\bm{\sigma}_\text{out}$}, which are \imp{unknown}.
+		\end{itemize}
+
+		At each stop $i$, we have
+
+		\begin{itemize}
+			\item $\rho^\text{in}_i = \sigma^\text{in}_i + x^\text{B}_{\bullet i}$,
+			\item $\rho^\text{out}_i = \sigma^\text{out}_i + x^\text{B}_{i \bullet}$,
+		\end{itemize}
+
+		where $x^\text{B}_{\bullet i}$ is the \imp{transfer flow entering the node $i$} and $x^\text{B}_{i \bullet}$ the \imp{transfer flow leaving the node $i$}. \\
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{The flow in/out lines/network}
+		\begin{center}
+			\hspace{0.1cm} \includegraphics[width=0.8\textwidth]{img/flow_balance.png}
+		\end{center}
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{The flow in/out lines/network}
+		\begin{center}
+			\includegraphics[width=0.9\textwidth]{img/flow_measured.png}
+		\end{center}
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{The flow in/out lines/network}
+		The flow entering/leaving the lines, $\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{out}$, are \imp{known}. \\
+		\vspace{0.4cm}
+		If we know \imp{transfer flow on edges}, i.e.
+$\mathbf{X}_\text{B} = (x^\text{B}_{ij})$, we can \imp{compute the flow entering/leaving the network}, $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{in}$.
+\\
+		\vspace{0.4cm}
+		The flow entering/leaving the lines, $\rho^\text{in}_i$ and $\rho^\text{out}_i$, can also acts as \imp{constraints on the in/out transfer flow}, $x^\text{B}_{\bullet i}$ and $x^\text{B}_{i \bullet}$.\\
+		\vspace{0.4cm}
+		When there are no transfers on $i$, we have $\rho^\text{in}_i = \sigma^\text{in}_i$ and  $\rho^\text{out}_i = \sigma^\text{out}_i$. \\
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Flow behavior}
+
+		The second difficulty is that there are generally \imp{multiple routes} to reach node $t$ from node $s$. This can be solved by making a new assumption. \\
+		\vspace{0.4cm}
+		In the multiple lines network, we will suppose that passengers take \imp{shortest-paths} in order to reach node $t$ from node $s$. \\
+		\vspace{0.4cm}
+		If multiple shortest-paths exists between $s$ and $t$, the passenger flow is \imp{divided equally} among them.
+
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Flow behavior}
+
+		This assumption unlocks a very useful property.
+If we have an \imp{origin-destination matrix} $\mathbf{N} = (n_{st})$, we can compute the $(n \times n)$ \imp{flow matrix on edges} $\mathbf{X} = (x_{ij})$.
+\\
+		\vspace{0.6cm}
+		\begin{columns}
+			\small
+			\begin{column}{0.7\textwidth}
+				$$
+					\hspace{1cm} \mathbf{N} = \left( \begin{array}{ccccc}
+					0 & 13055 & 243 & \cdots & 144 \\
+					3498 & 0 & 24429 & \cdots & 7523 \\
+					\vdots & \vdots & \ddots & \ddots & \vdots \\
+					1508 &  & \cdots & 0 & 5093 \\
+					8903 & 6343 & \cdots & 53 & 0 
+					\end{array} \right) \rightarrow 
+				$$ 
+			\end{column}
+			\begin{column}{0.6\textwidth}
+				\hspace{0.5cm} \includegraphics[width=0.5\textwidth]{img/flow_computed.png}
+			\end{column}
+		\end{columns}
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Algorithm outline}
+
+		We now begin to make an outline for an \imp{iterative Algorithm}. There will be 4 steps at each iteration: \\
+		\footnotesize
+		\begin{align}
+			\left. \begin{array}{r} 
+				{\color{cyan} \text{OD affinity matrix } \mathbf{S}} \\
+				{\color{cyan} \text{Flow in the network } \bm{\sigma}_\text{in}} \\
+				{\color{cyan} \text{Flow out the network } \bm{\sigma}_\text{out}}
+			\end{array} \right\}  &\overset{IPF}{\longrightarrow} \text{OD matrix } \mathbf{N} \\
+			\begin{array}{r} ^
+			\text{OD matrix } \mathbf{N}
+			\end{array} &\overset{SP}{\longrightarrow} \text{Edge transfer flow } \mathbf{X}^\text{B} \\
+			\left. \begin{array}{r} 
+			\text{Edge transfer flow } \mathbf{X}^\text{B} \\
+			\text{Flow in lines } \bm{\rho}_\text{in} \\
+			\text{Flow out lines } \bm{\rho}_\text{out}
+			\end{array} \right\}  &\overset{\text{Constraints}}{\longrightarrow} 
+			\left\{ \begin{array}{l}
+			\text{Corrected edge transfer flow } \widetilde{\mathbf{X}}^\text{B} \\
+			{\color{cyan} \text{Flow in the network } \bm{\sigma}_\text{in}} \\
+			{\color{cyan} \text{Flow out the network } \bm{\sigma}_\text{out}}
+			\end{array} \right. \\
+			\left. \begin{array}{r} 
+			\text{Edge transfer flow } \mathbf{X}^\text{B} \\
+			 \text{Corrected edge transfer flow } \widetilde{\mathbf{X}}^\text{B}
+			\end{array} \right\}  &\overset{\text{Affinity update}}{\longrightarrow} {\color{cyan} \text{OD affinity matrix } \mathbf{S}}
+		\end{align}	
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Step 1: iterative proportional fitting}
+
+		At the beginning of the algorithm, we have to set 
+		\begin{itemize}
+			\item An \imp{initial flow in the network }. We can set it to $\bm{\sigma}^\text{init}_\text{in} = \bm{\rho}_\text{in}$,
+			\item An \imp{initial flow out the network }. We can set it to $\bm{\sigma}^\text{init}_\text{out} = \bm{\rho}_\text{out}$,
+			\item An \imp{initial affinity matrix between origin-destination}, $\mathbf{S}^\text{init}$.
+		\end{itemize}
+
+		The initial affinity matrix $\mathbf{S}^\text{init} = (s^\text{init}_{st})$ is crafted in order to have:
+		\begin{itemize}
+			\item $s^\text{init}_{st} = 1$ if $s$ is a \imp{valid trajectory} for using the network.
+			\item $s^\text{init}_{st} = 0$ otherwise.
+		\end{itemize}
+
+		It now possible to obtain origin-destination matrix $\mathbf{N}$ with \imp{iterative proportional fitting}.
+
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Step 2: shortest-paths flow}
+
+		In this step, we use the assumption that passengers use \imp{shortest-paths} in the network to obtain flow on edges, and in particular, \imp{flow on transfer edges}:
+		$$
+			\mathbf{N} \longrightarrow \mathbf{X}^\text{B}
+		$$ 
+		\vspace{0.6cm}
+		\begin{columns}
+			\small
+			\begin{column}{0.7\textwidth}
+				$$
+				\hspace{1cm} \mathbf{N} = \left( \begin{array}{ccccc}
+				0 & 13055 & 243 & \cdots & 144 \\
+				3498 & 0 & 24429 & \cdots & 7523 \\
+				\vdots & \vdots & \ddots & \ddots & \vdots \\
+				1508 &  & \cdots & 0 & 5093 \\
+				8903 & 6343 & \cdots & 53 & 0 
+				\end{array} \right) \rightarrow 
+				$$ 
+			\end{column}
+			\begin{column}{0.6\textwidth}
+				\hspace{0.5cm} \includegraphics[width=0.5\textwidth]{img/flow_computed.png}
+			\end{column}
+		\end{columns}
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Step 3: corrected transfer flow}
+
+		This transfer flow $\mathbf{X}^\text{B}$ could be used to update $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$, with
+
+		\begin{itemize}
+			\item $\sigma^\text{in}_i = \rho^\text{in}_i - x^\text{B}_{\bullet i}$,
+			\item $\sigma^\text{out}_i = \rho^\text{out}_i - x^\text{B}_{i \bullet}$,
+		\end{itemize}
+		However, there is \imp{no guarantee that the flow will not exceed limits given by $\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{in}$}.
+		\begin{center}
+			\includegraphics[width=0.7\textwidth]{img/flow_constrained.png}
+		\end{center}
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Step 3: corrected transfer flow}
+
+		For each $x^\text{B}_{ij}$, we use $\rho^\text{out}_i$ and $\rho^\text{in}_j$ in order to compute a \imp{corrected transfer flow} $\widetilde{x}_{ij}^\text{B} \leq x_{ij}^\text{B}$.
+There are multiple choices:
+
+		\begin{enumerate}
+			\item \imp{Constraint thresholds} could be reachable.
+			\item We can set a \imp{percentage limit} for the transfer flow among the flow in/out of the lines.
+			\item We can set a \imp{soft limit} to the transfer flow, with, e.g. an exponential law.
+		\end{enumerate}
+
+		When this corrected flow is computed, $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ can be \imp{updated} with $\sigma^\text{in}_i = \rho^\text{in}_i - \widetilde{x}^\text{B}_{\bullet i}$ and $\sigma^\text{out}_i = \rho^\text{out}_i - \widetilde{x}^\text{B}_{i \bullet}$.
+
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Step 4: Affinity update}
+
+		What about the \imp{excess flow on transfer edges}, $x^\text{B}_{ij} - \widetilde{x}^\text{B}_{ij}$ ? \\
+		\vspace{0.4cm}
+		Note that we only affected \imp{margins distribution}, but we need also need a way to reduce \imp{particular dependencies} between some $s$ and $t$. \\
+		\vspace{0.4cm}
+		Having a \imp{shortest-path structure}, we know which couples $s, t$ are responsible for the excess flow on edge $(i, j)$.
+		\begin{center}
+			\includegraphics[width=0.6\textwidth]{img/excess_flow.png}
+		\end{center}
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Step 4: Affinity update}
+
+		On each transfer edge, we can compute the \imp{proportion of allowed flow}:
+		$$
+			p^\text{allowed}_{ij} = \frac{\widetilde{x}_{ij}^\text{B}}{x_{ij}^\text{B}}
+		$$
+		Each transfer edge will then "send a signal" to all couples $(s,t)$ using this edge, in order for them to \imp{reduce their affinities $s_{st}$}. \\
+		\vspace{0.4cm}
+		Each couple $(s,t)$, will recieve a \imp{list} of proportion of allowed flow, from all transfer edges on its shortest-path.
+The \imp{update factor for the affinity $s_{st}$} is constructed by using the minimum allowed flow received.
+		$$
+			s^\text{new}_{st} = s^\text{old}_{st} \cdot (1 - \min\{p^\text{allowed}_{i_1, j_1}, \ldots, p^\text{allowed}_{i_k, j_k} \})
+		$$
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\section{Results}
+
+	%------------------------------------------------------------------
+
 	\begin{frame}[allowframebreaks]{References}

### 2022-04-26 presentation_bb/presentation_bb.tex
Presentation in pre-final form

@@
 \title{Estimation of flow trajectories in a multiple lines network}
-\subtitle{Case studies with \textit{transports publics de la région lausannoise} (tl) data}
+\subtitle{Experiments with \textit{transports publics de la région lausannoise} (tl) data}
 \date{}
@@
 
-	\begin{frame}{The problematic}
+	\begin{frame}{Problematic}
 
@@
 
-	\begin{frame}{The problematic}
+	\begin{frame}
 
@@
 		\vspace{0.4cm}
-		We search a $(n \times n)$ \imp{origin-destination matrix} $\mathbf{N} = (n_{st})$ where components represents
+		We search a $(n \times n)$ \imp{origin-destination matrix $\mathbf{N} = (n_{st})$} where components represents
 		$$
-			n_{st} = \text{"\imp{the number of passengers entering line at $s$ and leaving at $t$}"}.
+			{\color{cyan} n_{st} = \text{"the number of passengers entering line at $s$ and leaving at $t$"}}.
 		$$
@@
 
-	The same solution can be obtained with the \imp{iterative proportional fitting (IPF)} algorithm. Let 
+	The same solution can be obtained with the \imp{iterative proportional fitting (IPF)} algorithm [Bishop et al., 1975]. Let 
 	\begin{enumerate}
@@
 
-		In our context, it means that if we define an \imp{origin-destination affinity matrix} $\mathbf{S} = (s_{st})$ with 
+		In our context, it means that if we define an \imp{origin-destination affinity matrix $\mathbf{S} = (s_{st})$} with 
 			$$
@@
 			$$
-		we can find the \imp{maximum entropy solution with iterative proportional fitting}, i.e we find $\mathbf{a} = (a_s)$ and $\mathbf{b} = (b_t)$, such that $n_{ij} = a_s b_t s_{st}$ verify:
+		we can find the \imp{maximum entropy solution with iterative proportional fitting}, i.e., we can find $\mathbf{a} = (a_s)$ and $\mathbf{b} = (b_t)$, such that $n_{ij} = a_s b_t s_{st}$ verifies
 		\begin{enumerate}
@@
 		\vspace{0.4cm}
-		The relationship $n_{ij} = a_s b_t s_{st}$ can be seen as a \imp{gravity model} (still to investigate).
+		The relationship $n_{st} = a_s b_t s_{st}$ can be seen as a \imp{gravity model} (still to investigate).
 	\end{frame}
@@
 		\begin{center}
-			\includegraphics[width=0.5\textwidth]{img/edge_type2.png}
+			\includegraphics[width=0.45\textwidth]{img/edge_type2.png}
 		\end{center}
 
-		We cannot use a Markov chain modeling in this case, but the \imp{iterative proportional fitting} approach is still promising. However, there are \imp{additional difficulties}.
+		It is not possible to use a Markov chain modeling in this case, but the \imp{iterative proportional fitting} approach is still valid. However, there are \imp{additional difficulties}.
 
@@
 
+		\begin{align*}
+			\rho^\text{in}_i &= \sigma^\text{in}_i + x^\text{B}_{\bullet i}, \\
+			\rho^\text{out}_i &= \sigma^\text{out}_i + x^\text{B}_{i \bullet},
+		\end{align*}
+
+		where 
 		\begin{itemize}
-			\item $\rho^\text{in}_i = \sigma^\text{in}_i + x^\text{B}_{\bullet i}$,
-			\item $\rho^\text{out}_i = \sigma^\text{out}_i + x^\text{B}_{i \bullet}$,
+			\item $x^\text{B}_{\bullet i}$ is the \imp{transfer flow entering the node $i$} and
+			\item $x^\text{B}_{i \bullet}$ is the \imp{transfer flow leaving the node $i$}.
 		\end{itemize}
-
-		where $x^\text{B}_{\bullet i}$ is the \imp{transfer flow entering the node $i$} and $x^\text{B}_{i \bullet}$ the \imp{transfer flow leaving the node $i$}. \\
 	\end{frame}
@@
 		\vspace{0.4cm}
-		The flow entering/leaving the lines, $\rho^\text{in}_i$ and $\rho^\text{out}_i$, can also acts as \imp{constraints on the in/out transfer flow}, $x^\text{B}_{\bullet i}$ and $x^\text{B}_{i \bullet}$.\\
+		The flow entering/leaving the lines, $\rho^\text{in}_i$ and $\rho^\text{out}_i$, acts as \imp{constraints on the in/out transfer flow}, $x^\text{B}_{\bullet i}$ and $x^\text{B}_{i \bullet}$.\\
 		\vspace{0.4cm}
@@
 
-		We now begin to make an outline for an \imp{iterative Algorithm}. There will be 4 steps at each iteration: \\
+		Let us make an outline for the \imp{iterative Algorithm}. There is \imp{4 steps} at each iteration: \\
 		\footnotesize
@@
 			\end{array} \right\}  &\overset{IPF}{\longrightarrow} \text{OD matrix } \mathbf{N} \\
-			\begin{array}{r} ^
+			\begin{array}{r}
 			\text{OD matrix } \mathbf{N}
-			\end{array} &\overset{SP}{\longrightarrow} \text{Edge transfer flow } \mathbf{X}^\text{B} \\
+			\end{array} &\overset{SP}{\longrightarrow} \text{Transfer flow } \mathbf{X}^\text{B} \\
 			\left. \begin{array}{r} 
-			\text{Edge transfer flow } \mathbf{X}^\text{B} \\
+			\text{Transfer flow } \mathbf{X}^\text{B} \\
 			\text{Flow in lines } \bm{\rho}_\text{in} \\
@@
 			\left\{ \begin{array}{l}
-			\text{Corrected edge transfer flow } \widetilde{\mathbf{X}}^\text{B} \\
+			\text{Allowed transfer flow } \widetilde{\mathbf{X}}^\text{B} \\
 			{\color{cyan} \text{Flow in the network } \bm{\sigma}_\text{in}} \\
@@
 			\left. \begin{array}{r} 
-			\text{Edge transfer flow } \mathbf{X}^\text{B} \\
-			 \text{Corrected edge transfer flow } \widetilde{\mathbf{X}}^\text{B}
+			\text{Transfer flow } \mathbf{X}^\text{B} \\
+			 \text{Allowed transfer flow } \widetilde{\mathbf{X}}^\text{B}
 			\end{array} \right\}  &\overset{\text{Affinity update}}{\longrightarrow} {\color{cyan} \text{OD affinity matrix } \mathbf{S}}
@@
 
-		It now possible to obtain origin-destination matrix $\mathbf{N}$ with \imp{iterative proportional fitting}.
+		It now possible to obtain a first origin-destination matrix $\mathbf{N}$ with \imp{iterative proportional fitting}.
 
@@
 
-	\begin{frame}{Step 3: corrected transfer flow}
+	\begin{frame}{Step 3: allowed transfer flow}
 
-		For each $x^\text{B}_{ij}$, we use $\rho^\text{out}_i$ and $\rho^\text{in}_j$ in order to compute a \imp{corrected transfer flow} $\widetilde{x}_{ij}^\text{B} \leq x_{ij}^\text{B}$.
+		For each $x^\text{B}_{ij}$, we use $\rho^\text{out}_i$ and $\rho^\text{in}_j$ in order to compute a \imp{allowed transfer flow} $\widetilde{x}_{ij}^\text{B} \leq x_{ij}^\text{B}$.
 There are multiple choices:
@@
 			\item We can set a \imp{percentage limit} for the transfer flow among the flow in/out of the lines.
-			\item We can set a \imp{soft limit} to the transfer flow, with, e.g. an exponential law.
+			\item We can set a \imp{soft limit} to the transfer flow, with, e.g., an exponential law.
 		\end{enumerate}
 
-		When this corrected flow is computed, $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ can be \imp{updated} with $\sigma^\text{in}_i = \rho^\text{in}_i - \widetilde{x}^\text{B}_{\bullet i}$ and $\sigma^\text{out}_i = \rho^\text{out}_i - \widetilde{x}^\text{B}_{i \bullet}$.
-
+		When this corrected flow is computed, $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ can be \imp{updated} with 
+		\begin{align*}
+			\sigma^\text{in}_i &= \rho^\text{in}_i - \widetilde{x}^\text{B}_{\bullet i}, \\	\sigma^\text{out}_i &= \rho^\text{out}_i - \widetilde{x}^\text{B}_{i \bullet}.
+		\end{align*}
 	\end{frame}
@@
 
-		What about the \imp{excess flow on transfer edges}, $x^\text{B}_{ij} - \widetilde{x}^\text{B}_{ij}$ ? \\
+		What about the \imp{excess flow on transfer edges}, i.e., $x^\text{B}_{ij} - \widetilde{x}^\text{B}_{ij}$ ? \\
 		\vspace{0.4cm}
-		Note that we only affected \imp{margins distribution}, but we need also need a way to reduce \imp{particular dependencies} between some $s$ and $t$. \\
+		Note that we only updated \imp{margins distribution}, but we need also need a way to reduce \imp{dependencies} between particular $s$ and $t$. \\
 		\vspace{0.4cm}
-		Having a \imp{shortest-path structure}, we know which couples $s, t$ are responsible for the excess flow on edge $(i, j)$.
+		Having flow following \imp{shortest-paths}, we know which couples $s, t$ are "responsible" for the excess flow on edge $(i, j)$.
 		\begin{center}
@@
 		$$
-		Each transfer edge will then "send a signal" to all couples $(s,t)$ using this edge, in order for them to \imp{reduce their affinities $s_{st}$}. \\
+		Each transfer edge will then "send a signal" to all couples $s,t$ using this edge on their shortest-paths, in order to \imp{reduce their affinities $s_{st}$}. \\
 		\vspace{0.4cm}
-		Each couple $(s,t)$, will recieve a \imp{list} of proportion of allowed flow, from all transfer edges on its shortest-path.
+		Each couple $s,t$, will recieve a \imp{list of proportion of allowed flow}, from all transfer edges on its shortest-path.
 The \imp{update factor for the affinity $s_{st}$} is constructed by using the minimum allowed flow received.
 		$$
-			s^\text{new}_{st} = s^\text{old}_{st} \cdot (1 - \min\{p^\text{allowed}_{i_1, j_1}, \ldots, p^\text{allowed}_{i_k, j_k} \})
+			s^\text{new}_{st} = \min\{p^\text{allowed}_{i_1, j_1}, \ldots, p^\text{allowed}_{i_k, j_k} \} \cdot s^\text{old}_{st}
 		$$
@@
 
-	\section{Results}
+	\begin{frame}{Step 4: Affinity update}
+		\begin{center}
+			\includegraphics[width=0.7\textwidth]{img/allowed_flow.png} \\
+			In this example, $s^\text{new}_{st} = 0.3 \cdot s^\text{old}_{st}$
+		\end{center}
+	\end{frame}
 
@@
 
-	\begin{frame}[allowframebreaks]{References}
+	\section{Results (demo)}
 
-		\bibliography{demo}
-		\bibliographystyle{abbrv}
+	%------------------------------------------------------------------
+
+	\section{Conclusion}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Conclusion}
+
+		As it is still a work in progress, the conclusion will take the form of a to-do list:
+		\begin{itemize}
+			\item \imp{Optimizing} speed/memory (almost done).
+			\item Running the algorithm on \imp{all lines}.
+			\item Using carefully crafted \imp{OD affinities}. 
+			\item Finding \imp{applications}.
+			\item Finding a \imp{proof of convergence}. 
+			\item Finding if the solution is \imp{unique} (doubt it).
+			\item Making links with \imp{gravity model}, \imp{transportation problem}, ...
+			\item \imp{Adapting} the algorithm to closely related problems.
+		\end{itemize}
 
@@
 
+	%------------------------------------------------------------------
+
+	\begin{frame}
+
+		\begin{center}
+			\huge Thank you for your attention ! \\
+			Questions ? \\
+			\vspace{1cm}
+			\small
+			(for real this time)
+		\end{center}
+
+	\end{frame}
+
+	%------------------------------------------------------------------
+
 \end{document}

### 2022-04-26 multiple_lines/2_compute_OD_matrix.R

@@ -36,7 +36,7 @@ prop_limit = 0.2
 # epsilon
 epsilon = 1e-40
 # max iteration
-max_it = 200
+max_it = 50
 
 #--------------------------------
 # Process

### 2022-04-26 multiple_lines/3_results_mapping.R

@@ -84,7 +84,7 @@ stops_sel_shp = stops_shp %>%
 
 # --- Creating big shapefiles from results  
 
-id_stop = 50
+id_stop = 1
 
 # Getting the ingoing and outgoing flow for each stops
 init_stop = stop_names[id_stop]

### 2022-04-27 presentation_bb/presentation_bb.tex
Presentation finished

@@
 
-		The \imp{tl dataset}, used by Romain Loup for his PhD:
+		The \imp{tl dataset}, exploited by Romain Loup for his PhD:
 
@@
 					\item Pedestrian time,
-					\item Distance.
+					\item Euclidean distance.
 				\end{itemize} 
@@
 
-		This dataset offers multiple axes of research. In this presentation, we will focus on one question: \vspace{0.4cm}
-
+		This dataset offers multiple axes of research. In this presentation, we focus on one question: \\
+		\vspace{0.6cm}
 		\begin{quotation}
@@
 		\end{quotation}
-
+		\normalsize
+		\vspace{0.4cm}
 		Short answer: {\large \imp{No}}.
@@
 
-	\begin{frame}{The problematic}
+	\begin{frame}{Problematic}
 
-		Exact trajectories are impossible to know, but, with additional assumptions, we can \imp{estimate} them. \\
+		Exact trajectories are impossible to know, but, with additional assumptions, we can produce  \imp{estimations} of them. \\
 		\vspace{0.4cm}
-		We will divide this problematic into two parts:
+		We divide this problematic into two parts:
 		\begin{itemize}
-			\item The estimation of trajectories on a \imp{single line}.
-			\item The estimation of trajectories on the \imp{multiple lines network}.
+			\item Estimation of trajectories on a \imp{single line}.
+			\item Estimation of trajectories on the \imp{multiple lines network}.
 		\end{itemize}
@@
 	\begin{frame}{Formal problem definition}
-		Let a line (in one direction), which have $n$ stops, indexed by line order.
-Let $\bm{\rho}_\text{in} = (\rho^\text{in}_s)$ and $\bm{\rho}_\text{out} = (\rho^\text{out}_t)$ be two vectors representing, respectively, the \imp{passengers entering and leaving lines at each stop}.
+		Let a line (in one direction), which have $n$ stops, indexed regarding line order.
+Let \imp{$\bm{\rho}_\text{in} = (\rho^\text{in}_s)$} and \imp{$\bm{\rho}_\text{out} = (\rho^\text{out}_t)$} be two vectors representing, respectively, the \imp{passengers entering and leaving the line at each stop}.
 \\
@@
 		\small
-		It reads:
+		It reads
 		$$
@@
 
-		In fact, we already know that some components are null:
+		In fact, we already know that some components are null
 		$$
@@
 
-		In this form, the problem is \imp{ill posed}, because it has multiple solutions. An exemple of solution is to make passengers follow a \imp{first in, first out (FIFO)} scheme. \\
+		In this form, the problem is \imp{ill posed}, because it has multiple acceptable solutions. An exemple of solution is to make passengers follow a \imp{first in, first out (FIFO)} scheme. \\
 		\vspace{0.4cm}
@@
 			$$
-		we can find the \imp{maximum entropy solution with iterative proportional fitting}, i.e., we can find $\mathbf{a} = (a_s)$ and $\mathbf{b} = (b_t)$, such that $n_{ij} = a_s b_t s_{st}$ verifies
+		we can find the \imp{same maximum entropy solution with iterative proportional fitting}, i.e., we can find $\mathbf{a} = (a_s)$ and $\mathbf{b} = (b_t)$, such that $n_{ij} = a_s b_t s_{st}$ verifies
 		\begin{enumerate}
@@
 
-		In this problem, we will have to use the whole multiple lines network, which is composed of \imp{line edges} and \imp{transfer edges}.
+		In this problem, the whole multiple lines network is used. Let us recall that it is composed of \imp{line edges} and \imp{transfer edges}.
 
@@
 
+	\begin{frame}{Flow behavior}
+
+		The first difficulty is that there are generally \imp{multiple routes} to go from node $s$ to $t$. This can be solved by making a new assumption. \\
+		\vspace{0.4cm}
+		In the multiple lines network, we suppose that passengers take \imp{shortest-paths} in order to reach node $t$ from node $s$. \\
+		\vspace{0.4cm}
+		If multiple shortest-paths exists between $s$ and $t$, the passenger flow is \imp{divided equally} among them.
+
+	\end{frame}
+
+	%------------------------------------------------------------------
+
+	\begin{frame}{Flow behavior}
+
+		This assumption unlocks a very useful property.
+If we have an \imp{origin-destination matrix} $\mathbf{N} = (n_{st})$, we can compute the $(n \times n)$ \imp{flow matrix on edges $\mathbf{X} = (x_{ij})$}.
+\\
+		\vspace{0.6cm}
+		\begin{columns}
+			\small
+			\begin{column}{0.7\textwidth}
+				$$
+				\hspace{1cm} \mathbf{N} = \left( \begin{array}{ccccc}
+				0 & 13055 & 243 & \cdots & 144 \\
+				3498 & 0 & 24429 & \cdots & 7523 \\
+				\vdots & \vdots & \ddots & \ddots & \vdots \\
+				1508 &  & \cdots & 0 & 5093 \\
+				8903 & 6343 & \cdots & 53 & 0 
+				\end{array} \right) \rightarrow 
+				$$ 
+			\end{column}
+			\begin{column}{0.6\textwidth}
+				\hspace{0.5cm} \includegraphics[width=0.5\textwidth]{img/flow_computed.png}
+			\end{column}
+		\end{columns}
+	\end{frame}
+
+	%------------------------------------------------------------------
+
 	\begin{frame}{The flow in/out lines/network}
 
-		To begin with, we have to distinguish between
+		A second problem is that we have to distinguish between
 
 		\begin{itemize}
-			\item The passengers \imp{entering and leaving lines} at each stops, represented by vectors \imp{$\bm{\rho}_\text{in}$} and \imp{$\bm{\rho}_\text{out}$} and 
-			\item The passengers \imp{entering and leaving the network} at each stops, represented by vectors \imp{$\bm{\sigma}_\text{in}$} and \imp{$\bm{\sigma}_\text{out}$}, which are \imp{unknown}.
+			\item The passengers \imp{entering and leaving lines} at each stop, represented by vectors \imp{$\bm{\rho}_\text{in}$} and \imp{$\bm{\rho}_\text{out}$}, which are \imp{measured}, and 
+			\item The passengers \imp{entering and leaving the network} at each stop, represented by vectors \imp{$\bm{\sigma}_\text{in}$} and \imp{$\bm{\sigma}_\text{out}$}, which are \imp{unknown}.
 		\end{itemize}
@@
 	\begin{frame}{The flow in/out lines/network}
-		The flow entering/leaving the lines, $\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{out}$, are \imp{known}. \\
-		\vspace{0.4cm}
 		If we know \imp{transfer flow on edges}, i.e.
@@
 		\vspace{0.4cm}
-		When there are no transfers on $i$, we have $\rho^\text{in}_i = \sigma^\text{in}_i$ and  $\rho^\text{out}_i = \sigma^\text{out}_i$. \\
-	\end{frame}
-
-	%------------------------------------------------------------------
-
-	\begin{frame}{Flow behavior}
-
-		The second difficulty is that there are generally \imp{multiple routes} to reach node $t$ from node $s$. This can be solved by making a new assumption. \\
-		\vspace{0.4cm}
-		In the multiple lines network, we will suppose that passengers take \imp{shortest-paths} in order to reach node $t$ from node $s$. \\
-		\vspace{0.4cm}
-		If multiple shortest-paths exists between $s$ and $t$, the passenger flow is \imp{divided equally} among them.
-
-	\end{frame}
-
-	%------------------------------------------------------------------
-
-	\begin{frame}{Flow behavior}
-
-		This assumption unlocks a very useful property.
-If we have an \imp{origin-destination matrix} $\mathbf{N} = (n_{st})$, we can compute the $(n \times n)$ \imp{flow matrix on edges} $\mathbf{X} = (x_{ij})$.
-\\
-		\vspace{0.6cm}
-		\begin{columns}
-			\small
-			\begin{column}{0.7\textwidth}
-				$$
-					\hspace{1cm} \mathbf{N} = \left( \begin{array}{ccccc}
-					0 & 13055 & 243 & \cdots & 144 \\
-					3498 & 0 & 24429 & \cdots & 7523 \\
-					\vdots & \vdots & \ddots & \ddots & \vdots \\
-					1508 &  & \cdots & 0 & 5093 \\
-					8903 & 6343 & \cdots & 53 & 0 
-					\end{array} \right) \rightarrow 
-				$$ 
-			\end{column}
-			\begin{column}{0.6\textwidth}
-				\hspace{0.5cm} \includegraphics[width=0.5\textwidth]{img/flow_computed.png}
-			\end{column}
-		\end{columns}
+		When there are no transfers on $i$, we have $\sigma^\text{in}_i = \rho^\text{in}_i$ and  $\sigma^\text{out}_i = \rho^\text{out}_i$. \\
 	\end{frame}
@@
 
-		Let us make an outline for the \imp{iterative Algorithm}. There is \imp{4 steps} at each iteration: \\
+		Let us make an outline for the \imp{iterative Algorithm}. There are \imp{4 steps} at each iteration: \\
 		\footnotesize
@@
 			\text{OD matrix } \mathbf{N}
-			\end{array} &\overset{SP}{\longrightarrow} \text{Transfer flow } \mathbf{X}^\text{B} \\
+			\end{array} &\overset{SP}{\longrightarrow} \text{Transfer flow } \mathbf{X}_\text{B} \\
 			\left. \begin{array}{r} 
@@
 			\left\{ \begin{array}{l}
-			\text{Allowed transfer flow } \widetilde{\mathbf{X}}^\text{B} \\
+			\text{Allowed transfer flow } \widetilde{\mathbf{X}}_\text{B} \\
 			{\color{cyan} \text{Flow in the network } \bm{\sigma}_\text{in}} \\
@@
 			\left. \begin{array}{r} 
-			\text{Transfer flow } \mathbf{X}^\text{B} \\
-			 \text{Allowed transfer flow } \widetilde{\mathbf{X}}^\text{B}
+			\text{Transfer flow } \mathbf{X}_\text{B} \\
+			 \text{Allowed transfer flow } \widetilde{\mathbf{X}}_\text{B}
 			\end{array} \right\}  &\overset{\text{Affinity update}}{\longrightarrow} {\color{cyan} \text{OD affinity matrix } \mathbf{S}}
@@
 
-		The initial affinity matrix $\mathbf{S}^\text{init} = (s^\text{init}_{st})$ is crafted in order to have:
+		The initial affinity matrix $\mathbf{S}^\text{init} = (s^\text{init}_{st})$ is crafted in order to have
 		\begin{itemize}
-			\item $s^\text{init}_{st} = 1$ if $s$ is a \imp{valid trajectory} for using the network.
-			\item $s^\text{init}_{st} = 0$ otherwise.
+			\item $s^\text{init}_{st} = 1$, if going from $s$ to $t$ is a \imp{valid trajectory} for using the network,
+			\item $s^\text{init}_{st} = 0$, otherwise.
 		\end{itemize}
@@
 		$$
-			\mathbf{N} \longrightarrow \mathbf{X}^\text{B}
+			\mathbf{N} \longrightarrow \mathbf{X}_\text{B}
 		$$ 
@@
 
-		This transfer flow $\mathbf{X}^\text{B}$ could be used to update $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$, with
-
-		\begin{itemize}
-			\item $\sigma^\text{in}_i = \rho^\text{in}_i - x^\text{B}_{\bullet i}$,
-			\item $\sigma^\text{out}_i = \rho^\text{out}_i - x^\text{B}_{i \bullet}$,
-		\end{itemize}
-		However, there is \imp{no guarantee that the flow will not exceed limits given by $\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{in}$}.
+		This transfer flow $\mathbf{X}_\text{B}$ could be used to update $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$, with 
+		\begin{align*}
+			\sigma^\text{in}_i &= \rho^\text{in}_i - x^\text{B}_{\bullet i}, \\
+			\sigma^\text{out}_i &= \rho^\text{out}_i - x^\text{B}_{i \bullet}.
+		\end{align*}
+		However, there is \imp{no guarantee that the transfer flow do not exceed limits given by $\bm{\rho}_\text{in}$ and $\bm{\rho}_\text{in}$}.
 		\begin{center}
@@
 
-		For each $x^\text{B}_{ij}$, we use $\rho^\text{out}_i$ and $\rho^\text{in}_j$ in order to compute a \imp{allowed transfer flow} $\widetilde{x}_{ij}^\text{B} \leq x_{ij}^\text{B}$.
+		For each $x^\text{B}_{ij}$, we use $\rho^\text{out}_i$ and $\rho^\text{in}_j$ in order to compute a \imp{allowed transfer flow  $\widetilde{x}_{ij}^\text{B}$}, with  $\widetilde{x}_{ij}^\text{B} \leq x_{ij}^\text{B}$.
 There are multiple choices:
@@
 
-		When this corrected flow is computed, $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ can be \imp{updated} with 
+		When this allowed flow is computed, $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ can be \imp{updated} with 
 		\begin{align*}
@@
 		\vspace{0.4cm}
-		Note that we only updated \imp{margins distribution}, but we need also need a way to reduce \imp{dependencies} between particular $s$ and $t$. \\
+		Note that we only updated \imp{margins distribution}, but we also need a way to reduce \imp{dependencies} between particular $s$ and $t$. \\
 		\vspace{0.4cm}
@@
 		$$
-		Each transfer edge will then "send a signal" to all couples $s,t$ using this edge on their shortest-paths, in order to \imp{reduce their affinities $s_{st}$}. \\
+		Each transfer edge $(i,j)$ transmits this proportion $p^\text{allowed}_{ij}$ to all couples $s,t$ that use this edge.
+Each couple $s,t$ receiving a reduced proportion have to \imp{reduce its affinity $s_{st}$}.
+\\
 		\vspace{0.4cm}
-		Each couple $s,t$, will recieve a \imp{list of proportion of allowed flow}, from all transfer edges on its shortest-path.
+		Each couple $s,t$ receives a \imp{list of proportion of allowed flow}, from all transfer edges on its shortest-path.
 The \imp{update factor for the affinity $s_{st}$} is constructed by using the minimum allowed flow received.
@@
 
-		As it is still a work in progress, the conclusion will take the form of a to-do list:
+		As it is still a work in progress, the conclusion takes the form of a to-do list:
 		\begin{itemize}
@@
 			\item Running the algorithm on \imp{all lines}.
+			\item Testing multiple \imp{initial conditions} and \imp{hyperparameter values}.  
+			\item Finding \imp{applications}.
+			\item Making links with \imp{gravity model}, \imp{transportation problem}, ...
 			\item Using carefully crafted \imp{OD affinities}.
-			\item Finding \imp{applications}.
-			\item Finding a \imp{proof of convergence}. 
-			\item Finding if the solution is \imp{unique} (doubt it).
-			\item Making links with \imp{gravity model}, \imp{transportation problem}, ...
+			\item Finding \imp{proof/conditions of convergence}.
 			\item \imp{Adapting} the algorithm to closely related problems.

### 2023-01-10 paper_complex_networks/paper_complex_networks_v1.tex
Add files via upload

%%
\documentclass{llncs}
%
\usepackage{mathptmx}       % selects Times Roman as basic font
\usepackage{helvet}         % selects Helvetica as sans-serif font
\usepackage{courier}        % selects Courier as typewriter font
\usepackage{type1cm}        % activate if the above 3 fonts are not available on your system
\usepackage{makeidx}         % allows index generation
\usepackage{graphicx}        % standard LaTeX graphics tool when including figure files
\usepackage{multicol}        % used for the two-column index
\usepackage[bottom]{footmisc}% places footnotes at page bottom
\usepackage{subfig}
\usepackage{amsfonts}
\usepackage[cmex10]{amsmath}
\usepackage{float}

\usepackage{bm} 
\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{caption}

\begin{document}

\title{Estimation of flow trajectories in a multi-lines network}
%
\titlerunning{Multiple lines network}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Guillaume Guex, Romain Loup \and
	François Bavaud}
%
\authorrunning{Guex et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Guillaume Guex, Romain Loup,François Bavaud}
%
\institute{Université de Lausanne, CH-1015,  Lausanne, Suisse\\
    \email{guillaume.guex@unil.ch},
	\email{romain.loup@unil.ch},
	\email{francois.bavaud@unil.ch}}

\maketitle

\section{Introduction}

\section{Notations and formalism}
\subsection{Lines and junctions}
\label{Lines and junctions}
Consider a transportation network made of bus lines numbered $\ell=1,\ldots, q$, of respective lengths (number of stops) $l_\ell$.  Opposite lines, that is parallel lines running in the back and forth directions are considered as distinct.

The $l=\sum_{\ell=1}^ql_\ell$ bus stops constitute the nodes of the transportation network. Each stop $i=1,\ldots,l$ belongs to a single bus line, and defines a unique next or forward stop $F(i)$ (unless $i$ is the line terminus) and a unique backward stop $B(i)$ (unless $i$ is the line start), both on the same line.

Junctions (super-stops), denoted $S=1,\ldots, L$, are defined as cliques (equivalence classes) of stops which can be mutually reached from one another within walking distance. True junctions 
involve at least two lines, while pseudo-junctions comprise isolated stops $i$, possibly together with their opposite stop (noted $i\text{-}$) on the parallel line running in the opposite direction.

In summary, each stop $i$ belongs to a single line $\ell(i)$ and a single junction $S(i)$. Lines and junctions define two partitions of the network nodes, with $q$, respectively $L$ blocks.

\subsection{Line edges, transfer edges and trips}
\label{Line edges, transfer edges and trips}
Two sorts of oriented edges are involved in the transportation network: 
\begin{enumerate}
  \item[$\bullet$] intra-line edges $(i,j)=(i,F(i))$ belonging to a single line  $\ell(i)=\ell(j)$
  \item[$\bullet$] inter-line or transfer edges $(i,j)$ connecting different lines $\ell(i)\neq \ell(j)$, taking place at some junction $S(i)=S(j)$. 
  \end{enumerate}
A $st$-trip, noted $[s,t]$, consists of entering into the network at stop $s$, and leaving the network at $t$, by following the shortest-path (i.e. achieving the minimum distance,  minimum time, or  minimum cost), supposed unique, leading to $s$ from $t$.

The succession of edges $(ij)$ belonging to the $st$-trip, noted $(ij)\in [s,t]$, is unique. Define the edge-trip incidence matrix as
\begin{equation}
\label{edgetrip}
\chi_{ij}^{st} = \begin{cases}
  1    & \text{if $(ij)\in [s,t]$}, \\
  0    & \text{otherwise}.
\end{cases}
\end{equation}
A $st$-trip always starts with the edge $(s,F(s))$, and finishes with $(B(t),t)$. Transfers can occur in-between, but never at the beginning nor at the end of the trip.

\vspace*{0.1cm}

Similarly, $ST$-trips between junctions, noted $[S,T]$, consist of the supposedly unique shortest $st$-trip $[s,t]$ among all  $s\in S$ and $t\in T$, followed by a traveller entering the network as junction $S$, and leaving it at junction $T$. The corresponding edge-trip incidence matrix is 
\begin{equation}
\label{edgetripjunction}
\tilde{\chi}_{ij}^{ST} = \begin{cases}
  1    & \text{if $(ij)\in [S,T]$}, \\
  0    & \text{otherwise}.
\end{cases}
\end{equation}
Unicity of the shortest-path implies that a passenger wishing to travel from $S$ to $T$ chooses the unique stop $\sigma(S,T)\in S$  as initial node (on line $\ell(\sigma)$), and the unique node $\tau(S,T)\in T$ as final node (on line $\ell(\tau)$).

\subsection{Transportation flows}
\label{Transportation flows}
Let  $x_{ij}$ count the number of travelers using edge $(ij)$ in a given period, such as a given hour, day, week or  year.  The edge flow $x_{ij}$ is denoted by $y_{ij}$ for an intra-line edge $(i,j)$, and 
by $z_{ij}$ for a transfer edge $(i,j)$. By construction, $x_{ij}=y_{ij}+z_{ij}$, where $y_{ij}\,  z_{ij}=0$.

\vspace*{0.1cm}

Let $a_i$, respectively $b_i$, the number of passengers embarking, respectively disembarking at stop $i$. By construction, 
\begin{equation}
\label{bilan1ligne}
\begin{cases}
 y_{i,F(i)}=a_i \text{\,  and } b_i=0   & \text{if $i$ is a line start}, \\
y_{B(i),i}=b_i \text{\,  and } a_i=0   & \text{if $i$ is a line terminus}, \\
 y_{i,F(i)}=y_{B(i),i}+a_i-b_i     & \text{otherwise}.
\end{cases}
\end{equation}
Also, $\mathbf{a}$ and $\mathbf{b}$ must be consistent, in the sense that $A_i\ge B_i$, where $A_i$ (respectively $B_i$) is the cumulated number of embarked 
(resp. disembarked) passengers on the line under consideration, recursively defined as $A_{F(i)}=A_i+a_i$ (resp. $B_{F(i)}=B_i+b_i$). Moreover,  $A_i=B_i$ at a terminal line stop $i$. This common value yields  the total number of passengers transported by the line.

\vspace*{0.1cm}

Let the transportation flow $n_{st}$ denote the number of passengers following an $st$-trip, that is entering the network at $s$ and leaving the network at $t$ by using the shortest path.

Also, 
at the junction level, let $\tilde{n}_{ST}$ denote the number of passengers following an $ST$-trip. As implied by the end of section (\ref{Line edges, transfer edges and trips}), 
\begin{equation}
\label{equivnntilde}
\tilde{n}_{ST}=\sum_{s\in S}\sum_{t\in T}n_{st} \qquad\qquad\qquad n_{st}=\sum_{ST}\tilde{n}_{ST}\: \delta_{s,\sigma(S,T)}\: \delta_{t,\tau(S,T)}
\end{equation}
Definitions (\ref{edgetrip}) and (\ref{edgetripjunction}) yield
\begin{equation}
\label{equationGG}
x_{ij}=\sum_{st}\chi_{ij}^{st}\:  n_{st}=\sum_{ST}\tilde{\chi}_{ij}^{ST}\:  \tilde{n}_{ST}
\end{equation}
Among the passengers embarking in $i$, some transfer from another line, and some others enter into the network: 
\begin{equation}
\label{entrer}
a_i=z_{\bullet i}+n_{i\bullet}
\end{equation}
where  ``$\bullet$" denotes the summation over the replaced index, as in $n_{i\bullet}=\sum_{j=1}^l n_{ij}$. Similarly, among the passengers disembarking in $i$, some transfer to another line, and some others leave the network: 
\begin{equation}
\label{sortir}
b_i=z_{i\bullet}+n_{\bullet i}
\end{equation}
By construction
\begin{displaymath}
a_{\bullet}=b_{\bullet}=z_{\bullet\bullet}+n_{\bullet\bullet}
\end{displaymath}
where $n_{\bullet\bullet}$ counts the number of passengers, and $z_{\bullet\bullet}$ counts the number of transfers. $z_{\bullet\bullet}/n_{\bullet\bullet}$  is the average number of transfers per passenger.

\vspace*{0.1cm}

As explained in section \ref{Lines and junctions}, transfers can only occur at junctions, that is $z_{ij}>0$ implies $S(i)=S(j)$. As $z_{ii}=0$ (no traveller is supposed to disembark and re-embark later at the same stop), and $z_{ii\text{-}}=0$ (where $i\text{-}$ is the stop opposite to $i$, when existing), transfers can occur at true junctions only.

\vspace*{0.1cm}

At the junction level, let 
\begin{displaymath}
\tilde{a}_I=\sum_{i\in I} a_i\enspace,\enspace\tilde{b}_I=\sum_{i\in I} b_i\enspace,\enspace\tilde{z}_{\bullet I}=\sum_{i\in I} z_{\bullet i}\enspace,\enspace\tilde{z}_{I\bullet}=\sum_{i\in I} z_{i\bullet}\enspace,\enspace\tilde{n}_{I\bullet}=\sum_{i\in I} n_{i\bullet}\enspace,\enspace\tilde{n}_{\bullet I}=\sum_{i\in I} n_{\bullet i}
\end{displaymath}
 respectively denote the number of embarking (disembarking)  passengers at junction~$I$; of embarking passengers after a transfer and of disembarking passengers before a transfer at  $I$; of passengers starting (ending) their trip at $I$. Then 
\begin{equation}
\label{junctionLevel}
\tilde{a}_I=\tilde{z}_{\bullet I}+\tilde{n}_{I\bullet}\qquad\qquad \tilde{b}_I=\tilde{z}_{I\bullet}+\tilde{n}_{\bullet I}\qquad\qquad \tilde{z}_{\bullet I}=\tilde{z}_{I\bullet}\le \min(\tilde{a}_I,\tilde{b}_I)\enspace. 
\end{equation}

\subsection{Statement of the problem}
Automatic passenger counters measure the number of passengers entering and leaving buses at each stop [Boyle, 1998], that is $\mathbf{a}$ and $\mathbf{b}$. Also, the geometry of the network permits to derive the edge-trip incidence matrix $\bm{\chi}$ (\ref{edgetrip}), as well as the partition $\mathcal{S}$  of stops nested within junctions.

Intra-line edge flows $\mathbf{Y}=(y_{ij})$ can be determined by (\ref{bilan1ligne}), but transfer edge flows $\mathbf{Z}=(z_{ij})$ are, here and typically, unknown. The objective is to estimate the $l\times l$ transportation flow $\mathbf{N}=(n_{st})$ at the stop level, or equivalently  the $L\times L$  transportation flow  $\tilde{\mathbf{N}}=(\tilde{n}_{ST})$ at the junction level. Many consistent solutions coexist in general, even for a single line with no transferts (section \ref{Single line}).

This issue of incompletely observed data can be tackled by the maximum entropy formalism: let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution). Let $\mathcal{F}$ consist of the empirical distributions $\mathbf{f}$ satisfying $R$ linear constraints of the form $\sum_{st}f_{st}\, o^r_{st}=\bar{o}^r$ for $r=1,\ldots, R$. Then the minimum Kullback-Leibler divergence 
\begin{displaymath}
\min_{\mathbf{f}\in\mathcal{F}}\;  \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}}
\end{displaymath}
is reached for the maximum entropy solution  
\begin{equation}
\label{MaxEnt}
f^0_{st}=\frac{1}{Z(\bm{\lambda})}\: g_{st}\exp(-\sum_{r=1}^R \lambda_r o^r_{st})\qquad\qquad Z(\bm{\lambda})=\sum_{st}g_{st}\exp(-\sum_{r=1}^R \lambda_r o^r_{st})
\end{equation}
where the Lagrange multipliers $\bm{\lambda}$ are determined  so as to satisfy the $R$ linear constraints (**see e.g. Bavaud and *** and references therein).

\section{Single line}
\label{Single line}
Let  $i=1,\ldots, l$ enumerate the bus stops in increasing order,  i.e. $F(i)=i+1$. Define $g_{st}=\frac{1}{(l-1)(l-2)}\chi(s<t)$ (where $\chi(.)$ denotes the 0/1  indicator function) as the maximally uniform prior, reflecting only the unidirectional nature of trips. 
Constraints $n_{r\bullet}=a_r$ (resp. $n_{\bullet r}=b_r$), correspond to $o^r_{st}=
\delta_{sr}$ (resp. $o^r_{st}=\delta_{tr}$) in (\ref{MaxEnt}), which finally yields, after reparametrization, the maximum entropy flow reads
\begin{equation}
\label{nosignle}
n^0_{st}=I(s<t)\, c_s\, d_t \qquad\qquad\mbox{where}\quad \sum_{s<t}c_s\, d_t=n_{\bullet\bullet}=a_{\bullet}=b_\bullet
\end{equation}
In addition, the (dis-)embarking constraints  yield
\begin{equation}
\label{(dis embarking constraints}
a_i=c_i \sum_{t>i}d_t=c_i\, D_i 
\qquad\qquad\qquad
b_j=d_j\sum_{s<j}c_s=d_j \, C_j  
\end{equation}
where $D_i=\sum_{t>i}d_t$ and $C_j=\sum_{s<j}c_s$. This maximum entropy solution constitutes one possible consistent transportation flow among many others, such as the ``first in, first out" (FIFO) flow. Interestingly enough, (\ref{nosignle}) is reminiscent of the so-called gravity flows of quantitative Geography (**ref**) $n_{st}=h_{st}\, c_s\, d_t $, but with a purely asymmetric ``distance deterrence function" $h_{st}=I(s<t)$.  Also, 
(\ref{nosignle}) shows that  the conditional probability to exit on $t$, given an entrance on $s$, is $w_{st}=n^0_{st}/n^0_{s\bullet}=I(s<t)\, d_t/D_s$.

\

*** ici la (les) figure de l'exemple  ``starting from Maladière, Riant-Cour, Dapples"... ? ***

\

*** ici l'équivalence avec l'approche chaîne de Markov de Guillaume ? ***

\section{Multiple lines}

*** $a_{\bullet}=b_{\bullet}$ (where ``$\bullet$" denotes summation over the replaced index).

\newpage

\section*{Previous submission to complex networks 2022}

%
\section*{Statement of the problem}
%

Automatic passenger counters measure the number of passengers entering and leaving buses at each stop \cite{boyle_passenger_1998}. Given this information, can we estimate the complete trajectories of passengers within the entire multi-line network? This communication attempts to propose an estimation of all passenger trajectories in the multi-line network with an algorithm based on iterative proportional fitting (IPF) \cite{bishop_discrete_2007}.

The exploited dataset is provided by the Lausanne Transportation Agency (tl) in Switzerland. The dataset includes 42 lines of buses (or subways) and more than 1361 stops, including 497 clusters of stops (Fig. \ref{fig:map-lausanne}),  carrying around 115 million passengers in 2019. Each stop refers to a single directed line, and return lines are considered as distinct.  In addition to  \emph{line edges},  it is possible to construct pedestrian \emph{transfer edges} (Fig. \ref{fig:network-lausanne}) to make the graph unilaterally connected by considering, e.g., clusters of stops connected.

\noindent \begin{minipage}[c]{0.5\textwidth}
   \begin{figure}[H]
      \centering
      \includegraphics[width=6cm]{img/stop_area_c.png}
      \caption{Cluster of stops} 
      \label{fig:map-lausanne}
   \end{figure}
\end{minipage}%
\begin{minipage}[c]{0.5\textwidth}
   \begin{figure}[H]
      \centering
      \includegraphics[width=4cm]{img/edge_type3.png}
      \caption{The multi-line network structure} 
      \label{fig:network-lausanne}
   \end{figure}
\end{minipage}

$\mbox{}$

Knowing only the network structure and the number of passengers embarking and disembarking at each stop, how can we infer the most probable passenger trajectories in the network? Before examining the general multi-line network problem,  we address the estimation of trajectories on a single line.

%
\section{Single line}
%

Consider a one-directional line with $n$ stops indexed regarding the order found in the line.  Let $\mathbf{a}_\text{in} = (a_{s}^\text{in})$ and $\mathbf{a}_\text{out} = (a_{t}^\text{out})$ be two vectors representing, respectively, the passengers entering and leaving the line at each stop. The goal is to estimate the $(n \times n)$ origin-destination matrix $\mathbf{N} = (n_{st})$, where $n_{st}$ represents the number of passengers entering the line at $s$ and leaving at $t$, subject to constraints $n_{s\bullet} = a^\text{in}_s$ and $n_{\bullet t} = a^\text{out}_t$.  Among many feasible solutions, arguably the most elegant one is the maximum entropy solution, which can be derived  from different means:  (i) passenger flows can be modelled using a Markovian assumption, which translates by assuming that every passenger has the same probability to continue the trip after having travelled at last one stop ; (ii) an iterative proportional fitting algorithm can be performed, starting with an initial origin-destination affinity matrix $\mathbf{S} = (s_{st})$, defined as the upper triangular $n\times n$ matrix filled with $1$, and then iterated to satisfy the margin constraints given by $\mathbf{a}_\text{in}$ and $\mathbf{a}_\text{out}$. Both approaches give the same solution, but only the latter remains pertinent in the multi-line problem.

%
\section{Multi-lines}
%

In the multi-line problem, a passenger can transfer from a line to another. The problem cannot be tackled with Markov chain modelling anymore, which generate unrealistic random trajectories. Instead, we will assume that passengers follow shortest paths. Starting from an origin-destination matrix $\mathbf{N} = (n_{st})$, where $s$ denotes the stop at which a passenger \emph{enters into the network} (and not simply enters a particular line), and $t$ denotes the stop where the passenger \emph{leaves definitively the network}, this shortest paths assumption allows us to compute the flow matrix on edges $\mathbf{X} = (x_{ij})$. The latter decomposes into the within-line flow and the transfer flow, i.e., $\mathbf{X}=\mathbf{X}_\text{W}+\mathbf{X}_\text{B}$. 
Moreover, in the multi-line problem, we also have to distinguish between:
\begin{itemize}
	\item[$\bullet$]  passengers who enter and leave bus lines at each stop, represented by vectors $\mathbf{a}_\text{in}$ and $\mathbf{a}_\text{out}$, which are \emph{measured},
	\item[$\bullet$] and passengers who enter and leave the network at each stop $i$, represented by the \emph{unknown} quantities $n_{\bullet i}$ and $n_{i \bullet}$.
\end{itemize}
By construction, 
\begin{align}
		a^\text{in}_i &=n_{i\bullet} + x^\text{B}_{i\bullet},  &
		a^\text{out}_i &= n_{\bullet i} + x^\text{B}_{\bullet i},
\end{align}

Using these two constraints, along with the shortest paths assumption and iterative proportional fitting, we propose the following iterative algorithm in order to find $\mathbf{N}$ from measured $\mathbf{a}_\text{in}$ and $\mathbf{a}_\text{out}$. \\

\noindent \fbox{\begin{minipage}{\textwidth}

\textbf{Initialisation:} $\mathbf{S}^{(0)}$ is filled with 1 excepted for aberrant origin-destination pairs (such as $t$ beeing a previous stop of the same line as $s$). The margins of $\mathbf{N}$ are fixed as $\mathbf{n}^{(0)}_{\mbox{\scriptsize in}} = \mathbf{a}_\text{in}$ and $\mathbf{n}^{(0)}_{\mbox{\scriptsize out}} = \mathbf{a}_\text{out}$.

\textbf{Step 1, Iterative proportional fitting:}
We use IPF to compute $\mathbf{N}^{(r)}$ starting from $\mathbf{S}^{(r)}$, such that margin constraints, defined by $\mathbf{n}^{(r)}_\text{in}$ and $\mathbf{n}^{(r)}_\text{out}$ are satisfied.

\textbf{Step 2, Shortest paths flow:}
Using shortest paths information, we compute $\mathbf{X}^{(r)}_\text{B}$ from $\mathbf{N}^{(r)}$.

\textbf{Step 3, Affinity and margin update:}
$\mathbf{S}^{(r+1)}$, $\mathbf{n}^{(r+1)}_\text{in}$ and $\mathbf{n}^{(r+1)}_\text{out}$ are updated in order to respect constraints defined by (1).
\end{minipage}}

Step 1, 2, and 3 are iterated until convergence, giving an admissible solution to the problem.
%
\section{A small example}
%

As an illustration, an estimated solution proposed by the algorithm on a restricted network made of four lines only is depicted on Fig. \ref{fig:example-map}. A total of $n_{\bullet \bullet} = 16,837,494$ passengers using this network is estimated by the algorithm. The red circle on the bottom left represents the start of the trip $s$ and the size of the circles at stops $t$ represents the estimated number of passengers terminating their trip at $t$. In this example, the majority of passengers exit the network on the same initial embarkment line. A small fraction of them takes another line.

Table \ref{fig:example-edges} represents the estimated ten most frequented transfer edges. The code of the stop represents the number of the line, the direction and a condensed name of its stop cluster. The third column gives the number (in thousands) of passengers transferring through this edge.

\noindent \begin{minipage}[c]{0.6\textwidth}
   \begin{figure}[H]
      \includegraphics[width=7cm]{img/example_map.png}
      \caption{Example map} 
      \label{fig:example-map}
   \end{figure}
\end{minipage}%
\begin{minipage}[c]{0.43\textwidth}
      %\centering
      \begin{table}[H]
		\begin{tabular}{l|l|r}
		\multicolumn{1}{c|}{\textbf{From stop}} & \multicolumn{1}{c|}{\textbf{To stop}} & \multicolumn{1}{c}{\textbf{Count}} \\ \hline
		{\scriptsize S7\_A\_SF\_O}			& {\scriptsize S9\_A\_SF\_O}			& 192k		\\ \hline
		{\scriptsize S9\_R\_CH\_E}			& {\scriptsize S6\_A\_CH\_E}			& 187k		\\ \hline
		{\scriptsize S7\_A\_SF\_O}		& {\scriptsize S8\_R\_SF\_S}				& 135k			\\ \hline
		{\scriptsize S6\_R\_CH\_O}			& {\scriptsize S9\_A\_CH\_O}			& 135k			\\ \hline
		{\scriptsize S8\_A\_GTE\_N}		& {\scriptsize S9\_R\_GTE\_E}			& 103k		\\ \hline
		{\scriptsize S9\_R\_B-AIR\_C}		& {\scriptsize S8\_A\_B-AIR\_N}		& 99k			\\ \hline
		{\scriptsize S9\_A\_SF\_O}			& {\scriptsize S7\_A\_SF\_O}			& 88k			\\ \hline
		{\scriptsize S8\_R\_B-AIR\_D}		& {\scriptsize S6\_A\_B-AIR\_C}		& 87k			\\ \hline
		{\scriptsize S6\_R\_SF\_O}			& {\scriptsize S8\_A\_SF\_O}			& 86k			\\ \hline
		{\scriptsize S9\_A\_GTE\_O}		& {\scriptsize S8\_R\_GTE\_S}		& 84k                                        
		\end{tabular}
\vspace*{0.2cm}	\caption{List of the ten most\\ frequented transfer edges} 
	\label{fig:example-edges}
	\end{table}
\end{minipage}

\vspace*{0.2cm}

The current work performs computer-intensive simulations of flow over the entire network (1361 stops), permitting to extract usual network indices (centrality, betweenness...) characterizing both the stops \emph{and} the lines. In parallel, the computational effects of various fine tuning calibration parameters used in the algorithm are investigated.

\bibliographystyle{apalike}
\bibliography{paper_complex_networks}

\end{document}

### 2023-01-13 multiple_lines/local_functions_v2.R
toy example graphics

@@ -571,7 +571,9 @@ compute_x_from_n = function(n_mat, edge_ref, sp_ref, p_mat){
 # - None, only plotting
 #-------------------------------------------------------------------------------
 
-plot_flow_graph = function(adj, n_mat, layout=NULL){
+plot_flow_graph = function(adj, n_mat, layout=NULL, main=NULL){
+  
+  igraph_options(annotate.plot=TRUE)
   
   df_line = as_data_frame(graph_from_adjacency_matrix(adj, weighted = TRUE))
   df_line["type"] = "line"
@@ -589,14 +591,13 @@ plot_flow_graph = function(adj, n_mat, layout=NULL){
   n_edges = length(E(g_tot))
   
   std_weights = E(g_tot)$weight[E(g_tot)$type == "flow"]
-  std_weights = std_weights - min(std_weights) 
-  std_weights = std_weights / max(std_weights) * 0.9 + 0.1
+  std_weights = std_weights / max(std_weights)
   
   color_green_fn = colorRamp(c("white", "forestgreen"))
   color_red_fn = colorRamp(c("white", "red"))
   color_blue_fn = colorRamp(c("white", "blue"))
   
-  edge_sizes = (E(g_tot)$type == "line") * 3
+  edge_sizes = (E(g_tot)$type == "line") * 0.5
   edge_sizes[E(g_tot)$type == "flow"] = std_weights * 2
   edge_curves = (E(g_tot)$type == "line") * 0.2
   edge_colors = rep("black", n_edges)
@@ -622,8 +623,8 @@ plot_flow_graph = function(adj, n_mat, layout=NULL){
   in_out_colors = setNames(in_out_colors, names(in_prop))
   
   plot(g_tot, layout=layout, edge.color=edge_colors, edge.width=edge_sizes,
-       edge.curved=edge_curves, edge.arrow.size=0.5, vertex.shape="pie", 
-       vertex.pie=pie_prop, vertex.pie.color=in_out_colors)
+       edge.curved=edge_curves, edge.arrow.size=0.1, vertex.shape="pie", 
+       vertex.pie=pie_prop, vertex.pie.color=in_out_colors, main=main)
 }
 
 #-------------------------------------------------------------------------------

### 2023-01-13 multiple_lines/testing_toy.R

@@ -19,7 +19,7 @@ sp_ref = edge_ref_p_mat_sp_ref(adj)[[3]]
 
 # Testing with a case
 
-set.seed(9)
+#set.seed(2)
 n_real = n_multin(paths, 100)
 n_real = n_real / sum(n_real)
 x_btw = compute_x_from_n(n_real, edge_ref, sp_ref, p_mat)$x_btw
@@ -30,7 +30,7 @@ rho_out = colSums(n_real) + rowSums(x_btw)
 
 epsilon = 1e-10
 conv_thres_if = 0.00001
-prop_limit = 0.2
+prop_limit = 0.3
 
 # INIT
 
@@ -40,102 +40,123 @@ sigma_out = rho_out
 
 # ------- ALGO (to iterate)
 
-# 1) Computing if
-
-b = rep(1, ncol(s_it_mat))
-converge_if = F
-while(!converge_if){
-  b_old = b
-  a = (sigma_in + epsilon) / colSums(t(s_it_mat + epsilon) * b)
-  b = (sigma_out + epsilon) / colSums((s_it_mat + epsilon) * a)
-  if(sum(abs(b_old - b)) < conv_thres_if){
-    converge_if = T
-  }
-}
-n_mat = t(b * t(a * (s_it_mat + epsilon)))
-
-# 2) Compute the flow
-
-n = length(rho_in)
-sp_order_ref = mapply(function(i, j) i+n*(j-1), sp_ref[, 1], sp_ref[, 2])
-where_edge_btw = as.logical(edge_ref[, 3])
-p_btw_mat = p_mat[, where_edge_btw]
-edge_btw_ref = edge_ref[where_edge_btw, 1:2]
-
-sp_flow_vec = as.vector(n_mat)[sp_order_ref]
-# Compute the flow on edges
-btw_edge_flow = as.vector(t(p_btw_mat) %*% sp_flow_vec)
-# Compute the in-out between flow on nodes 
-x_btw = sparseMatrix(i=edge_btw_ref[, 1], j=edge_btw_ref[, 2], 
-                     x=btw_edge_flow, dims=c(n, n))
-
-# 3) flow on nodes
-
-node_in_btw = colSums(x_btw)
-node_out_btw = rowSums(x_btw)
-
-# 4) limits on nodes
-
-# Copy the vector
-allowed_in_btw = node_in_btw
-allowed_out_btw = node_out_btw
-# Replace the overflowing value with the limit
[…]

### 2023-01-13 paper_complex_networks/paper_complex_networks_v1.tex
add another statement of the problem

@@
 
+\subsection{Statement of the problem (version 2, added by GG)}
+Let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution).
+Also suppose we know the margins distribution of the real transportation flow $n^\star_{st}$, i.e.
+$c_s = n^\star_{s \bullet} / n^\star_{\bullet \bullet}$ and $d_t = n^\star_{\bullet t} / n^\star_{\bullet \bullet}$.
+We are interested in finding
+\begin{align}
+	\label{constr_MaxEnt}
+	\min_{\mathbf{f}\in\mathcal{F}} &\; \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}}, \notag \\
+	s.t. &\; \sum_t f_{st} = c_s, \notag \\
+	&\; \sum_s f_{st} = d_t.
+\end{align}
+The Lagragian is
+\begin{equation}
+	\label{Lag}
+	L = \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}} - \sum_s \lambda_s (c_s - \sum_t f_{st}) - \sum_t \mu_t (d_t - \sum_s f_{st}),
+\end{equation}
+which gives, after deriving,
+\begin{equation}
+	\label{Sol}
+	f_{st} = \alpha_s \beta_t g_{st} \qquad \text{with } \alpha_s := \exp(- 1 - \lambda_s) \text{, } \beta_t := \exp(- \mu_t).
+\end{equation}
+Using constraints in (\ref{constr_MaxEnt}), we find
+\begin{equation}
+	\label{Sol_LagMult}
+	\alpha_s = \frac{c_s}{\sum_t \beta_t g_{st}}, \qquad \beta_t = \frac{d_t}{\sum_s \alpha_s g_{st}}, 
+\end{equation}
+which gives an \emph{iterative fitting algorithm}
+\begin{equation}
+	\label{Iterative fitting}
+	\alpha^{(\iota)}_s = \frac{c_s}{\sum_t \beta^{(\iota)}_t g_{st}}, \qquad \beta^{(\iota + 1)}_t = \frac{d_t}{\sum_s \alpha^{(\iota)}_s g_{st}}, 
+\end{equation}
+
 \section{Single line}

### 2023-01-13 paper_complex_networks/paper_complex_networks_v1.tex
update notes

@@
 Let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution).
-Also suppose we know the margins distribution of the real transportation flow $n^\star_{st}$, i.e.
-$c_s = n^\star_{s \bullet} / n^\star_{\bullet \bullet}$ and $d_t = n^\star_{\bullet t} / n^\star_{\bullet \bullet}$.
+Also suppose a prior guess of row and margin constraints for $f_{st}$, respectively $c_s$ and $d_t$, which has to verify
+\begin{equation}
+	\frac{c_i}{c_j} = \frac{a_i}{a_j} \qquad \frac{d_i}{d_j} = \frac{b_i}{b_j}  \qquad \forall i, j \text{ without junction.} \\
+\end{equation}
+Setting these gives us
+\begin{equation}
+	n_{\bullet \bullet} = \frac{c_i}{a_i} = \frac{d_i}{b_i} \qquad \forall i \text{ without junction.}  
+\end{equation}
 We are interested in finding
@@
 \begin{equation}
-	\label{Lag}
-	L = \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}} - \sum_s \lambda_s (c_s - \sum_t f_{st}) - \sum_t \mu_t (d_t - \sum_s f_{st}),
+	L = \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}} - \sum_s \lambda_s (c_s - \sum_t f_{st}) - \sum_t \mu_t (d_t - \sum_s f_{st}), \notag
 \end{equation}
-which gives, after deriving,
+which gives, after deriving and setting to zero,
 \begin{equation}

### 2023-01-14 paper_complex_networks/paper_complex_networks_v1.tex
update article

--- 2023-01-10 paper_complex_networks/paper_complex_networks_v1.tex
@@
 
+\subsection{Statement of the problem (version 2, added by GG)}
+Let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution).
+Also suppose a prior guess of row and margin constraints for $f_{st}$, respectively $\alpha_s$ and $\beta_t$, which has to verify
+\begin{equation}
+	\label{init_margin}
+	\frac{\alpha_i}{\alpha_j} = \frac{a_i}{a_j} \qquad \frac{\beta_i}{\beta_j} = \frac{b_i}{b_j}  \qquad \forall i, j \text{ which are not true junctions.} \\
+\end{equation}
+Setting these gives us
+\begin{equation}
+	\label{flow_sum}
+	n_{\bullet \bullet} = \frac{a_i}{\alpha_i} = \frac{b_i}{\beta_i} \qquad \forall i \text{ which are not true junctions.}  
+\end{equation}
+We are interested in finding
+\begin{align}
+	\label{constr_MaxEnt}
+	\min_{\mathbf{f}\in\mathcal{F}} &\; \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}}, \notag \\
+	s.t. &\; \sum_t f_{st} = \alpha_s, \notag \\
+	&\; \sum_s f_{st} = \beta_t.
+\end{align}
+The Lagragian is
+\begin{equation}
+	L = \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}} - \sum_s \lambda_s (\alpha_s - \sum_t f_{st}) - \sum_t \mu_t (\beta_t - \sum_s f_{st}), \notag
+\end{equation}
+which gives, after deriving and setting to zero,
+\begin{equation}
+	\label{Sol}
+	f_{st} = \phi_s \psi_t g_{st} \qquad \text{with } \phi_s := \exp(- 1 - \lambda_s) \text{, } \psi_t := \exp(- \mu_t).
+\end{equation}
+Using constraints in (\ref{constr_MaxEnt}), we find
+\begin{equation}
+	\label{Sol_LagMult}
+	\phi_s = \frac{\alpha_s}{\sum_t \psi_t g_{st}}, \qquad \psi_t = \frac{\beta_t}{\sum_s \phi_s g_{st}}, 
+\end{equation}
+which gives an \emph{iterative fitting algorithm}. Starting with any $\psi^{(0)}_t > 0$, we do
+\begin{equation}
+	\label{Iterative fitting}
+	\phi^{(\iota)}_s = \frac{\alpha_s}{\sum_t \psi^{(\iota)}_t g_{st}}, \qquad \psi^{(\iota + 1)}_t = \frac{\beta_t}{\sum_s \phi^{(\iota)}_s g_{st}}, 
+\end{equation}
+
+\subsection{Margin distributions update (added by GG)}
+
+With equation (\ref{flow_sum}), we can obtain the flow associated to $f_{st}$
+\begin{equation}
+	\label{flow_from_distrib}
+	n_{st} = n_{\bullet \bullet} f_{st}, 
+\end{equation}
+and with the edge-trip incidence matrix $\bm{\chi}$ (\ref{edgetrip}), we obtain the intra-line edge flows $\mathbf{Y}=(y_{ij})$ and the transfer edge flows $\mathbf{Z}=(z_{ij})$. \\
+
+We now have to make sure constraints (\ref{entrer}) and (\ref{sortir}) are verified. The new margins distribution are set to 
+\begin{align}
+	\widetilde{\alpha}_s &= \frac{\min(0, a_s - z_{\bullet s})}{\sum_{s'} \min(0, a_{s'} - z_{\bullet {s'}})} \label{alpha_update} \\
+	\widetilde{\beta}_t &= \frac{\min(0, b_t - z_{t \bullet})}{\sum_{t'} \min(0, b_{t'} - z_{{t'} \bullet})} \label{beta_update}
+\end{align}
+
+\subsection{Prior distribution update (added by GG)}
+
+Transfer edges $(i, j)$ where $z_{i \bullet} > b_i$ or $z_{\bullet j} > a_j$ typically denote an overflowing tansfer flow regarding the $g_{st}$ prior distribution.
+The latter must be adjusted in order to have $\mathbf{Z}$ comptible with constraints.
+For any edge $(i, j)$, let us compute the \emph{flow ratio} $r_{ij}$, with
+\begin{equation}
+	\label{flow_ratio}
+	r_{ij} = \max \left(1, \frac{z_{i \bullet}}{b_i}, \frac{z_{\bullet j}}{a_j} \right),
+\end{equation}
+where $r_{ij} > 1$ typically denotes an overflow through edge $(i, j)$.
+This flow ratio can be traced back to origin-destination couples $s, t$ in order to compute the \emph{orgin-destination flow ratio} $\bar{r}_{st}$
+\begin{equation}
+	\label{st_flow_ratio}
+	\bar{r}_{st} = \max_{ij} \chi_{ij}^{st} r_{ij},
+\end{equation}
+where again $\bar{r}_{st} > 1$ denotes an overflow between $s$ and $t$. To ajdust the flow, we can devide the previous flow by this ratio
+\begin{equation}
+	\label{update_flow}
+	\widetilde{n}_{st} =\frac{n_{st}}{\bar{r}_{st}}
+\end{equation}
+And the new prior distribution should follow 
+\begin{equation}
+	\label{update_distrib}
+	\widetilde{g}_{st} = \frac{\left( \frac{\widetilde{n}_{st}}{\phi_s \psi_t} \right)}{\sum_{s',t'} \left( \frac{\widetilde{n}_{s',t'}}{\phi_{s'} \psi_{t'}} \right)}
+\end{equation}
+
+With the new prior distribution $\widetilde{g}_{st}$ and the new margin distributions $\widetilde{\alpha}_s$, $\widetilde{\beta}_t$, we can begin a new iteration step by going through 2.5 -2.7 again.
+
 \section{Single line}

### 2023-01-14 multiple_lines/local_functions_v2.R
begin the new algorithm

@@ -427,7 +427,188 @@ compute_origin_destination = function(rho_in, rho_out, edge_ref, sp_ref, p_mat,
         node_in_btw[allowed_in_btw > node_in_btw]
       allowed_out_btw[allowed_out_btw > node_out_btw] = 
         node_out_btw[allowed_out_btw > node_out_btw]
-    # Else, set the limit to a crisp limit 
+      # Else, set the limit to a crisp limit 
+    } else {
+      # Copy the vector
+      allowed_in_btw = node_in_btw
+      allowed_out_btw = node_out_btw
+      # Replace the overflowing value with the limit
+      allowed_in_btw[allowed_in_btw > rho_in*(1 - prop_limit)] =
+        rho_in[allowed_in_btw > rho_in*(1 - prop_limit)]*(1 - prop_limit)
+      allowed_out_btw[allowed_out_btw > rho_out*(1 - prop_limit)] =
+        rho_out[allowed_out_btw > rho_out*(1 - prop_limit)]*(1 - prop_limit)
+    }
+    
+    # --- Compute the allowed flow on edge
+    
+    # Compute the proportion of allowed flow
+    p_allowed_in = allowed_in_btw / (node_in_btw + epsilon)
+    p_allowed_out = allowed_out_btw / (node_out_btw + epsilon)
+    p_allowed_in[p_allowed_in == 0] = 1
+    p_allowed_out[p_allowed_out == 0] = 1
+    # Compute the allowed flow on edges
+    p_allowed_min = mapply(
+      function(i, j) min(p_allowed_out[i], p_allowed_in[j]), 
+      edge_btw_ref[ ,1], edge_btw_ref[,2])
+    allowed_edge_flow = p_allowed_min * btw_edge_flow
+    
+    # --- Update the flow entering and leaving the network
+    
+    # Set the allowed flow in a matrix
+    allowed_flow_mat = sparseMatrix(i=edge_btw_ref[, 1], j=edge_btw_ref[, 2],
+                                    x=allowed_edge_flow, dims=c(n, n))
+    # Update with margins
+    sigma_in = rho_in - colSums(allowed_flow_mat)
+    sigma_out = rho_out - rowSums(allowed_flow_mat)
+    
+    # --- Compute the reducing factor and update the affinity matrix
+    
+    # Compute the excess proportion of flow
+    p_to_red = (btw_edge_flow - allowed_edge_flow) / (btw_edge_flow + epsilon)
[…]

### 2023-02-02 multiple_lines/local_functions.R
Add toy example network creation and computation

@@ -994,3 +994,92 @@ n_multin = function(paths, n_passagers, p_pos = rep(1/sum(paths), sum(paths)),
   n_drawn[n_drawn == 1] = drawn
   return(n_drawn)
 }
+
+#-------------------------------------------------------------------------------
+# Build and compute toy example
+#-------------------------------------------------------------------------------
+# Each line crosses all other lines
+
+
+
+# Build the toy network according to th number of lines
+# - choose number of lines                    - nb_lines
+# - choose number of stops                    - nb_stops = nb_lines + 1
+
+network_prop = function(nb_lines, nb_stops){
+  # Intermediate outputs
+  ########## 1. ##########
+  # Stops' list
+  stops = stops_list(nb_lines, nb_stops)
+  ########## 2. ##########
+  name_stops = name_stops_function(stops)
+  ########## 3. ##########
+  # adj = adj_function(stops, name_stops)
+  adj_list = adj_function2(stops, name_stops)
+  adj = adj_list$adj_w + adj_list$adj_b
+  ########## 3. ##########
+  # paths = paths_function(nb_lines*nb_stops*2, name_stops, cross_stop)
+  line_mbrshps = interaction(name_stops$V1, name_stops$V2)
+  tour_mbrshps = name_stops$V1
+  sp_data = build_sp_data(line_mbrshps, tour_mbrshps, adj_list$adj_w, 
+                          adj_list$adj_b)
+  edge_ref = sp_data$edge_ref
+  sp_edge_link = sp_data$sp_edge_link
+  sp_ref = sp_data$sp_ref
+  paths = 1*as.matrix(sparseMatrix(i=sp_ref[, 1], j=sp_ref[, 2], 
+                                   dims=c(nb_lines*nb_stops*2, nb_lines*nb_stops*2)))
+  
+  return(list(stops=stops, name_stops=name_stops, adj=adj, edge_ref=edge_ref,
+              sp_edge_link=sp_edge_link, sp_ref=sp_ref, paths=paths))
+}
+
+# Compute the toy network according to following parameters
+# - conv threshold for iterative fitting      - conv_thres_if
+# - conv threshold                            - conv_thres_algo
+# - epsilon                                   - epsilon
+# - max nb of iterations                      - max_it
+# - print iterations                          - display_it = F
+# - number of iterations                      - n_test = 50
+# - prop_limit betw. 0 and 1 (can be a list)  - hyper_par
+# nb passengers into the network              - n_passengers
+
[…]

### 2023-02-02 multiple_lines/5_1_visualisation.R
Visualisations

@@ -1,0 +1,77 @@
+#-------------------------------------------------------------------------------
+#-------------------------------------------------------------------------------
+# COMPUTE TOY EXAMPLE
+# Visualisation outputs
+#-------------------------------------------------------------------------------
+#-------------------------------------------------------------------------------
+
+#--------------------------------
+# Head
+#--------------------------------
+
+# Set working directory path
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Libraries
+#--------------------------------
+library(dplyr)
+library(ggplot2)
+library(gridExtra)
+library(reshape2)
+
+# Load functions
+source("local_functions.R")
+
+# The folder containing pre-processed data
+data_folder = "multilines_data/preprocessed_data/loop_toy_example"
+
+# Output folder for results
+out_folder = "results/loop_toy_example"
+
+#--------------------------------
+# Network parameters
+#--------------------------------
+# Choose number of lines
+nb_lines = 3
+# Choose number of stops
+nb_stops = nb_lines + 1
+
+#--------------------------------
+# Algorithm parameters
+#--------------------------------
+
+# Conv threshold for iterative fitting
+conv_thres_if = 0.0001
+# Conv threshold
+conv_thres_algo = 0.0001
+# epsilon
+epsilon = 1e-40
+# max iteration
+max_it = 1000
+# print iterations
+display_it = F
+# number of iterations
+n_test = 10
+# prop limit
+hyper_par = c(0.01, 0.1, 0.3, 0.5, 0.7, 0.9, 0.99)
+# number of passengers
+n_passengers = 2000
+
+#--------------------------------
+# Output and visualizations
+#--------------------------------
+
+# Network properties output
+#--------------------------
+network_prop = network_prop(nb_lines, nb_stops)
+# Unlist and save variables
+list2env(network_prop, .GlobalEnv)
+
+# Alogorithm output
+#------------------
+res_mat = compute_toy(hyper_par, n_test, paths, n_passengers, edge_ref, sp_ref,
+                      sp_edge_link, conv_thres_algo, conv_thres_if, max_it,
+                      display_it)
+
+# Results according to the prop limit parameter
+mean_test = apply(res_mat, 2, mean)

### 2023-02-02 multiple_lines/local_functions.R

@@ -1033,6 +1033,7 @@ network_prop = function(nb_lines, nb_stops){
               sp_edge_link=sp_edge_link, sp_ref=sp_ref, paths=paths))
 }
 
+
 # Compute the toy network according to following parameters
 # - conv threshold for iterative fitting      - conv_thres_if
 # - conv threshold                            - conv_thres_algo

### 2023-02-02 multiple_lines/5_1_visualisation.R
First example output

@@ -1,9 +1,12 @@
 #-------------------------------------------------------------------------------
 #-------------------------------------------------------------------------------
 # COMPUTE TOY EXAMPLE
-# Visualisation outputs
+# Visualisation outputs (3)
 #-------------------------------------------------------------------------------
 #-------------------------------------------------------------------------------
+# - Histogram according to several prop limit parameter
+# - Scatter plot according to several prop limit parameter
+# - Network drawing
 
 #--------------------------------
 # Head
@@ -75,3 +78,43 @@ res_mat = compute_toy(hyper_par, n_test, paths, n_passengers, edge_ref, sp_ref,
 
 # Results according to the prop limit parameter
 mean_test = apply(res_mat, 2, mean)
+mean_test = as.data.frame(mean_test)
+rownames(mean_test) = as.character(hyper_par)
+
+# Best hyperparameter
+paste0("Best parameter: ",rownames(mean_test)[which.min(mean_test$mean_test)])
+
+
+# Histograms
+df = as.data.frame(res_mat)
+rownames(df) = as.character(paste("it",1:n_test))
+colnames(df) = as.character(paste("par",hyper_par))
+dfcolnames = colnames(df)
+
+out = list()
+for (i in 1:length(hyper_par)){
+  x = df[,i]
+  if (i == which.min(mean_test$mean_test)) {
+    out[[i]] <- ggplot(data.frame(x), aes(x)) + geom_histogram(bins = 20, color="black", fill="#addd8e") + xlim(min(df), max(df)) + labs(x = paste0("Best: ", dfcolnames[i], " Mean = ", round(mean_test$mean_test[i]*100)/100))
+  }
+  else {
+    out[[i]] <- ggplot(data.frame(x), aes(x)) + geom_histogram(bins = 20, color="black") + xlim(min(df), max(df)) + labs(x = paste0(dfcolnames[i], " Mean = ", round(mean_test$mean_test[i]*100)/100))
+  }
+}
+do.call(grid.arrange, out)
+
+# Scatter plot
+mdf = df
+mdf$par = row.names(mdf)
+mdf = melt(mdf, id=c("par"))
+
+ggplot(mdf, aes(x=variable, y=value, color=par)) +
+  # geom_smooth(show.legend = FALSE, fullrange=TRUE) +
+  geom_point(show.legend = FALSE) +
+  labs(x ="Parameter", y = "% of error") +
[…]

### 2023-02-02 multiple_lines/5_2_visualisation.R
Update visu

@@ -1,0 +1,102 @@
+#-------------------------------------------------------------------------------
+#-------------------------------------------------------------------------------
+# COMPUTE TOY EXAMPLE
+# Visualisation output
+#-------------------------------------------------------------------------------
+#-------------------------------------------------------------------------------
+# - Histogram of mean error, according to the number of passengers travelling
+#   into the network (only 1 hyper parameter)
+
+#--------------------------------
+# Head
+#--------------------------------
+
+# Set working directory path
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Libraries
+#--------------------------------
+library(dplyr)
+library(ggplot2)
+library(gridExtra)
+library(reshape2)
+
+# Load functions
+source("local_functions.R")
+
+# The folder containing pre-processed data
+data_folder = "multilines_data/preprocessed_data/loop_toy_example"
+
+# Output folder for results
+out_folder = "results/loop_toy_example"
+
+#--------------------------------
+# Network parameters
+#--------------------------------
+# Choose number of lines
+nb_lines = 3
+# Choose number of stops
+nb_stops = nb_lines + 1
+
+#--------------------------------
+# Algorithm parameters
+#--------------------------------
+
+# Conv threshold for iterative fitting
+conv_thres_if = 0.0001
+# Conv threshold
+conv_thres_algo = 0.0001
+# epsilon
+epsilon = 1e-40
+# max iteration
+max_it = 1000
+# print iterations
+display_it = F
+# number of iterations
+n_test = 10
+# prop limit
+hyper_par = 0.1
+# number of passengers
+n_passengers = 2000
+
+#--------------------------------
+# Output and visualizations
+#--------------------------------
+
+# Network properties output
+#--------------------------
+network_prop = network_prop(nb_lines, nb_stops)
+# Unlist and save variables
+list2env(network_prop, .GlobalEnv)
+
+# Create a data frame with all different number of passengers into the network
+res_mat_passengers = c()
+seq_passengers = seq(from = 200, to = 4000, by = 200)
+
+for (i in seq_passengers) {
+  res_mat = compute_toy(hyper_par, n_test, paths, i, edge_ref, sp_ref,
+                        sp_edge_link, conv_thres_algo, conv_thres_if, max_it,
[…]

### 2023-02-02 multiple_lines/5_3_visualisation.R

@@ -1,0 +1,98 @@
+#-------------------------------------------------------------------------------
+#-------------------------------------------------------------------------------
+# COMPUTE TOY EXAMPLE
+# Visualisation output
+#-------------------------------------------------------------------------------
+#-------------------------------------------------------------------------------
+# - Histogram of mean error, according to the number of lines into the network
+#   (only 1 hyper parameter)
+
+#--------------------------------
+# Head
+#--------------------------------
+
+# Set working directory path
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Libraries
+#--------------------------------
+library(dplyr)
+library(ggplot2)
+library(gridExtra)
+library(reshape2)
+
+# Load functions
+source("local_functions.R")
+
+
+# The folder containing pre-processed data
+data_folder = "multilines_data/preprocessed_data/loop_toy_example"
+
+# Output folder for results
+out_folder = "results/loop_toy_example"
+
+#--------------------------------
+# Network parameters
+#--------------------------------
+# Choose number of lines
+# below
+
+#--------------------------------
+# Algorithm parameters
+#--------------------------------
+
+# Conv threshold for iterative fitting
+conv_thres_if = 0.0001
+# Conv threshold
+conv_thres_algo = 0.0001
+# epsilon
+epsilon = 1e-40
+# max iteration
+max_it = 1000
+# print iterations
+display_it = F
+# number of iterations
+n_test = 5
+# prop limit
+hyper_par = 0.3
+# number of passengers
+n_passengers = 2000
+# number of cores
+mc.cores = detectCores() - 2
+
+#--------------------------------
+# Output and visualizations
+#--------------------------------
+
+# Create a data frame with different number of line into the network
+res_mat_line = c()
+seq_lines = 2:8
+
+for (i in seq_lines) {
+  nb_lines = i
+  nb_stops = i + 1
+  # Network properties output
+  network_prop_res = network_prop(nb_lines, nb_stops, mc.cores)
+  # Unlist and save variables
+  list2env(network_prop_res, .GlobalEnv)
+  
+  res_mat = compute_toy(hyper_par, n_test, paths, n_passengers, edge_ref, sp_ref,
+                        sp_edge_link, conv_thres_algo, conv_thres_if, max_it,
+                        display_it)
+  
+  colnames(res_mat)[1] <- paste("Lines:", i)
[…]

### 2023-02-02 multiple_lines/local_functions.R

@@ -1006,7 +1006,7 @@ n_multin = function(paths, n_passagers, p_pos = rep(1/sum(paths), sum(paths)),
 # - choose number of lines                    - nb_lines
 # - choose number of stops                    - nb_stops = nb_lines + 1
 
-network_prop = function(nb_lines, nb_stops){
+network_prop = function(nb_lines, nb_stops, mc.cores=NULL){
   # Intermediate outputs
   ########## 1. ##########
   # Stops' list
@@ -1021,8 +1021,14 @@ network_prop = function(nb_lines, nb_stops){
   # paths = paths_function(nb_lines*nb_stops*2, name_stops, cross_stop)
   line_mbrshps = interaction(name_stops$V1, name_stops$V2)
   tour_mbrshps = name_stops$V1
-  sp_data = build_sp_data(line_mbrshps, tour_mbrshps, adj_list$adj_w, 
-                          adj_list$adj_b)
+  if(is.null(mc.cores)){
+    sp_data = build_sp_data(line_mbrshps, tour_mbrshps, adj_list$adj_w, 
+                            adj_list$adj_b)
+  } else {
+    sp_data = build_sp_data_mc(line_mbrshps, tour_mbrshps, adj_list$adj_w, 
+                            adj_list$adj_b, mc.cores=mc.cores)
+  }
+  
   edge_ref = sp_data$edge_ref
   sp_edge_link = sp_data$sp_edge_link
   sp_ref = sp_data$sp_ref

### 2023-02-02 latex_files/notes_multiples_lines_GG/latex/Notes_trajet.tex
reorganize repository

--- 2022-04-12 multiple_lines/notes/latex/Notes_trajet.tex
@@
 Each node belongs to only one line, i.e.
-$\mathcal{V} = \bigcup_{k=1}^p \mathcal{V}_k$ and $\mathcal{V}_k \cap \mathcal{V}_l = \emptyset, \, \forall k,l$, where $\mathcal{V}_k$ represents the set of nodes in line $k$.
+$\mathcal{V} = \bigcup_{k=1}^p \mathcal{V}_k$ and $\mathcal{V}_k \cap \mathcal{V}_l = \emptyset, \, \forall k\neq l$, where $\mathcal{V}_k$ represents the set of nodes in line $k$.
 The edge set $\mathcal{E}$, can also be decomposed with
 \begin{equation}
-\mathcal{E} = \mathcal{E}_\text{W} \cup \mathcal{E}_\text{B}, \qquad \mathcal{E}_\text{W} := \bigcup_{k=1}^p \mathcal{E}_k, \qquad \mathcal{E}_\text{W} \cap \mathcal{E}_\text{B} = \emptyset, \quad \mathcal{E}_k \cap \mathcal{E}_l = \emptyset, \quad \forall k, l.
+\mathcal{E} = \mathcal{E}_\text{W} \cup \mathcal{E}_\text{B}, \qquad \mathcal{E}_\text{W} := \bigcup_{k=1}^p \mathcal{E}_k, \qquad \mathcal{E}_\text{W} \cap \mathcal{E}_\text{B} = \emptyset, \quad \mathcal{E}_k \cap \mathcal{E}_l = \emptyset, \quad \forall k\neq l.
 \end{equation}
@@
 p_{\alpha \beta} = \begin{cases}
-\text{P}_{st}(i, j) & \text{if } \alpha = t + n(s - 1) \text{ and }  \beta = j + n(i - 1), \\
+\text{P}_{st}(i, j) & \text{if } \alpha = s + n(t - 1) \text{ and }  \beta = i + n(j - 1), \\
 0 & \text{otherwise}.
@@
 		\tilde{x}^{\text{B,in},(\tau)}_i &= \rho_i^\text{in} \phi \left( \frac{x^{\text{B,in},(\tau)}_i}{\rho_i^\text{in}} \right), \\
-		\tilde{x}^{\text{B,in},(\tau)}_i &= \rho_i^\text{out} \phi \left( \frac{x^{\text{B,out},(\tau)}_i}{\rho_i^\text{out}} \right),
+		\tilde{x}^{\text{B,out},(\tau)}_i &= \rho_i^\text{out} \phi \left( \frac{x^{\text{B,out},(\tau)}_i}{\rho_i^\text{out}} \right),
 	\end{align}
@@
 	\end{equation}
-	where $\odot$ designate the Hadamard (component-wise) product of matrices. 
+	where $\odot$ designates the Hadamard (component-wise) product of matrices. 
 \end{enumerate}

### 2023-02-02 latex_files/notes_single_line/notes_FB/MaxEnt_une_ligne_v2.tex
reorganize repository

--- 2021-07-22 notes/notes_FB/MaxEnt_une_ligne_v3.tex
@@
 
- \section{Estimation des param. $a$, $b$ dans $n_{ij}=a_ib_j I(i<j)$}
-On a $n_{i\bullet}=x_i$ et $n_{i\bullet }=y_j$. Soit 
-\begin{equation}
-\label{ }
-s:=\min_i \{i | x_i>0\}
-\qquad\qquad
-t:=\max_j \{j | y_j>0\}
-\end{equation}
-Par construction, $1\le s<t\le l$. On suppose que $n_{ij}$ est irréductible (pas de tronçon à vide); sinon il faut considérer chaque tronçon plein séparément.  
-De fait, l'expression $n_{ij}=a_ib_j I(i<j)$ présuppose explicitement l'irréductibilité.
-
-\
-
-En particulier, 
-$a_i=0$ pour $i<s$ et $i=l$; $b_j=0$ pour $j>t$ et $j=1$; 
-$a_s>0$ et $b_t>$. Alors, pour $ s\le i <j\le t$, 
-\begin{equation}
-\label{ }
-n_{sj}=a_sb_j\qquad\qquad n_{it}=a_ib_t\qquad\qquad n_{st}=a_sb_t
-\end{equation}
-et donc
-\begin{equation}
-\label{ }
-\mbox{\fbox{$a_ib_j=\frac{n_{it}n_{sj}}{n_{st}}$}}
-\end{equation}
-
 \end{document}

### 2023-02-02 latex_files/notes_trajet_FB_2023/Notes_trajet_FB_2023.tex
reorganize repository

\documentclass{article}

\usepackage{arxivMod}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}		% Can be removed after putting your text content
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{doi}

\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
%\RequirePackage[authoryear]{natbib}
%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
%\RequirePackage{graphicx}

%\textheight = 615pt
%\textwidth = 420pt

%% Personal packages
\RequirePackage{bm}
\usepackage{xcolor}

\theoremstyle{plain}
%\newtheorem{axiom}{Axiom}  % consignes
%\newtheorem{claim}[axiom]{Claim} % consignes
%\newtheorem{theorem}{Theorem}[section]% consignes 
%\newtheorem{corollary}{Corollary}[section]% consignes 
%\newtheorem{lemma}[theorem]{Lemma} % consignes 
\newtheorem{axiom}{Axiom}
\newtheorem{claim}[axiom]{Claim}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}

%% Personal newcommands
\newcommand{\lo}{\overline{\bm{\phi}\vphantom{\bm{\alpha}}}}
\newcommand{\ld}{\overline{\bm{\phi}^2}}
\newcommand{\lt}{\overline{\bm{\phi}^3}}
\renewcommand{\lq}{\overline{\bm{\phi}^4}}

\newcommand{\mo}{\overline{\bm{\gamma}\vphantom{\bm{\phi}}}}
\newcommand{\md}{\overline{\bm{\gamma}^2}}
\newcommand{\mt}{\overline{\bm{\gamma}^3}}
\newcommand{\mq}{\overline{\bm{\gamma}^4}}

\newcommand{\loc}{\overline{\bm{\phi}_c}}
\newcommand{\ldc}{\overline{\bm{\phi}_c^2}}
\newcommand{\ltc}{\overline{\bm{\phi}_c^3}}
\newcommand{\lqc}{\overline{\bm{\phi}_c^4}}

\newcommand{\moc}{\overline{\bm{\gamma}_c\vphantom{\bm{\phi}^\varphi}}}
\newcommand{\mdc}{\overline{\bm{\gamma}_c^2}}
\newcommand{\mtc}{\overline{\bm{\gamma}_c^3}}
\newcommand{\mqc}{\overline{\bm{\gamma}_c^4}}

\newcommand{\bbGamma}{{\mathpalette\makebbGamma\relax}}
\newcommand{\makebbGamma}[2]{%
  \raisebox{\depth}{\scalebox{1}[-1]{$\mathsurround=0pt#1\mathbb{L}$}}}

\title{Notes trajets TL, 2023}

%\date{September 9, 1985}	% Here you can change the date presented in the paper title
%\date{} 					% Or removing it

\author{\rm  F. Bavaud
}
% Uncomment to remove the date
\date{}

% Uncomment to override  the `A preprint' in the header
%\renewcommand{\headeright}{Technical Report}
%\renewcommand{\undertitle}{Technical Report}
\renewcommand{\shorttitle}{Notes trajets TL, 2023}

%%% Add PDF metadata to help others organize their library
%%% Once the PDF is generated, you can check the metadata with
%%% $ pdfinfo template.pdf
\hypersetup{
pdftitle={Weighted multidimensional scaling from network affinities},
pdfsubject={q-bio.NC, q-bio.QM},
pdfauthor={Fran\c{c}ois Bavaud},
pdfkeywords={First keyword, Second keyword, More},
}

\begin{document}
\maketitle

\section{Notations}
\subsection{Lignes et flux}
Les arrêts de bus $i=1,\ldots,n$, formant l'ensemble $V$ des noeuds du réseau, sont associés à l'une des lignes TL unidirectionnelles $l=1,\ldots,m$.

On note par $l[i]$  la ligne sur laquelle est située l'arrêt $i$.

On note par $\alpha(l)$ l'arrêt de départ de la ligne $l$, et par $\omega(l)$ l'arrêt terminus de la ligne $l$.

 On note par $F(i)$ (forward) l'arrêt suivant. La fonction est bien définie sauf $i=\omega(l[i])$ (i.e si $i$ est le terminus de ligne), auquel cas $F[i]=\emptyset$.

 On note par $B(i)$ (forward) l'arrêt précédent. La fonction est bien définie sauf $i=\alpha(l[i])$ (i.e si $i$ est le départ de ligne),  auquel cas $B[i]=\emptyset$.

Les flux de passagers $x_{ij}$ de $i$ vers  $j$ sont de deux types:  (1) intra-ligne  ssi  $j=F(i)$ (ce qui implique $l[i]=l[j]$), noté alors $y_{ij}$;  (2) inter-ligne (transfert péderstre) ssi $l[i]\neq l[j]$, noté alors $z_{ij}$. Comme un flux est soit de l'un, soit de l'autre type, on a 
\begin{displaymath}
x_{ij}=y_{ij}+z_{ij}
\end{displaymath}

\section{Données}
Les données sont constituées 
\begin{itemize}
  \item[(a)] des montées $a_i$ à l'arrêt $i$, avec $a_i=0$ si $i=\omega(l[i])$ (i.e si $i$ est un terminus de ligne) 
  \item[(b)]  des descentes $b_i$ à l'arrêt $i$, avec $b_i=0$ si  $i=\alpha(l[i])$ (i.e si $i$ est un départ de ligne) 
  \item[(c)]  du {\em tenseur d'intermédiarité} (betweenness tensor) $\chi_{ij}^{st}$, défini comme
\begin{equation}
\label{ }
\begin{cases}
 \chi_{ij}^{st}=1     & \text{si  le plus court chemin de $s$ vers $t$ (supposé unique) passe par l'arrête $ij$}, \\
 \chi_{ij}^{st}=0         & \text{sinon}.
\end{cases}
\end{equation}
Il possède $n^4$ composantes. Dans un second temps, on pourra relaxer l'hypothèse d'un chemin le plus court unique de $s$ à $t$, et remplacer $\chi_{ij}^{st}$ par une fonction (fortement) pénalisée par la longueur du chemin.
\end{itemize}
Le support $S_Z$ des arrêtes $ij$ de transfert, i.e.  l'ensemble des arrêtes $ij$ pour lesquelles il est possible que $z_{ij}>0$, est tel que
\begin{displaymath}
\max_{st} \chi_{ij}^{st}=1 
\end{displaymath}
i.e. si l'arrête $ij$ est située sur un plus court chemin d'au moins une entrée $s$ et une sortie $t$ du réseau.

Les transferts possibles $z_{ij}>0$  impliquent que $i$ et $j$ soient suffisamment proches, i.e. appartiennent à un même {\em super-stop} ou {\em crossing} $c=1,\ldots,r$. On note par $V_c$ ($c=1,\ldots, r)$ l'ensemble des noeuds appartenant au super-stop $c$. Un noeud $i$ ne peut appartenir à deux super-stops, mais il peut n'appartenir à aucun super-stop. On note par $V_0$ l'ensemble des noeuds n'appartenant à aucun super-stop. Par construction, $V$ est partitionnné comme
 $V=\{V_0,V_1,\ldots, V_r\}$. Aussi, $z_{ij}>0$ n'est possible que si $i$ et $j$ appartiennent au même super-stop (et $i\neq j$)  et ainsi 
 \begin{displaymath}
S_Z=\bigcup_{c=1}^r (V_c\times V_c)\quad \setminus\quad  \bigcup_{i=1}^n (\{i\}\times \{i\})
\end{displaymath}

Le support $S_Y$ des arrêtes $ij$ intra-ligne,  i.e.  l'ensemble des arrêtes $ij$ pour lesquelles il est possible que $y_{ij}>0$, est constitué des paires telles que  $j=F(i)$.

\subsection{Proportion de transferts}
On a aussi pour tout $i$ qui n'est ni terminus ni départ. 
\begin{equation}
\label{bilan1ligne}
y_{i,F(i)}=y_{B(i),i}+a_i-b_i
\end{equation}
ce qui permet de déterminer les flux intra-ligne $y_{ij}$.

Parmi les passagers montant en $i$, certains viennent d'une autre ligne, d'autres entrent dans le réseau: 
\begin{equation}
\label{entrer}
a_i=z_{\bullet i}+n_{i\bullet}
\end{equation}
Parmi les passagers descendant en $i$, certains commutent vers une autre ligne, d'autres sortent du réseau: 
\begin{equation}
\label{sortir}
b_i=z_{i\bullet}+n_{\bullet i}
\end{equation}
Naturellement 
\begin{displaymath}
a_{\bullet}=b_{\bullet}=z_{\bullet\bullet}+n_{\bullet\bullet}
\end{displaymath}
La quantité $n_{\bullet\bullet}$ compte le nombre de personnes utilisant le réseau. La quantité $z_{\bullet\bullet}$ compte le nombre total de transferts (zéro, un, deux, trois...  transferts par personne) effectués par les utilisateurs du réseau. La proportion moyenne de transferts par utilisateur est $z_{\bullet\bullet}/n_{\bullet\bullet}$ .

\subsection{Flux compatibles}
Les flux $\mathbf{X}=(\mathbf{Y},\mathbf{Z})$ sont compatibles avec les montées-descentes $(\mathbf{a},\mathbf{b})$ ssi
\begin{equation}
\label{compatibilite}
y_{i,F(i)}=y_{B(i),i}+a_i-b_i
\qquad\qquad 
z_{\bullet i}\le a_i 
\qquad\qquad 
z_{i\bullet}\le b_i
\end{equation}

\section{Inférence}
\subsection{Identité de GG et centralité} 
On considère toutes les paires possibles $(s,t)\in V^2$, où $s$ est l'arrêt (le noeud)  d'entrée d'un passager dans le réseau, et $t$ est l'arrêt (le noeud)  de sortie de ce  passager dans le réseau. Il suffit de considérer $s\neq t$. 
On veut estimer $n_{st}$, le nombre de passagers entrant dans le réseau en $s$, et sortant du réseau en $t$. On a l'identité de GG
\begin{equation}
\label{identiteGG}
x_{ij}=y_{ij}+z_{ij}=\sum_{st}\chi_{ij}^{st}\, n_{st}
\end{equation}
C'est seulement par la quantité $\chi_{ij}^{st}$ que les trajectoires  $z_{ii}$ ou comme $z_{ii-}$ sont interdites (section \ref{casdev}).

On a $x_{ij}=y_{ij}+z_{ij}$, où les deux termes ne peuvent être non-nuls simultanément: $y_{ij}$ est possiblement positif ssi $j=F(i)$, et $z_{ij}$ est possiblement positif ssi $i,j\in I$ (même jonction) avec $i\neq j$.

Un chemin (d'utilisation du réseau) $st$ commence toujours par une arrête intra $(s, F(s))$,  et se termine toujours par une arrête intra $(B(t),t)$. Des arrêtes de transferts peuvent être empruntées dans les parties {\em intermédiaires} du chemin.  On définit
\begin{equation}
\label{inter}
\begin{cases}
 \gamma_i^{st}:= 1     & \text{si  le plus court chemin de $s$ vers $t$  passe par le stop $i$ ($i=s$ et $i=t$ possibles),} \\
 \gamma_i^{st}=0         & \text{sinon}.
\end{cases}
\end{equation}
Alors 
\begin{displaymath}
 \gamma_i^{st}=\chi_{i\bullet}^{st}+\delta_{it}=\chi_{\bullet i}^{st}+\delta_{is}
\end{displaymath}
On pose $\Gamma_i:=\sum_{st} \gamma_i^{st}\, n_{st}$, qui est un indice de {\em centralité d'intermédiarité} du stop $i$. 
En sommant (\ref{identiteGG}), on a 
\begin{equation}
\label{SumCol}
y_{i\bullet}+z_{i\bullet}=\sum_{st} \gamma_i^{st}\, n_{st}-\sum_sn_{si}=\Gamma_i-n_{\bullet i}
\end{equation}
De même, 
\begin{equation}
\label{SumRow}
y_{\bullet i}+z_{\bullet i}=\sum_{st} \gamma_i^{st}\, n_{st}-\sum_tn_{it}=\Gamma_i-n_{i\bullet}
\end{equation}
Mais $b_i=z_{i\bullet}+n_{\bullet i}$, et donc (\ref{SumCol}) s'écrit 
\begin{equation}
\label{Gammab}
\Gamma_i=y_{i\bullet}+b_i=y_{i,F(i)}+b_i
\end{equation}
De même, $a_i=z_{\bullet i}+n_{i\bullet}$,  et donc  (\ref{SumRow}) s'écrit 
\begin{equation}
\label{GammaA}
\Gamma_i=y_{\bullet i}+a_i=y_{B(i),i}+a_i
\end{equation}
En particulier, $y_{i,F(i)}=y_{B(i),i}+a_i-b_i$, comme il se doit.

Si $i$ est le début de ligne, alors $y_{B(i),i}=0$ et $\Gamma_i=a_i$. Si 
$i$ est le terminus, alors $y_{i,F(i)}=0$ et $\Gamma_i=b_i$.

\subsubsection{(*) Cas possiblement déviants}
\label{casdev}
Pour $z_{ii}$ : on a que $\chi_{ii}^{st}=0$: l'``arrête" $ii$ serait empruntée par  un voyageur descendant en $i$, puis remontant en $i$ (par le/un bus d'après), possible mais certainement pas le plus court chemin: irrationnel. Sauf si on hésite à poursuivre et qu'on décide quand même de poursuivre après réflexion. Ou que, voyageur sans billet, des contrôleurs s'apprêtent à monter en $i$, on y descend précipitamment et attend le prochain bus: finalement très rationnel, possiblement...

Pour $z_{ii-}$ : passer à l'arrêt opposé (on suppose que $B(i)-=F(i-)$ (i.e. que les lignes étaient parallèles ``avant l'arrêt $i$") ne peut pas être le plus  court chemin: si le voyageur descend à $i$, il a passé au moins par $B(i)$, et s'il monte en $i-$, il passe au moins par $F(i-)$, Or, quel que soient $s$ et $t$, 
\begin{displaymath}
d_{sB(i)}+d_{B(i)i}+d_{ii-}+d_{i-F(i-)}+d_{F(i-)t}\; >\; d_{sB(i)}+d_{B(i)B(i)-}+d_{F(i-)t}
\end{displaymath}
car 
\begin{displaymath}
 d_{B(i)i}+d_{ii-}+d_{i-F(i-)}\; >\;  d_{B(i)B(i)-}
\end{displaymath}
où $d_{ii-}$ est le coût de traverser la route en $i$, et $d_{B(i)B(i)-}$ celui de traverser la route l'arrêt d'avant. Mais s'il est épouvantable de traverser la route en $i$ (camions, pas de passage sécurisé), ou s'il est embarrassant de traverser la route en $i$ (par exemple sous les yeux d'une connaissance possiblement dans les parages qui nous croyait à l'étranger, ou en arrêt maladie), alors de nouveau il peut être préférable de descendre un arrêt plus loin puis de rebrousser chemin: de nouveau,  finalement très rationnel, possiblement...

\subsection{Independence assumption on stops}
Assume $n_{st}^{(r)}=\frac{n^{(r)}_{s\bullet}n^{(r)}_{\bullet t}}{n^{(r)}_{\bullet\bullet}}$. Compute $\mathbf{X}^{(r+1)}=\bm{\chi}\mathbf{N}^{(r)}$, then compute
\begin{equation}
\label{IterInd}
n_{i\bullet}^{(r+1)}=a_i-z_{\bullet i}^{(r+1)}
\qquad\qquad
n_{\bullet i}^{(r+1)}=b_i-z_{i\bullet}^{(r+1)}
\qquad\qquad
n_{st}^{(r+1)}=\frac{n^{(r+1)}_{s\bullet}n^{(r+1)}_{\bullet t}}{n^{(r+1)}_{\bullet\bullet}}
\end{equation}
Start from some $n^{(0)}_{s\bullet}$ and $n^{(0)}_{\bullet t}$ with coinciding totals (such as $n^{(0)}_{s\bullet}=a_s$ and $n^{(0)}_{\bullet t}=b_t$, i.e. no transfers $\mathbf{Z}^{(0)}=\mathbf{0}$), and iterate (\ref{IterInd}) until convergence.

Il faudrait en plus s'assurer de la compatibilité $a_i\ge z_{\bullet i}^{(r+1)}$ et $b_i\ge z_{i\bullet}^{(r+1)}$ à chaque étape.

\subsubsection{*** contre-argument? Mais non***}
*** Mais en fait $a_i-z_{\bullet i}^{(r)}=n_{i\bullet}^{(r)}$, et donc la première identité (\ref{IterInd}) s'écrit $n_{i\bullet}^{(r+1)}=n_{i\bullet}^{(r)}$. De même, 
$b_i-z_{i\bullet}^{(r)}=n_{\bullet i}^{(r)}$ et donc $n_{\bullet i}^{(r+1)}=n_{\bullet i}^{(r)}$. En conséquence, toute solution initiale est déjà un point fixe, il n'y aucune évolution / convergence. Si on part avec une condition initiale sans transferts, toutes les lignes sont de facto déconnectées, avec des flux $n_{st}^{(0)}$ {\em purement mono-lignes},  et ainsi aucun transfert $z_{ij}>0$ ne va jamais émerger de $\mathbf{X}^{(r)}=\bm{\chi}\mathbf{N}^{(r)}$ ... $\frown$. Il faudrait essayer travailler au niveau de {\em jonctions} aux flux indépendants.

\subsection{Independence assumption on junctions}
Each stop $i$ belongs to a single line $\ell(i)$, with a successor stop $F(i)$ always defined unless $i$ is the terminus of the line.

Also, each stop  $i$ belongs to a single super-stop junction $I[i]$, including ``bare junctions" where no transfers take place.  Define the edge-(junction-)path incidence matrix
\begin{equation}
\label{ }
\begin{cases}
 \chi_{ij}^{ST}=1     & \text{si  le plus court chemin (supposé unique)  de la jonction $S$ vers la jonction $T$ passe par l'arrête $ij$}, \\
 \chi_{ij}^{ST}=0         & \text{sinon}.
\end{cases}
\end{equation}
Un chemin (d'utilisation du réseau) $ST$ commence toujours par l'arrête intra $(i, F(i))$ avec $i\in S$, et se termine toujours par une arrête intra $(B(j),j)$ avec $j\in T$. Des arrêtes de transferts peuvent être empruntées dans les parties {\em intermédiaires} du chemin.

On indique, systématiquement, des jonctions par des majuscules, dont les indices {\em inférieurs} (i.e. pas comme dans $ \chi_{ij}^{ST}$) dénotent des agrégations (sommations) de stops dans les jonctions. Par exemple: 
\begin{displaymath}
n_{ST}=\sum_{s\in S; t\in T}n_{st}\qquad\qquad y_{\bullet S}=\sum_{s\in S} y_{\bullet s}\qquad\qquad\Gamma_S=\sum_{i\in S}\Gamma_i
\end{displaymath}
L'identité de GG devient 
\begin{equation}
\label{GGalt}
x_{ij}=y_{ij}+z_{ij}=\sum_{ST}\chi_{ij}^{ST}\, n_{ST}
\end{equation}
On a 
 $ \chi_{ij}^{ST}\neq \sum_{s\in S; t\in T}\chi_{ij}^{st}$. L'identité (\ref{GGalt}) diffère de (\ref{identiteGG}) qui s'écrit
\begin{displaymath}
x_{ij}=y_{ij}+z_{ij}=\sum_{st}\chi_{ij}^{st}\, n_{st}=\sum_{ST}\sum_{s\in S; t\in T}\chi_{ij}^{st}\, n_{st}
\end{displaymath}
L'itération commence par le calcul de $\mathbf{X}^{(r+1)}=\bm{\chi}\mathbf{N}^{(r)}$ selon  (\ref{GGalt}), puis par le calcul de l'analogue des identités (\ref{IterInd}) qui deviennent 
\begin{equation}
\label{IterIndAlt}
n_{S\bullet}^{(r+1)}=a_S-z_{\bullet S}^{(r+1)}
\qquad\qquad
n_{\bullet T}^{(r+1)}=b_T-z_{T\bullet}^{(r+1)}
\qquad\qquad
n_{ST}^{(r+1)}=\frac{n^{(r+1)}_{S\bullet}n^{(r+1)}_{\bullet T}}{n^{(r+1)}_{\bullet\bullet}}
\end{equation}
A noter que $z_{\bullet S}=z_{S\bullet}$ quel que soit $\mathbf{N}$. Ainsi, on a toujours $n_{S\bullet}-n_{\bullet S}=a_S-b_S$ (les flux  $\mathbf{N}$ sont généralement marginalement inhomogènes), avec $a_{\bullet}=b_{\bullet}$ évidemment.

\subsection{Une seule ligne}
On veut trouver $\mathbf{N}=(n_{st})$ (qu'on peut normaliser et considérer comme une distribution de probabilité) qui maximise l'entropie sous les contraintes $n_{s\bullet}=a_s$ et  $n_{\bullet t}=b_t$, i.e. $\sum_{st}n_{st}\delta_{si}=a_i$ et $\sum_{st}n_{st}\delta_{tj}=b_j$, avec le prior théorique $f^M_{st}=\frac{2}{l(l-1)}I(s<t)$. Avec les multiplicateurs de Lagrange (intégrant la normalisation / le total), la solution est de la forme 
\begin{displaymath}
n_{st}=I(s<t)\prod_{i}\exp(-\lambda_i \delta_{si})\prod_{j}\exp(-\mu_j \delta_{tj})=I(s<t)\,c_s\, d_t \qquad\qquad \sum_{s<t}c_s\, d_t=n_{\bullet\bullet}=a_{\bullet}=b_\bullet
\end{displaymath}
Les contraintes sont que 
\begin{displaymath}
a_i=\sum_{st}I(s<t)\,c_s\, d_t \, \delta_{si}=c_i \sum_{t>i}d_t=c_i\, D_i \qquad\qquad D_i :=\sum_{t>i}d_t
\end{displaymath}
\begin{displaymath}
b_j=\sum_{st}I(s<t)\,c_s\, d_t \, \delta_{tj}=d_j\sum_{s<j}c_s=d_j \, C_j \qquad\qquad C_j:=\sum_{s<j}c_s
\end{displaymath}
Soit $m_{st}$ le nombre de voyageurs partant de $s$ et toujours présents en $t$, et soit $\rho_j$ la probabilité de descendre en $j$. On a 
\begin{displaymath}
n_{st}=m_{st}\rho_t \qquad\qquad n_{s,t+1}=m_{st}(1-\rho_t)\rho_{t+1}\qquad\qquad \frac{n_{s,t+1}}{n_{st}}=\frac{(1-\rho_t)\rho_{t+1}}{\rho_t}=\frac{d_{t+1}}{d_t}
\end{displaymath}
Ainsi 
\begin{displaymath}
\frac{d_{t+2}}{d_{t+1}}=\frac{(1-\rho_{t+1})\rho_{t+2}}{\rho_{t+1}}
\qquad\qquad \frac{d_{t+2}}{d_{t}}=\frac{d_{t+2}}{d_{t+1}}\frac{d_{t+1}}{d_{t}}=\frac{(1-\rho_{t+1})\rho_{t+2}}{\rho_{t+1}}\frac{(1-\rho_t)\rho_{t+1}}{\rho_t}=
\rho_{t+2}(1-\rho_{t+1})(1-\rho_{t})\frac{1}{\rho_{t}}
\end{displaymath}
\begin{displaymath}
\frac{d_{t+q}}{d_{t}}=
\rho_{t+q}(1-\rho_{t+q-1})(1-\rho_{t+q-2})\ldots(1-\rho_{t+1}) (1-\rho_{t})\frac{1}{\rho_{t}}
\end{displaymath}
\begin{displaymath}
d_l+d_{l-1}=d_2[\rho_l(1-\rho_{l-1})+\rho_{l-1}](1-\rho_{l-2})(1-\rho_{l-3})\ldots(1-\rho_{3}) (1-\rho_{2})\frac{1}{\rho_{2}}
\end{displaymath}
\begin{displaymath}
D_i=\sum_{t>i}d_t
\end{displaymath}
Soit $z_{st}$ la proportion de voyageurs présents dans le bus en $s$ et toujours présents en $t>s$. Soit $\tau_s$ la probabilité de monter en $s$. Alors
$f_{st}=\tau_sz_{st}\rho_t$

\newpage

\section{*** OLD JUNK}

\subsection{Une seule ligne}
On ordonne les arrêts $i=1,\ldots, M$ de façon croissante dans le sens du parcours (i.e. $F(i)=i+1$). 
Soit $A_i=\sum_{j\le i}a_j$, respectivement $B_i=\sum_{j\le i}b_j$, le nombre cumulé de montées, respectivement de descentes, jusqu'en $i$. Il faut que $A_i\ge B_i$ et 
$a_{\bullet}=b_{\bullet}$.

Alors $y_{i,F(i)}=A_i-B_i$, et $y_{B(i),i}=(A_i-a_i)-(B_i-b_i)=y_{i,F(i)}-a_i+b_i$, 	qui satisfait (\ref{bilan1ligne}) comme il se doit.

On a $\chi_{ij}^{st}=I(j=i+1)\, I(s<t)\, I(s\le i)\, I(j\le t)=I(j=i+1)\,  I(s\le i)\, I(j\le t)$

\

Soit  $n_{st}=c_sd_t I(s<t)$. En substituant dans (\ref{identiteGG}), on a 
\begin{displaymath}
y_{ij}=I(j=i+1)\sum_{s\le i}\sum_{t\ge j}c_sd_t 
\end{displaymath}

\subsubsection{Maxent}
Let $f_{st}^D:=\frac{n_{st}}{n_{\bullet\bullet}}$ be the empirical distribution, incompletely observed: only $n_{s\bullet}=a_s$ and $n_{\bullet}=b_t$ are known (entailing
$n_{\bullet\bullet}=a_{\bullet}=b_{\bullet}$). Consider the ``semi-uniform" prior $f_{st}^M=\frac{I(s<t)}{\frac12 M(M-1)}$ (where $M$ is the number of stops of the single line), assigning a constant probability to any $st$-trip provided $s<t$.

\

The MaxEnt solution $\tilde{f}_{st}^D$ minimizing $K(f^D|| f^M)=\sum_{st}f_{st}^D\ln \frac{f_{st}^D}{f_{st}^M}$ under the constraints 
$\sum_t f_{s't}^D=\sum_{st} f_{st}^DI(s=s')=\frac{a_{s'}}{a_\bullet}$ and $\sum_s f_{st}^D=\sum_{st'} f_{st'}^DI(t=t')=\frac{b_t}{b_\bullet}$ is of the form 
\begin{displaymath}
\tilde{f}_{st}^D=\frac{f_{st}^M \exp(\phi_s)\exp(\gamma_t)}{Z}
\end{displaymath}
that is 
\begin{displaymath}
\tilde{n}_{st}^D=n_{\bullet\bullet} \tilde{f}_{st}^D=c_sd_t I(s<t)\qquad\qquad\mbox{avec}\quad \sum_{st}c_sd_t I(s<t)=n_{\bullet\bullet}
\end{displaymath}
with the constraints
\begin{equation}
\label{ConstS}
c_s \sum_{t>s}d_t=a_s\qquad\qquad d_t \sum_{s<t}c_s =b_t
\end{equation}
On a $a_M=0$ et $b_1=0$. Aussi, pour tout $t=1,\ldots, M$, 
\begin{displaymath}
A_t:=\sum_{s=1}^t a_s\quad \ge \quad B_t:=\sum_{s=1}^t b_s
\end{displaymath}

\subsubsection{Interprétation markovienne de GG}
On peut monter que la  solution MaxEnt est de la forme 
\begin{equation}
\label{ }
\tilde{n}_{st}=I(s<t)\, a_{\bullet}\,  p_s^{\mbox{\tiny in}}c_{s+1}\ldots c_{t-1} p_t^{\mbox{\tiny out}}
\qquad\qquad \tilde{n}_{s,s+1}=a_{\bullet} p_s^{\mbox{\tiny in}} p_{s+1}^{\mbox{\tiny out}}
\end{equation}
avec $p_s^{\mbox{\tiny in}}=\frac{a_s}{a_{\bullet}}$,  \quad $p_t^{\mbox{\tiny out}}=\frac{b_t}{y_{t-1,t}}$ et $c_i=1-p_i^{\mbox{\tiny out}}$. On a $p_M^{\mbox{\tiny out}}=\frac{b_M}{y_{M-1,M}}=1$ (terminus, tout le monde descend).

*** preuve ***

\subsubsection{Identité de GG}

\subsubsection{Maxent for multilines}
A {\em junction $I$} (not "superstop") is a set of disjoint set $I$ of stops "not too far apart", on which transfers can take place. Any transfer $ij$ is allowed for $i,j\in I$, with the exception of $i=j$ (unboarding at $i$ followed by boarding at the same stop) and $j=i_-$, where $i_-$ denotes the stop "opposite to $i$", that is crossing the road to take the line opposite to

\

$n_{\bullet\bullet}$ is not known. Its maximal value is reached under no transfers, that is when each boarding is an entrance in the network, and when each unboarding an exit. That is 
\begin{displaymath}
n_{\bullet\bullet}\le a_{\bullet}=b_{\bullet}
\end{displaymath}

On the other direction, $n_{\bullet\bullet}$ is minimum when transfers are maximal: that is, one would like, for every super-stop $I$  to match each unboarding traveller with a boarding traveller, leaving $I$ traveller-free. Yet  superstop total boardings and unboardings  do not match in general: 
\begin{displaymath}
a_I=\sum_{i\in I}a_i\qquad\qquad b_I=\sum_{i\in I}b_i\qquad\qquad a_I\neq b_I\mbox{ in general.}
\end{displaymath}
By construction 
\begin{equation}
\label{sortir2}
b_I=z_{I\bullet}+n_{\bullet I}
\qquad\qquad a_I=z_{\bullet I}+n_{I\bullet}
\end{equation}
But $\mathbf{Z}$ is a directed sum of flows concentrated on junctions $I$, and zero outside junctions (denoted $0$), i.e. $\mathbf{Z}=\mathbf{Z}^0\bigoplus_I \mathbf{Z}^I $. Hence 
\begin{displaymath}
z_{I\bullet}=z_{\bullet I}\qquad\qquad  n_{I\bullet}-n_{\bullet I}=a_I-b_I
\end{displaymath}
Note that 
\begin{equation}
\label{ }
a_{\bullet}=a_{0}+\sum_{I}a_I=b_{\bullet}=b_{0}+\sum_{I}b_I
\end{equation}
where $a_{0}$ is the number of travellers boarding outside junctions, and $b_{0}$ is the number of travellers un-boarding outside junctions.

\

A maximum amount of $\min(a_I,b_I)$ can thus be transferred at (supstop) junction $I$, and 
\begin{equation}
\label{333}
n_{\bullet\bullet}\ge  a_{\bullet}-\sum_I \min(a_I,b_I)=\frac12(a_0+b_0+\sum_I |a_I-b_I|)
\end{equation}
Constraints such that no transfers between opposite directions take place are not taken into account, so the lower bound (\ref{333}) is too small in some cases (when the activity on the other lines is too small).

\

Geography defines junctions as disjoint set of stops "not too far apart", and forbids transfers between opposite directions by imposing that transfer $i\to j$ must necessarily take place on shortest $st$ paths, that is edge $ij$ belongs to the $st$ shortest path (symmetric in composition for bidirectional lines), that is iff 
$\chi_{ij}^{st}=1$. But there is no $z_{st}$ such that  $\chi_{ij}^{st}=1$ : if $i$ and $j$ are "opposite stops": boarding on $s$ "before $i$" (some reference direction) and un-boarding at $t$ "after $j$" is never on the shortest path, unless $s=i$ and $j=t$, in which case there is no transfer.

\

*** pas de transferts possibles: fantaisie de lignes de bus entièrement disconnectées...

\

Prior: $f^{M}_{st}=\frac{1}{z} I(t\neq s)I(t\neq s_-)$, where $s_-$ is the bus stop "opposite" to $s$: one assumes twin, bidirectional bus lines). In particuler, one assumes that each stop can be reached

As a consequence, each bus top $s$  (belonging to a junction of not) possesses an unique  "opposite" bus stop $s_-$. Hence $z=\sum_{st} I(t\neq s)I(t\neq s_-)=M(M-2)$, where $M$ is the total number of bus stops, namely 
\begin{displaymath}
M=\sum_{l=1}^L M_l
\end{displaymath}
where $M_l$ is the number of bus stops at line $l$, and the number of lines $L$ is even (twin, bidirectional bus lines).

\subsection{MaxEnt multilines, revisited} 
L'identité de GG s'écrit 
\begin{equation}
\label{identiteGGrev}
x_{ij}=\sum_{st}\chi_{ij}^{st}\, n_{st}=\sum_{[st]\ni (ij)}n_{st}
\end{equation}
où $(ij)$ est une arrête (orientée) du réseau, $[st]$ le plus court chemin (orienté, supposé unique) allant de $s$ à $t$. On note $(ij)\in[st]$ ou $[st]\ni (ij)$ le fait que le chemin $[st]$ passe par l'arrête $(ij)$, i.e. que l'arrête $(ij)$ est sur le chemin $[st]$. L'expression $\sum_{[st]\ni (ij)}\ldots$ est la somme sur tous les chemins $[st]$ passant par $(ij)$ (fixé).

\

La solution MaxEnt pour $n_{st}$ (convertible en une probabilité jointe) avec un prior uniforme et les contraintes  (\ref{identiteGGrev}) (convertibles en moyennes) est 
\begin{equation}
\label{solMaxENT}
n_{st}=\prod_{(ij)}\exp(-\gamma_{ij}\: \chi_{ij}^{st}) =\prod_{(ij)\in[st]}\phi_{ij}
\end{equation}
i.e. un produit de termes consécutifs sur les arrêtes de $[st]$.

\

En particulier, si $(s,F(s))\in [s,t]$ (i.e. $F(s)$ est sur la ligne amenant vers $t$ depuis $s$), alors \fbox{$n_{st}=\phi_{(s,F(s))}\: n_{F(s),t}$} : vision intra-ligne, on reste dans le bus, on était dans le bus en $s$, on y était soit monté, soit resté depuis $B(s)$. En fait, chaque stop $s$ appartient à une ligne, et donc $F(s)$ est toujours défini. Aussi, 
si quelqu'en monte en $s$, il reste au moins jusqu'en $F(s)$. S'il descend en $s$, il ne peut pas être entré dans le réseau en $s$.

\

Si en plus on assume l'indépendance $n_{st}=n_{\bullet\bullet}\rho_s\rho_t$, alors de ce qui précède $\rho_{s}=\phi_{(s,F(s))}\: \rho_{F(s)}$, et donc 
$\phi_{(s,F(s))}=\frac{\rho_{s}}{\rho_{F(s)}}$. Si l'on généralise à (toutes, de deux sortes? ) les arrêtes, on a \fbox{$\phi_{ij}=\frac{c_i}{c_{j}}$} et

\

By the way: on est en fait vraiment intéressé par $n_{ST}$ ($S,T$ = jonctions "vraies" ou stops isolés) plutôt que par $n_{st}$.

\

Let $M$ be the number of edges (within + transfer), and let $N$ be the number of stops (nodes); so $N^2$ is the number of paths. The $M\times N^2$ matrix 
$\bm{\chi}=(\chi_{(ij)[st]})\equiv(\chi_{ij}^{st})$  is the {\em edge-path incidence matrix}. Identity {\ref{identiteGGrev}) reads $\mathbf{X}=\bm{\chi}\mathbf{N}$. Identity 
(\ref{solMaxENT}) reads $\log n_{st}=\sum_{(ij)\in[st]}\log \phi_{ij}=\: -\sum_{(ij)} \gamma_{ij}\, \chi_{(ij)[st]} $, that is 
$\log \mathbf{N}=-\bm{\chi}^\top \log \bm{\Gamma}$

\section{EM approach}
Initial: no transfers?  Minimize $n_{\bullet\bullet} K(f^D||f^M)$ ?

\

If $n_{st}$ is given, then $y_{ij}$ (observed) and $z_{ij}$ (unobserved) follow from $x_{ij}=\sum_{st}\chi_{ij}^{st}\, n_{st}$

\

But $\mathbf{Y}$ must obey $y_{i,F(i)}=y_{B(i),i}+a_i-b_i$

\

If $y_{ij}$  and $z_{ij}$ are given, how to infer $n_{st}$?

\end{document}

### 2023-02-02 latex_files/paper_complex_networks_old/paper_complex_networks_v1.tex
reorganize repository

--- 2023-01-14 paper_complex_networks/paper_complex_networks_v1.tex
@@
 
-\subsection{Margin distributions update (added by GG)}
-
 With equation (\ref{flow_sum}), we can obtain the flow associated to $f_{st}$
@@
 \end{equation}
-and with the edge-trip incidence matrix $\bm{\chi}$ (\ref{edgetrip}), we obtain the intra-line edge flows $\mathbf{Y}=(y_{ij})$ and the transfer edge flows $\mathbf{Z}=(z_{ij})$. \\
+and with the edge-trip incidence matrix $\bm{\chi}$ (\ref{edgetrip}), we obtain the intra-line edge flows $\mathbf{Y}=(y_{ij})$ and the transfer edge flows $\mathbf{Z}=(z_{ij})$.
 
-We now have to make sure constraints (\ref{entrer}) and (\ref{sortir}) are verified. The new margins distribution are set to 
+\subsection{Margin distributions update (added by GG)}
+
+Let us define the hyperparameter $0 \geq \theta > 1$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the network at each stop}.
+Note that we could set a different hyperparmeter for each node, and for embarkations and disembarkations, but without addition information, we will restrain to this simpler case.
+This hyperparamter means that
 \begin{align}
-	\widetilde{\alpha}_s &= \frac{\min(0, a_s - z_{\bullet s})}{\sum_{s'} \min(0, a_{s'} - z_{\bullet {s'}})} \label{alpha_update} \\
-	\widetilde{\beta}_t &= \frac{\min(0, b_t - z_{t \bullet})}{\sum_{t'} \min(0, b_{t'} - z_{{t'} \bullet})} \label{beta_update}
+	z_{\bullet s} &< (1 - \theta) a_s \\
+	z_{t \bullet} &< (1 - \theta) b_t
+\end{align}
+and with constraints (\ref{entrer}) and (\ref{sortir}), the update of margins distribution must be
+\begin{align}
+	\widetilde{\alpha}_s &= \frac{\min(\theta a_s, a_s - z_{\bullet s})}{\sum_{s'} \min(\theta a_s, a_{s'} - z_{\bullet {s'}})} \label{alpha_update} \\
+	\widetilde{\beta}_t &= \frac{\min(\theta b_t, b_t - z_{t \bullet})}{\sum_{t'} \min(\theta b_t, b_{t'} - z_{{t'} \bullet})} \label{beta_update}
 \end{align}
@@
 
-Transfer edges $(i, j)$ where $z_{i \bullet} > b_i$ or $z_{\bullet j} > a_j$ typically denote an overflowing tansfer flow regarding the $g_{st}$ prior distribution.
+Transfer edges $(i, j)$ where $z_{i \bullet} > (1 - \theta)b_i$ or $z_{\bullet j} > (1 - \theta)a_j$ typically denote an overflowing tansfer flow regarding the $g_{st}$ prior distribution.
 The latter must be adjusted in order to have $\mathbf{Z}$ comptible with constraints.
@@
 	\label{flow_ratio}
-	r_{ij} = \max \left(1, \frac{z_{i \bullet}}{b_i}, \frac{z_{\bullet j}}{a_j} \right),
+	r_{ij} = \max \left(1, \frac{z_{i \bullet}}{(1 - \theta)b_i}, \frac{z_{\bullet j}}{(1 - \theta)a_j} \right),
 \end{equation}

### 2023-02-02 1_build_network_data.R
test new organization

@@ -9,7 +9,7 @@
 #--------------------------------
 
 # Set working directory path
-# setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
 # Load functions
 source("local_functions.R")
@@ -20,21 +20,21 @@ source("local_functions.R")
 
 # --- TEST 6789
 
-# # The file containing the dataset
-# line_data_file = "multilines_data/formatted_data/test_6789/bus_df.csv"
-# # The file containing the pedestrian distance between stops
-# d_ped_file = "multilines_data/formatted_data/test_6789/ped_time.csv"
-# # Out folder
-# out_folder = "multilines_data/preprocessed_data/test_mc_6789"
+# The file containing the dataset
+line_data_file = "multilines_data/formatted_data/test_6789/bus_df.csv"
+# The file containing the pedestrian distance between stops
+d_ped_file = "multilines_data/formatted_data/test_6789/ped_time.csv"
+# Out folder
+out_folder = "multilines_data/preprocessed_data/test_mc_6789"
 
 # --- ALL LINES
-
-# The file containing the dataset
-line_data_file = "multilines_data/formatted_data/all_lines/bus_df.csv"
-# The file containing the pedestrian distance between stops
-d_ped_file = "multilines_data/formatted_data/all_lines/ped_time.csv"
-# Out folder
-out_folder = "multilines_data/preprocessed_data/all_lines"
+# 
+# # The file containing the dataset
+# line_data_file = "multilines_data/formatted_data/all_lines/bus_df.csv"
+# # The file containing the pedestrian distance between stops
+# d_ped_file = "multilines_data/formatted_data/all_lines/ped_time.csv"
+# # Out folder
+# out_folder = "multilines_data/preprocessed_data/all_lines"
 
 # Pedestrian threshold for connecting stops
 ped_t_thres = 120

### 2023-02-02 latex_files/applied_network_science_article/article_v0.tex
add latex template for the final article

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%% loading packages, author definitions

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass{bmcart}

\usepackage{mathptmx}       % selects Times Roman as basic font
\usepackage{helvet}         % selects Helvetica as sans-serif font
\usepackage{courier}        % selects Courier as typewriter font
\usepackage{type1cm}        % activate if the above 3 fonts are not available on your system
\usepackage{makeidx}         % allows index generation
\usepackage{graphicx}        % standard LaTeX graphics tool when including figure files
\usepackage{multicol}        % used for the two-column index
\usepackage[bottom]{footmisc} % places footnotes at page bottom
\usepackage{subfig}
\usepackage{amsfonts}
\usepackage[cmex10]{amsmath}
\usepackage{float}
\usepackage[utf8]{inputenc}

\usepackage{bm} 
\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{caption}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\def\includegraphic{}
\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

%%% Begin ...
\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Research}

% Document information

\title{Estimation of flow trajectories in a multi-lines network}

\author[
  addressref={aff1},                   
  corref={aff1},                       
  email={gguex@unil.ch}
]{\inits{G.G.}\fnm{Guillaume} \snm{Guex}}
\author[
  addressref={aff2},
  email={rloup@unil.ch}
]{\inits{R.L.}\fnm{Romain} \snm{Loup}}
\author[
addressref={aff1, aff2},
email={fbavaud@unil.ch}
]{\inits{F.B.}\fnm{François} \snm{Bavaud}}

\address[id=aff1]{%                           
  \orgdiv{Departement of Language and Information Sciences},             
  \orgname{University of Lausanne},          
  \city{Lausanne},                              
  \cny{Switzerland}                                   
}
\address[id=aff2]{%
  \orgdiv{Institute of Geography and Sustainability},
  \orgname{University of Lausanne},          
  \city{Lausanne},                       
  \cny{Switzerland} 
}

\end{fmbox}

% The Abstract begins here

\begin{abstractbox}

\begin{abstract} % abstract
Blabla
\end{abstract}

% Keywords begin here

\begin{keyword}
\kwd{sample}
\kwd{article}
\kwd{author}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}

\end{frontmatter}

% Main Body

%% --------------------------------- INTRODUCTION

\section{Introduction}

\subsection{Context}

\subsection{Statement of the problem}

\subsection{Related Works}

%% --------------------------------- FORMALISM

\section{Formalism}

\subsection{Notations}

\subsection{Problem}

\subsection{Solution}

%% --------------------------------- CASE STUDIES

\section{Case Studies}

\subsection{Toy Examples}

\subsection{Real Example}

%% --------------------------------- CONCLUSION

\section{Conclusion}

\section*{Appendix}
Text for this section\ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{backmatter}

\section*{Acknowledgements}%% if any
Text for this section\ldots

\section*{Funding}%% if any
Text for this section\ldots

\section*{Abbreviations}%% if any
Text for this section\ldots

\section*{Availability of data and materials}%% if any
Text for this section\ldots

\section*{Ethics approval and consent to participate}%% if any
Text for this section\ldots

\section*{Competing interests}
The authors declare that they have no competing interests.

\section*{Consent for publication}%% if any
Text for this section\ldots

\section*{Authors' contributions}
Text for this section \ldots

\section*{Authors' information}%% if any
Text for this section\ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{bmc_article}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
%\nocite{label}

% or include bibliography directly:
% \begin{thebibliography}
% \bibitem{b1}
% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\section*{Figures}
  \begin{figure}[h!]
  \caption{Sample figure title}
\end{figure}

\begin{figure}[h!]
  \caption{Sample figure title}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Sample table title. This is where the description of the table should go}
  \begin{tabular}{cccc}
    \hline
    & B1  &B2   & B3\\ \hline
    A1 & 0.1 & 0.2 & 0.3\\
    A2 & ... & ..  & .\\
    A3 & ..  & .   & .\\ \hline
  \end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- Sample additional file title}
    Additional file descriptions text (including details of how to
    view the file, if it is in a non-standard format or the file extension).  This might
    refer to a multi-page table or a figure.

  \subsection*{Additional file 2 --- Sample additional file title}
    Additional file descriptions text.

\end{backmatter}
\end{document}

### 2023-02-06 2_compute_OD_matrix.R
add a limit to iterations in iterative fitting

@@ -9,7 +9,7 @@
 #--------------------------------
 
 # Set working directory path
-setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+# setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
 # Load functions
 source("local_functions.R")
@@ -33,7 +33,7 @@ data_folder = "multilines_data/preprocessed_data/all_lines"
 out_folder = "results/all_lines"
 
 # Conv threshold for iterative fitting
-conv_thres_if = 1e-8
+conv_thres_if = 1e-5
 # Conv threshold
 conv_thres_algo = 1e-5
 # proportional limit 

### 2023-02-06 local_functions.R

@@ -428,7 +428,7 @@ compute_origin_destination =
   function(flow_l_in, flow_l_out, edge_ref, sp_ref, sp_edge_link, 
            s_mat=NULL, min_p_ntwk=0.1, conv_thres_if=1e-5,
            conv_thres_algo=1e-5, epsilon=1e-40,
-           max_it=200, display_it=T){
+           max_it=200, max_it_if=200, display_it=T){
   
   # --- Get the network structure 
   
@@ -487,7 +487,8 @@ compute_origin_destination =
     
     psi = rep(1, ncol(g_ref))
     converge_if = F
-    while(!converge_if){
+    it_if = 1
+    while(!converge_if & it_if <= max_it_if){
       # Saving old b results
       psi_old = psi
       # Compute new a and b
@@ -497,6 +498,7 @@ compute_origin_destination =
       if(sum(abs(psi_old - psi)) < conv_thres_if){
         converge_if = T
       }
+      it_if = it_if + 1
     }
     
     # Building f_mat

### 2023-02-06 local_functions.R
modify function

@@ -498,7 +498,6 @@ compute_origin_destination =
       if(sum(abs(psi_old - psi)) < conv_thres_if){
         converge_if = T
       }
-      it_if = it_if + 1
     }
     
     # Building f_mat

### 2023-02-06 latex_files/applied_network_science_article/article_v0.tex
update article
just a test
copying paper_complex_networks_v1.tex

@@
 
-%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
+%\documentclass[twocolumn]{bmcart}% uncomment this for two column layout and comment line below
 \documentclass{bmcart}
@@
 
-\def\includegraphic{}
-\def\includegraphics{}
+%\def\includegraphic{}
+%\def\includegraphics{}
 
@@
 
+%% --------------------------------- OLD IMPORT
+
+\section{Notations and formalism}
+\subsection{Lines and junctions}
+\label{Lines and junctions}
+Consider a transportation network made of bus lines numbered $\ell=1,\ldots, q$, of respective lengths (number of stops) $l_\ell$.
+Opposite lines, that is parallel lines running in the back and forth directions are considered as distinct.
+
+The $l=\sum_{\ell=1}^ql_\ell$ bus stops constitute the nodes of the transportation network.
+Each stop $i=1,\ldots,l$ belongs to a single bus line, and defines a unique next or forward stop $F(i)$ (unless $i$ is the line terminus) and a unique backward stop $B(i)$ (unless $i$ is the line start), both on the same line.
+
+Junctions (super-stops), denoted $S=1,\ldots, L$, are defined as cliques (equivalence classes) of stops which can be mutually reached from one another within walking distance. True junctions 
+involve at least two lines, while pseudo-junctions comprise isolated stops $i$, possibly together with their opposite stop (noted $i\text{-}$) on the parallel line running in the opposite direction.
+
+In summary, each stop $i$ belongs to a single line $\ell(i)$ and a single junction $S(i)$. Lines and junctions define two partitions of the network nodes, with $q$, respectively $L$ blocks.
+
+\subsection{Line edges, transfer edges and trips}
+\label{Line edges, transfer edges and trips}
+Two sorts of oriented edges are involved in the transportation network: 
+\begin{enumerate}
+  \item[$\bullet$] intra-line edges $(i,j)=(i,F(i))$ belonging to a single line  $\ell(i)=\ell(j)$
+  \item[$\bullet$] inter-line or transfer edges $(i,j)$ connecting different lines $\ell(i)\neq \ell(j)$, taking place at some junction $S(i)=S(j)$. 
+  \end{enumerate}
+A $st$-trip, noted $[s,t]$, consists of entering into the network at stop $s$, and leaving the network at $t$, by following the shortest-path (i.e.
+achieving the minimum distance,  minimum time, or  minimum cost), supposed unique, leading to $s$ from $t$.
+
+The succession of edges $(ij)$ belonging to the $st$-trip, noted $(ij)\in [s,t]$, is unique. Define the edge-trip incidence matrix as
+\begin{equation}
+\label{edgetrip}
+\chi_{ij}^{st} = \begin{cases}
+  1    & \text{if $(ij)\in [s,t]$}, \\
+  0    & \text{otherwise}.
+\end{cases}
+\end{equation}
+A $st$-trip always starts with the edge $(s,F(s))$, and finishes with $(B(t),t)$. Transfers can occur in-between, but never at the beginning nor at the end of the trip.
+
+\vspace*{0.1cm}
+
+Similarly, $ST$-trips between junctions, noted $[S,T]$, consist of the supposedly unique shortest $st$-trip $[s,t]$ among all  $s\in S$ and $t\in T$, followed by a traveller entering the network as junction $S$, and leaving it at junction $T$.
+The corresponding edge-trip incidence matrix is
+\begin{equation}
+\label{edgetripjunction}
+\tilde{\chi}_{ij}^{ST} = \begin{cases}
+  1    & \text{if $(ij)\in [S,T]$}, \\
+  0    & \text{otherwise}.
+\end{cases}
+\end{equation}
+Unicity of the shortest-path implies that a passenger wishing to travel from $S$ to $T$ chooses the unique stop $\sigma(S,T)\in S$  as initial node (on line $\ell(\sigma)$), and the unique node $\tau(S,T)\in T$ as final node (on line $\ell(\tau)$).
+
+\subsection{Transportation flows}
+\label{Transportation flows}
+Let  $x_{ij}$ count the number of travelers using edge $(ij)$ in a given period, such as a given hour, day, week or  year.
+The edge flow $x_{ij}$ is denoted by $y_{ij}$ for an intra-line edge $(i,j)$, and
+by $z_{ij}$ for a transfer edge $(i,j)$. By construction, $x_{ij}=y_{ij}+z_{ij}$, where $y_{ij}\,  z_{ij}=0$.
+
+\vspace*{0.1cm}
+
+Let $a_i$, respectively $b_i$, the number of passengers embarking, respectively disembarking at stop $i$. By construction, 
+\begin{equation}
+\label{bilan1ligne}
+\begin{cases}
+ y_{i,F(i)}=a_i \text{\,  and } b_i=0   & \text{if $i$ is a line start}, \\
+y_{B(i),i}=b_i \text{\,  and } a_i=0   & \text{if $i$ is a line terminus}, \\
+ y_{i,F(i)}=y_{B(i),i}+a_i-b_i     & \text{otherwise}.
+\end{cases}
+\end{equation}
+Also, $\mathbf{a}$ and $\mathbf{b}$ must be consistent, in the sense that $A_i\ge B_i$, where $A_i$ (respectively $B_i$) is the cumulated number of embarked 
+(resp.
+disembarked) passengers on the line under consideration, recursively defined as $A_{F(i)}=A_i+a_i$ (resp.
+$B_{F(i)}=B_i+b_i$).
+Moreover,  $A_i=B_i$ at a terminal line stop $i$.
+This common value yields  the total number of passengers transported by the line.
+
+\vspace*{0.1cm}
+
+Let the transportation flow $n_{st}$ denote the number of passengers following an $st$-trip, that is entering the network at $s$ and leaving the network at $t$ by using the shortest path.
+
+Also, 
+at the junction level, let $\tilde{n}_{ST}$ denote the number of passengers following an $ST$-trip. As implied by the end of section (\ref{Line edges, transfer edges and trips}), 
+\begin{equation}
+\label{equivnntilde}
+\tilde{n}_{ST}=\sum_{s\in S}\sum_{t\in T}n_{st} \qquad\qquad\qquad n_{st}=\sum_{ST}\tilde{n}_{ST}\: \delta_{s,\sigma(S,T)}\: \delta_{t,\tau(S,T)}
+\end{equation}
+Definitions (\ref{edgetrip}) and (\ref{edgetripjunction}) yield
+\begin{equation}
+\label{equationGG}
+x_{ij}=\sum_{st}\chi_{ij}^{st}\:  n_{st}=\sum_{ST}\tilde{\chi}_{ij}^{ST}\:  \tilde{n}_{ST}
+\end{equation}
+Among the passengers embarking in $i$, some transfer from another line, and some others enter into the network: 
+\begin{equation}
+\label{entrer}
+a_i=z_{\bullet i}+n_{i\bullet}
+\end{equation}
+where  ``$\bullet$" denotes the summation over the replaced index, as in $n_{i\bullet}=\sum_{j=1}^l n_{ij}$.
+Similarly, among the passengers disembarking in $i$, some transfer to another line, and some others leave the network:
+\begin{equation}
+\label{sortir}
+b_i=z_{i\bullet}+n_{\bullet i}
+\end{equation}
+By construction
+\begin{displaymath}
+a_{\bullet}=b_{\bullet}=z_{\bullet\bullet}+n_{\bullet\bullet}
+\end{displaymath}
+where $n_{\bullet\bullet}$ counts the number of passengers, and $z_{\bullet\bullet}$ counts the number of transfers.
+$z_{\bullet\bullet}/n_{\bullet\bullet}$  is the average number of transfers per passenger.
+
+\vspace*{0.1cm}
+
+As explained in section \ref{Lines and junctions}, transfers can only occur at junctions, that is $z_{ij}>0$ implies $S(i)=S(j)$.
+As $z_{ii}=0$ (no traveller is supposed to disembark and re-embark later at the same stop), and $z_{ii\text{-}}=0$ (where $i\text{-}$ is the stop opposite to $i$, when existing), transfers can occur at true junctions only.
+
+\vspace*{0.1cm}
+
+At the junction level, let 
+\begin{displaymath}
+\tilde{a}_I=\sum_{i\in I} a_i\enspace,\enspace\tilde{b}_I=\sum_{i\in I} b_i\enspace,\enspace\tilde{z}_{\bullet I}=\sum_{i\in I} z_{\bullet i}\enspace,\enspace\tilde{z}_{I\bullet}=\sum_{i\in I} z_{i\bullet}\enspace,\enspace\tilde{n}_{I\bullet}=\sum_{i\in I} n_{i\bullet}\enspace,\enspace\tilde{n}_{\bullet I}=\sum_{i\in I} n_{\bullet i}
+\end{displaymath}
+ respectively denote the number of embarking (disembarking)  passengers at junction~$I$; of embarking passengers after a transfer and of disembarking passengers before a transfer at  $I$; of passengers starting (ending) their trip at $I$.
+Then
+\begin{equation}
+\label{junctionLevel}
+\tilde{a}_I=\tilde{z}_{\bullet I}+\tilde{n}_{I\bullet}\qquad\qquad \tilde{b}_I=\tilde{z}_{I\bullet}+\tilde{n}_{\bullet I}\qquad\qquad \tilde{z}_{\bullet I}=\tilde{z}_{I\bullet}\le \min(\tilde{a}_I,\tilde{b}_I)\enspace.
+\end{equation}
+
+\subsection{Statement of the problem}
+Automatic passenger counters measure the number of passengers entering and leaving buses at each stop [Boyle, 1998], that is $\mathbf{a}$ and $\mathbf{b}$.
+Also, the geometry of the network permits to derive the edge-trip incidence matrix $\bm{\chi}$ (\ref{edgetrip}), as well as the partition $\mathcal{S}$  of stops nested within junctions.
+
+Intra-line edge flows $\mathbf{Y}=(y_{ij})$ can be determined by (\ref{bilan1ligne}), but transfer edge flows $\mathbf{Z}=(z_{ij})$ are, here and typically, unknown.
+The objective is to estimate the $l\times l$ transportation flow $\mathbf{N}=(n_{st})$ at the stop level, or equivalently  the $L\times L$  transportation flow  $\tilde{\mathbf{N}}=(\tilde{n}_{ST})$ at the junction level.
+Many consistent solutions coexist in general, even for a single line with no transferts (section \ref{Single line}).
+
+This issue of incompletely observed data can be tackled by the maximum entropy formalism: let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution).
+Let $\mathcal{F}$ consist of the empirical distributions $\mathbf{f}$ satisfying $R$ linear constraints of the form $\sum_{st}f_{st}\, o^r_{st}=\bar{o}^r$ for $r=1,\ldots, R$.
+Then the minimum Kullback-Leibler divergence
+\begin{displaymath}
+\min_{\mathbf{f}\in\mathcal{F}}\;  \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}}
+\end{displaymath}
+is reached for the maximum entropy solution  
+\begin{equation}
+\label{MaxEnt}
+f^0_{st}=\frac{1}{Z(\bm{\lambda})}\: g_{st}\exp(-\sum_{r=1}^R \lambda_r o^r_{st})\qquad\qquad Z(\bm{\lambda})=\sum_{st}g_{st}\exp(-\sum_{r=1}^R \lambda_r o^r_{st})
+\end{equation}
+where the Lagrange multipliers $\bm{\lambda}$ are determined  so as to satisfy the $R$ linear constraints (**see e.g. Bavaud and *** and references therein).
+
+\subsection{Statement of the problem (version 2, added by GG)}
+Let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution).
+Also suppose a prior guess of row and margin constraints for $f_{st}$, respectively $\alpha_s$ and $\beta_t$, which has to verify
+\begin{equation}
+	\label{init_margin}
+	\frac{\alpha_i}{\alpha_j} = \frac{a_i}{a_j} \qquad \frac{\beta_i}{\beta_j} = \frac{b_i}{b_j}  \qquad \forall i, j \text{ which are not true junctions.} \\
+\end{equation}
+Setting these gives us
+\begin{equation}
+	\label{flow_sum}
+	n_{\bullet \bullet} = \frac{a_i}{\alpha_i} = \frac{b_i}{\beta_i} \qquad \forall i \text{ which are not true junctions.}  
+\end{equation}
+We are interested in finding
+\begin{align}
+	\label{constr_MaxEnt}
+	\min_{\mathbf{f}\in\mathcal{F}} &\; \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}}, \notag \\
+	s.t. &\; \sum_t f_{st} = \alpha_s, \notag \\
+	&\; \sum_s f_{st} = \beta_t.
+\end{align}
+The Lagragian is
+\begin{equation}
+	L = \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}} - \sum_s \lambda_s (\alpha_s - \sum_t f_{st}) - \sum_t \mu_t (\beta_t - \sum_s f_{st}), \notag
+\end{equation}
+which gives, after deriving and setting to zero,
+\begin{equation}
+	\label{Sol}
+	f_{st} = \phi_s \psi_t g_{st} \qquad \text{with } \phi_s := \exp(- 1 - \lambda_s) \text{, } \psi_t := \exp(- \mu_t).
+\end{equation}
+Using constraints in (\ref{constr_MaxEnt}), we find
+\begin{equation}
+	\label{Sol_LagMult}
+	\phi_s = \frac{\alpha_s}{\sum_t \psi_t g_{st}}, \qquad \psi_t = \frac{\beta_t}{\sum_s \phi_s g_{st}}, 
+\end{equation}
+which gives an \emph{iterative fitting algorithm}. Starting with any $\psi^{(0)}_t > 0$, we do
+\begin{equation}
+	\label{Iterative fitting}
+	\phi^{(\iota)}_s = \frac{\alpha_s}{\sum_t \psi^{(\iota)}_t g_{st}}, \qquad \psi^{(\iota + 1)}_t = \frac{\beta_t}{\sum_s \phi^{(\iota)}_s g_{st}}, 
+\end{equation}
+
+With equation (\ref{flow_sum}), we can obtain the flow associated to $f_{st}$
+\begin{equation}
+	\label{flow_from_distrib}
+	n_{st} = n_{\bullet \bullet} f_{st}, 
+\end{equation}
+and with the edge-trip incidence matrix $\bm{\chi}$ (\ref{edgetrip}), we obtain the intra-line edge flows $\mathbf{Y}=(y_{ij})$ and the transfer edge flows $\mathbf{Z}=(z_{ij})$.
+
+\subsection{Margin distributions update (added by GG)}
+
+Let us define the hyperparameter $0 \geq \theta > 1$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the network at each stop}.
+Note that we could set a different hyperparmeter for each node, and for embarkations and disembarkations, but without addition information, we will restrain to this simpler case.
+This hyperparamter means that
+\begin{align}
+	z_{\bullet s} &< (1 - \theta) a_s \\
+	z_{t \bullet} &< (1 - \theta) b_t
+\end{align}
+and with constraints (\ref{entrer}) and (\ref{sortir}), the update of margins distribution must be
+\begin{align}
+	\widetilde{\alpha}_s &= \frac{\min(\theta a_s, a_s - z_{\bullet s})}{\sum_{s'} \min(\theta a_s, a_{s'} - z_{\bullet {s'}})} \label{alpha_update} \\
+	\widetilde{\beta}_t &= \frac{\min(\theta b_t, b_t - z_{t \bullet})}{\sum_{t'} \min(\theta b_t, b_{t'} - z_{{t'} \bullet})} \label{beta_update}
+\end{align}
+
+\subsection{Prior distribution update (added by GG)}
+
+Transfer edges $(i, j)$ where $z_{i \bullet} > (1 - \theta)b_i$ or $z_{\bullet j} > (1 - \theta)a_j$ typically denote an overflowing tansfer flow regarding the $g_{st}$ prior distribution.
+The latter must be adjusted in order to have $\mathbf{Z}$ comptible with constraints.
+For any edge $(i, j)$, let us compute the \emph{flow ratio} $r_{ij}$, with
+\begin{equation}
+	\label{flow_ratio}
+	r_{ij} = \max \left(1, \frac{z_{i \bullet}}{(1 - \theta)b_i}, \frac{z_{\bullet j}}{(1 - \theta)a_j} \right),
+\end{equation}
+where $r_{ij} > 1$ typically denotes an overflow through edge $(i, j)$.
+This flow ratio can be traced back to origin-destination couples $s, t$ in order to compute the \emph{orgin-destination flow ratio} $\bar{r}_{st}$
+\begin{equation}
+	\label{st_flow_ratio}
+	\bar{r}_{st} = \max_{ij} \chi_{ij}^{st} r_{ij},
+\end{equation}
+where again $\bar{r}_{st} > 1$ denotes an overflow between $s$ and $t$. To ajdust the flow, we can devide the previous flow by this ratio
+\begin{equation}
+	\label{update_flow}
+	\widetilde{n}_{st} =\frac{n_{st}}{\bar{r}_{st}}
+\end{equation}
+And the new prior distribution should follow 
+\begin{equation}
+	\label{update_distrib}
+	\widetilde{g}_{st} = \frac{\left( \frac{\widetilde{n}_{st}}{\phi_s \psi_t} \right)}{\sum_{s',t'} \left( \frac{\widetilde{n}_{s',t'}}{\phi_{s'} \psi_{t'}} \right)}
+\end{equation}
+
+With the new prior distribution $\widetilde{g}_{st}$ and the new margin distributions $\widetilde{\alpha}_s$, $\widetilde{\beta}_t$, we can begin a new iteration step by going through 2.5 -2.7 again.
+
+\section{Single line}
+\label{Single line}
+Let  $i=1,\ldots, l$ enumerate the bus stops in increasing order,  i.e.
+$F(i)=i+1$.
+Define $g_{st}=\frac{1}{(l-1)(l-2)}\chi(s<t)$ (where $\chi(.)$ denotes the 0/1  indicator function) as the maximally uniform prior, reflecting only the unidirectional nature of trips.
+Constraints $n_{r\bullet}=a_r$ (resp. $n_{\bullet r}=b_r$), correspond to $o^r_{st}=
+\delta_{sr}$ (resp. $o^r_{st}=\delta_{tr}$) in (\ref{MaxEnt}), which finally yields, after reparametrization, the maximum entropy flow reads
+\begin{equation}
+\label{nosignle}
+n^0_{st}=I(s<t)\, c_s\, d_t \qquad\qquad\mbox{where}\quad \sum_{s<t}c_s\, d_t=n_{\bullet\bullet}=a_{\bullet}=b_\bullet
+\end{equation}
+In addition, the (dis-)embarking constraints  yield
+\begin{equation}
+\label{(dis embarking constraints}
+a_i=c_i \sum_{t>i}d_t=c_i\, D_i 
+\qquad\qquad\qquad
+b_j=d_j\sum_{s<j}c_s=d_j \, C_j  
+\end{equation}
+where $D_i=\sum_{t>i}d_t$ and $C_j=\sum_{s<j}c_s$.
+This maximum entropy solution constitutes one possible consistent transportation flow among many others, such as the ``first in, first out" (FIFO) flow.
+Interestingly enough, (\ref{nosignle}) is reminiscent of the so-called gravity flows of quantitative Geography (**ref**) $n_{st}=h_{st}\, c_s\, d_t $, but with a purely asymmetric ``distance deterrence function" $h_{st}=I(s<t)$.
+Also,
+(\ref{nosignle}) shows that  the conditional probability to exit on $t$, given an entrance on $s$, is $w_{st}=n^0_{st}/n^0_{s\bullet}=I(s<t)\, d_t/D_s$.
+
+\
+
+*** ici la (les) figure de l'exemple  ``starting from Maladière, Riant-Cour, Dapples"... ? ***
+
+\
+
+*** ici l'équivalence avec l'approche chaîne de Markov de Guillaume ? ***
+
+\section{Multiple lines}
+
+*** $a_{\bullet}=b_{\bullet}$ (where ``$\bullet$" denotes summation over the replaced index).
+
+\newpage
+
+\section*{Previous submission to complex networks 2022}
+
+%
+\section*{Statement of the problem}
+%
+
+Automatic passenger counters measure the number of passengers entering and leaving buses at each stop \cite{boyle_passenger_1998}.
+Given this information, can we estimate the complete trajectories of passengers within the entire multi-line network?
+This communication attempts to propose an estimation of all passenger trajectories in the multi-line network with an algorithm based on iterative proportional fitting (IPF) \cite{bishop_discrete_2007}.
+
+The exploited dataset is provided by the Lausanne Transportation Agency (tl) in Switzerland.
+The dataset includes 42 lines of buses (or subways) and more than 1361 stops, including 497 clusters of stops (Fig.
+\ref{fig:map-lausanne}),  carrying around 115 million passengers in 2019.
+Each stop refers to a single directed line, and return lines are considered as distinct.
+In addition to  \emph{line edges},  it is possible to construct pedestrian \emph{transfer edges} (Fig.
+\ref{fig:network-lausanne}) to make the graph unilaterally connected by considering, e.g., clusters of stops connected.
+
+\noindent \begin{minipage}[c]{0.5\textwidth}
+   \begin{figure}[H]
+      \centering
+      \includegraphics[width=6cm]{img/stop_area_c.png}
+      \caption{Cluster of stops} 
+      \label{fig:map-lausanne}
+   \end{figure}
+\end{minipage}%
+\begin{minipage}[c]{0.5\textwidth}
+   \begin{figure}[H]
+      \centering
+      \includegraphics[width=4cm]{img/edge_type3.png}
+      \caption{The multi-line network structure} 
+      \label{fig:network-lausanne}
+   \end{figure}
+\end{minipage}
+
+$\mbox{}$
+
+Knowing only the network structure and the number of passengers embarking and disembarking at each stop, how can we infer the most probable passenger trajectories in the network?
+Before examining the general multi-line network problem,  we address the estimation of trajectories on a single line.
+
+%
+\section{Single line}
+%
+
+Consider a one-directional line with $n$ stops indexed regarding the order found in the line.
+Let $\mathbf{a}_\text{in} = (a_{s}^\text{in})$ and $\mathbf{a}_\text{out} = (a_{t}^\text{out})$ be two vectors representing, respectively, the passengers entering and leaving the line at each stop.
+The goal is to estimate the $(n \times n)$ origin-destination matrix $\mathbf{N} = (n_{st})$, where $n_{st}$ represents the number of passengers entering the line at $s$ and leaving at $t$, subject to constraints $n_{s\bullet} = a^\text{in}_s$ and $n_{\bullet t} = a^\text{out}_t$.
+Among many feasible solutions, arguably the most elegant one is the maximum entropy solution, which can be derived  from different means:  (i) passenger flows can be modelled using a Markovian assumption, which translates by assuming that every passenger has the same probability to continue the trip after having travelled at last one stop ; (ii) an iterative proportional fitting algorithm can be performed, starting with an initial origin-destination affinity matrix $\mathbf{S} = (s_{st})$, defined as the upper triangular $n\times n$ matrix filled with $1$, and then iterated to satisfy the margin constraints given by $\mathbf{a}_\text{in}$ and $\mathbf{a}_\text{out}$.
+Both approaches give the same solution, but only the latter remains pertinent in the multi-line problem.
+
+%
+\section{Multi-lines}
+%
+
+In the multi-line problem, a passenger can transfer from a line to another.
+The problem cannot be tackled with Markov chain modelling anymore, which generate unrealistic random trajectories.
+Instead, we will assume that passengers follow shortest paths.
+Starting from an origin-destination matrix $\mathbf{N} = (n_{st})$, where $s$ denotes the stop at which a passenger \emph{enters into the network} (and not simply enters a particular line), and $t$ denotes the stop where the passenger \emph{leaves definitively the network}, this shortest paths assumption allows us to compute the flow matrix on edges $\mathbf{X} = (x_{ij})$.
+The latter decomposes into the within-line flow and the transfer flow, i.e., $\mathbf{X}=\mathbf{X}_\text{W}+\mathbf{X}_\text{B}$.
+Moreover, in the multi-line problem, we also have to distinguish between:
+\begin{itemize}
+	\item[$\bullet$]  passengers who enter and leave bus lines at each stop, represented by vectors $\mathbf{a}_\text{in}$ and $\mathbf{a}_\text{out}$, which are \emph{measured},
+	\item[$\bullet$] and passengers who enter and leave the network at each stop $i$, represented by the \emph{unknown} quantities $n_{\bullet i}$ and $n_{i \bullet}$.
+\end{itemize}
+By construction, 
+\begin{align}
+		a^\text{in}_i &=n_{i\bullet} + x^\text{B}_{i\bullet},  &
+		a^\text{out}_i &= n_{\bullet i} + x^\text{B}_{\bullet i},
+\end{align}
+
+Using these two constraints, along with the shortest paths assumption and iterative proportional fitting, we propose the following iterative algorithm in order to find $\mathbf{N}$ from measured $\mathbf{a}_\text{in}$ and $\mathbf{a}_\text{out}$.
+\\
+
+\noindent \fbox{\begin{minipage}{\textwidth}
+
+\textbf{Initialisation:} $\mathbf{S}^{(0)}$ is filled with 1 excepted for aberrant origin-destination pairs (such as $t$ beeing a previous stop of the same line as $s$).
+The margins of $\mathbf{N}$ are fixed as $\mathbf{n}^{(0)}_{\mbox{\scriptsize in}} = \mathbf{a}_\text{in}$ and $\mathbf{n}^{(0)}_{\mbox{\scriptsize out}} = \mathbf{a}_\text{out}$.
+
+\textbf{Step 1, Iterative proportional fitting:}
+We use IPF to compute $\mathbf{N}^{(r)}$ starting from $\mathbf{S}^{(r)}$, such that margin constraints, defined by $\mathbf{n}^{(r)}_\text{in}$ and $\mathbf{n}^{(r)}_\text{out}$ are satisfied.
+
+\textbf{Step 2, Shortest paths flow:}
+Using shortest paths information, we compute $\mathbf{X}^{(r)}_\text{B}$ from $\mathbf{N}^{(r)}$.
+
+\textbf{Step 3, Affinity and margin update:}
+$\mathbf{S}^{(r+1)}$, $\mathbf{n}^{(r+1)}_\text{in}$ and $\mathbf{n}^{(r+1)}_\text{out}$ are updated in order to respect constraints defined by (1).
+\end{minipage}}
+
+Step 1, 2, and 3 are iterated until convergence, giving an admissible solution to the problem.
+%
+\section{A small example}
+%
+
+As an illustration, an estimated solution proposed by the algorithm on a restricted network made of four lines only is depicted on Fig.
+\ref{fig:example-map}.
+A total of $n_{\bullet \bullet} = 16,837,494$ passengers using this network is estimated by the algorithm.
+The red circle on the bottom left represents the start of the trip $s$ and the size of the circles at stops $t$ represents the estimated number of passengers terminating their trip at $t$.
+In this example, the majority of passengers exit the network on the same initial embarkment line.
+A small fraction of them takes another line.
+
+Table \ref{fig:example-edges} represents the estimated ten most frequented transfer edges.
+The code of the stop represents the number of the line, the direction and a condensed name of its stop cluster.
+The third column gives the number (in thousands) of passengers transferring through this edge.
+
+\noindent \begin{minipage}[c]{0.6\textwidth}
+   \begin{figure}[H]
+      \includegraphics[width=7cm]{img/example_map.png}
+      \caption{Example map} 
+      \label{fig:example-map}
+   \end{figure}
+\end{minipage}%
+\begin{minipage}[c]{0.43\textwidth}
+      %\centering
+      \begin{table}[H]
+		\begin{tabular}{l|l|r}
+		\multicolumn{1}{c|}{\textbf{From stop}} & \multicolumn{1}{c|}{\textbf{To stop}} & \multicolumn{1}{c}{\textbf{Count}} \\ \hline
+		{\scriptsize S7\_A\_SF\_O}			& {\scriptsize S9\_A\_SF\_O}			& 192k		\\ \hline
+		{\scriptsize S9\_R\_CH\_E}			& {\scriptsize S6\_A\_CH\_E}			& 187k		\\ \hline
+		{\scriptsize S7\_A\_SF\_O}		& {\scriptsize S8\_R\_SF\_S}				& 135k			\\ \hline
+		{\scriptsize S6\_R\_CH\_O}			& {\scriptsize S9\_A\_CH\_O}			& 135k			\\ \hline
+		{\scriptsize S8\_A\_GTE\_N}		& {\scriptsize S9\_R\_GTE\_E}			& 103k		\\ \hline
+		{\scriptsize S9\_R\_B-AIR\_C}		& {\scriptsize S8\_A\_B-AIR\_N}		& 99k			\\ \hline
+		{\scriptsize S9\_A\_SF\_O}			& {\scriptsize S7\_A\_SF\_O}			& 88k			\\ \hline
+		{\scriptsize S8\_R\_B-AIR\_D}		& {\scriptsize S6\_A\_B-AIR\_C}		& 87k			\\ \hline
+		{\scriptsize S6\_R\_SF\_O}			& {\scriptsize S8\_A\_SF\_O}			& 86k			\\ \hline
+		{\scriptsize S9\_A\_GTE\_O}		& {\scriptsize S8\_R\_GTE\_S}		& 84k                                        
+		\end{tabular}
+\vspace*{0.2cm}	\caption{List of the ten most\\ frequented transfer edges} 
+	\label{fig:example-edges}
+	\end{table}
+\end{minipage}
+
+\vspace*{0.2cm}
+
+The current work performs computer-intensive simulations of flow over the entire network (1361 stops), permitting to extract usual network indices (centrality, betweenness...) characterizing both the stops \emph{and} the lines.
+In parallel, the computational effects of various fine tuning calibration parameters used in the algorithm are investigated.
+
 %% --------------------------------- INTRODUCTION
@@
 
-\subsection{Real Example}
+\subsection{Real Data}
 

### 2023-02-06 latex_files/applied_network_science_article/article_v0.tex
starting modifications

@@
 \section{Notations and formalism}
-\subsection{Lines and junctions}
+\subsection{Lines, stops  and junctions}
 \label{Lines and junctions}
@@
 The $l=\sum_{\ell=1}^ql_\ell$ bus stops constitute the nodes of the transportation network.
-Each stop $i=1,\ldots,l$ belongs to a single bus line, and defines a unique next or forward stop $F(i)$ (unless $i$ is the line terminus) and a unique backward stop $B(i)$ (unless $i$ is the line start), both on the same line.
+Each stop $i=1,\ldots,l$ belongs to {\em  a single bus line}, and defines a unique next or forward stop $F(i)$ (unless $i$ is the line terminus) and a unique backward stop $B(i)$ (unless $i$ is the line start), both on the same line.
 
-Junctions (super-stops), denoted $S=1,\ldots, L$, are defined as cliques (equivalence classes) of stops which can be mutually reached from one another within walking distance. True junctions 
-involve at least two lines, while pseudo-junctions comprise isolated stops $i$, possibly together with their opposite stop (noted $i\text{-}$) on the parallel line running in the opposite direction.
-
-In summary, each stop $i$ belongs to a single line $\ell(i)$ and a single junction $S(i)$. Lines and junctions define two partitions of the network nodes, with $q$, respectively $L$ blocks.
+Let $S_i$ denote the set of stops which can be reached from stop $i$ within walking distance, excluding $i$ itself.
+A stop $i$ is referred to as an {\em isolated stop} if $S_i=\emptyset$, and to as a {\em junction} otherwise.
 
@@
   \item[$\bullet$] intra-line edges $(i,j)=(i,F(i))$ belonging to a single line  $\ell(i)=\ell(j)$
-  \item[$\bullet$] inter-line or transfer edges $(i,j)$ connecting different lines $\ell(i)\neq \ell(j)$, taking place at some junction $S(i)=S(j)$. 
+  \item[$\bullet$] inter-line or transfer edges $(i,j)$ connecting different lines $\ell(i)\neq \ell(j)$, involving walks from junction $i$ to $j\in S_i$.
   \end{enumerate}
@@
 \vspace*{0.1cm}
-
-Similarly, $ST$-trips between junctions, noted $[S,T]$, consist of the supposedly unique shortest $st$-trip $[s,t]$ among all  $s\in S$ and $t\in T$, followed by a traveller entering the network as junction $S$, and leaving it at junction $T$.
-The corresponding edge-trip incidence matrix is
-\begin{equation}
-\label{edgetripjunction}
-\tilde{\chi}_{ij}^{ST} = \begin{cases}
-  1    & \text{if $(ij)\in [S,T]$}, \\
-  0    & \text{otherwise}.
-\end{cases}
-\end{equation}
-Unicity of the shortest-path implies that a passenger wishing to travel from $S$ to $T$ chooses the unique stop $\sigma(S,T)\in S$  as initial node (on line $\ell(\sigma)$), and the unique node $\tau(S,T)\in T$ as final node (on line $\ell(\tau)$).
 

### 2023-02-07 latex_files/applied_network_science_article/article_v0.tex
ajouter étoiles

@@
 
-Also, 
+*** Also, 
 at the junction level, let $\tilde{n}_{ST}$ denote the number of passengers following an $ST$-trip. As implied by the end of section (\ref{Line edges, transfer edges and trips}), 
@@
 \bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
-\bibliography{bmc_article}      % Bibliography file (usually '*.bib' )
+\bibliography{article_tl.bib}      % Bibliography file (usually '*.bib' )
 % for author-year bibliography (bmc-mathphys or spbasic)

### 2023-02-07 0_network_correction.R
Data set line correction

@@ -1,0 +1,43 @@
+#-------------------------------------------------------------------------------
+#-------------------------------------------------------------------------------
+# Data set lines correction
+#-------------------------------------------------------------------------------
+#-------------------------------------------------------------------------------
+# Delete transports lines with a high difference between in and out passengers
+
+# Set working directory path
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Read data
+line_df = read.csv("/multilines_data/preprocessed_data/all_lines/line_df.csv")
+
+# Sum passengers by line A and R
+sum_table = data.frame(line=numeric(), a_in=numeric(), r_in=numeric(),
+                       diff_a=numeric(), a_out=numeric(), r_out=numeric(),
+                       diff_r=numeric())
+for (i in unique(line_df$line_nbr)) {
+  line = paste0(i)
+  a_in = sum(ifelse(line_df$line_nbr == i &
+                      line_df$direction == "A", line_df$passengers_in, 0))
+  a_out = sum(ifelse(line_df$line_nbr == i & 
+                       line_df$direction == "A", line_df$passengers_out, 0))
+  r_in = sum(ifelse(line_df$line_nbr == i & 
+                      line_df$direction == "R", line_df$passengers_in, 0))
+  r_out = sum(ifelse(line_df$line_nbr == i & 
+                       line_df$direction == "R", line_df$passengers_out, 0))
+  
+  diff_a = a_in/a_out
+  diff_r = r_in/r_out
+  
+  new_row <- data.frame(line=line, a_in=a_in,a_out=a_out, diff_a=diff_a,
+                        r_in=r_in, r_out=r_out, diff_r=diff_r)
+  sum_table <- rbind(sum_table, new_row)
+}
+
+
+# Show lines which have more than 30% on difference in/out or only one direction
+# or don't have way back
+invalid_lines = sum_table[(sum_table$diff_a < 0.3 | sum_table$diff_a > 1.3) | 
+                            (sum_table$diff_r < 0.7 | sum_table$diff_r > 1.3)| 
+                            is.na(sum_table$diff_r) | is.na(sum_table$diff_a), ]
+

### 2023-02-07 0_network_correction.R
correction

@@ -9,7 +9,7 @@
 setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
 # Read data
-line_df = read.csv("/multilines_data/preprocessed_data/all_lines/line_df.csv")
+line_df = read.csv("multilines_data/preprocessed_data/all_lines/line_df.csv")
 
 # Sum passengers by line A and R
 sum_table = data.frame(line=numeric(), a_in=numeric(), r_in=numeric(),

### 2023-02-08 latex_files/applied_network_science_article/article_v0.tex
commit ****
rewriting method

@@
 Let the transportation flow $n_{st}$ denote the number of passengers following an $st$-trip, that is entering the network at $s$ and leaving the network at $t$ by using the shortest path.
-
-*** Also, 
-at the junction level, let $\tilde{n}_{ST}$ denote the number of passengers following an $ST$-trip. As implied by the end of section (\ref{Line edges, transfer edges and trips}), 
-\begin{equation}
-\label{equivnntilde}
-\tilde{n}_{ST}=\sum_{s\in S}\sum_{t\in T}n_{st} \qquad\qquad\qquad n_{st}=\sum_{ST}\tilde{n}_{ST}\: \delta_{s,\sigma(S,T)}\: \delta_{t,\tau(S,T)}
-\end{equation}
-Definitions (\ref{edgetrip}) and (\ref{edgetripjunction}) yield
+One gets from (\ref{edgetrip})
 \begin{equation}
 \label{equationGG}
-x_{ij}=\sum_{st}\chi_{ij}^{st}\:  n_{st}=\sum_{ST}\tilde{\chi}_{ij}^{ST}\:  \tilde{n}_{ST}
+x_{ij}=\sum_{st}\chi_{ij}^{st}\:  n_{st}
 \end{equation}
@@
 
-As explained in section \ref{Lines and junctions}, transfers can only occur at junctions, that is $z_{ij}>0$ implies $S(i)=S(j)$.
-As $z_{ii}=0$ (no traveller is supposed to disembark and re-embark later at the same stop), and $z_{ii\text{-}}=0$ (where $i\text{-}$ is the stop opposite to $i$, when existing), transfers can occur at true junctions only.
+As explained in section \ref{Lines and junctions}, transfers can only occur at junctions, that is $z_{ij}>0$ implies $j\in S_i$.
+In particular,  $z_{ii}=0$ : no traveller is supposed to disembark and re-embark later at the same stop.
 
-\vspace*{0.1cm}
-
-At the junction level, let 
-\begin{displaymath}
-\tilde{a}_I=\sum_{i\in I} a_i\enspace,\enspace\tilde{b}_I=\sum_{i\in I} b_i\enspace,\enspace\tilde{z}_{\bullet I}=\sum_{i\in I} z_{\bullet i}\enspace,\enspace\tilde{z}_{I\bullet}=\sum_{i\in I} z_{i\bullet}\enspace,\enspace\tilde{n}_{I\bullet}=\sum_{i\in I} n_{i\bullet}\enspace,\enspace\tilde{n}_{\bullet I}=\sum_{i\in I} n_{\bullet i}
-\end{displaymath}
- respectively denote the number of embarking (disembarking)  passengers at junction~$I$; of embarking passengers after a transfer and of disembarking passengers before a transfer at  $I$; of passengers starting (ending) their trip at $I$.
-Then
-\begin{equation}
-\label{junctionLevel}
-\tilde{a}_I=\tilde{z}_{\bullet I}+\tilde{n}_{I\bullet}\qquad\qquad \tilde{b}_I=\tilde{z}_{I\bullet}+\tilde{n}_{\bullet I}\qquad\qquad \tilde{z}_{\bullet I}=\tilde{z}_{I\bullet}\le \min(\tilde{a}_I,\tilde{b}_I)\enspace.
-\end{equation}
-
-\subsection{Statement of the problem}
+\subsection{Statement of the problem and solution method}
 Automatic passenger counters measure the number of passengers entering and leaving buses at each stop [Boyle, 1998], that is $\mathbf{a}$ and $\mathbf{b}$.
-Also, the geometry of the network permits to derive the edge-trip incidence matrix $\bm{\chi}$ (\ref{edgetrip}), as well as the partition $\mathcal{S}$  of stops nested within junctions.
+Also, the geometry of the network permits to derive the edge-trip incidence matrix $\bm{\chi}$ defined in (\ref{edgetrip}).
 
 Intra-line edge flows $\mathbf{Y}=(y_{ij})$ can be determined by (\ref{bilan1ligne}), but transfer edge flows $\mathbf{Z}=(z_{ij})$ are, here and typically, unknown.
-The objective is to estimate the $l\times l$ transportation flow $\mathbf{N}=(n_{st})$ at the stop level, or equivalently  the $L\times L$  transportation flow  $\tilde{\mathbf{N}}=(\tilde{n}_{ST})$ at the junction level.
+The objective is to estimate the $l\times l$ transportation flow $\mathbf{N}=(n_{st})$.
 Many consistent solutions coexist in general, even for a single line with no transferts (section \ref{Single line}).
+This issue of incompletely observed data can be tackled by the maximum entropy formalism (*** ref***).
 
-This issue of incompletely observed data can be tackled by the maximum entropy formalism: let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution).
+Let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution). 
+Assuming some reasonable initial prior $g_{st}$, 
+\begin{itemize}
+\item[(1)] we shall first suppose that the empirical margins $\alpha_s=f_{s\bullet}$ and $\beta_t=f_{\bullet t}$ are known.  
+Then $f_{st}$ can be determined as the maximum entropy solution (section \ref{maxenso}), i.e.
+as the distribution closest to $g_{st}$ in the Kullback-Leibler divergence sense under the margin constraints
+\item[(2)] then (section \ref{marginup}), the margins will be updated to $\tilde{\alpha}_s$ and $\tilde{\beta}_t$   by requiring a minimum proportion $\theta\in (0,1)$ of passengers entering/leaving the network at each stop, as well as avoiding transfer overflow exceeding the embarking and disembarking counts at each stop
+ \item[(3)] finally (section \ref{priorup}), the prior will be updated to $\tilde{g}_{st}$ by contracting, if necessary, the priors $g_{st}$ associated to overflows. 
+\end{itemize}
+With the new prior distribution $\widetilde{g}_{st}$ and the new margin distributions $\widetilde{\alpha}_s$, $\widetilde{\beta}_t$, we can iterate the 
+ the above steps, until convergence. The only free parameter is $\theta$, which will be discussed in section *** . This solution method is reminiscent, but quite distinct of the ME algorithm ...***
+
+\subsubsection{***Maximum entropy solution} 
 Let $\mathcal{F}$ consist of the empirical distributions $\mathbf{f}$ satisfying $R$ linear constraints of the form $\sum_{st}f_{st}\, o^r_{st}=\bar{o}^r$ for $r=1,\ldots, R$.
@@
 
-\subsection{Statement of the problem (version 2, added by GG)}
+\subsubsection{Maximum entropy solution ***Statement of the problem (version 2, added by GG)}
+\label{maxenso}
 Let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution). 
@@
 
-\subsection{Margin distributions update (added by GG)}
+\subsubsection{Margin distributions update}
+\label{marginup}
 
@@
 
-\subsection{Prior distribution update (added by GG)}
-
+\subsubsection{Prior distribution update (added by GG)}
+\label{priorup}
 Transfer edges $(i, j)$ where $z_{i \bullet} > (1 - \theta)b_i$ or $z_{\bullet j} > (1 - \theta)a_j$ typically denote an overflowing tansfer flow regarding the $g_{st}$ prior distribution.
@@
 \end{equation}
-
-With the new prior distribution $\widetilde{g}_{st}$ and the new margin distributions $\widetilde{\alpha}_s$, $\widetilde{\beta}_t$, we can begin a new iteration step by going through 2.5 -2.7 again.
 

### 2023-02-09 latex_files/applied_network_science_article/article_v0.tex

@@
 \subsection{Statement of the problem and solution method}
-Automatic passenger counters measure the number of passengers entering and leaving buses at each stop [Boyle, 1998], that is $\mathbf{a}$ and $\mathbf{b}$.
+Automatic passenger counters measure the number of passengers entering and leaving buses at each stop [Boyle, 1998], that is $\mathbf{a}$ and $\mathbf{b}$, which provide the basic raw data of the present study, kindly provided by the Lausanne Transportation Agency (tl), after some preliminary, undocumented corrections (i.e.
+the components of $\mathbf{a}$ and $\mathbf{b}$ can be non-integer).
+It may also happen that raw data do not obey the necessary consistency condition  $a_\bullet=b_\bullet$, in which case we did rescale the embarking and disembarking counts as
+\begin{displaymath}
+\hat{a}_i=\varepsilon a_i \qquad\qquad \hat{b}_i=\frac{1}{\varepsilon} b_i \qquad\qquad\mbox{with}\quad \varepsilon=\sqrt{\frac{b_\bullet}{a_\bullet}}
+\end{displaymath}
+ensuring $\hat{\mathbf{a}}_\bullet=\hat{\mathbf{b}}_\bullet$, unless $\varepsilon> 2$ or $\varepsilon<\frac{1}{2}$ ***@R, @G : c'est à peu près cela??*** , in which case the corresponding line (which always turned out to be a temporary line with tiny counts) was simply disregarded and removed from the network.
+
 Also, the geometry of the network permits to derive the edge-trip incidence matrix $\bm{\chi}$ defined in (\ref{edgetrip}).
@@
 \item[(2)] then (section \ref{marginup}), the margins will be updated to $\tilde{\alpha}_s$ and $\tilde{\beta}_t$   by requiring a minimum proportion $\theta\in (0,1)$ of passengers entering/leaving the network at each stop, as well as avoiding transfer overflow exceeding the embarking and disembarking counts at each stop
- \item[(3)] finally (section \ref{priorup}), the prior will be updated to $\tilde{g}_{st}$ by contracting, if necessary, the priors $g_{st}$ associated to overflows. 
+ \item[(3)] finally (section \ref{priorup}), the prior will be updated to $\tilde{g}_{st}$ by shrinking, if necessary, the priors $g_{st}$ associated to overflows. 
 \end{itemize}
 With the new prior distribution $\widetilde{g}_{st}$ and the new margin distributions $\widetilde{\alpha}_s$, $\widetilde{\beta}_t$, we can iterate the 
- the above steps, until convergence. The only free parameter is $\theta$, which will be discussed in section *** . This solution method is reminiscent, but quite distinct of the ME algorithm ...***
+ the above steps, until convergence.
+The only free parameter is $\theta$, which will be discussed in section *** .
+*** This solution method is reminiscent, but quite distinct of the ME algorithm ...***
 
-\subsubsection{***Maximum entropy solution} 
-Let $\mathcal{F}$ consist of the empirical distributions $\mathbf{f}$ satisfying $R$ linear constraints of the form $\sum_{st}f_{st}\, o^r_{st}=\bar{o}^r$ for $r=1,\ldots, R$.
-Then the minimum Kullback-Leibler divergence
-\begin{displaymath}
-\min_{\mathbf{f}\in\mathcal{F}}\;  \sum_{st}f_{st}\log \frac{f_{st}}{g_{st}}
-\end{displaymath}
-is reached for the maximum entropy solution  
-\begin{equation}
-\label{MaxEnt}
-f^0_{st}=\frac{1}{Z(\bm{\lambda})}\: g_{st}\exp(-\sum_{r=1}^R \lambda_r o^r_{st})\qquad\qquad Z(\bm{\lambda})=\sum_{st}g_{st}\exp(-\sum_{r=1}^R \lambda_r o^r_{st})
-\end{equation}
-where the Lagrange multipliers $\bm{\lambda}$ are determined  so as to satisfy the $R$ linear constraints (**see e.g. Bavaud and *** and references therein).
-
-\subsubsection{Maximum entropy solution ***Statement of the problem (version 2, added by GG)}
+\subsubsection{Maximum entropy estimate of $st$-trips}
 \label{maxenso}
-Let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution). 
-Also suppose a prior guess of row and margin constraints for $f_{st}$, respectively $\alpha_s$ and $\beta_t$, which has to verify 
-\begin{equation}
-	\label{init_margin}
-	\frac{\alpha_i}{\alpha_j} = \frac{a_i}{a_j} \qquad \frac{\beta_i}{\beta_j} = \frac{b_i}{b_j}  \qquad \forall i, j \text{ which are not true junctions.} \\
-\end{equation}
-Setting these gives us
-\begin{equation}
-	\label{flow_sum}
-	n_{\bullet \bullet} = \frac{a_i}{\alpha_i} = \frac{b_i}{\beta_i} \qquad \forall i \text{ which are not true junctions.}  
-\end{equation}
-We are interested in finding
+As announced,  the proportion of $st$-trips $f_{st}=n_{st}/n_{\bullet\bullet}$ (empirical distribution) will be estimated from some prior guess $g_{st}$ (theoretical distribution) and 
+margin constraints $\alpha_s$ and $\beta_t$ for $f_{st}$ by maximum entropy, i.e. by 
+solving the problem 
 \begin{align}
@@
 \end{equation}
-which gives an \emph{iterative fitting algorithm}. Starting with any $\psi^{(0)}_t > 0$, we do
+which yields the following \emph{iterative fitting algorithm}: starting with some $\psi^{(0)}_t > 0$, one performs the iteration
 \begin{equation}
@@
 \end{equation}
+until convergence to $\phi_s$ and $\psi_t$ obeying (\ref{Sol_LagMult}).
 
-With equation (\ref{flow_sum}), we can obtain the flow associated to $f_{st}$
+In view of (\ref{entrer}) and {\ref{sortir}), the postulated margins must satisfy, for each isolated stop $i$
+\begin{equation}
+\label{ }
+\alpha_i=\frac{a_i}{n_{\bullet \bullet}}\qquad\qquad \beta_i=\frac{b_i}{n_{\bullet \bullet}}
+\end{equation}
+permitting to determine the total flow as $n_{\bullet \bullet}=\frac{a_i}{\alpha_i}$, or  $n_{\bullet \bullet}=\frac{b_i}{\beta_i}$ for any isolated stop $i$, and thus 
+the $st$-flow itself as 
 \begin{equation}
 	\label{flow_from_distrib}
-	n_{st} = n_{\bullet \bullet} f_{st}, 
+	n_{st} = n_{\bullet \bullet} f_{st}= n_{\bullet \bullet}\phi_s \psi_t g_{st} 
 \end{equation}
-and with the edge-trip incidence matrix $\bm{\chi}$ (\ref{edgetrip}), we obtain the intra-line edge flows $\mathbf{Y}=(y_{ij})$ and the transfer edge flows $\mathbf{Z}=(z_{ij})$.
+whose plugging into (\ref{equationGG}) yields the intra-line edge flows $\mathbf{Y}=(y_{ij})$ and the transfer edge flows $\mathbf{Z}=(z_{ij})$.
 
-\subsubsection{Margin distributions update}
+\subsubsection{Initialization of the prior and the margins}
+The geometry of the network permits to rule out forbidden $st$-paths, i.e.
+obeying  $\chi^{st}_{\bullet\bullet}=0$.
+Such forbidden $st$-paths consist of *** @G?
+donner une liste exhaustive de critères d'exclusion ici ?
+***  The initial prior was chosen as the uniform distribution on the remaining $c<l^2$ admissible $st$-paths, *** @G: que vaut $l$?
+que vaut $c$?
+***,  that is as
+\begin{equation*}
+g_{ij} = \begin{cases}
+  \frac{1}{c}    & \text{if $\chi^{st}_{\bullet\bullet}>0$}, \\
+  0    & \text{otherwise}.
+\end{cases}
+\end{equation*}
+The initial margins were chosen as  those of the flow without transfer, namely $\alpha_i=\frac{a_i}{a_{\bullet}}$ and $\beta_i=\frac{b_i}{b_\bullet}$ for all stops.
+
+\subsubsection{Updating  the margin distributions}
 \label{marginup}
-
-Let us define the hyperparameter $0 \geq \theta > 1$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the network at each stop}.
-Note that we could set a different hyperparmeter for each node, and for embarkations and disembarkations, but without addition information, we will restrain to this simpler case.
-This hyperparamter means that
+Let us define the hyperparameter $ \theta\in [0, 1)$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the network at each stop}.
+Note that we could set a different hyperparameter for each node, and differing for embarkations and disembarkations, but without addition information, we will restrain to this simpler case.
+Fixing this hyperparameter requires that
 \begin{align}
+\label{constz}
 	z_{\bullet s} &< (1 - \theta) a_s \\
@@
 \end{align}
-and with constraints (\ref{entrer}) and (\ref{sortir}), the update of margins distribution must be
+the violation of which constitutes transfer overflow. Hence requiring a minimal transfer yet avoiding overflow can be satisfied with the updating of margins ***
 \begin{align}
-	\widetilde{\alpha}_s &= \frac{\min(\theta a_s, a_s - z_{\bullet s})}{\sum_{s'} \min(\theta a_s, a_{s'} - z_{\bullet {s'}})} \label{alpha_update} \\
-	\widetilde{\beta}_t &= \frac{\min(\theta b_t, b_t - z_{t \bullet})}{\sum_{t'} \min(\theta b_t, b_{t'} - z_{{t'} \bullet})} \label{beta_update}
+	\widetilde{\alpha}_s &= \frac{\min(\theta a_s, a_s - z_{\bullet s})}{\sum_{s'} \min(\theta a_{s'}, a_{s'} - z_{\bullet {s'}})} \label{alpha_update} \\
+	\widetilde{\beta}_t &= \frac{\min(\theta b_t, b_t - z_{t \bullet})}{\sum_{t'} \min(\theta b_{t'}, b_{t'} - z_{{t'} \bullet})} \label{beta_update}\enspace. 
 \end{align}
 
-\subsubsection{Prior distribution update (added by GG)}
+\subsubsection{Updating the prior distribution}
 \label{priorup}

### 2023-02-09 latex_files/applied_network_science_article/article_v0.tex
end of chapter 1

@@
 \end{displaymath}
-ensuring $\hat{\mathbf{a}}_\bullet=\hat{\mathbf{b}}_\bullet$, unless $\varepsilon> 2$ or $\varepsilon<\frac{1}{2}$ ***@R, @G : c'est à peu près cela??*** , in which case the corresponding line (which always turned out to be a temporary line with tiny counts) was simply disregarded and removed from the network.
+ensuring $\hat{a}_\bullet=\hat{b}_\bullet$, unless $\varepsilon> 2$ or $\varepsilon<\frac{1}{2}$ ***@R, @G : c'est à peu près cela??*** , in which case the corresponding line (which always turned out to be a temporary line with small  counts) was simply disregarded and removed from the network.
 
@@
 Many consistent solutions coexist in general, even for a single line with no transferts (section \ref{Single line}).
-This issue of incompletely observed data can be tackled by the maximum entropy formalism (*** ref***).
+This issue of incompletely observed data can be tackled by the maximum entropy formalism \cite{jaynes1957information}, and has been often  in transportation modelling \cite{wilson1967statistical}  \cite{erlander1990gravity}.
+Its contemporary justification corresponds to the ``expectation step" of the EM algorithm \cite{dempster1977maximum}  \cite{bavaud2009information}.
 
@@
 
-In view of (\ref{entrer}) and {\ref{sortir}), the postulated margins must satisfy, for each isolated stop $i$
+In view of (\ref{entrer}) and (\ref{sortir}), the postulated margins must satisfy, for each isolated stop $i$
 \begin{equation}
@@
 \label{marginup}
-Let us define the hyperparameter $ \theta\in [0, 1)$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the network at each stop}.
-Note that we could set a different hyperparameter for each node, and differing for embarkations and disembarkations, but without addition information, we will restrain to this simpler case.
-Fixing this hyperparameter requires that
-\begin{align}
-\label{constz}
-	z_{\bullet s} &< (1 - \theta) a_s \\
-	z_{t \bullet} &< (1 - \theta) b_t
-\end{align}
-the violation of which constitutes transfer overflow. Hence requiring a minimal transfer yet avoiding overflow can be satisfied with the updating of margins ***
-\begin{align}
-	\widetilde{\alpha}_s &= \frac{\min(\theta a_s, a_s - z_{\bullet s})}{\sum_{s'} \min(\theta a_{s'}, a_{s'} - z_{\bullet {s'}})} \label{alpha_update} \\
-	\widetilde{\beta}_t &= \frac{\min(\theta b_t, b_t - z_{t \bullet})}{\sum_{t'} \min(\theta b_{t'}, b_{t'} - z_{{t'} \bullet})} \label{beta_update}\enspace. 
-\end{align}
+Let us define the hyperparameter $ \theta\in [0, 1)$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the network at each stop}, that is $n_{s\bullet}\ge \theta a_s$ and $n_{\bullet t}\ge \theta b_t$.
+Note that we could set a different hyperparameter for each node, and differing for embarkments and disembarkments, but without addition information, we will restrain to this simpler case.
+Identities (\ref{entrer}) and  (\ref{sortir}) then imply the inequalities
+\begin{displaymath}
+z_{\bullet s} \le (1 - \theta) a_s\qquad\qquad \qquad z_{t \bullet} \le  (1 - \theta) b_t
+\end{displaymath}
+the violation of which constitutes transfer overflow. Hence requiring a minimal transfer yet avoiding overflow can be granted with the following updating of margins
+\begin{equation}
+\label{alpha_beta_update}
+\widetilde{\alpha}_s = \frac{\min(\theta a_s, a_s - z_{\bullet s})}{\sum_{s'} \min(\theta a_{s'}, a_{s'} - z_{\bullet {s'}})}   \qquad \qquad \qquad
+	\widetilde{\beta}_t = \frac{\min(\theta b_t, b_t - z_{t \bullet})}{\sum_{t'} \min(\theta b_{t'}, b_{t'} - z_{{t'} \bullet})}  \enspace. 
+\end{equation}
 
@@
 \label{priorup}
-Transfer edges $(i, j)$ where $z_{i \bullet} > (1 - \theta)b_i$ or $z_{\bullet j} > (1 - \theta)a_j$ typically denote an overflowing tansfer flow regarding the $g_{st}$ prior distribution.
-The latter must be adjusted in order to have $\mathbf{Z}$ comptible with constraints.
-For any edge $(i, j)$, let us compute the \emph{flow ratio} $r_{ij}$, with
+Overflow occurs in transfer edge $(i, j)$ if  $z_{i \bullet} > (1 - \theta)b_i$ or $z_{\bullet j} > (1 - \theta)a_j$. To avoid it, components $g_{st}$ of the prior distribution 
+will be shrinked by a suitable ratio whenever edge flows $(i,j)\in [s,t]$ exhibit overflow.
+
+ For any edge $(i, j)$, let us compute the \emph{flow ratio} $r_{ij}$ as
 \begin{equation}
 	\label{flow_ratio}
-	r_{ij} = \max \left(1, \frac{z_{i \bullet}}{(1 - \theta)b_i}, \frac{z_{\bullet j}}{(1 - \theta)a_j} \right),
+	r_{ij} = \max \left(1, \frac{z_{i \bullet}}{(1 - \theta)b_i}, \frac{z_{\bullet j}}{(1 - \theta)a_j} \right)\: \ge\: 1\enspace ,
 \end{equation}
-where $r_{ij} > 1$ typically denotes an overflow through edge $(i, j)$.
-This flow ratio can be traced back to origin-destination couples $s, t$ in order to compute the \emph{orgin-destination flow ratio} $\bar{r}_{st}$
+where $r_{ij} > 1$ denotes an overflow through edge $(i, j)$. For a given origin-destination $[s,t]$, define the \emph{orgin-destination flow ratio} $\bar{r}_{st}$ 
+as the largest $r_{ij}$ among edge flows $(i,j)\in [s,t]$, that is as 
 \begin{equation}
 	\label{st_flow_ratio}
-	\bar{r}_{st} = \max_{ij} \chi_{ij}^{st} r_{ij},
+	\bar{r}_{st} = \max_{ij} \chi_{ij}^{st} r_{ij}\: \ge\: 1\enspace ,
 \end{equation}
-where again $\bar{r}_{st} > 1$ denotes an overflow between $s$ and $t$. To ajdust the flow, we can devide the previous flow by this ratio
+where again $\bar{r}_{st} > 1$ denotes an overflow between $s$ and $t$. To adjust the flow, we shall divide the previous flow by this ratio
 \begin{equation}
@@
 \end{equation}
-And the new prior distribution should follow 
+and define the new prior distribution  as
 \begin{equation}
 	\label{update_distrib}
-	\widetilde{g}_{st} = \frac{\left( \frac{\widetilde{n}_{st}}{\phi_s \psi_t} \right)}{\sum_{s',t'} \left( \frac{\widetilde{n}_{s',t'}}{\phi_{s'} \psi_{t'}} \right)}
+	\widetilde{g}_{st} = \frac{\left( \frac{\widetilde{n}_{st}}{\phi_s \psi_t} \right)}{\sum_{s',t'} \left( \frac{\widetilde{n}_{s',t'}}{\phi_{s'} \psi_{t'}} \right)}\enspace. 
 \end{equation}
+where $\phi_s$ and $\psi_t$ are the values (\ref{Sol_LagMult}) obtained in the previous maximum entropy step.
 

### 2023-02-09 latex_files/applied_network_science_article/article_v0.tex
rewriting the method (in progress)

@@
 
-%% --------------------------------- OLD IMPORT
+%% --------------------------------- INTRODUCTION
+
+\section{Introduction}
+
+\subsection{Context}
+
+\subsection{Statement of the problem}
+
+\subsection{Related Works}
+
+%% --------------------------------- NOTATIONS AND FORMALISM
 
@@
 	\label{flow_ratio}
-	r_{ij} = \max \left(1, \frac{z_{i \bullet}}{(1 - \theta)b_i}, \frac{z_{\bullet j}}{(1 - \theta)a_j} \right)\: \ge\: 1\enspace ,
+	r_{ij} = \max \left(1, \frac{z_{i \bullet}}{(1 - \theta)b_i}, \frac{z_{\bullet j}}{(1 - \theta)a_j} \right)\: \ge\: 1\enspace,
 \end{equation}
@@
 	\label{st_flow_ratio}
-	\bar{r}_{st} = \max_{ij} \chi_{ij}^{st} r_{ij}\: \ge\: 1\enspace ,
+	\bar{r}_{st} = \max_{ij} \chi_{ij}^{st} r_{ij}\: \ge\: 1\enspace.
 \end{equation}
-where again $\bar{r}_{st} > 1$ denotes an overflow between $s$ and $t$. To adjust the flow, we shall divide the previous flow by this ratio
+By construction, $\bar{r}_{st} > 1$ denotes an overflow on some transfer edge between $s$ and $t$. To adjust the flow, we shall divide the previous flow by this ratio
 \begin{equation}
@@
 
-\section{Single line}
+\subsection{The case of a single line}
 \label{Single line}
+A ``network" made of a single line contains no transfers, and flow estimates can be obtained at once by the maximum entropy step only.
+
 Let  $i=1,\ldots, l$ enumerate the bus stops in increasing order,  i.e.
 $F(i)=i+1$.
-Define $g_{st}=\frac{1}{(l-1)(l-2)}\chi(s<t)$ (where $\chi(.)$ denotes the 0/1  indicator function) as the maximally uniform prior, reflecting only the unidirectional nature of trips.
-Constraints $n_{r\bullet}=a_r$ (resp. $n_{\bullet r}=b_r$), correspond to $o^r_{st}=
-\delta_{sr}$ (resp. $o^r_{st}=\delta_{tr}$) in (\ref{MaxEnt}), which finally yields, after reparametrization, the maximum entropy flow reads
+The initial prior is simply $g_{st}=\frac{I(s<t)}{(l-1)(l-2)}$ (where $I(.)$ denotes the 0/1  indicator function), reflecting only the unidirectional nature of trips.
+The margins of the empirical distribution $f_{st}$ as well as the total flow are know, namely
+\begin{displaymath}
+\alpha_s=\frac{a_s}{a_\bullet}\qquad\qquad\qquad \beta_t=\frac{b_t}{b_\bullet}\qquad\qquad\qquad n_{\bullet\bullet}=a_{\bullet}=b_\bullet
+\end{displaymath}
+and (\ref{Sol}) yields, after reparameterization, maximum entropy flows of the form
 \begin{equation}
 \label{nosignle}
-n^0_{st}=I(s<t)\, c_s\, d_t \qquad\qquad\mbox{where}\quad \sum_{s<t}c_s\, d_t=n_{\bullet\bullet}=a_{\bullet}=b_\bullet
+n_{st}=I(s<t)\, c_s\, d_t \qquad\qquad\mbox{where}\quad \sum_{s}c_s\, \sum_{t>s}d_t=n_{\bullet\bullet}
 \end{equation}
-In addition, the (dis-)embarking constraints  yield
+where the constraints (\ref{Sol_LagMult}) equivalently express as 
 \begin{equation}
-\label{(dis embarking constraints}
-a_i=c_i \sum_{t>i}d_t=c_i\, D_i 
+\label{dis embarking constraints}
+c_s=\frac{a_s}{\sum_{t>s}d_t}
 \qquad\qquad\qquad
-b_j=d_j\sum_{s<j}c_s=d_j \, C_j  
+d_t=\frac{b_t}{\sum_{s<t}c_s}
 \end{equation}
-where $D_i=\sum_{t>i}d_t$ and $C_j=\sum_{s<j}c_s$.
+to be solved by iterative fitting.
+
+The form (\ref{nosignle}) for the flows is reminiscent of the {\em gravity flows} \cite{wilson1967statistical}  \cite{erlander1990gravity}, where 
+$c_s$ is the {\em push factor}, $d_t$ is the {\em pull factor}, and $I(s<t)$ the {\em distance deterrence function}. Yet, instead of  
+being symmetric in $s,t$ and decreasing with the distance $|s-t|$, the distance deterrence function is here asymmetric due to line orientation, but otherwise constant.
+This constancy entails the following Markovian behavior of flows: let $m_{st}$ be the number of travelers embarking on $s$ and still present at stop $t>s$, and $\rho_s$ the probability to disembark at $s$.
+Then, by definition and (\ref{nosignle}), ****
+\begin{displaymath}
+n_{st}=m_{st}\rho_t \qquad\qquad n_{s,t+1}=m_{st}(1-\rho_t)\rho_{t+1}\qquad\qquad \frac{n_{s,t+1}}{n_{st}}=\frac{(1-\rho_t)\rho_{t+1}}{\rho_t}=\frac{d_{t+1}}{d_t}
+\end{displaymath}
+Ainsi 
+\begin{displaymath}
+\frac{d_{t+2}}{d_{t+1}}=\frac{(1-\rho_{t+1})\rho_{t+2}}{\rho_{t+1}}
+\qquad\qquad \frac{d_{t+2}}{d_{t}}=\frac{d_{t+2}}{d_{t+1}}\frac{d_{t+1}}{d_{t}}=\frac{(1-\rho_{t+1})\rho_{t+2}}{\rho_{t+1}}\frac{(1-\rho_t)\rho_{t+1}}{\rho_t}=
+\rho_{t+2}(1-\rho_{t+1})(1-\rho_{t})\frac{1}{\rho_{t}}
+\end{displaymath}
+\begin{displaymath}
+\frac{d_{t+q}}{d_{t}}=
+\rho_{t+q}(1-\rho_{t+q-1})(1-\rho_{t+q-2})\ldots(1-\rho_{t+1}) (1-\rho_{t})\frac{1}{\rho_{t}}
+\end{displaymath}
+\begin{displaymath}
+d_l+d_{l-1}=d_2[\rho_l(1-\rho_{l-1})+\rho_{l-1}](1-\rho_{l-2})(1-\rho_{l-3})\ldots(1-\rho_{3}) (1-\rho_{2})\frac{1}{\rho_{2}}
+\end{displaymath}
+\begin{displaymath}
+D_i=\sum_{t>i}d_t
+\end{displaymath}
+Soit $z_{st}$ la proportion de voyageurs présents dans le bus en $s$ et toujours présents en $t>s$. Soit $\tau_s$ la probabilité de monter en $s$. Alors
+$f_{st}=\tau_sz_{st}\rho_t$
+
+\
+
+\
+
+*****
+
+where $D_s=\sum_{t>s}d_t$ and $C_t=\sum_{s<t}c_s$.
 This maximum entropy solution constitutes one possible consistent transportation flow among many others, such as the ``first in, first out" (FIFO) flow.
@@
 
-\section{Multiple lines}
+%% --------------------------------- CASE STUDIES
 
-*** $a_{\bullet}=b_{\bullet}$ (where ``$\bullet$" denotes summation over the replaced index).
+\section{Case Studies}
+
+\subsection{Toy Examples}
+
+\subsection{Real Data}
+
+%% --------------------------------- CONCLUSION
+
+\section{Conclusion}
 
@@
 
-%% --------------------------------- INTRODUCTION
-
-\section{Introduction}
-
-\subsection{Context}
-
-\subsection{Statement of the problem}
-
-\subsection{Related Works}
-
-%% --------------------------------- FORMALISM
-
-\section{Formalism}
-
-\subsection{Notations}
-
-\subsection{Problem}
-
-\subsection{Solution}
-
-%% --------------------------------- CASE STUDIES
-
-\section{Case Studies}
-
-\subsection{Toy Examples}
-
-\subsection{Real Data}
-
-%% --------------------------------- CONCLUSION
-
-\section{Conclusion}
-
 \section*{Appendix}

### 2023-02-10 latex_files/applied_network_science_article/article_v0.tex
Finished (for now) with chapter2: Notations and formalism

@@
 $F(i)=i+1$.
-The initial prior is simply $g_{st}=\frac{I(s<t)}{(l-1)(l-2)}$ (where $I(.)$ denotes the 0/1  indicator function), reflecting only the unidirectional nature of trips.
-The margins of the empirical distribution $f_{st}$ as well as the total flow are know, namely
+The initial prior is simply $g_{st}=\frac{I(s<t)}{(l-1)(l-2)}$ (where $I(.)$ denotes the 0/1  indicator function), capturing solely the unidirectional nature of trips.
+The margins of the empirical distribution $f_{st}$, as well as the total flow, are here known :
 \begin{displaymath}
@@
 \end{equation}
-where the constraints (\ref{Sol_LagMult}) equivalently express as 
+where (setting $D_s:=\sum_{t>s}d_t$ and $C_t:=\sum_{s<t}c_s$) the constraints (\ref{Sol_LagMult}) equivalently express as 
 \begin{equation}
 \label{dis embarking constraints}
-c_s=\frac{a_s}{\sum_{t>s}d_t}
+c_s=\frac{a_s}{\sum_{t>s}d_t}=\frac{a_s}{D_s}
 \qquad\qquad\qquad
-d_t=\frac{b_t}{\sum_{s<t}c_s}
+d_t=\frac{b_t}{\sum_{s<t}c_s}=\frac{b_t}{C_t}
 \end{equation}
@@
 
-The form (\ref{nosignle}) for the flows is reminiscent of the {\em gravity flows} \cite{wilson1967statistical}  \cite{erlander1990gravity}, where 
+Interestingly enough, the form (\ref{nosignle}) for the flows is reminiscent of the {\em gravity flows} of quantitative Geography \cite{wilson1967statistical}  \cite{erlander1990gravity} \cite{bavaud2002quasi}, where
 $c_s$ is the {\em push factor}, $d_t$ is the {\em pull factor}, and $I(s<t)$ the {\em distance deterrence function}. Yet, instead of  
-being symmetric in $s,t$ and decreasing with the distance $|s-t|$, the distance deterrence function is here asymmetric due to line orientation, but otherwise constant.
-This constancy entails the following Markovian behavior of flows: let $m_{st}$ be the number of travelers embarking on $s$ and still present at stop $t>s$, and $\rho_s$ the probability to disembark at $s$.
-Then, by definition and (\ref{nosignle}), ****
+being symmetric in $s,t$ and decreasing with the distance $|s-t|$, the distance deterrence function is here asymmetric due to the line orientation, but otherwise constant.
+
+This constancy entails the following Markovian behaviour for flows: let $m_{st}$ be the number of travelers embarking at stop $s$ and still inside the bus at stop $t>s$, and let $\rho_{st}$ the probability that travelers embarking at $s$ will disembark at $t$.
+By (\ref{nosignle}),
 \begin{displaymath}
-n_{st}=m_{st}\rho_t \qquad\qquad n_{s,t+1}=m_{st}(1-\rho_t)\rho_{t+1}\qquad\qquad \frac{n_{s,t+1}}{n_{st}}=\frac{(1-\rho_t)\rho_{t+1}}{\rho_t}=\frac{d_{t+1}}{d_t}
+m_{st}=\sum_{u\ge t}n_{su}=c_s \sum_{u\ge t} I(s<u)\, d_u =c_s (d_t+D_t) 
 \end{displaymath}
-Ainsi 
+The empirical estimate of $\rho_{st}$ is given by the proportion, among the travelers embarking at $s$ and 
+ still present at $t>s$,  of travelers disembarking at $t$, that is 
 \begin{displaymath}
-\frac{d_{t+2}}{d_{t+1}}=\frac{(1-\rho_{t+1})\rho_{t+2}}{\rho_{t+1}}
-\qquad\qquad \frac{d_{t+2}}{d_{t}}=\frac{d_{t+2}}{d_{t+1}}\frac{d_{t+1}}{d_{t}}=\frac{(1-\rho_{t+1})\rho_{t+2}}{\rho_{t+1}}\frac{(1-\rho_t)\rho_{t+1}}{\rho_t}=
-\rho_{t+2}(1-\rho_{t+1})(1-\rho_{t})\frac{1}{\rho_{t}}
+\rho_{st}=\frac{n_{st}}{m_{st}}=\frac{c_s\,  d_t}{c_s\,  (d_t+D_t)}=\frac{d_t}{d_t+D_t}=\frac{b_t}{C_t\, (d_t+D_t)}\le 1
 \end{displaymath}
-\begin{displaymath}
-\frac{d_{t+q}}{d_{t}}=
-\rho_{t+q}(1-\rho_{t+q-1})(1-\rho_{t+q-2})\ldots(1-\rho_{t+1}) (1-\rho_{t})\frac{1}{\rho_{t}}
-\end{displaymath}
-\begin{displaymath}
-d_l+d_{l-1}=d_2[\rho_l(1-\rho_{l-1})+\rho_{l-1}](1-\rho_{l-2})(1-\rho_{l-3})\ldots(1-\rho_{3}) (1-\rho_{2})\frac{1}{\rho_{2}}
-\end{displaymath}
-\begin{displaymath}
-D_i=\sum_{t>i}d_t
-\end{displaymath}
-Soit $z_{st}$ la proportion de voyageurs présents dans le bus en $s$ et toujours présents en $t>s$. Soit $\tau_s$ la probabilité de monter en $s$. Alors
-$f_{st}=\tau_sz_{st}\rho_t$
+which depends on $t$ only: it  finally  turns out that the disembarkment probability $\rho_t=\frac{d_t}{d_t+D_t}$ at $t$ is {\em independent} of the embarkment stop $s$.
+Said otherwise, a traveler embarking at any stop $s$ (and thus necessarily in the bus at $F(s)=s+1$) experiences the {\em same disembarkment probability} at each  stop $t$.
+
+This Markov property, enjoyed by maximum-entropic flows, contrasts other possible solutions, such as  the ``first in, first out" (FIFO) flows (homogenizing the 
+traveled distances among users) or the ``last in, first out" (LIFO) flows (tending to generate maximally  contrasted traveled distances).
 
@@
 
-\
-
-*****
-
-where $D_s=\sum_{t>s}d_t$ and $C_t=\sum_{s<t}c_s$.
-This maximum entropy solution constitutes one possible consistent transportation flow among many others, such as the ``first in, first out" (FIFO) flow.
-Interestingly enough, (\ref{nosignle}) is reminiscent of the so-called gravity flows of quantitative Geography (**ref**) $n_{st}=h_{st}\, c_s\, d_t $, but with a purely asymmetric ``distance deterrence function" $h_{st}=I(s<t)$.
-Also,
-(\ref{nosignle}) shows that  the conditional probability to exit on $t$, given an entrance on $s$, is $w_{st}=n^0_{st}/n^0_{s\bullet}=I(s<t)\, d_t/D_s$.
+***  @G, ici davantage de développements sur l'approche Markovienne, ainsi que le joli schéma ?  ***
 
@@
 
-*** ici la (les) figure de l'exemple  ``starting from Maladière, Riant-Cour, Dapples"... ? ***
-
-\
-
-*** ici l'équivalence avec l'approche chaîne de Markov de Guillaume ? ***
+*** @G, @R, ici la (les) figure de l'exemple  ``starting from Maladière, Riant-Cour, Dapples"... ? ***
 

### 2023-02-11 latex_files/applied_network_science_article/article_v0.tex
writing two paragraphs in the intro

@@
 \section{Introduction}
+Transportation networks determine our mobility, require a considerable amount of planning and resources, and elicit much hopes and critics. 
+They also constitute an endless source of inspiration in formal modeling and optimization, as attested in operations research (classical optimal transportation, maximum flow problem), quantitative geography and spatial econometrics (spatial navigation, multimodality, gravity models for flows), and machine learning (recent developments in regularized optimal transportation, such as color transfer or images interpolation; see e.g.
+\cite{peyre2019computational}).
+
+This contribution adresses a simple, yet central issue in public transportation networks: given a network made of many bus lines, how can one estimate the real trips made by the travelers, on the sole basis of  the embarkment and disembarkment counts for each bus stop?
+Transfer of travelers between bus lines constitute the missing information, and their defendable evaluation requests some methodological reflexion and experimentation.
+
+***
 
@@
 \end{displaymath}
-which depends on $t$ only: it  finally  turns out that the disembarkment probability $\rho_t=\frac{d_t}{d_t+D_t}$ at $t$ is {\em independent} of the embarkment stop $s$.
-Said otherwise, a traveler embarking at any stop $s$ (and thus necessarily in the bus at $F(s)=s+1$) experiences the {\em same disembarkment probability} at each  stop $t$.
+which depends on $t$ only: it appears that the disembarkment probability $\rho_t=\frac{d_t}{d_t+D_t}$ at $t$ is {\em independent} of the embarkment stop $s$.
+Said otherwise, a traveler embarking at any stop $s$ (and thus necessarily in the bus at $F(s)=s+1$) experiences the {\em same disembarkment probability} at each further stop $t>s$.
 

### 2023-02-11 latex_files/applied_network_science_article/article_v0.tex
writing the beginning of the abstract, introduction and theory

@@
 
-\title{Estimation of flow trajectories in a multi-lines network}
+\title{Estimation of flow trajectories in a multi-lines transportation network}
 
@@
 \address[id=aff1]{%                           
-  \orgdiv{Departement of Language and Information Sciences},             
+  \orgdiv{Department of Language and Information Sciences},             
   \orgname{University of Lausanne},          
@@
 \begin{abstract} % abstract
-Blabla
+Characterizing a public transportation network, such as an urban multi-lines bus network, requires the origin-destination trip counts during a given period.
+Yet, if automatic counting makes the embarkment (boarding) and disembarkment (alighting) counts at each bus stop known, it often happens that pedestrian transfers between stops are unknown, and this contribution proposes a three-steps procedure for estimating the missing information, involving maximum entropy and iterative fitting.
+** à poursuivre ***
 \end{abstract}
@@
 \begin{keyword}
-\kwd{sample}
-\kwd{article}
-\kwd{author}
+\kwd{multiline bus network}
+\kwd{origin-destination flows}
+\kwd{boarding and alighting counts}
+\kwd{transit flows}
+\kwd{maximum entropy estimation}
 \end{keyword}
@@
 \section{Introduction}
-Transportation networks determine our mobility, require a considerable amount of planning and resources, and elicit much hopes and critics. 
+% Alternative title: Estimating transit flows from boarding and alighting counts
+Transportation networks determine our mobility, require a considerable amount of planning and resources, and elicit much public hopes and critics. 
 They also constitute an endless source of inspiration in formal modeling and optimization, as attested in operations research (classical optimal transportation, maximum flow problem), quantitative geography and spatial econometrics (spatial navigation, multimodality, gravity models for flows), and machine learning (recent developments in regularized optimal transportation, such as color transfer or images interpolation; see e.g.
@@
 
-This contribution adresses a simple, yet central issue in public transportation networks: given a network made of many bus lines, how can one estimate the real trips made by the travelers, on the sole basis of  the embarkment and disembarkment counts for each bus stop?
-Transfer of travelers between bus lines constitute the missing information, and their defendable evaluation requests some methodological reflexion and experimentation.
+This contribution adresses a straightforward, yet central issue in public transportation networks: given a network made of many bus lines, how can one estimate the real trips made by the travelers, on the sole basis of  the embarkment (boarding) counts and disembarkment (alighting) counts for each bus stop?
+Although estimating origin-destination flows is a central issue in transportation modeling (see e.g \cite{bell1997transportation} \cite{hazelton2000estimation} \cite{ashok2002estimation} \cite{cui2006bus} and references therein), the specific problem addressed in this contribution seems, to the best of our knowledge, original.
 
-***
+Pedestrian transfers of travelers between bus lines here constitute the missing information, whose principled evaluation require some methodological reflexion and experimentation.
+Section \ref{notforma} introduces the notations and the formalism, as well as the statement of the problem and the iterative solution method, which consist of three consecutive steps: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior (section \ref{maxenso}); an update of the marginal flows to avoid transfer overflow (section \ref{marginup}); and an update of the prior distribution  (section \ref{priorup}) by shrinking the components responsible for overflow.
 
-\subsection{Context}
+The fist step only is required for solving the single line case (section \ref{Single line}), naturally much simpler but  yet  not trivial, and exhibiting a disembarking probability independent of the embarking stop (Markov property).
 
-\subsection{Statement of the problem}
-
-\subsection{Related Works}
+Cases studies are presented in section \ref{casestudies} *** à poursuivre ***
 
@@
 \section{Notations and formalism}
+\label{notforma}
 \subsection{Lines, stops  and junctions}
@@
 This issue of incompletely observed data can be tackled by the maximum entropy formalism \cite{jaynes1957information}, and has been often  in transportation modelling \cite{wilson1967statistical}  \cite{erlander1990gravity}.
-Its contemporary justification corresponds to the ``expectation step" of the EM algorithm \cite{dempster1977maximum}  \cite{bavaud2009information}.
 
@@
 Then $f_{st}$ can be determined as the maximum entropy solution (section \ref{maxenso}), i.e.
-as the distribution closest to $g_{st}$ in the Kullback-Leibler divergence sense under the margin constraints
+as the distribution closest to $g_{st}$ in the Kullback-Leibler divergence sense under the margin constraints, to be calibrated by an iterative fitting inner loop
 \item[(2)] then (section \ref{marginup}), the margins will be updated to $\tilde{\alpha}_s$ and $\tilde{\beta}_t$   by requiring a minimum proportion $\theta\in (0,1)$ of passengers entering/leaving the network at each stop, as well as avoiding transfer overflow exceeding the embarking and disembarking counts at each stop
@@
 With the new prior distribution $\widetilde{g}_{st}$ and the new margin distributions $\widetilde{\alpha}_s$, $\widetilde{\beta}_t$, we can iterate the 
- the above steps, until convergence.
-The only free parameter is $\theta$, which will be discussed in section *** .
-*** This solution method is reminiscent, but quite distinct of the ME algorithm ...***
+ the above steps, until convergence. The only free parameter is $\theta$, which will be discussed in section *** .
+
+ *** The above iterative solution method is somewhat reminiscent of the EM algorithm.
+As a matter of fact, the first  step  (maximum entropy) exactly correspond to the ``expectation step" of the EM algorithm (see e.g.
+\cite{dempster1977maximum}  \cite{bavaud2009information}), but steps two and three, aiming at calibrating the parameters $\bm{\alpha}$, $\bm{\beta}$ and $\mathbf{g}$, do not follow the maximum likelihood rationale of the ``maximisation step".
 
@@
 
-Interestingly enough, the form (\ref{nosignle}) for the flows is reminiscent of the {\em gravity flows} of quantitative Geography \cite{wilson1967statistical}  \cite{erlander1990gravity} \cite{bavaud2002quasi}, where
+Interestingly enough, the form (\ref{nosignle}) for the flows is reminiscent of the {\em gravity flows} of quantitative Geography \cite{wilson1967statistical}  \cite{erlander1990gravity} \cite{bavaud2002quasi} \cite{Thomas-Agnan2021}, where
 $c_s$ is the {\em push factor}, $d_t$ is the {\em pull factor}, and $I(s<t)$ the {\em distance deterrence function}. Yet, instead of  
@@
 \section{Case Studies}
+\label{casestudies}
 

### 2023-02-12 latex_files/applied_network_science_article/article_v0.tex
writing the beginning of the abstract, introduction and theory

@@
 
-This contribution adresses a straightforward, yet central issue in public transportation networks: given a network made of many bus lines, how can one estimate the real trips made by the travelers, on the sole basis of  the embarkment (boarding) counts and disembarkment (alighting) counts for each bus stop?
-Although estimating origin-destination flows is a central issue in transportation modeling (see e.g \cite{bell1997transportation} \cite{hazelton2000estimation} \cite{ashok2002estimation} \cite{cui2006bus} and references therein), the specific problem addressed in this contribution seems, to the best of our knowledge, original.
+This contribution adresses a straightforward, yet central question  in public transportation networks: given a network made of many bus lines, how can one estimate the real trips made by the travelers, on the sole basis of  the embarkment (boarding) counts and disembarkment (alighting) counts for each bus stop?
+Although estimating origin-destination flows is a much addressed issue in transportation modeling (see e.g \cite{bell1997transportation} \cite{hazelton2000estimation} \cite{ashok2002estimation} \cite{cui2006bus} and references therein), the specific problem addressed in this contribution seems, to the best of our knowledge, original.
 

### 2023-02-13 latex_files/applied_network_science_article/article_v0.tex
discussed changes of Monday made

@@
 \begin{displaymath}
-\hat{a}_i=\varepsilon a_i \qquad\qquad \hat{b}_i=\frac{1}{\varepsilon} b_i \qquad\qquad\mbox{with}\quad \varepsilon=\sqrt{\frac{b_\bullet}{a_\bullet}}
+\hat{a}_i=(1-\frac{a_\bullet-b_\bullet}{a_\bullet+b_\bullet})\, a_i 
+ \qquad\qquad\qquad
+  \hat{b}_i=(1+\frac{a_\bullet-b_\bullet}{a_\bullet+b_\bullet})\,  b_i 
 \end{displaymath}
-ensuring $\hat{a}_\bullet=\hat{b}_\bullet$, unless $\varepsilon> 2$ or $\varepsilon<\frac{1}{2}$ ***@R, @G : c'est à peu près cela??*** , in which case the corresponding line (which always turned out to be a temporary line with small  counts) was simply disregarded and removed from the network.
+ensuring $\hat{a}_\bullet=\hat{b}_\bullet=2a_\bullet b_\bullet/(a_\bullet+b_\bullet)$.
+However, strongly unbalanced lines such that  $|a_\bullet-b_\bullet|/a_\bullet > 0.3$ or $|a_\bullet-b_\bullet|/b_\bullet > 0.3$  (which always turned out to be temporary lines with small  counts) were simply disregarded and removed from the network.
 
@@
 
-\subsection{The case of a single line}
+\subsection{Markov property for a single line}
 \label{Single line}
@@
 $F(i)=i+1$.
-The initial prior is simply $g_{st}=\frac{I(s<t)}{(l-1)(l-2)}$ (where $I(.)$ denotes the 0/1  indicator function), capturing solely the unidirectional nature of trips.
+The initial prior is simply $g_{st}=c\; I(s<t)$ and captures solely the unidirectional nature of trips,  where $I(.)$ denotes the 0/1  indicator function and $c=\frac{1}{(l-1)(l-2)}$.
 The margins of the empirical distribution $f_{st}$, as well as the total flow, are here known :
 \begin{displaymath}
-\alpha_s=\frac{a_s}{a_\bullet}\qquad\qquad\qquad \beta_t=\frac{b_t}{b_\bullet}\qquad\qquad\qquad n_{\bullet\bullet}=a_{\bullet}=b_\bullet
+\alpha_s=\frac{a_s}{a_\bullet}\qquad\qquad\qquad \beta_t=\frac{b_t}{b_\bullet}\qquad\qquad\qquad n_{\bullet\bullet}=a_{\bullet}=b_\bullet\enspace. 
 \end{displaymath}
-and (\ref{Sol}) yields, after reparameterization, maximum entropy flows of the form
+Following  (\ref{Sol}) maximum entropy flows are of the form
 \begin{equation}
 \label{nosignle}
-n_{st}=I(s<t)\, c_s\, d_t \qquad\qquad\mbox{where}\quad \sum_{s}c_s\, \sum_{t>s}d_t=n_{\bullet\bullet}
+n_{st}= n_{\bullet\bullet}\,  c\, I(s<t)\, \phi_s\,  \psi_t 
 \end{equation}
-where (setting $D_s:=\sum_{t>s}d_t$ and $C_t:=\sum_{s<t}c_s$) the constraints (\ref{Sol_LagMult}) equivalently express as 
+where (setting $\Psi_s:=\sum_{t>s}\psi_t$ and $\Phi_t:=\sum_{s<t}c\phi_s$) the constraints (\ref{Sol_LagMult}) equivalently read
 \begin{equation}
 \label{dis embarking constraints}
-c_s=\frac{a_s}{\sum_{t>s}d_t}=\frac{a_s}{D_s}
+\phi_s=\frac{\alpha_s}{c\, \sum_{t>s}\psi_t}=\frac{a_s}{n_{\bullet\bullet}\,  c\, \Psi_s}
 \qquad\qquad\qquad
-d_t=\frac{b_t}{\sum_{s<t}c_s}=\frac{b_t}{C_t}
+\psi_t=\frac{\beta_t}{c\, \sum_{s<t}\phi_s}=\frac{b_t}{n_{\bullet\bullet}\, c\, \Phi_t}
 \end{equation}
@@
 Interestingly enough, the form (\ref{nosignle}) for the flows is reminiscent of the {\em gravity flows} of quantitative Geography \cite{wilson1967statistical}  \cite{erlander1990gravity} \cite{bavaud2002quasi} \cite{Thomas-Agnan2021}, where
-$c_s$ is the {\em push factor}, $d_t$ is the {\em pull factor}, and $I(s<t)$ the {\em distance deterrence function}. Yet, instead of  
+$\phi_s$ is the {\em push factor}, $\psi_t$ is the {\em pull factor}, and $I(s<t)$ the {\em distance deterrence function}. Yet, instead of  
 being symmetric in $s,t$ and decreasing with the distance $|s-t|$, the distance deterrence function is here asymmetric due to the line orientation, but otherwise constant.
@@
 \begin{displaymath}
-m_{st}=\sum_{u\ge t}n_{su}=c_s \sum_{u\ge t} I(s<u)\, d_u =c_s (d_t+D_t) 
+m_{st}=\sum_{u\ge t}n_{su}=n_{\bullet\bullet}\, c\, \phi_s  \sum_{u\ge t} I(s<u)\, \psi_u =n_{\bullet\bullet}\, c\,  \phi_s (\psi_t+\Psi_t) 
 \end{displaymath}
@@
 \begin{displaymath}
-\rho_{st}=\frac{n_{st}}{m_{st}}=\frac{c_s\,  d_t}{c_s\,  (d_t+D_t)}=\frac{d_t}{d_t+D_t}=\frac{b_t}{C_t\, (d_t+D_t)}\le 1
+\rho_{st}=\frac{n_{st}}{m_{st}}=\frac{n_{\bullet\bullet}\, c\,  \phi_s\,   \psi_t}{n_{\bullet\bullet}\, c\,  \phi_s (\psi_t+\Psi_t)}=\frac{\psi_t}{\psi_t+\Psi_t}\le 1 
 \end{displaymath}
-which depends on $t$ only: it appears that the disembarkment probability $\rho_t=\frac{d_t}{d_t+D_t}$ at $t$ is {\em independent} of the embarkment stop $s$.
+which depends on $t$ only: it appears that the disembarkment probability $\rho_t=\frac{\psi_t}{\psi_t+\Psi_t}$ at $t$ is {\em independent} of the embarkment stop $s$.
 Said otherwise, a traveler embarking at any stop $s$ (and thus necessarily in the bus at $F(s)=s+1$) experiences the {\em same disembarkment probability} at each further stop $t>s$.
@@
 traveled distances among users) or the ``last in, first out" (LIFO) flows (tending to generate maximally  contrasted traveled distances).
-
-\
-
-***  @G, ici davantage de développements sur l'approche Markovienne, ainsi que le joli schéma ?  ***
-
-\
-
-*** @G, @R, ici la (les) figure de l'exemple  ``starting from Maladière, Riant-Cour, Dapples"... ? ***
 

### 2023-02-13 latex_files/applied_network_science_article/article_v0.tex
update algorithm in article

--- 2023-02-12 latex_files/applied_network_science_article/article_v0.tex
@@
 \usepackage{caption}
+
+\usepackage{algpseudocode}
+\usepackage{algorithm}
 
@@
 
+\subsection{Algorithm pseudo-code}
+\label{algo}
+\begin{algorithm}[H]
+	\caption{Compute the transportation flow matrix $\mathbf{N} = (n_{st})$ knowing the edge-trip incidence matrix $\bm{\chi} = (\chi_{ij}^{st})$, the embarking flow $\mathbf{a}$, the disambarking flow $\mathbf{b}$, the index of an isolated source node $\tilde{s}$, and the minimum proportion of passengers entering/leaving the network hyperparameter $\theta$.}
+	\label{algo1}
+	\begin{algorithmic}[1]
+		\State $g_{st} \leftarrow I(\chi_{\bullet \bullet}^{st} > 0) / \sum_{s't'}I(\chi_{\bullet \bullet}^{s't'} > 0), \; \forall s,t$ \Comment{{\footnotesize Initialize the prior distribution.}}
+		\State $\alpha_s \leftarrow a_s / a_\bullet ,\; \beta_t \leftarrow b_t / b_\bullet, \; \forall s,t$ \Comment{{\footnotesize Initialize the margin distributions.}}
+		\State $\epsilon \leftarrow 10^{-40}$ \Comment{{\footnotesize Fix a small quantity}}
+		\While{$\mathbf{N} = (n_{st})$ has not converge} \Comment{{\footnotesize Main loop.}}
+			\State $\psi_t \leftarrow 1, \; \forall t$ \Comment{{\footnotesize Initialize the lagragian multiplier for destinations.}}
+			\While{$\bm{\psi} = (\psi_t)$ has not converge} \Comment{{\footnotesize Iterative fitting loop.}}
+				\State $\phi_s \leftarrow \alpha_s / (\sum_t \psi_t g_{st} + \epsilon), \; \forall s$
+				\State $\psi_t \leftarrow \beta_t / (\sum_s \phi_s g_{st} + \epsilon), \; \forall t$
+			\EndWhile
+
+			\State $n_{st} \leftarrow \frac{a_{\tilde{s}}}{\alpha_{\tilde{s}}} \phi_s \psi_t g_{st}, \; \forall s,t$ \Comment{{\footnotesize Compute the transportation flow.}}
+
+		\EndWhile
+		\State \Return $\mathbf{Z}$
+	\end{algorithmic}
+\end{algorithm}
+
 %% --------------------------------- CASE STUDIES

### 2023-02-13 multilines_data/formatted_data/all_lines_old/remove_temp_lines.R

@@ -8,7 +8,7 @@ bus_df = read.csv("bus_df.csv")
 ped_time_mat = read.csv("ped_time.csv", header=F)
 
 # Line numbers to remove
-lines_to_remove = c(23, 29, 36, 41, 48, 54, 60, 81, 82)
+lines_to_remove = c(23, 29, 36, 41, 48, 60, 69, 81, 82)
 
 # Indices to remove
 to_remove_ids = which(bus_df$line_nbr %in% lines_to_remove)

### 2023-02-13 utilities/data_processing/network_correction.R

@@ -5,11 +5,8 @@
 #-------------------------------------------------------------------------------
 # Delete transports lines with a high difference between in and out passengers
 
-# Set working directory path
-setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
-
 # Read data
-line_df = read.csv("multilines_data/preprocessed_data/all_lines/line_df.csv")
+line_df = read.csv("multilines_data/formatted_data/all_lines_old/bus_df.csv")
 
 # Sum passengers by line A and R
 sum_table = data.frame(line=numeric(), a_in=numeric(), r_in=numeric(),
@@ -28,9 +25,12 @@ for (i in unique(line_df$line_nbr)) {
   
   diff_a = a_in/a_out
   diff_r = r_in/r_out
+  diff_tot_a = round((a_in - a_out) / (a_in + a_out + 1e-40), 4)
+  diff_tot_r = round((r_in - r_out) / (r_in + r_out + 1e-40), 4)
   
   new_row <- data.frame(line=line, a_in=a_in,a_out=a_out, diff_a=diff_a,
-                        r_in=r_in, r_out=r_out, diff_r=diff_r)
+                        r_in=r_in, r_out=r_out, diff_r=diff_r, 
+                        diff_tot_a=diff_tot_a, diff_tot_r=diff_tot_r)
   sum_table <- rbind(sum_table, new_row)
 }
 
@@ -40,4 +40,13 @@ for (i in unique(line_df$line_nbr)) {
 invalid_lines = sum_table[(sum_table$diff_a < 0.3 | sum_table$diff_a > 1.3) | 
                             (sum_table$diff_r < 0.7 | sum_table$diff_r > 1.3)| 
                             is.na(sum_table$diff_r) | is.na(sum_table$diff_a), ]
+
+
+threshold = 0.15
+min_count = 10000
+invalid_lines_2 = sum_table[abs(sum_table$diff_tot_a) > threshold | 
+                              abs(sum_table$diff_tot_r) > threshold | 
+                              (((sum_table$a_in+sum_table$a_out)/2 < min_count) & 
+                               ((sum_table$r_in+sum_table$r_out)/2 < min_count)), ]
+
 

### 2023-02-13 latex_files/applied_network_science_article/article_v0.tex

--- 2023-02-13 latex_files/applied_network_science_article/article_v0.tex
@@
 
+***  @G, ici davantage de développements sur l'approche Markovienne, ainsi que le joli schéma ?  ***
+
+\
+
+*** @G, @R, ici la (les) figure de l'exemple  ``starting from Maladière, Riant-Cour, Dapples"... ? ***
+
+\subsection{Algorithm pseudo-code}
+\label{algo}
+\begin{algorithm}[H]
+	\caption{Compute the transportation flow matrix $\mathbf{N} = (n_{st})$ knowing the edge-trip incidence matrix $\bm{\chi} = (\chi_{ij}^{st})$, the embarking flow $\mathbf{a}$, the disambarking flow $\mathbf{b}$, the index of an isolated source node $\tilde{s}$, and the minimum proportion of passengers entering/leaving the network hyperparameter $\theta$.}
+	\label{algo1}
+	\begin{algorithmic}[1]
+		\State $g_{st} \leftarrow I(\chi_{\bullet \bullet}^{st} > 0) / \sum_{s't'}I(\chi_{\bullet \bullet}^{s't'} > 0), \; \forall s,t$ \Comment{{\footnotesize Initialize the prior distribution.}}
+		\State $\alpha_s \leftarrow a_s / a_\bullet ,\; \beta_t \leftarrow b_t / b_\bullet, \; \forall s,t$ \Comment{{\footnotesize Initialize the margin distributions.}}
+		\State $\epsilon \leftarrow 10^{-40}$ \Comment{{\footnotesize Fix a small quantity}}
+		\While{$\mathbf{N} = (n_{st})$ has not converge} \Comment{{\footnotesize Main loop.}}
+			\State $\psi_t \leftarrow 1, \; \forall t$ \Comment{{\footnotesize Initialize the lagragian multiplier for destinations.}}
+			\While{$\bm{\psi} = (\psi_t)$ has not converge} \Comment{{\footnotesize Iterative fitting loop.}}
+				\State $\phi_s \leftarrow \alpha_s / (\sum_t \psi_t g_{st} + \epsilon), \; \forall s$
+				\State $\psi_t \leftarrow \beta_t / (\sum_s \phi_s g_{st} + \epsilon), \; \forall t$
+			\EndWhile
+
+			\State $n_{st} \leftarrow \frac{a_{\tilde{s}}}{\alpha_{\tilde{s}}} \phi_s \psi_t g_{st}, \; \forall s,t$ \Comment{{\footnotesize Compute the transportation flow.}}
+
+		\EndWhile
+		\State \Return $\mathbf{Z}$
+	\end{algorithmic}
+\end{algorithm}
+
 %% --------------------------------- CASE STUDIES

### 2023-02-13 latex_files/applied_network_science_article/article_v0.tex
put pseudocode for the algorithm

--- 2023-02-13 latex_files/applied_network_science_article/article_v0.tex
@@
 
+\subsubsection{Algorithm pseudocode}
+
+\begin{algorithm}[H]
+	\caption{Compute the transportation flow matrix $\mathbf{N} = (n_{st})$ knowing the edge-trip incidence matrix $\bm{\chi} = (\chi_{ij}^{st})$, the set of transfer edges $\mathcal{T}$, the embarking flow $\mathbf{a}$, the disambarking flow $\mathbf{b}$, the index of an isolated source node $\tilde{s}$, and the minimum proportion of passengers entering/leaving the network $\theta$.}
+	\label{algo1}
+	\begin{algorithmic}[1]
+		\State $g_{st} \leftarrow I(\chi_{\bullet \bullet}^{st} > 0) / \sum_{s't'}I(\chi_{\bullet \bullet}^{s't'} > 0), \; \forall s,t$ \Comment{{\footnotesize Initialize the prior distribution}}
+
+		\State $\alpha_s \leftarrow a_s / a_\bullet, \; \forall s,t$ \Comment{{\footnotesize Initialize the network ingoing distribution}}
+
+		\State $\beta_t \leftarrow b_t / b_\bullet, \; \forall t$ \Comment{{\footnotesize Initialize the network outgoing distribution}}
+
+		\State $\epsilon \leftarrow 10^{-40}$ \Comment{{\footnotesize Fix a small quantity}}
+
+		\While{$\mathbf{N} = (n_{st})$ has not converge} \Comment{{\footnotesize Main loop}}
+
+		\State $\psi_t \leftarrow 1, \; \forall t$
+
+		\While{$\bm{\psi} = (\psi_t)$ has not converge} \Comment{{\footnotesize Iterative fitting loop}}
+
+		\State $\phi_s \leftarrow \alpha_s / (\sum_t \psi_t g_{st} + \epsilon), \; \forall s$
+		\State $\psi_t \leftarrow \beta_t / (\sum_s \phi_s g_{st} + \epsilon), \; \forall t$
+
+		\EndWhile
+
+		\State $n_{st} \leftarrow \frac{a_{\tilde{s}}}{\alpha_{\tilde{s}}} \phi_s \psi_t g_{st}, \; \forall s,t$ \Comment{{\footnotesize Compute the transportation flow}}
+
+		\State $z_{ij} \leftarrow I((i,j) \in \mathcal{T})\sum_{st} \chi_{ij}^{st} n_{st}, \; \forall i,j$
+
+		\State $\alpha_s \leftarrow \frac{\min(\theta a_s, a_s - z_{\bullet s})}{\sum_{s'} \min(\theta a_{s'}, a_{s'} - z_{\bullet {s'}})}, \; \forall s$ \Comment{{\footnotesize Update the network ingoing distribution}}
+
+		\State $\beta_t \leftarrow \frac{\min(\theta b_t, b_t - z_{t \bullet})}{\sum_{t'} \min(\theta b_{t'}, b_{t'} - z_{{t'} \bullet})}, \; \forall t$ \Comment{{\footnotesize Update the network outgoing distribution}}
+
+		\State $r_{ij} \leftarrow \max \left(1, \frac{z_{i \bullet}}{(1 - \theta)b_i}, \frac{z_{\bullet j}}{(1 - \theta)a_j} \right), \; \forall i,j$
+
+		\State $\bar{r}_{st} \leftarrow \max_{ij} \chi_{ij}^{st} r_{ij}, \; \forall s,t$
+
+		\State $\widetilde{g}_{st} \leftarrow \frac{\left( \frac{n_{st}}{\phi_s \psi_t \bar{r}_{st} + \epsilon} \right)}{\sum_{s',t'} \left( \frac{n_{s't'}}{\phi_{s'} \psi_{t'}\bar{r}_{s't'} + \epsilon} \right) + \epsilon}, \; \forall s,t$ \Comment{{\footnotesize Update the prior distribution}}
+
+		\EndWhile
+		\State \Return $\mathbf{N} = (n_{st})$
+	\end{algorithmic}
+\end{algorithm}
+
 \subsection{Markov property for a single line}
@@
 traveled distances among users) or the ``last in, first out" (LIFO) flows (tending to generate maximally  contrasted traveled distances).
+
+***  @G, ici davantage de développements sur l'approche Markovienne, ainsi que le joli schéma ?  ***
+
+\
+
+*** @G, @R, ici la (les) figure de l'exemple  ``starting from Maladière, Riant-Cour, Dapples"... ? ***
 

### 2023-02-13 latex_files/applied_network_science_article/article_v0.tex
remove previous submission in article

@@
 
-		\State $\alpha_s \leftarrow a_s / a_\bullet, \; \forall s,t$ \Comment{{\footnotesize Initialize the network ingoing distribution}}
+		\State $\alpha_s \leftarrow a_s / a_\bullet, \; \forall s$ \Comment{{\footnotesize Initialize the network ingoing distribution}}
 
@@
 \newpage
-
-\section*{Previous submission to complex networks 2022}
-
-%
-\section*{Statement of the problem}
-%
-
-Automatic passenger counters measure the number of passengers entering and leaving buses at each stop \cite{boyle_passenger_1998}.
-Given this information, can we estimate the complete trajectories of passengers within the entire multi-line network?
-This communication attempts to propose an estimation of all passenger trajectories in the multi-line network with an algorithm based on iterative proportional fitting (IPF) \cite{bishop_discrete_2007}.
-
-The exploited dataset is provided by the Lausanne Transportation Agency (tl) in Switzerland.
-The dataset includes 42 lines of buses (or subways) and more than 1361 stops, including 497 clusters of stops (Fig.
-\ref{fig:map-lausanne}),  carrying around 115 million passengers in 2019.
-Each stop refers to a single directed line, and return lines are considered as distinct.
-In addition to  \emph{line edges},  it is possible to construct pedestrian \emph{transfer edges} (Fig.
-\ref{fig:network-lausanne}) to make the graph unilaterally connected by considering, e.g., clusters of stops connected.
-
-\noindent \begin{minipage}[c]{0.5\textwidth}
-   \begin{figure}[H]
-      \centering
-      \includegraphics[width=6cm]{img/stop_area_c.png}
-      \caption{Cluster of stops} 
-      \label{fig:map-lausanne}
-   \end{figure}
-\end{minipage}%
-\begin{minipage}[c]{0.5\textwidth}
-   \begin{figure}[H]
-      \centering
-      \includegraphics[width=4cm]{img/edge_type3.png}
-      \caption{The multi-line network structure} 
-      \label{fig:network-lausanne}
-   \end{figure}
-\end{minipage}
-
-$\mbox{}$
-
-Knowing only the network structure and the number of passengers embarking and disembarking at each stop, how can we infer the most probable passenger trajectories in the network?
-Before examining the general multi-line network problem,  we address the estimation of trajectories on a single line.
-
-%
-\section{Single line}
-%
-
-Consider a one-directional line with $n$ stops indexed regarding the order found in the line.
-Let $\mathbf{a}_\text{in} = (a_{s}^\text{in})$ and $\mathbf{a}_\text{out} = (a_{t}^\text{out})$ be two vectors representing, respectively, the passengers entering and leaving the line at each stop.
-The goal is to estimate the $(n \times n)$ origin-destination matrix $\mathbf{N} = (n_{st})$, where $n_{st}$ represents the number of passengers entering the line at $s$ and leaving at $t$, subject to constraints $n_{s\bullet} = a^\text{in}_s$ and $n_{\bullet t} = a^\text{out}_t$.
-Among many feasible solutions, arguably the most elegant one is the maximum entropy solution, which can be derived  from different means:  (i) passenger flows can be modelled using a Markovian assumption, which translates by assuming that every passenger has the same probability to continue the trip after having travelled at last one stop ; (ii) an iterative proportional fitting algorithm can be performed, starting with an initial origin-destination affinity matrix $\mathbf{S} = (s_{st})$, defined as the upper triangular $n\times n$ matrix filled with $1$, and then iterated to satisfy the margin constraints given by $\mathbf{a}_\text{in}$ and $\mathbf{a}_\text{out}$.
-Both approaches give the same solution, but only the latter remains pertinent in the multi-line problem.
-
-%
-\section{Multi-lines}
-%
-
-In the multi-line problem, a passenger can transfer from a line to another.
-The problem cannot be tackled with Markov chain modelling anymore, which generate unrealistic random trajectories.
-Instead, we will assume that passengers follow shortest paths.
-Starting from an origin-destination matrix $\mathbf{N} = (n_{st})$, where $s$ denotes the stop at which a passenger \emph{enters into the network} (and not simply enters a particular line), and $t$ denotes the stop where the passenger \emph{leaves definitively the network}, this shortest paths assumption allows us to compute the flow matrix on edges $\mathbf{X} = (x_{ij})$.
-The latter decomposes into the within-line flow and the transfer flow, i.e., $\mathbf{X}=\mathbf{X}_\text{W}+\mathbf{X}_\text{B}$.
-Moreover, in the multi-line problem, we also have to distinguish between:
-\begin{itemize}
-	\item[$\bullet$]  passengers who enter and leave bus lines at each stop, represented by vectors $\mathbf{a}_\text{in}$ and $\mathbf{a}_\text{out}$, which are \emph{measured},
-	\item[$\bullet$] and passengers who enter and leave the network at each stop $i$, represented by the \emph{unknown} quantities $n_{\bullet i}$ and $n_{i \bullet}$.
-\end{itemize}
-By construction, 
-\begin{align}
-		a^\text{in}_i &=n_{i\bullet} + x^\text{B}_{i\bullet},  &
-		a^\text{out}_i &= n_{\bullet i} + x^\text{B}_{\bullet i},
-\end{align}
-
-Using these two constraints, along with the shortest paths assumption and iterative proportional fitting, we propose the following iterative algorithm in order to find $\mathbf{N}$ from measured $\mathbf{a}_\text{in}$ and $\mathbf{a}_\text{out}$.
-\\
-
-\noindent \fbox{\begin{minipage}{\textwidth}
-
-\textbf{Initialisation:} $\mathbf{S}^{(0)}$ is filled with 1 excepted for aberrant origin-destination pairs (such as $t$ beeing a previous stop of the same line as $s$).
-The margins of $\mathbf{N}$ are fixed as $\mathbf{n}^{(0)}_{\mbox{\scriptsize in}} = \mathbf{a}_\text{in}$ and $\mathbf{n}^{(0)}_{\mbox{\scriptsize out}} = \mathbf{a}_\text{out}$.
-
-\textbf{Step 1, Iterative proportional fitting:}
-We use IPF to compute $\mathbf{N}^{(r)}$ starting from $\mathbf{S}^{(r)}$, such that margin constraints, defined by $\mathbf{n}^{(r)}_\text{in}$ and $\mathbf{n}^{(r)}_\text{out}$ are satisfied.
-
-\textbf{Step 2, Shortest paths flow:}
-Using shortest paths information, we compute $\mathbf{X}^{(r)}_\text{B}$ from $\mathbf{N}^{(r)}$.
-
-\textbf{Step 3, Affinity and margin update:}
-$\mathbf{S}^{(r+1)}$, $\mathbf{n}^{(r+1)}_\text{in}$ and $\mathbf{n}^{(r+1)}_\text{out}$ are updated in order to respect constraints defined by (1).
-\end{minipage}}
-
-Step 1, 2, and 3 are iterated until convergence, giving an admissible solution to the problem.
-%
-\section{A small example}
-%
-
-As an illustration, an estimated solution proposed by the algorithm on a restricted network made of four lines only is depicted on Fig.
-\ref{fig:example-map}.
-A total of $n_{\bullet \bullet} = 16,837,494$ passengers using this network is estimated by the algorithm.
-The red circle on the bottom left represents the start of the trip $s$ and the size of the circles at stops $t$ represents the estimated number of passengers terminating their trip at $t$.
-In this example, the majority of passengers exit the network on the same initial embarkment line.
-A small fraction of them takes another line.
-
-Table \ref{fig:example-edges} represents the estimated ten most frequented transfer edges.
-The code of the stop represents the number of the line, the direction and a condensed name of its stop cluster.
-The third column gives the number (in thousands) of passengers transferring through this edge.
-
-\noindent \begin{minipage}[c]{0.6\textwidth}
-   \begin{figure}[H]
-      \includegraphics[width=7cm]{img/example_map.png}
-      \caption{Example map} 
-      \label{fig:example-map}
-   \end{figure}
-\end{minipage}%
-\begin{minipage}[c]{0.43\textwidth}
-      %\centering
-      \begin{table}[H]
-		\begin{tabular}{l|l|r}
-		\multicolumn{1}{c|}{\textbf{From stop}} & \multicolumn{1}{c|}{\textbf{To stop}} & \multicolumn{1}{c}{\textbf{Count}} \\ \hline
-		{\scriptsize S7\_A\_SF\_O}			& {\scriptsize S9\_A\_SF\_O}			& 192k		\\ \hline
-		{\scriptsize S9\_R\_CH\_E}			& {\scriptsize S6\_A\_CH\_E}			& 187k		\\ \hline
-		{\scriptsize S7\_A\_SF\_O}		& {\scriptsize S8\_R\_SF\_S}				& 135k			\\ \hline
-		{\scriptsize S6\_R\_CH\_O}			& {\scriptsize S9\_A\_CH\_O}			& 135k			\\ \hline
-		{\scriptsize S8\_A\_GTE\_N}		& {\scriptsize S9\_R\_GTE\_E}			& 103k		\\ \hline
-		{\scriptsize S9\_R\_B-AIR\_C}		& {\scriptsize S8\_A\_B-AIR\_N}		& 99k			\\ \hline
-		{\scriptsize S9\_A\_SF\_O}			& {\scriptsize S7\_A\_SF\_O}			& 88k			\\ \hline
-		{\scriptsize S8\_R\_B-AIR\_D}		& {\scriptsize S6\_A\_B-AIR\_C}		& 87k			\\ \hline
-		{\scriptsize S6\_R\_SF\_O}			& {\scriptsize S8\_A\_SF\_O}			& 86k			\\ \hline
-		{\scriptsize S9\_A\_GTE\_O}		& {\scriptsize S8\_R\_GTE\_S}		& 84k                                        
-		\end{tabular}
-\vspace*{0.2cm}	\caption{List of the ten most\\ frequented transfer edges} 
-	\label{fig:example-edges}
-	\end{table}
-\end{minipage}
-
-\vspace*{0.2cm}
-
-The current work performs computer-intensive simulations of flow over the entire network (1361 stops), permitting to extract usual network indices (centrality, betweenness...) characterizing both the stops \emph{and} the lines.
-In parallel, the computational effects of various fine tuning calibration parameters used in the algorithm are investigated.
 

### 2023-02-15 latex_files/applied_network_science_article/article_v0.tex
specify line $\elle$ in 2.4

@@
 the components of $\mathbf{a}$ and $\mathbf{b}$ can be non-integer).
-It may also happen that raw data do not obey the necessary consistency condition  $a_\bullet=b_\bullet$, in which case we did rescale the embarking and disembarking counts as
+It may also happen that, on some lines $\ell$,  raw data do not obey the necessary consistency condition  $a_\bullet^\ell=b_\bullet^\ell$ (where the latter quantities denote the total embarkments and disembarkments on line $\ell$), in which case we did rescale the embarking and disembarking line counts as
 \begin{displaymath}
-\hat{a}_i=(1-\frac{a_\bullet-b_\bullet}{a_\bullet+b_\bullet})\, a_i 
+\hat{a}_i=(1-\frac{a_\bullet^\ell-b_\bullet^\ell}{a_\bullet^\ell+b_\bullet^\ell})\, a_i 
  \qquad\qquad\qquad
-  \hat{b}_i=(1+\frac{a_\bullet-b_\bullet}{a_\bullet+b_\bullet})\,  b_i 
+  \hat{b}_i=(1+\frac{a_\bullet^\ell-b_\bullet^\ell}{a_\bullet^\ell+b_\bullet^\ell})\,  b_i 
 \end{displaymath}
-ensuring $\hat{a}_\bullet=\hat{b}_\bullet=2a_\bullet b_\bullet/(a_\bullet+b_\bullet)$.
-However, strongly unbalanced lines such that  $|a_\bullet-b_\bullet|/a_\bullet > 0.3$ or $|a_\bullet-b_\bullet|/b_\bullet > 0.3$  (which always turned out to be temporary lines with small  counts) were simply disregarded and removed from the network.
+ensuring $\hat{a}_\bullet^\ell=\hat{b}_\bullet^\ell=2a_\bullet^\ell b_\bullet^\ell/(a_\bullet^\ell+b_\bullet^\ell)$.
+However, strongly unbalanced lines such that  $|a_\bullet^\ell-b_\bullet|^\ell/a_\bullet^\ell > 0.3$ or $|a_\bullet^\ell-b_\bullet^\ell|/b_\bullet^\ell > 0.3$  (which always turned out to be temporary lines with small  counts) were simply disregarded and line $\ell$ removed from the network.
 

### 2023-02-15 latex_files/applied_network_science_article/article_v0.tex
graph for toy examples computed

@@
 
+Case studies are divided in two sections.
+In the first section, we test the algorithm on toy examples, which are artifical networks where the transportation flow $n_{st}$ is randomly drawn.
+These examples enable some kind of validation of the algorithm, as "real" transportation flows are known and can be compared to the solutions given by our method.
+This setup differs from the second section, which is dedicated to applying the algorithm to the real case of the public transportation network of the city of Lausanne (tl), where embarkement and disembarkement flows are measured but transportation flows are unknown.
+This second case study shows that the algorithm is applicable on large, real datasets and can gives insights to transportation network planners.
+
 \subsection{Toy Examples}
+
+\subsubsection{Construction}
+
+All constructed toy examples are built following the same approach, which aims at being simple but somewhat realistic.
+We fix a number of \emph{line tours} $p \geq 2$, each of which is constituted of a forward line and a backward line, for a total of $q = 2p$ lines.
+Every line has a starting and ending node, which are isolated node, and posses $p-1$ intermediary nodes which allows transfers to the other tour lines, giving a total of $n = 2p(p + 1)$ nodes in the network.
+Examples of these toy networks can be found in Figure \ref{toy_example_plots}.
+
+\begin{figure}
+	\label{toy_example_plots}
+	\includegraphics[width=0.3\textwidth]{fig/toy_2_display.pdf}
+	\includegraphics[width=0.3\textwidth]{fig/toy_3_display.pdf}
+	\includegraphics[width=0.3\textwidth]{fig/toy_4_display.pdf}
+	\caption{3 toy examples, with the number of tours $p \in \{2,3,4\}$.
+Tours are in the same color, line have a unique letter, and each stop a unique combination of a letter and a number.
+Clusters of nodes represent positions where transfer between tours are possible.}
+\end{figure}
+
+\begin{figure}
+	\label{iteration_plots}
+	\includegraphics[width=0.98\textwidth]{fig/iterations.pdf}
+	\caption{Blabla}
+\end{figure}
 

### 2023-02-15 4.1_plot_toys.R

@@ -18,38 +18,79 @@ source("local_functions.R")
 #--------------------------------
 
 # Result folder
-result_folder = "results/iteration_plots"
+result_folder = "results/toy_example_plots"
 
 # Choose number of lines
 n_lines_vec = c(2, 3, 4)
+seed_vec = c(0, 2, 11)
+theta_vec = c(0, -0.08,-0.051)
+
+# Choose a seed
 
 #--------------------------------
 # Code
 #--------------------------------
 
 # --- loop
-
-n_lines = n_lines_vec[1]
-n_stops = n_lines + 1
-
-# --- Network creation
-
-# Create the network
-network_data = network_prop(n_lines, n_stops)
-adj = network_data$adj
-
-# Stop names 
-if(nb_lines == 2){
-  stop_names = c("A1", "A2", "A3", 
-                 "B1", "B2", "B3", 
-                 "C1", "C2", "C3", 
-                 "D1", "D2", "D3")
-} else {
-  stop_names = NULL
+for(i in 1:length(n_lines_vec)){
+  
+  n_lines = n_lines_vec[i]
+  n_stops = n_lines + 1
+  set.seed(seed_vec[i])
+  theta = theta_vec[i]
+  
+  # --- Network creation
+  
+  # Create the network
+  network_data = network_prop(n_lines, n_stops)
+  adj = network_data$adj
+  
+  # Stop names 
+  letters = c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J", "K", "L")
+  line_letters = letters[1:(2*n_lines)]
+  stop_numbers = 1:n_stops
+  stop_names = paste0(rep(line_letters, each=n_stops), 
+                      rep(stop_numbers, 2*n_lines))
+  colnames(adj) = stop_names
+  rownames(adj) = stop_names
+  
+  # Make the graph
+  g = graph_from_adjacency_matrix(adj)
+  
+  # Make the colors
+  new_palette = palette(rainbow(n_lines))
+  stop_colors = rep(new_palette, each=2*n_stops)
+  
+  # Make the colors
+  new_palette = palette(rainbow(n_lines))
+  stop_colors = rep(new_palette, each=2*n_stops)
+  
+  # Layout
+  if(n_lines == 2){
+    layout = matrix(c(1, 2,
+                      2, 2,
+                      2, 1,
+                      3, 1,
+                      3, 2,
+                      4, 2,
[…]

### 2023-02-15 4.2_plot_toy_iterations.R

@@ -18,10 +18,10 @@ source("local_functions.R")
 #--------------------------------
 
 # Result folder
-result_folder = "results/iteration_plots"
+result_folder = "results/toy_example_plots"
 
 # Choose number of lines
-nb_lines = 4
+nb_lines = 2
 # Choose number of stops
 nb_stops = nb_lines + 1
 # Choose number of passengers in the network (with normal distribution)
@@ -123,15 +123,16 @@ if(nb_lines == 2){
 }
 
 
-e_line_size = 0.2
+e_line_size = 0.02
 e_flow_size = 3
-v_label_size = 1
+v_size = 20
+v_label_size = 0.00001
 
-pdf(paste0(result_folder, "/iterations_2.pdf"))
+pdf(paste0(result_folder, "/iterations.pdf"))
 par(mfrow=c(2,3))
 plot_flow_graph(adj, n_real, layout, v_names=stop_names, main="Real", 
                 e_line_size=e_line_size, e_flow_size=e_flow_size, 
-                v_label_size=v_label_size)
+                v_label_size=v_label_size, v_size=v_size)
 
 for(it in it_list){
   n_mat = n_mat_list[[it]]
@@ -147,7 +148,7 @@ for(it in it_list){
                               format(error, digits=3),"\nMargins error=", 
                               round(error_m, digits=3)), 
                   e_line_size=e_line_size, e_flow_size=e_flow_size, 
-                  v_label_size=v_label_size)
+                  v_label_size=v_label_size, v_size=v_size)
 }
 dev.off()
 

### 2023-02-15 local_functions.R

@@ -705,7 +705,7 @@ compute_x_from_n = function(n_mat, edge_ref, sp_ref, sp_edge_link){
 
 plot_flow_graph = function(adj, n_mat, layout=NULL, v_names=NULL, 
                            main=NULL, e_line_size=0.2, e_flow_size=2, 
-                           v_label_size=2){
+                           v_label_size=2, v_size=20){
   
   # Number of edges
   n = dim(adj)[1]
@@ -775,10 +775,10 @@ plot_flow_graph = function(adj, n_mat, layout=NULL, v_names=NULL,
   # Plot graph
   par(mar=c(0,0,1.2,0)+.1)
   plot(g_tot, layout=layout, edge.color=edge_colors, edge.width=edge_sizes,
-       edge.curved=edge_curves, edge.arrow.size=edge_sizes*3, 
-       edge.arrow.width=edge_sizes*3, vertex.shape="pie", vertex.pie=pie_prop, 
+       edge.curved=edge_curves, edge.arrow.size=0.5, 
+       edge.arrow.width=1.5, vertex.shape="pie", vertex.pie=pie_prop, 
        vertex.pie.color=in_out_colors, main=main, vertex.label.cex=v_label_size, 
-       vertex.label.color="black")
+       vertex.label.color="black", vertex.size=v_size)
   if(!is.null(main)){
     title(main)
   }

### 2023-02-16 5.3_visualisation.R
Passenger toy example

@@ -12,7 +12,7 @@
 #--------------------------------
 
 # Set working directory path
-setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+# setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
 # Libraries
 #--------------------------------
@@ -34,8 +34,7 @@ out_folder = "results/loop_toy_example"
 # Network parameters
 #--------------------------------
 # Choose number of lines
-nb_lines = 3
-nb_lines = 2:3
+nb_lines = 2:8
 # Choose number of stops
 nb_stops = nb_lines + 1
 
@@ -54,11 +53,9 @@ max_it = 1000
 # print iterations
 display_it = F
 # number of iterations
-n_test = 3
+n_test = 10
 # prop limit
 hyper_par = 0.1
-# number of passengers
-n_passengers = 1000
 
 #--------------------------------
 # Output and visualizations
@@ -66,110 +63,63 @@ n_passengers = 1000
 
 # Network properties output
 #--------------------------
-network_prop = network_prop(nb_lines, nb_stops)
+# network_prop_list = network_prop(nb_lines, nb_stops)
 # Unlist and save variables
-list2env(network_prop, .GlobalEnv)
+# list2env(network_prop_list, .GlobalEnv)
 
 # Create a data frame with all different number of passengers into the network
-res_mat_passengers = c()
+
+res_mean = c()
 seq_passengers = seq(from = 200, to = 4000, by = 200)
 
-for (i in seq_passengers) {
-  res_mat = compute_toy(hyper_par, n_test, paths, i, edge_ref, sp_ref,
-                        sp_edge_link, conv_thres_algo, conv_thres_if, max_it,
-                        display_it)
-  colnames(res_mat)[1] <- paste("Nb:", i)
-  res_mat_passengers = cbind(res_mat_passengers, res_mat)
-}
-
-
-# Add the mean and the standard deviation
-mean_pass = as.data.frame(colMeans(res_mat_passengers))
-sd_pass = apply(res_mat_passengers, 2, sd)
-sd_pass = sd_pass/sqrt(n_test)
-mean_pass = cbind(seq_passengers, mean_pass, sd_pass)
-colnames(mean_pass) <- c("Passengers","mean_error", "sd_error")
-
-
-
-
-
-
-
-
-
-
-
-
-
-
[…]

### 2023-02-16 5.4_visualisation.R

@@ -12,7 +12,7 @@
 #--------------------------------
 
 # Set working directory path
-setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+# setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
 # Libraries
 #--------------------------------
@@ -107,7 +107,7 @@ for (j in 1:length(hyper_par)) {
 # write.csv(mean_line, "mean_line_param_small.csv", row.names = F)
 
 # Read saved data
-mean_line = read.csv("mean_line_param.csv")
+mean_line = read.csv("results/5_toy_ex_outputs/mean_line_param.csv")
 # Delete parm. 0
 mean_line = mean_line[-(1:7),]
 

### 2023-02-16 latex_files/applied_network_science_article/article_v0.tex
time conflict

@@
 This setup differs from the second section, which is dedicated to applying the algorithm to the real case of the public transportation network of the city of Lausanne (tl), where embarkement and disembarkement flows are measured but transportation flows are unknown.
-This second case study shows that the algorithm is applicable on large, real datasets and can gives insights to transportation network planners.
+This second case study shows that the algorithm is applicable on large, real datasets and can gives insights about passengers probable routes in the network.
 
@@
 
-\subsubsection{Construction}
+\subsubsection{Examples construction}
+\label{example_construction}
 
@@
 \begin{figure}
-	\label{toy_example_plots}
 	\includegraphics[width=0.3\textwidth]{fig/toy_2_display.pdf}
@@
 	\caption{3 toy examples, with the number of tours $p \in \{2,3,4\}$.
-Tours are in the same color, line have a unique letter, and each stop a unique combination of a letter and a number.
+Tours are displayed in the same color, line have a unique letter, and each stop a unique combination of a letter and a number.
 Clusters of nodes represent positions where transfer between tours are possible.}
+	\label{toy_example_plots}
 \end{figure}
+
+Allowed origin-destination pairs $(s,t)$ are constructed with the same set of rules found in (REF), but without considering the exclusion criterion based on the pedestrian distance, as nodes are not spatially located.
+The graph shortest-path structure for the allowed origin-destination pairs is stored in the edge-trip incidence matrix $\bm{\chi}$.
+
+A transportation flow $\mathbf{N}_\text{ref} = (n^\text{ref}_{st})$ is drawn by setting a fixed number of passengers $n^\text{ref}_{\bullet \bullet}$, and each passenger is assigned randomly to a $(s, t)$ pair drawn uniformly among all allowed origin-destination pairs.
+From this reference transportation flow $\mathbf{N}_\text{ref}$, using the edge-trip incidence matrix $\bm{\chi}$ and equation (\ref{equationGG}), we can compute flow on edges $\mathbf{X}_\text{ref}$ and, in turn, the number of passengers embarking $\mathbf{a}_\text{ref}$ and the number of passengers disembarking $\mathbf{b}_\text{ref}$ at each stop.
+
+\subsubsection{Error measurements}
+\label{error_measures}
+Using the network structure, $\mathbf{a}_\text{ref}$, $\mathbf{b}_\text{ref}$, we can obtain a estimation of the real flow through the algorithm, noted $\mathbf{N} = (n_{st})$.
+There are two types of dissimilarity measure between the real data and the solution proposed by the algorithm: (1) how much $\mathbf{N}$ differs from $\mathbf{N}_\text{ref}$, and (2) how well constraints defined by the number of passengers embarking $\mathbf{a}_\text{ref}$ and disembarking $\mathbf{b}_\text{ref}$ are respected.
+The first dissimilarity is measured through the \emph{mean transportation error}, denoted by $\text{MTE}(\mathbf{N})$, and computed with
+\begin{equation}
+	\text{MTE}(\mathbf{N}) = \sum_{st} \frac{n^\text{ref}_{st}}{n^\text{ref}_{\bullet \bullet}} \frac{\lvert n_{st} - n^\text{ref}_{st}\lvert}{n^\text{ref}_{st}} = \frac{\sum_{st} \lvert n_{st} - n^\text{ref}_{st}\lvert}{n^\text{ref}_{\bullet \bullet}}
+	\label{MTE}
+\end{equation} 
+and the second one with the \emph{mean margin error}, noted $\text{MME}(\mathbf{N})$, obtained with
+\begin{align}
+	\text{MME}(\mathbf{N}) &= \frac{1}{2} \sum_{i} \frac{a^\text{ref}_i}{a^\text{ref}_\bullet} \frac{\lvert z_{\bullet i} + n_{i \bullet} - a^\text{ref}_i \lvert}{a^\text{ref}_i} + \frac{1}{2} \sum_{i} \frac{b^\text{ref}_i}{b^\text{ref}_\bullet} \frac{\lvert z_{i \bullet} + n_{\bullet i} - b^\text{ref}_i \lvert}{a^\text{ref}_i} \notag \\
+	&= \frac{\sum_i (\lvert z_{\bullet i} + n_{i \bullet} - a^\text{ref}_i \lvert + \lvert z_{i \bullet} + n_{\bullet i} - b^\text{ref}_i \lvert)}{2n^\text{ref}_{\bullet \bullet}}
+	\label{MME}
+\end{align} 
+where $z_{ij} = I((i,j) \in \mathcal{T})\sum_{st} \chi_{ij}^{st} n_{st}$ is the flow on transfer edges $\mathcal{T}$.
+
+Note that, by construction, the mean margin error should be null when the algorithm converge.
+However, it can be informative to track down this error along iterations and, in some practical cases where the network is large, the algorithm convergence criterion is reached without having margin constraints perfectly respected.
+
+\subsubsection{Algorithm iterations}
+\label{example_construction}
 

### 2023-02-16 4.1_plot_toys.R

@@ -1,0 +1,96 @@
+#-------------------------------------------------------------------------------
+#
+# Preprocess data and save it
+#
+#-------------------------------------------------------------------------------
+
+#--------------------------------
+# Head
+#--------------------------------
+
+#------ Set working directory path
+# setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+source("local_functions.R")
+
+#--------------------------------
+# Parameters
+#--------------------------------
+
+# Result folder
+result_folder = "results/toy_example_plots"
+
+# Choose number of lines
+n_lines_vec = c(2, 3, 4)
+seed_vec = c(0, 2, 11)
+theta_vec = c(0, -0.08,-0.051)
+
+# Choose a seed
+
+#--------------------------------
+# Code
+#--------------------------------
+
+# --- loop
+for(i in 1:length(n_lines_vec)){
+  
+  n_lines = n_lines_vec[i]
+  n_stops = n_lines + 1
+  set.seed(seed_vec[i])
+  theta = theta_vec[i]
+  
+  # --- Network creation
+  
+  # Create the network
+  network_data = network_prop(n_lines, n_stops)
+  adj = network_data$adj
+  
+  # Stop names 
+  letters = c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J", "K", "L")
+  line_letters = letters[1:(2*n_lines)]
+  stop_numbers = 1:n_stops
+  stop_names = paste0(rep(line_letters, each=n_stops), 
+                      rep(stop_numbers, 2*n_lines))
+  colnames(adj) = stop_names
+  rownames(adj) = stop_names
+  
+  # Make the graph
+  g = graph_from_adjacency_matrix(adj)
+  
+  # Make the colors
+  new_palette = palette(rainbow(n_lines))
+  stop_colors = rep(new_palette, each=2*n_stops)
+  
+  # Make the colors
+  new_palette = palette(rainbow(n_lines))
+  stop_colors = rep(new_palette, each=2*n_stops)
+  
+  # Layout
+  if(n_lines == 2){
+    layout = matrix(c(1, 2,
+                      2, 2,
+                      2, 1,
+                      3, 1,
+                      3, 2,
+                      4, 2,
+                      4, 3,
+                      3, 3,
+                      3, 4,
+                      2, 4,
+                      2, 3,
[…]

### 2023-02-16 4.2_plot_toy_iterations.R

@@ -1,0 +1,155 @@
+#-------------------------------------------------------------------------------
+#
+# Preprocess data and save it
+#
+#-------------------------------------------------------------------------------
+
+#--------------------------------
+# Head
+#--------------------------------
+
+#------ Set working directory path
+# setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+source("local_functions.R")
+
+#--------------------------------
+# Parameters
+#--------------------------------
+
+# Result folder
+result_folder = "results/toy_example_plots"
+
+# Choose number of lines
+nb_lines = 2
+# Choose number of stops
+nb_stops = nb_lines + 1
+# Choose number of passengers in the network (with normal distribution)
+nb_passengers = 100
+# Total number of stops *2 (back and forth)
+nb_stops_tot = nb_lines*nb_stops*2
+
+# Conv threshold for iterative fitting
+conv_thres_if = 1e-6
+# Conv threshold
+conv_thres_algo = 1e-8
+# proportional limit 
+min_p_ntwk = 0.1
+# epsilon
+epsilon = 1e-40
+# max iterations
+max_it = 16
+# max iterations for iterative fitting
+max_it_if = 1000
+
+# Seed
+set.seed(10)
+
+# It list
+it_list = c(2, 4, 7, 11, 16)
+
+#--------------------------------
+# Code
+#--------------------------------
+
+# --- Network creation
+
+# Stop names 
+if(nb_lines == 2){
+  stop_names = c("A1", "A2", "A3", 
+                 "B1", "B2", "B3", 
+                 "C1", "C2", "C3", 
+                 "D1", "D2", "D3")
+} else {
+  stop_names = NULL
+}
+
+# Create the network
+network_data = network_prop(nb_lines, nb_stops)
+edge_ref = network_data$edge_ref
+sp_ref = network_data$sp_ref
+sp_edge_link = network_data$sp_edge_link
+paths = network_data$paths
+adj = network_data$adj
+
+# --- Flow creation
+
+n_real = n_multin(paths, nb_passengers)
+x_btw = compute_x_from_n(n_real, edge_ref, sp_ref, sp_edge_link)$x_btw
+flow_l_in = rowSums(n_real) + colSums(x_btw)
+flow_l_out = colSums(n_real) + rowSums(x_btw)
+
+# --- Run the algo 
+
+n_mat_list = compute_origin_destination(flow_l_in,
+                                        flow_l_out,
+                                        edge_ref,
[…]

### 2023-02-16 4_plot_iterations_toy.R

@@ -1,150 +1,0 @@
-#-------------------------------------------------------------------------------
-#
-# Preprocess data and save it
-#
-#-------------------------------------------------------------------------------
-
-#--------------------------------
-# Head
-#--------------------------------
-
-#------ Set working directory path
-# setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
-
-source("local_functions.R")
-
-#--------------------------------
-# Parameters
-#--------------------------------
-
-# Result folder
-result_folder = "results/iteration_plots"
-
-# Choose number of lines
-nb_lines = 2
-# Choose number of stops
-nb_stops = nb_lines + 1
-# Choose number of passengers in the network (with normal distribution)
-nb_passengers = 100
-# Total number of stops *2 (back and forth)
-nb_stops_tot = nb_lines*nb_stops*2
-
-# Conv threshold for iterative fitting
-conv_thres_if = 1e-6
-# Conv threshold
-conv_thres_algo = 1e-8
-# proportional limit 
-min_p_ntwk = 0.1
-# epsilon
-epsilon = 1e-40
-# max iterations
-max_it = 16
-# max iterations for iterative fitting
-max_it_if = 1000
-
-# Seed
-set.seed(10)
-
-# It list
-it_list = c(2, 4, 7, 11, 16)
-
-#--------------------------------
-# Code
-#--------------------------------
-
-# --- Network creation
-
-# Stop names 
-stop_names = c("A1", "A2", "A3", 
-               "B1", "B2", "B3", 
-               "C1", "C2", "C3", 
-               "D1", "D2", "D3")
-
-# Create the network
-network_prop = network_prop(nb_lines, nb_stops)
-edge_ref = network_prop$edge_ref
-sp_ref = network_prop$sp_ref
-sp_edge_link = network_prop$sp_edge_link
-paths = network_prop$paths
-adj = network_prop$adj
-colnames(adj) = stops
-rownames(adj) = stops
-
-# --- Flow creation
-
-n_real = n_multin(paths, nb_passengers)
-n_real = n_real / sum(n_real)
-x_btw = compute_x_from_n(n_real, edge_ref, sp_ref, sp_edge_link)$x_btw
-flow_l_in = rowSums(n_real) + colSums(x_btw)
-flow_l_out = colSums(n_real) + rowSums(x_btw)
-
-# --- Run the algo 
-
-n_mat_list = compute_origin_destination(flow_l_in,
-                                        flow_l_out,
-                                        edge_ref,
[…]

### 2023-02-16 5.4_visualisation.R

@@ -12,7 +12,7 @@
 #--------------------------------
 
 # Set working directory path
-setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+# setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
 # Libraries
 #--------------------------------
@@ -54,10 +54,14 @@ display_it = F
 # number of iterations
 n_test = 50
 # prop limit
-# hyper_par = c(0.1, 0.3, 0.5, 0.7)
-hyper_par = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1)
+# hyper_par = c(0, 0.01, 0.02, 0.03, 0.04, 0.05, 0.06, 0.07, 0.08, 0.09, 0.1,
+#               0.12, 0.14, 0.16, 0.18, 0.2,
+#               0.25, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1)
+hyper_par = seq(0.001,0.009, 0.001)
+
 # number of passengers
-n_passengers = 2000
+# n_passengers = 2000
+nb_pass_stop = 5
 # number of cores
 mc.cores = detectCores() - 2
 # lines
@@ -75,9 +79,12 @@ for (j in 1:length(hyper_par)) {
     nb_lines = i
     nb_stops = i + 1
     # Network properties output
-    network_prop_res = network_prop(nb_lines, nb_stops, mc.cores)
+    network_prop_res = network_prop(nb_lines, nb_stops)
     # Unlist and save variables
     list2env(network_prop_res, .GlobalEnv)
+    
+    # Passengers depend on the number of stops
+    n_passengers = sum(paths)*nb_pass_stop
     
     res_mat = compute_toy(hyper_par[j], n_test, paths, n_passengers, edge_ref,
                           sp_ref, sp_edge_link, conv_thres_algo, conv_thres_if, 
@@ -90,14 +97,23 @@ for (j in 1:length(hyper_par)) {
   # Add the mean and the standard deviation
   mean_line = as.data.frame(colMeans(res_mat_line))
   sd_line = apply(res_mat_line, 2, sd)
-  # sd_line = sd_line/sqrt(n_test)
+  sd_line = sd_line/sqrt(n_test)
   mean_line = cbind(seq_lines, mean_line, sd_line)
   mean_line$param = row.names(mean_line)
   colnames(mean_line) <- c("Lines", "mean_error", "sd_error","param")
[…]

### 2023-02-16 local_functions.R

@@ -704,8 +704,8 @@ compute_x_from_n = function(n_mat, edge_ref, sp_ref, sp_edge_link){
 #-------------------------------------------------------------------------------
 
 plot_flow_graph = function(adj, n_mat, layout=NULL, v_names=NULL, 
-                           main=NULL, e_line_size=1.5, e_flow_size=2, 
-                           v_label_size=2){
+                           main=NULL, e_line_size=0.2, e_flow_size=2, 
+                           v_label_size=2, v_size=20){
   
   # Number of edges
   n = dim(adj)[1]
@@ -775,10 +775,10 @@ plot_flow_graph = function(adj, n_mat, layout=NULL, v_names=NULL,
   # Plot graph
   par(mar=c(0,0,1.2,0)+.1)
   plot(g_tot, layout=layout, edge.color=edge_colors, edge.width=edge_sizes,
-       edge.curved=edge_curves, edge.arrow.size=edge_sizes*3, 
-       edge.arrow.width=edge_sizes*3, vertex.shape="pie", vertex.pie=pie_prop, 
+       edge.curved=edge_curves, edge.arrow.size=0.5, 
+       edge.arrow.width=1.5, vertex.shape="pie", vertex.pie=pie_prop, 
        vertex.pie.color=in_out_colors, main=main, vertex.label.cex=v_label_size, 
-       vertex.label.color="black")
+       vertex.label.color="black", vertex.size=v_size)
   if(!is.null(main)){
     title(main)
   }
@@ -864,6 +864,7 @@ adj_function = function(stops, name_stops){
 # Crossing on different stops
 adj_function2 = function(stops, name_stops){
   nb_stops_tot = length(stops)
+  nb_lines = length(unique(name_stops[, 1]))
   nb_stops = dim(name_stops[name_stops[, 1] == name_stops[1, 1], ])[1] / 2
   adj = matrix(0, nrow = nb_stops_tot, ncol = nb_stops_tot)
   colnames(adj) = rownames(adj) = stops
@@ -1115,7 +1116,6 @@ compute_toy = function(hyper_par, n_test, paths, n_passengers, edge_ref, sp_ref,
   for (i in 1:n_test) {
     set.seed(i)
     paths_passengers = n_multin(paths, n_passengers, epsilon=epsilon)
-    paths_passengers = paths_passengers / sum(paths_passengers)
     x_btw = compute_x_from_n(paths_passengers, edge_ref, sp_ref, 
                              sp_edge_link)$x_btw
[…]

### 2023-02-16 latex_files/applied_network_science_article/article_v0.tex
short dataset description

@@
 \subsection{Real Data}
+\subsubsection{Data set construction}
+The original dataset used in this contribution consists of over 13 million rows of data over more than 1200 stop on 35 bus lines.
+Each data row represents a stop by a bus to embark and disembark passengers.
+The dataset used for further analysis is summed by stop over the year 2019.
 

### 2023-02-16 latex_files/applied_network_science_article/article_v0.tex
typo

--- 2023-02-16 latex_files/applied_network_science_article/article_v0.tex
@@
 \subsection{Real Data}
+\subsubsection{Dataset construction}
+The original dataset used in this contribution consists of over 13 million rows of data over more than 1200 stop on 35 bus lines.
+Each data row represents a stop by a bus to embark and disembark passengers.
+The dataset used for further analysis is summed by stop over the year 2019.
 

### 2023-02-21 latex_files/applied_network_science_article/article_v0.tex
toy results

@@
 We fix a number of \emph{line tours} $p \geq 2$, each of which is constituted of a forward line and a backward line, for a total of $q = 2p$ lines.
-Every line has a starting and ending node, which are isolated node, and posses $p-1$ intermediary nodes which allows transfers to the other tour lines, giving a total of $n = 2p(p + 1)$ nodes in the network.
+Every line has a starting and ending node, which are isolated nodes, and posses $p-1$ intermediary nodes which allows transfers to the other tour lines, giving a total of $n = 2p(p + 1)$ nodes in the network.
 Examples of these toy networks can be found in Figure \ref{toy_example_plots}.
@@
 \end{figure}
+
+\begin{figure}
+	\includegraphics[width=0.47\textwidth]{fig/toy_MTE_passengers_theta01.pdf}
+	\includegraphics[width=0.47\textwidth]{fig/toy_MTE_tours.pdf}
+	\caption{\emph{Mean transportation error} (MTE), according to the number of tours $p \in \{2,3,4,5,6,7,8\}$ and resp.
+to the number of passengers into the network per line tour and the minimum proportion $\theta =0.1$ of passengers entering/leaving the network at each stop with a proportional number of passengers to the shortest paths of each line tour.}
+	\label{toy_MTE}
+\end{figure}
+
+By running the algorithm for line tours $p \in \{2,3,4,5,6,7,8\}$, we observe that MTE depends on different parameters.
+On the graph on the left in Figure \ref{toy_MTE}, the number of passengers inserted in the network has an influence on MTE.
+The higher the number of passengers, the smaller the error is because passengers are better distributed throughout the larger network.
+On the right in Figure \ref{toy_MTE}, the aim is to understand how the proportion $\theta\in (0,1)$ responds to different line tours.
+Dots represent local minimum MTE for each tour.
 

### 2023-02-21 5.3_visualisation.R

@@ -71,6 +71,8 @@ hyper_par = 0.1
 
 res_mean = c()
 seq_passengers = seq(from = 200, to = 4000, by = 200)
+seq_passengers = seq(from = 5000, to = 20000, by = 1000)
+
 
 for (j in 1:length(nb_lines)) {
   
@@ -102,6 +104,9 @@ for (j in 1:length(nb_lines)) {
 }
 
 # write.csv(res_mean, "results/5_toy_ex_outputs/res_mean_passengers.csv", row.names = F)
+res_mean0 = read.csv("results/5_toy_ex_outputs/res_mean_passengers_theta01.csv")
+res_mean2 = rbind(res_mean0,res_mean)
+
 
 ### Best graph according to the number of passengers into the network
 # only 1/3 error bar
@@ -115,10 +120,10 @@ for (j in 1:length(nb_lines)) {
 #   labs(title = paste0(n_test, " iterations, ", expression(theta), ": ", hyper_par), x = "Passengers into the network", y = "Mean error")
 
 # Plot the results 2
-ggplot(data=res_mean) +
+ggplot(data=res_mean2) +
   geom_ribbon(aes(x=passengers, ymin=mean_error-sd_error/sqrt(n_test), 
                   ymax=mean_error+sd_error/sqrt(n_test), group=line, 
                   fill=factor(line)), alpha=0.2) +
   geom_line(aes(x=passengers, y=mean_error, group=line, color=factor(line))) +
-  labs(x="Nb passengers", y="MTE") +
-  labs(color = "Nb tours", fill = "Nb tours")
+  labs(x="Number of passengers", y="MTE") +
+  labs(color = "p", fill = "Nb tours")

### 2023-02-21 latex_files/applied_network_science_article/article_v0.tex
write article

--- 2023-02-16 latex_files/applied_network_science_article/article_v0.tex
@@
 \label{Lines and junctions}
-Consider a transportation network made of bus lines numbered $\ell=1,\ldots, q$, of respective lengths (number of stops) $l_\ell$.
+Consider a \emph{transportation network} made of \emph{lines} numbered $\ell=1,\ldots, q$, of respective lengths (number of stops) $l_\ell$.
 Opposite lines, that is parallel lines running in the back and forth directions are considered as distinct.
 
-The $l=\sum_{\ell=1}^ql_\ell$ bus stops constitute the nodes of the transportation network.
-Each stop $i=1,\ldots,l$ belongs to {\em  a single bus line}, and defines a unique next or forward stop $F(i)$ (unless $i$ is the line terminus) and a unique backward stop $B(i)$ (unless $i$ is the line start), both on the same line.
+The $l=\sum_{\ell=1}^ql_\ell$ \emph{stops} constitute the nodes of the transportation network.
+Each stop $i=1,\ldots,l$ belongs to a single line, and defines a unique next or forward stop $F(i)$ (unless $i$ is the line terminus) and a unique backward stop $B(i)$ (unless $i$ is the line start), both on the same line.
 
-Let $S_i$ denote the set of stops which can be reached from stop $i$ within walking distance, excluding $i$ itself.
-A stop $i$ is referred to as an {\em isolated stop} if $S_i=\emptyset$, and to as a {\em junction} otherwise.
+Let $S_i$ denote the \emph{set of stops which can be reached from stop $i$ outside lines connection} (with, e.g., an acceptable walking distance), excluding $i$ itself.
+A stop $i$ is referred to as an \emph{isolated stop} if $S_i=\emptyset$, and to as a \emph{junction} otherwise.
 
-\subsection{Line edges, transfer edges and trips}
+\subsection{Edges, trips, and the incidence matrix}
 \label{Line edges, transfer edges and trips}
+
 Two sorts of oriented edges are involved in the transportation network: 
 \begin{enumerate}
-  \item[$\bullet$] intra-line edges $(i,j)=(i,F(i))$ belonging to a single line  $\ell(i)=\ell(j)$
-  \item[$\bullet$] inter-line or transfer edges $(i,j)$ connecting different lines $\ell(i)\neq \ell(j)$, involving walks from junction $i$ to $j\in S_i$.
+  \item[$\bullet$] \emph{intra-line edges} $(i,j)=(i,F(i))$ belonging to a single line  $\ell(i)=\ell(j)$
+  \item[$\bullet$] \emph{inter-line} or \emph{transfer edges} $(i,j)$ connecting different lines $\ell(i)\neq \ell(j)$, involving walks from junction $i$ to $j\in S_i$.
+The \emph{set of transfer edges} is denoted by $T$.
 \end{enumerate}
-A $st$-trip, noted $[s,t]$, consists of entering into the network at stop $s$, and leaving the network at $t$, by following the shortest-path (i.e.
+
+A \emph{$st$-trip}, noted $[s,t]$, consists of entering into the network at stop $s$, and leaving the network at $t$, by following the shortest-path (i.e.
 achieving the minimum distance,  minimum time, or  minimum cost), supposed unique, leading to $s$ from $t$.
 
-The succession of edges $(ij)$ belonging to the $st$-trip, noted $(ij)\in [s,t]$, is unique. Define the edge-trip incidence matrix as
+The succession of edges $(i, j)$ belonging to the $st$-trip, noted $(i,j)\in [s,t]$, is unique. Define the \emph{edge-trip incidence matrix} as
 \begin{equation}
@@
 \end{equation}
-A $st$-trip always starts with the edge $(s,F(s))$, and finishes with $(B(t),t)$. Transfers can occur in-between, but never at the beginning nor at the end of the trip.
+
+Note that we can also forbid some aberrant trips across the network, for example, trips $[s, t]$ where $(s, t)$ is a transfer edge (making this trip do not actually use the line network).
+The \emph{set of permitted trips} across the network is denoted by $P$, and can be defined regarding some conditions.
 
@@
 \label{Transportation flows}
-Let  $x_{ij}$ count the number of travelers using edge $(ij)$ in a given period, such as a given hour, day, week or  year.
+Let  $x_{ij}$ count the \emph{number of travelers using edge $(i,j)$} in a given period, such as a given hour, day, week or  year.
 The edge flow $x_{ij}$ is denoted by $y_{ij}$ for an intra-line edge $(i,j)$, and
@@
 
-Let $a_i$, respectively $b_i$, the number of passengers embarking, respectively disembarking at stop $i$. By construction, 
+Let $a_i$, respectively $b_i$, the \emph{number of passengers embarking}, respectively \emph{disembarking} at stop $i$. By construction, 
 \begin{equation}
@@
 \end{equation}
-Also, $\mathbf{a}$ and $\mathbf{b}$ must be consistent, in the sense that $A_i\ge B_i$, where $A_i$ (respectively $B_i$) is the cumulated number of embarked 
+Also, $\mathbf{a}$ and $\mathbf{b}$ must be consistent, in the sense that $A_i\ge B_i$, where $A_i$ (respectively $B_i$) is the \emph{cumulated number of embarked 
 (resp.
-disembarked) passengers on the line under consideration, recursively defined as $A_{F(i)}=A_i+a_i$ (resp.
+disembarked) passengers} on the line under consideration, recursively defined as $A_{F(i)}=A_i+a_i$ (resp.
 $B_{F(i)}=B_i+b_i$).
@@
 
-Let the transportation flow $n_{st}$ denote the number of passengers following an $st$-trip, that is entering the network at $s$ and leaving the network at $t$ by using the shortest path.
+Let the \emph{transportation flow} $n_{st}$ denotes the number of passengers following an $st$-trip, that is entering the network at $s$ and leaving the network at $t$ by using the shortest-path.
 One gets from (\ref{edgetrip})
@@
 
-As explained in section \ref{Lines and junctions}, transfers can only occur at junctions, that is $z_{ij}>0$ implies $j\in S_i$.
+As explained in section \ref{Lines and junctions}, transfers can only occur at junctions, that is $z_{ij}>0$ implies $(i, j) \in T$.
 In particular,  $z_{ii}=0$ : no traveller is supposed to disembark and re-embark later at the same stop.
@@
 \subsection{Statement of the problem and solution method}
-Automatic passenger counters measure the number of passengers entering and leaving buses at each stop [Boyle, 1998], that is $\mathbf{a}$ and $\mathbf{b}$, which provide the basic raw data of the present study, kindly provided by the Lausanne Transportation Agency (tl), after some preliminary, undocumented corrections (i.e.
-the components of $\mathbf{a}$ and $\mathbf{b}$ can be non-integer).
-It may also happen that, on some lines $\ell$,  raw data do not obey the necessary consistency condition  $a_\bullet^\ell=b_\bullet^\ell$ (where the latter quantities denote the total embarkments and disembarkments on line $\ell$), in which case we did rescale the embarking and disembarking line counts as
-\begin{displaymath}
-\hat{a}_i=(1-\frac{a_\bullet^\ell-b_\bullet^\ell}{a_\bullet^\ell+b_\bullet^\ell})\, a_i 
- \qquad\qquad\qquad
-  \hat{b}_i=(1+\frac{a_\bullet^\ell-b_\bullet^\ell}{a_\bullet^\ell+b_\bullet^\ell})\,  b_i 
-\end{displaymath}
-ensuring $\hat{a}_\bullet^\ell=\hat{b}_\bullet^\ell=2a_\bullet^\ell b_\bullet^\ell/(a_\bullet^\ell+b_\bullet^\ell)$.
-However, strongly unbalanced lines such that  $|a_\bullet^\ell-b_\bullet|^\ell/a_\bullet^\ell > 0.3$ or $|a_\bullet^\ell-b_\bullet^\ell|/b_\bullet^\ell > 0.3$  (which always turned out to be temporary lines with small  counts) were simply disregarded and line $\ell$ removed from the network.
-
-Also, the geometry of the network permits to derive the edge-trip incidence matrix $\bm{\chi}$ defined in (\ref{edgetrip}).
+\emph{Automatic passenger counters} measure the number of passengers entering and leaving lines at each stop [Boyle, 1998], that is $\mathbf{a}$ and $\mathbf{b}$, which provide the basic raw data of the present study, kindly provided by the Lausanne Transportation Agency (tl) for the case study in section \ref{real_data}.
+We will suppose here that this data obeys the necessary consistency condition $a_\bullet^\ell=b_\bullet^\ell$ (where the latter quantities denote the total embarkments and disembarkments on line $\ell$), even if, in real case studies, a rescaling must usually be performed to balance in and out-flows on each lines.
 
 Intra-line edge flows $\mathbf{Y}=(y_{ij})$ can be determined by (\ref{bilan1ligne}), but transfer edge flows $\mathbf{Z}=(z_{ij})$ are, here and typically, unknown.
-The objective is to estimate the $l\times l$ transportation flow $\mathbf{N}=(n_{st})$.
+The objective is to estimate the $l\times l$ transportation flow matrix $\mathbf{N}=(n_{st})$.
 Many consistent solutions coexist in general, even for a single line with no transferts (section \ref{Single line}).
-This issue of incompletely observed data can be tackled by the maximum entropy formalism \cite{jaynes1957information}, and has been often  in transportation modelling \cite{wilson1967statistical}  \cite{erlander1990gravity}.
+This issue of incompletely observed data can be tackled by the maximum entropy formalism \cite{jaynes1957information}, which has often been the case in transportation modelling researches \cite{wilson1967statistical}  \cite{erlander1990gravity}.
 
-Let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the proportion of $st$-trips (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution). 
+Let $f_{st}=n_{st}/n_{\bullet\bullet}$ be the \emph{distribution of $st$-trips} (empirical distribution) and let $g_{st}$ be some prior guess on its shape (theoretical distribution). 
 Assuming some reasonable initial prior $g_{st}$, 
@@
 as the distribution closest to $g_{st}$ in the Kullback-Leibler divergence sense under the margin constraints, to be calibrated by an iterative fitting inner loop
-\item[(2)] then (section \ref{marginup}), the margins will be updated to $\tilde{\alpha}_s$ and $\tilde{\beta}_t$   by requiring a minimum proportion $\theta\in (0,1)$ of passengers entering/leaving the network at each stop, as well as avoiding transfer overflow exceeding the embarking and disembarking counts at each stop
+\item[(2)] then (section \ref{marginup}), the margins will be updated to $\tilde{\alpha}_s$ and $\tilde{\beta}_t$   by requiring a \emph{minimum proportion} $\theta\in (0,1)$ of passengers entering/leaving the network at each stop, as well as avoiding transfer overflow exceeding the embarking and disembarking counts at each stop
  \item[(3)] finally (section \ref{priorup}), the prior will be updated to $\tilde{g}_{st}$ by shrinking, if necessary, the priors $g_{st}$ associated to overflows. 
@@
 With the new prior distribution $\widetilde{g}_{st}$ and the new margin distributions $\widetilde{\alpha}_s$, $\widetilde{\beta}_t$, we can iterate the 
- the above steps, until convergence. The only free parameter is $\theta$, which will be discussed in section *** .
+ the above steps, until convergence. The only free parameter is $\theta$, whose effect is studied on toy examples in section \ref{toy_examples}.
 
- *** The above iterative solution method is somewhat reminiscent of the EM algorithm.
+The above iterative solution method is somewhat reminiscent of the EM algorithm.
 As a matter of fact, the first  step  (maximum entropy) exactly correspond to the ``expectation step" of the EM algorithm (see e.g.
-\cite{dempster1977maximum}  \cite{bavaud2009information}), but steps two and three, aiming at calibrating the parameters $\bm{\alpha}$, $\bm{\beta}$ and $\mathbf{g}$, do not follow the maximum likelihood rationale of the ``maximisation step".
+\cite{dempster1977maximum}  \cite{bavaud2009information}), but steps two and three, aiming at calibrating parameters $\alpha_s$, $\beta_t$ and $g_{st}$, do not follow the maximum likelihood rationale of the ``maximisation step".
+Pseudocode of the algorithm is shown with Algorithm \ref{algo1}.
 
@@
 \subsubsection{Initialization of the prior and the margins}
-The geometry of the network permits to rule out forbidden $st$-paths, i.e.
-obeying  $\chi^{st}_{\bullet\bullet}=0$.
-Such forbidden $st$-paths consist of *** @G?
-donner une liste exhaustive de critères d'exclusion ici ?
-***  The initial prior was chosen as the uniform distribution on the remaining $c<l^2$ admissible $st$-paths, *** @G: que vaut $l$?
-que vaut $c$?
-***,  that is as
+The geometry of the network permits to define the set of permitted $st$-trips across the network denoted by $P$.
+The initial prior was chosen as the uniform distribution on admissible paths, that is as
 \begin{equation*}
-g_{ij} = \begin{cases}
-  \frac{1}{c}    & \text{if $\chi^{st}_{\bullet\bullet}>0$}, \\
+g_{st} = \begin{cases}
+  \frac{1}{\vert P \vert}    & [s, t] \in $P$, \\
   0    & \text{otherwise}.
@@
 \label{marginup}
-Let us define the hyperparameter $ \theta\in [0, 1)$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the network at each stop}, that is $n_{s\bullet}\ge \theta a_s$ and $n_{\bullet t}\ge \theta b_t$.
+Let us define the hyperparameter $ \theta\in (0, 1)$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the network at each stop}, that is $n_{s\bullet}\ge \theta a_s$ and $n_{\bullet t}\ge \theta b_t$.
 Note that we could set a different hyperparameter for each node, and differing for embarkments and disembarkments, but without addition information, we will restrain to this simpler case.
@@
 \end{displaymath}
-the violation of which constitutes transfer overflow. Hence requiring a minimal transfer yet avoiding overflow can be granted with the following updating of margins
+the violation of which constitutes transfer overflow. Hence requiring a maximal transfer yet avoiding overflow can be granted with the following updating of margins
 \begin{equation}
@@
 \label{priorup}
-Overflow occurs in transfer edge $(i, j)$ if  $z_{i \bullet} > (1 - \theta)b_i$ or $z_{\bullet j} > (1 - \theta)a_j$. To avoid it, components $g_{st}$ of the prior distribution 
-will be shrinked by a suitable ratio whenever edge flows $(i,j)\in [s,t]$ exhibit overflow.
-
+Overflow occurs in transfer edge $(i, j)$ if  $z_{i \bullet} > (1 - \theta)b_i$ or $z_{\bullet j} > (1 - \theta)a_j$.
+To avoid it, components $g_{st}$ of the prior distribution will be shrinked by a suitable ratio whenever edge flows $(i,j)\in [s,t]$ exhibit overflow.
 For any edge $(i, j)$, let us compute the \emph{flow ratio} $r_{ij}$ as
 \begin{equation}
-	\label{flow_ratio}
-	r_{ij} = \max \left(1, \frac{z_{i \bullet}}{(1 - \theta)b_i}, \frac{z_{\bullet j}}{(1 - \theta)a_j} \right)\: \ge\: 1\enspace,
+	r_{ij} = \max \left(1, \frac{z_{i \bullet}}{(1 - \theta)b_i}, \frac{z_{\bullet j}}{(1 - \theta)a_j} \right)\: \ge\: 1\enspace, \label{flow_ratio}
 \end{equation}
@@
 
-\subsubsection{Algorithm pseudocode}
-
-\begin{algorithm}[H]
-	\caption{Compute the transportation flow matrix $\mathbf{N} = (n_{st})$ knowing the edge-trip incidence matrix $\bm{\chi} = (\chi_{ij}^{st})$, the set of transfer edges $\mathcal{T}$, the embarking flow $\mathbf{a}$, the disambarking flow $\mathbf{b}$, the index of an isolated source node $\tilde{s}$, and the minimum proportion of passengers entering/leaving the network $\theta$.}
+\begin{algorithm}[h]
+	\caption{Compute the transportation flow matrix $\mathbf{N} = (n_{st})$ knowing the edge-trip incidence matrix $\bm{\chi} = (\chi_{ij}^{st})$, the set of transfer edges $T$, the set of permitted trips $P$, the embarking flow $\mathbf{a}$, the disambarking flow $\mathbf{b}$, the index of an isolated source node $\tilde{s}$, and the minimum proportion of passengers entering/leaving the network $\theta$.}
 	\label{algo1}
 	\begin{algorithmic}[1]
-		\State $g_{st} \leftarrow I(\chi_{\bullet \bullet}^{st} > 0) / \sum_{s't'}I(\chi_{\bullet \bullet}^{s't'} > 0), \; \forall s,t$ \Comment{{\footnotesize Initialize the prior distribution}}
+		\State $g_{st} \leftarrow I([s, t] \in P) / \vert P \vert, \; \forall s,t$ \Comment{{\footnotesize Initialize the prior distribution}}
 
@@
 
-		\State $z_{ij} \leftarrow I((i,j) \in \mathcal{T})\sum_{st} \chi_{ij}^{st} n_{st}, \; \forall i,j$
+		\State $z_{ij} \leftarrow I((i,j) \in T)\sum_{st} \chi_{ij}^{st} n_{st}, \; \forall i,j$
 
@@
 
-***  @G, ici davantage de développements sur l'approche Markovienne, ainsi que le joli schéma ?  ***
-
-\
-
-*** @G, @R, ici la (les) figure de l'exemple  ``starting from Maladière, Riant-Cour, Dapples"... ? ***
-
 %% --------------------------------- CASE STUDIES
@@
 Case studies are divided in two sections.
-In the first section, we test the algorithm on toy examples, which are artifical networks where the transportation flow $n_{st}$ is randomly drawn.
+In the first section, we test the algorithm on toy examples, which are artificial networks where the transportation flow $n_{st}$ is randomly drawn.
 These examples enable some kind of validation of the algorithm, as "real" transportation flows are known and can be compared to the solutions given by our method.
-This setup differs from the second section, which is dedicated to applying the algorithm to the real case of the public transportation network of the city of Lausanne (tl), where embarkement and disembarkement flows are measured but transportation flows are unknown.
+This setup differs from the second section, which is dedicated to applying the algorithm to the real case of the public transportation network of the city of Lausanne (tl), where embankment and disembarkment flows are measured but transportation flows are unknown.
 This second case study shows that the algorithm is applicable on large, real datasets and can gives insights about passengers probable routes in the network.
 
-\subsection{Toy Examples}
+\subsection{Error measurements}
+\label{error_measures}
 
-\subsubsection{Examples construction}
-\label{example_construction}
-
-All constructed toy examples are built following the same approach, which aims at being simple but somewhat realistic.
-We fix a number of \emph{line tours} $p \geq 2$, each of which is constituted of a forward line and a backward line, for a total of $q = 2p$ lines.
-Every line has a starting and ending node, which are isolated node, and posses $p-1$ intermediary nodes which allows transfers to the other tour lines, giving a total of $n = 2p(p + 1)$ nodes in the network.
-Examples of these toy networks can be found in Figure \ref{toy_example_plots}.
-
-\begin{figure}
-	\includegraphics[width=0.3\textwidth]{fig/toy_2_display.pdf}
-	\includegraphics[width=0.3\textwidth]{fig/toy_3_display.pdf}
-	\includegraphics[width=0.3\textwidth]{fig/toy_4_display.pdf}
-	\caption{3 toy examples, with the number of tours $p \in \{2,3,4\}$.
-Tours are displayed in the same color, line have a unique letter, and each stop a unique combination of a letter and a number.
-Clusters of nodes represent positions where transfer between tours are possible.}
-	\label{toy_example_plots}
-\end{figure}
-
-Allowed origin-destination pairs $(s,t)$ are constructed with the same set of rules found in (REF), but without considering the exclusion criterion based on the pedestrian distance, as nodes are not spatially located.
-The graph shortest-path structure for the allowed origin-destination pairs is stored in the edge-trip incidence matrix $\bm{\chi}$.
-
-A transportation flow $\mathbf{N}_\text{ref} = (n^\text{ref}_{st})$ is drawn by setting a fixed number of passengers $n^\text{ref}_{\bullet \bullet}$, and each passenger is assigned randomly to a $(s, t)$ pair drawn uniformly among all allowed origin-destination pairs.
-From this reference transportation flow $\mathbf{N}_\text{ref}$, using the edge-trip incidence matrix $\bm{\chi}$ and equation (\ref{equationGG}), we can compute flow on edges $\mathbf{X}_\text{ref}$ and, in turn, the number of passengers embarking $\mathbf{a}_\text{ref}$ and the number of passengers disembarking $\mathbf{b}_\text{ref}$ at each stop.
-
-\subsubsection{Error measurements}
-\label{error_measures}
-Using the network structure, $\mathbf{a}_\text{ref}$, $\mathbf{b}_\text{ref}$, we can obtain a estimation of the real flow through the algorithm, noted $\mathbf{N} = (n_{st})$.
-There are two types of dissimilarity measure between the real data and the solution proposed by the algorithm: (1) how much $\mathbf{N}$ differs from $\mathbf{N}_\text{ref}$, and (2) how well constraints defined by the number of passengers embarking $\mathbf{a}_\text{ref}$ and disembarking $\mathbf{b}_\text{ref}$ are respected.
+In all case studies, we obtain a estimation of the transportation flow with the algorithm, noted $\mathbf{N} = (n_{st})$, starting from the real embarkment flow $\mathbf{a}_\text{ref}$ and disembarkment flow $\mathbf{b}_\text{ref}$.
+In toy examples, we also have access to the real transportation flow $\mathbf{N}_\text{ref}$.
+There are two types of dissimilarity measures between the data and the solution proposed by the algorithm: (1) if we have access to $\mathbf{N}_\text{ref}$, how much $\mathbf{N}$ differs from it, and (2) how well constraints defined by $\mathbf{a}_\text{ref}$ and $\mathbf{b}_\text{ref}$ are respected.
 The first dissimilarity is measured through the \emph{mean transportation error}, denoted by $\text{MTE}(\mathbf{N})$, and computed with
@@
 \end{align} 
-where $z_{ij} = I((i,j) \in \mathcal{T})\sum_{st} \chi_{ij}^{st} n_{st}$ is the flow on transfer edges $\mathcal{T}$.
+where $z_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} n_{st}$ is the flow on transfer edges $T$.
 
-Note that, by construction, the mean margin error should be null when the algorithm converge.
+Note that, by construction, the mean margin error should be null when the algorithm converges.
 However, it can be informative to track down this error along iterations and, in some practical cases where the network is large, the algorithm convergence criterion is reached without having margin constraints perfectly respected.
 
-\subsubsection{Algorithm iterations}
+\subsection{Toy Examples}
+\label{toy_examples}
+
+\subsubsection{Examples construction}
 \label{example_construction}
 
-\begin{figure}
+All constructed toy examples are built following the same approach, which aims at being simple but somewhat realistic.
+We fix a number of \emph{line tours} $p \geq 2$, each of which is constituted of a forward line and a backward line, for a total of $q = 2p$ lines.
+Every line has a starting and ending node, which are isolated node, and posses $p-1$ intermediary nodes which allows transfers to the other tour lines, giving a total of $n = 2p(p + 1)$ nodes in the network.
+Examples of these toy networks can be found in Figure \ref{toy_example_plots}.
+
+\begin{figure}[h]
+	\includegraphics[width=0.3\textwidth]{fig/toy_2_display.pdf}
+	\includegraphics[width=0.3\textwidth]{fig/toy_3_display.pdf}
+	\includegraphics[width=0.3\textwidth]{fig/toy_4_display.pdf}
+	\caption{3 toy examples, with the number of tours $p \in \{2,3,4\}$.
+Tours are displayed in the same color, line have a unique letter, and each stop a unique combination of a letter and a number.
+Clusters of nodes represent positions where transfer between tours are possible.}
+	\label{toy_example_plots}
+\end{figure}
+
+In order to be realistic, permitted $st$-trips set $P$ is constructed considering all shortest-path between pair of nodes, excluding :
+\begin{itemize}
+\item $s$ and $t$ that are on the same line but with $t$ preceding $s$ in the line order.
+\item $s$ and $t$ that are on the same tour but opposite line.
+\item $s$ and $t$ whose shortest-path starts with a transfer edge, ends with a transfer edge, or possesses two (or more) consecutive transfer edges.
+\end{itemize}
+
+A transportation flow $\mathbf{N}_\text{ref} = (n^\text{ref}_{st})$ is drawn by setting a fixed number of passengers $n^\text{ref}_{\bullet \bullet}$, and each passenger is assigned randomly to a $(s, t)$ pair drawn uniformly among $P$.
+From this reference transportation flow $\mathbf{N}_\text{ref}$, using the edge-trip incidence matrix $\bm{\chi}$ and equation (\ref{equationGG}), we can compute flow on edges $\mathbf{X}_\text{ref}$ and, in turn, the number of passengers embarking $\mathbf{a}_\text{ref}$ and the number of passengers disembarking $\mathbf{b}_\text{ref}$ at each stop.
+
+\subsubsection{Algorithm iterations}
+\label{algorithm_iterations}
+
+First, we shows some algorithm iterations on a toy example with $p=2$, where $100$ passengers where drawn uniformly across the $\vert P \vert = 20$ possible st-trips.
+Some iterations of the algorithm with $\theta=0.1$, along with MTE and MME errors, are shown in Figure \ref{iteration_plots}.
+\begin{figure}[h]
 	\label{iteration_plots}
 	\includegraphics[width=0.98\textwidth]{fig/iterations.pdf}
-	\caption{Blabla}
+	\caption{Real transportation flow (green arrows) obtained by randomly drawing $100$ passengers on the graph toy example with $p=2$ line tours, along with iterations $1, 3, 6, 10, 15$ of the algorithm with $\theta=0.1$.
+MTE and MME errors are computed, and embarkment and disembarkment flows are represented respectively by the red and blue colors on nodes.}
 \end{figure}
+On this small example, we can see that the algorithm quickly find an estimation giving small MME error, but still give an MTE of 0.252.
+This result is due to the fact that only $100$ passengers are drawn, giving a large deviation compared to the optimally found solution which maximize the entropy.
+Interestingly, we see that a better result is found on iteration 6, but margins constraints at this point are not perfectly respected yet.
+
+\subsubsection{Algorithm iterations}
+\label{algorithm_iterations}
 
 \subsection{Real Data}
+\label{real_data}
+
+after some preliminary, undocumented corrections (i.e.
+the components of $\mathbf{a}$ and $\mathbf{b}$ can be non-integer).
+It may also happen that, on some lines $\ell$,  raw data do not obey the necessary consistency condition  $a_\bullet^\ell=b_\bullet^\ell$ (where the latter quantities denote the total embarkments and disembarkments on line $\ell$), in which case we did rescale the embarking and disembarking line counts as
+\begin{displaymath}
+	\hat{a}_i=(1-\frac{a_\bullet^\ell-b_\bullet^\ell}{a_\bullet^\ell+b_\bullet^\ell})\, a_i 
+	\qquad\qquad\qquad
+	\hat{b}_i=(1+\frac{a_\bullet^\ell-b_\bullet^\ell}{a_\bullet^\ell+b_\bullet^\ell})\,  b_i 
+\end{displaymath}
+ensuring $\hat{a}_\bullet^\ell=\hat{b}_\bullet^\ell=2a_\bullet^\ell b_\bullet^\ell/(a_\bullet^\ell+b_\bullet^\ell)$.
+However, strongly unbalanced lines such that  $|a_\bullet^\ell-b_\bullet|^\ell/a_\bullet^\ell > 0.3$ or $|a_\bullet^\ell-b_\bullet^\ell|/b_\bullet^\ell > 0.3$  (which always turned out to be temporary lines with small  counts) were simply disregarded and line $\ell$ removed from the network.
+
+Also, the geometry of the network permits to derive the edge-trip incidence matrix $\bm{\chi}$ defined in (\ref{edgetrip}).
+
 \subsubsection{Dataset construction}

### 2023-02-21 4.2_plot_toy_iterations.R

@@ -144,8 +144,8 @@ for(it in it_list){
   error_out = sum(abs(flow_l_out - flow_r_out)) / sum(flow_l_out)
   error_m = mean(c(error_in, error_out))
   plot_flow_graph(adj, n_mat, layout, v_names=stop_names, 
-                  main=paste0("\n\nIt=", it - 1, "\nError=", 
-                              format(error, digits=3),"\nMargins error=", 
+                  main=paste0("\n\nIt=", it - 1, "\nMTE=", 
+                              format(error, digits=3),"\nMME=", 
                               round(error_m, digits=3)), 
                   e_line_size=e_line_size, e_flow_size=e_flow_size, 
                   v_label_size=v_label_size, v_size=v_size)

### 2023-02-21 latex_files/applied_network_science_article/article_v0.tex

@@
 
-Let $S_i$ denote the \emph{set of stops which can be reached from stop $i$ outside lines connection} (with, e.g., an acceptable walking distance), excluding $i$ itself.
+Let $S_i$ denotes the \emph{set of stops which can be reached from stop $i$ outside lines connection} (with, e.g., an acceptable walking distance), excluding $i$ itself.
 A stop $i$ is referred to as an \emph{isolated stop} if $S_i=\emptyset$, and to as a \emph{junction} otherwise.
@@
 
-First, we shows some algorithm iterations on a toy example with $p=2$, where $100$ passengers where drawn uniformly across the $\vert P \vert = 20$ possible st-trips.
+First, we shows some algorithm iterations on a toy example with $p=2$, where $100$ passengers where drawn uniformly across the $\vert P \vert = 20$ possible $st$-trips.
 Some iterations of the algorithm with $\theta=0.1$, along with MTE and MME errors, are shown in Figure \ref{iteration_plots}.
@@
 \label{algorithm_iterations}
+
+\begin{figure}
+	\includegraphics[width=0.47\textwidth]{fig/toy_MTE_passengers_theta01.pdf}
+	\includegraphics[width=0.47\textwidth]{fig/toy_MTE_tours.pdf}
+	\caption{\emph{Mean transportation error} (MTE), according to the number of tours $p \in \{2,3,4,5,6,7,8\}$ and resp.
+to the number of passengers into the network per line tour and the minimum proportion $\theta =0.1$ of passengers entering/leaving the network at each stop with a proportional number of passengers to the shortest paths of each line tour.}
+	\label{toy_MTE}
+\end{figure}
+
+By running the algorithm for line tours $p \in \{2,3,4,5,6,7,8\}$, we observe that MTE depends on different parameters.
+On the graph on the left in Figure \ref{toy_MTE}, the number of passengers inserted in the network has an influence on MTE.
+The higher the number of passengers, the smaller the error is because passengers are better distributed throughout the larger network.
+On the right in Figure \ref{toy_MTE}, the aim is to understand how the proportion $\theta\in (0,1)$ responds to different line tours.
+Dots represent local minimum MTE for each tour.
 

### 2023-02-21 latex_files/applied_network_science_article/article_v0.tex
write article

@@
 
-Note that, by construction, the mean margin error should be null when the algorithm converges.
+Note that, by construction, the MME should be null when the algorithm converges.
 However, it can be informative to track down this error along iterations and, in some practical cases where the network is large, the algorithm convergence criterion is reached without having margin constraints perfectly respected.
@@
 
-\subsubsection{Examples construction}
+\subsubsection{Construction}
 \label{example_construction}
@@
 \begin{figure}[h]
-	\label{iteration_plots}
 	\includegraphics[width=0.98\textwidth]{fig/iterations.pdf}
@@
 MTE and MME errors are computed, and embarkment and disembarkment flows are represented respectively by the red and blue colors on nodes.}
+	\label{iteration_plots}
 \end{figure}
@@
 
-\subsubsection{Algorithm iterations}
-\label{algorithm_iterations}
+\subsubsection{MTE study}
+\label{mte_study}
+
+The main goal of toy examples, since we have access to the real transportation flow, is to study how the resulting $MTE$ behave regarding passenger $st$-trips distribution and hyperparameter $\theta$ on different network sizes.
 

### 2023-02-22 5.3_visualisation.R
new graphs

@@ -43,9 +43,9 @@ nb_stops = nb_lines + 1
 #--------------------------------
 
 # Conv threshold for iterative fitting
-conv_thres_if = 0.0001
+conv_thres_if = 1e-5
 # Conv threshold
-conv_thres_algo = 0.0001
+conv_thres_algo = 1e-8
 # epsilon
 epsilon = 1e-40
 # max iteration
@@ -55,7 +55,7 @@ display_it = F
 # number of iterations
 n_test = 10
 # prop limit
-hyper_par = 0.1
+hyper_par = 0.001
 
 #--------------------------------
 # Output and visualizations
@@ -70,9 +70,11 @@ hyper_par = 0.1
 # Create a data frame with all different number of passengers into the network
 
 res_mean = c()
-seq_passengers = seq(from = 200, to = 4000, by = 200)
-seq_passengers = seq(from = 5000, to = 20000, by = 1000)
+# seq_passengers = seq(from = 200, to = 4000, by = 200)
+seq_passengers = seq(from = 1000, to = 20000, by = 1000)
+seq_passengers = seq(from = 200, to = 800, by = 200)
 
+# rbind(seq_passengers,seq_passengers2)
 
 for (j in 1:length(nb_lines)) {
   
@@ -103,9 +105,10 @@ for (j in 1:length(nb_lines)) {
   res_mean = rbind(res_mean, mean_pass)
 }
 
-# write.csv(res_mean, "results/5_toy_ex_outputs/res_mean_passengers.csv", row.names = F)
-res_mean0 = read.csv("results/5_toy_ex_outputs/res_mean_passengers_theta01.csv")
-res_mean2 = rbind(res_mean0,res_mean)
+# write.csv(res_mean2, "results/5_toy_ex_outputs/res_mean_passengersNEWbig.csv", row.names = F)
+# res_mean_big = res_mean
+# res_mean0 = read.csv("results/5_toy_ex_outputs/res_mean_passengers_theta01.csv")
+# res_mean2 = rbind(res_mean,res_mean_big)
 
 
 ### Best graph according to the number of passengers into the network
@@ -120,10 +123,10 @@ res_mean2 = rbind(res_mean0,res_mean)
 #   labs(title = paste0(n_test, " iterations, ", expression(theta), ": ", hyper_par), x = "Passengers into the network", y = "Mean error")
 
 # Plot the results 2
-ggplot(data=res_mean2) +
+ggplot(data=res_mean) +
[…]

### 2023-02-22 5.4_visualisation.R

@@ -42,9 +42,9 @@ out_folder = "results/loop_toy_example"
 #--------------------------------
 
 # Conv threshold for iterative fitting
-conv_thres_if = 0.0001
+conv_thres_if = 1e-5
 # Conv threshold
-conv_thres_algo = 0.0001
+conv_thres_algo = 1e-8
 # epsilon
 epsilon = 1e-40
 # max iteration
@@ -140,4 +140,4 @@ ggplot(data=mean_line) +
   labs(x=expression(theta), y="MTE") +
   scale_x_continuous(breaks = seq(0, 1, 0.1)) +
   geom_point(data=min_max,aes(x=param, y=mean_error, color=factor(Lines)))+
-  labs(color = "Nb tours", fill = "Nb tours")
+  labs(color = "p", fill = "p")

### 2023-02-22 latex_files/applied_network_science_article/article_v0.tex
add graph to latex

@@
 
-The main goal of toy examples, since we have access to the real transportation flow, is to study how the resulting $MTE$ behave regarding passenger $st$-trips distribution and hyperparameter $\theta$ on different network sizes.
+The main goal of toy examples, since we have access to the real transportation flow, is to study how the resulting $MTE$ behave regarding passenger $st$-trips distribution and hyperparameter $\theta$, on different network sizes.
+The randomness of the $st$-trips distribution
 
 \begin{figure}
-	\includegraphics[width=0.47\textwidth]{fig/toy_MTE_passengers_theta01.pdf}
-	\includegraphics[width=0.47\textwidth]{fig/toy_MTE_tours.pdf}
+	\includegraphics[width=0.98\textwidth]{fig/MTE_passengers_theta01.pdf}
 	\caption{\emph{Mean transportation error} (MTE), according to the number of tours $p \in \{2,3,4,5,6,7,8\}$ and resp.
 to the number of passengers into the network per line tour and the minimum proportion $\theta =0.1$ of passengers entering/leaving the network at each stop with a proportional number of passengers to the shortest paths of each line tour.}
-	\label{toy_MTE}
+	\label{MTE_vs_passengers}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=0.98\textwidth]{fig/MTE_tours.pdf}
+	\caption{\emph{Mean transportation error} (MTE), according to the number of tours $p \in \{2,3,4,5,6,7,8\}$ and resp.
+to the number of passengers into the network per line tour and the minimum proportion $\theta =0.1$ of passengers entering/leaving the network at each stop with a proportional number of passengers to the shortest paths of each line tour.}
+	\label{MTE_vs_theta}
 \end{figure}

### 2023-02-22 4.2_plot_toy_iterations.R

@@ -18,7 +18,7 @@ source("local_functions.R")
 #--------------------------------
 
 # Result folder
-result_folder = "results/toy_example_plots"
+result_folder = "results/toy_example_plots_2"
 
 # Choose number of lines
 nb_lines = 2

### 2023-02-23 latex_files/applied_network_science_article/article_v0.tex
write article

@@
 as the distribution closest to $g_{st}$ in the Kullback-Leibler divergence sense under the margin constraints, to be calibrated by an iterative fitting inner loop
-\item[(2)] then (section \ref{marginup}), the margins will be updated to $\tilde{\alpha}_s$ and $\tilde{\beta}_t$   by requiring a \emph{minimum proportion} $\theta\in (0,1)$ of passengers entering/leaving the network at each stop, as well as avoiding transfer overflow exceeding the embarking and disembarking counts at each stop
+\item[(2)] then (section \ref{marginup}), the margins will be updated to $\tilde{\alpha}_s$ and $\tilde{\beta}_t$   by requiring a \emph{minimum proportion} $\theta\in [0,1)$ of passengers entering/leaving the network at each stop, as well as avoiding transfer overflow exceeding the embarking and disembarking counts at each stop
  \item[(3)] finally (section \ref{priorup}), the prior will be updated to $\tilde{g}_{st}$ by shrinking, if necessary, the priors $g_{st}$ associated to overflows. 
@@
 \end{equation*}
-The initial margins were chosen as  those of the flow without transfer, namely $\alpha_i=\frac{a_i}{a_{\bullet}}$ and $\beta_i=\frac{b_i}{b_\bullet}$ for all stops.
+The initial margins were chosen to initially match $g_{st}$, namely $\alpha_s=g_{s \bullet}$ and $\beta_t=g_{\bullet t}$ for all stops.
 
@@
 \label{marginup}
-Let us define the hyperparameter $ \theta\in (0, 1)$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the network at each stop}, that is $n_{s\bullet}\ge \theta a_s$ and $n_{\bullet t}\ge \theta b_t$.
+Let us define the hyperparameter $ \theta\in [0, 1)$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the network at each stop}, that is $n_{s\bullet}\ge \theta a_s$ and $n_{\bullet t}\ge \theta b_t$.
 Note that we could set a different hyperparameter for each node, and differing for embarkments and disembarkments, but without addition information, we will restrain to this simpler case.
@@
 In the first section, we test the algorithm on toy examples, which are artificial networks where the transportation flow $n_{st}$ is randomly drawn.
-These examples enable some kind of validation of the algorithm, as "real" transportation flows are known and can be compared to the solutions given by our method.
-This setup differs from the second section, which is dedicated to applying the algorithm to the real case of the public transportation network of the city of Lausanne (tl), where embankment and disembarkment flows are measured but transportation flows are unknown.
+These examples enable some kind of validation of the algorithm, as the "real" transportation flow is known and can be compared to solutions given by our method.
+This setup differs from the second section, which is dedicated to applying the algorithm to the real case of the public transportation network of the city of Lausanne (tl), where embankment and disembarkment flows are measured but the real transportation flow is unknown.
 This second case study shows that the algorithm is applicable on large, real datasets and can gives insights about passengers probable routes in the network.
@@
 \end{align} 
-where $z_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} n_{st}$ is the flow on transfer edges $T$.
+where $z_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} n_{st}$ is the flow on transfer edges $T$. Both errors can be interpreted as a weighted mean deviation percentage.
 
@@
 	\includegraphics[width=0.98\textwidth]{fig/iterations.pdf}
-	\caption{Real transportation flow (green arrows) obtained by randomly drawing $100$ passengers on the graph toy example with $p=2$ line tours, along with iterations $1, 3, 6, 10, 15$ of the algorithm with $\theta=0.1$.
+	\caption{Real transportation flow (green arrows) obtained by randomly drawing $100$ passengers on the graph toy example with $p=2$ line tours, along with iterations $1, 2, 4, 7, 10$ of the algorithm with $\theta=0.1$.
 MTE and MME errors are computed, and embarkment and disembarkment flows are represented respectively by the red and blue colors on nodes.}
@@
 On this small example, we can see that the algorithm quickly find an estimation giving small MME error, but still give an MTE of 0.252.
-This result is due to the fact that only $100$ passengers are drawn, giving a large deviation compared to the optimally found solution which maximize the entropy.
-Interestingly, we see that a better result is found on iteration 6, but margins constraints at this point are not perfectly respected yet.
+This result is due to the fact that only $50$ passengers are drawn, giving a large deviation compared to the optimally found solution which maximize the entropy.
+Interestingly, we see that a better result is found on iteration , but margins constraints at this point are not perfectly respected yet.
 
@@
 The main goal of toy examples, since we have access to the real transportation flow, is to study how the resulting $MTE$ behave regarding passenger $st$-trips distribution and hyperparameter $\theta$, on different network sizes.
-The randomness of the $st$-trips distribution
+
+The randomness of the $st$-trips distribution is controlled by the number of drawn passengers and we can see that the algorithm performs better if this number increases, as seen in Figure \ref{MTE_vs_passengers}.
+This behavior can be expected as the method is constructed to find the solution maximizing the entropy while respecting embarkment and disembarkment constraint.
+In fact, all MTE should converge to $0$ eventually, however this rate of convergence seems to be low.
+\\
+As for the hyperparameter study, found in Figure \ref{MTE_vs_theta}, we see that the optimal parameter seems to decreases as network size increases. However,
 
@@
 	\includegraphics[width=0.98\textwidth]{fig/MTE_passengers_theta01.pdf}
-	\caption{\emph{Mean transportation error} (MTE), according to the number of tours $p \in \{2,3,4,5,6,7,8\}$ and resp.
-to the number of passengers into the network per line tour and the minimum proportion $\theta =0.1$ of passengers entering/leaving the network at each stop with a proportional number of passengers to the shortest paths of each line tour.}
+	\caption{\emph{Mean transportation error} (MTE) vs the number of randomly drawn passengers, in toy examples with number of tours $p$ varying from 2 to 8.
+The hyperparameter was set to $\theta=0.001$ and $10$ different drawn where performed for each data point.}
 	\label{MTE_vs_passengers}
@@
 \end{figure}
-
-By running the algorithm for line tours $p \in \{2,3,4,5,6,7,8\}$, we observe that MTE depends on different parameters.
-On the graph on the left in Figure \ref{toy_MTE}, the number of passengers inserted in the network has an influence on MTE.
-The higher the number of passengers, the smaller the error is because passengers are better distributed throughout the larger network.
-On the right in Figure \ref{toy_MTE}, the aim is to understand how the proportion $\theta\in (0,1)$ responds to different line tours.
-Dots represent local minimum MTE for each tour.
 

### 2023-02-23 4.2_plot_toy_iterations.R

@@ -18,14 +18,14 @@ source("local_functions.R")
 #--------------------------------
 
 # Result folder
-result_folder = "results/toy_example_plots_2"
+result_folder = "results/toy_example_plots"
 
 # Choose number of lines
 nb_lines = 2
 # Choose number of stops
 nb_stops = nb_lines + 1
 # Choose number of passengers in the network (with normal distribution)
-nb_passengers = 100
+nb_passengers = 50
 # Total number of stops *2 (back and forth)
 nb_stops_tot = nb_lines*nb_stops*2
 
@@ -34,19 +34,19 @@ conv_thres_if = 1e-6
 # Conv threshold
 conv_thres_algo = 1e-8
 # proportional limit 
-min_p_ntwk = 0.1
+min_p_ntwk = 0.01
 # epsilon
 epsilon = 1e-40
 # max iterations
-max_it = 16
+max_it = 50
 # max iterations for iterative fitting
 max_it_if = 1000
 
 # Seed
-set.seed(10)
+set.seed(3)
 
 # It list
-it_list = c(2, 4, 7, 11, 16)
+it_list = c(1, 2, 4, 7, 10)
 
 #--------------------------------
 # Code
@@ -144,7 +144,7 @@ for(it in it_list){
   error_out = sum(abs(flow_l_out - flow_r_out)) / sum(flow_l_out)
   error_m = mean(c(error_in, error_out))
   plot_flow_graph(adj, n_mat, layout, v_names=stop_names, 
-                  main=paste0("\n\nIt=", it - 1, "\nMTE=", 
+                  main=paste0("\n\nIt=", it, "\nMTE=", 
                               format(error, digits=3),"\nMME=", 
                               round(error_m, digits=3)), 
                   e_line_size=e_line_size, e_flow_size=e_flow_size, 

### 2023-02-24 5.3_visu_passengers.R
visu passenger update

@@ -68,11 +68,9 @@ hyper_par = 0.001
 # list2env(network_prop_list, .GlobalEnv)
 
 # Create a data frame with all different number of passengers into the network
-
 res_mean = c()
-# seq_passengers = seq(from = 200, to = 4000, by = 200)
-seq_passengers = seq(from = 1000, to = 20000, by = 1000)
-seq_passengers = seq(from = 200, to = 800, by = 200)
+seq_passengers = c(100,500,1000,1500,2000,2500,3000,3500,4000,5000,6000,7500,
+                   10000,12500,15000,20000,30000,50000)
 
 # rbind(seq_passengers,seq_passengers2)
 
@@ -105,11 +103,7 @@ for (j in 1:length(nb_lines)) {
   res_mean = rbind(res_mean, mean_pass)
 }
 
-# write.csv(res_mean2, "results/5_toy_ex_outputs/res_mean_passengersNEWbig.csv", row.names = F)
-# res_mean_big = res_mean
-# res_mean0 = read.csv("results/5_toy_ex_outputs/res_mean_passengers_theta01.csv")
-# res_mean2 = rbind(res_mean,res_mean_big)
-
+write.csv(res_mean, "results/5_toy_ex_outputs/res_mean_passengers.csv", row.names = F)
 
 ### Best graph according to the number of passengers into the network
 # only 1/3 error bar
@@ -124,9 +118,10 @@ for (j in 1:length(nb_lines)) {
 
 # Plot the results 2
 ggplot(data=res_mean) +
-  geom_ribbon(aes(x=passengers, ymin=mean_error-sd_error/sqrt(n_test), 
-                  ymax=mean_error+sd_error/sqrt(n_test), group=line, 
+  geom_ribbon(aes(x=passengers, ymin=mean_error-2*sd_error/sqrt(n_test), 
+                  ymax=mean_error+2*sd_error/sqrt(n_test), group=line, 
                   fill=factor(line)), alpha=0.2) +
   geom_line(aes(x=passengers, y=mean_error, group=line, color=factor(line))) +
   labs(x="Number of passengers", y="MTE") +
   labs(color = "p", fill = "p")
+

### 2023-02-24 latex_files/applied_network_science_article/article_v0.tex
Dataset description

@@
 \subsubsection{Dataset construction}
-The original dataset used in this contribution consists of over 13 million rows of data over more than 1200 stop on 35 bus lines.
-Each data row represents a stop by a bus to embark and disembark passengers.
-The dataset used for further analysis is summed by stop over the year 2019.
+This research presents a dataset derived from the public transportation agency in Lausanne (tl), Switzerland.
+Automatic passenger counters were used to collect data on the number of passengers entering and leaving buses and metros at each stop.
+The dataset comprises 44 lines of metro and bus transportation, 1361 stops, and over 115 million passengers per year.
+Each round trip forms two separate lines, which often have stops that correspond to both the inbound and outbound directions, but this is not always the case.
+Some stops may be unique to either the inbound or outbound direction.
+The initial dataset contained 13 million data rows, which were aggregated to the year 2019, and the passenger data was filtered to include only passengers who boarded and disembarked at each stop.
+Lines with the most errors were removed, resulting in a final dataset of 35 transportation lines and 1216 stops.
+The threshold for invalid lines was set at a difference of 15\% between boarded and disembarked passenger counters.
 

### 2023-02-24 2_compute_OD_matrix.R
modified algorithm

@@ -35,9 +35,9 @@ out_folder = "results/all_lines"
 # Conv threshold for iterative fitting
 conv_thres_if = 1e-5
 # Conv threshold
-conv_thres_algo = 1e-8
+conv_thres_algo = 1e-7
 # proportional limit 
-min_p_ntwk = 0.001
+min_p_ntwk = 0.1
 # epsilon
 epsilon = 1e-40
 # max iterations

### 2023-02-24 4.2_plot_toy_iterations.R

@@ -32,7 +32,7 @@ nb_stops_tot = nb_lines*nb_stops*2
 # Conv threshold for iterative fitting
 conv_thres_if = 1e-6
 # Conv threshold
-conv_thres_algo = 1e-8
+conv_thres_algo = 1e-10
 # proportional limit 
 min_p_ntwk = 0.01
 # epsilon
@@ -46,7 +46,7 @@ max_it_if = 1000
 set.seed(3)
 
 # It list
-it_list = c(1, 2, 4, 7, 10)
+it_list = c(1, 2, 4, 7, 50)
 
 #--------------------------------
 # Code
@@ -128,12 +128,13 @@ e_flow_size = 3
 v_size = 20
 v_label_size = 0.00001
 
-pdf(paste0(result_folder, "/iterations.pdf"))
+pdf(paste0(result_folder, "/iterations_test.pdf"))
 par(mfrow=c(2,3))
 plot_flow_graph(adj, n_real, layout, v_names=stop_names, main="Real", 
                 e_line_size=e_line_size, e_flow_size=e_flow_size, 
                 v_label_size=v_label_size, v_size=v_size)
 
+it_list[it_list > length(n_mat_list)] = length(n_mat_list)
 for(it in it_list){
   n_mat = n_mat_list[[it]]
   x_r_btw = compute_x_from_n(n_mat, edge_ref, sp_ref, sp_edge_link)$x_btw

### 2023-02-24 local_functions.R

@@ -476,7 +476,227 @@ balance_flow_l_inout =
 # - iterations_df:    A dataframe containing iterations details.
 #-------------------------------------------------------------------------------
 
+
 compute_origin_destination = 
+  function(flow_l_in, flow_l_out, edge_ref, sp_ref, sp_edge_link, 
+           s_mat=NULL, min_p_ntwk=0.1, conv_thres_if=1e-5,
+           conv_thres_algo=1e-5, epsilon=1e-40,
+           max_it=200, max_it_if=400, display_it=T, return_it=F, 
+           return_details=F){
+    
+    # --- Get the network structure 
+    
+    # The number of stops 
+    n = length(flow_l_in)
+    # The reference of transfer edges
+    where_edge_btw = which(as.logical(edge_ref[, 3]))
+    # The list of in-out nodes for transfer edges 
+    edge_btw_ref = edge_ref[where_edge_btw, 1:2]
+    # The shortest-path - edge matrix retrained on transfer edges
+    sp_edge_btw_link = sp_edge_link[, where_edge_btw]
+    # The shortest-path order reference
+    sp_order_ref = mapply(function(i, j) i+n*(j-1), sp_ref[, 1], sp_ref[, 2])
+    
+    # If s_mat is not given, build it
+    if(is.null(s_mat)){
+      # The matrix of admissible sp 
+      s_mat = 1*as.matrix(sparseMatrix(i=sp_ref[, 1], j=sp_ref[, 2], 
+                                       dims=c(n, n)))
+    }
+    # A stop without transfer edges
+    trans_edge = edge_ref[edge_ref[,3] == 1, ]
+    node_ref = setdiff(1:n, unique(c(trans_edge[,1], trans_edge[,2])))[1]
+    
+    # --- Algorithm 
+    
+    # The st-distribution reference matrix
+    g_ref = s_mat / sum(s_mat)
+    # The initial value for f_mat (has no effect, only to create old variable)
+    f_mat = matrix(1e5, n, n)
+    # The sum in and out
+    sum_flow_l_in = sum(flow_l_in)
+    sum_flow_l_out = sum(flow_l_out)
+    # The evolving in and out distribution
+    # sigma_in = flow_l_in / sum_flow_l_in
+    # sigma_out = flow_l_out / sum_flow_l_out
+    sigma_in = rowSums(g_ref)
+    sigma_out = colSums(g_ref)
+    # A boolean for convergence
+    converge_algo = F
+    # Iteration counter
[…]

### 2023-02-24 latex_files/applied_network_science_article/article_v0.tex
write article

--- 2023-02-23 latex_files/applied_network_science_article/article_v0.tex
@@
 \begin{abstract} % abstract
-Characterizing a public transportation network, such as an urban multi-lines bus network, requires the origin-destination trip counts during a given period.
-Yet, if automatic counting makes the embarkment (boarding) and disembarkment (alighting) counts at each bus stop known, it often happens that pedestrian transfers between stops are unknown, and this contribution proposes a three-steps procedure for estimating the missing information, involving maximum entropy and iterative fitting.
-** à poursuivre ***
+Characterizing a public transportation network, such as an urban network with multiple lines, requires the origin-destination trip counts during a given period.
+Yet, if automatic counting makes the embarkment (boarding) and disembarkment (alighting) counts in vehicles known, it often happens that pedestrian transfers between lines are harder to track, and require costly and invasive devices (e.g., facial recognition system) to be estimated.
+In this contribution, we propose a method, based on maximum entropy and involving an iterative fitting procedure, which estimates the passenger flow between origins and destinations solely based on embarkment and disembarkment flows.
+Moreover, this method is flexible enough to provide an adaptable framework in case additional data is known, such as attraction poles between certain points in the network, or percentages of transferring passengers between some lines.
+This method is tested on toy examples, as well as with the data of the public transportation network of the city of Lausanne provided by its Transportation Agency (tl), and gives convincing estimations of the transportation flow.
 \end{abstract}
@@
 \kwd{maximum entropy estimation}
+\kwd{iterative proportional fitting}
 \end{keyword}
@@
 
-This contribution adresses a straightforward, yet central question  in public transportation networks: given a network made of many bus lines, how can one estimate the real trips made by the travelers, on the sole basis of  the embarkment (boarding) counts and disembarkment (alighting) counts for each bus stop?
+This contribution addresses a straightforward, yet central question  in public transportation networks: given a network made of many train, bus, subway, or tram lines, how can one estimate the real trips made by the travelers, on the sole basis of the embarkment (boarding) counts and disembarkment (alighting) counts in each vehicle?
 Although estimating origin-destination flows is a much addressed issue in transportation modeling (see e.g \cite{bell1997transportation} \cite{hazelton2000estimation} \cite{ashok2002estimation} \cite{cui2006bus} and references therein), the specific problem addressed in this contribution seems, to the best of our knowledge, original.
 
-Pedestrian transfers of travelers between bus lines here constitute the missing information, whose principled evaluation require some methodological reflexion and experimentation.
-Section \ref{notforma} introduces the notations and the formalism, as well as the statement of the problem and the iterative solution method, which consist of three consecutive steps: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior (section \ref{maxenso}); an update of the marginal flows to avoid transfer overflow (section \ref{marginup}); and an update of the prior distribution  (section \ref{priorup}) by shrinking the components responsible for overflow.
+Pedestrian transfers of travelers between different lines here constitute the missing information, which planners traditionally try to estimate using census, or more recently with costly and invasive devices located at station, such as mobile phone tracking or facial recognition systems.
+In this article, we take a different approach, which is to give the best estimation of transportation trajectories without additional data, using the principle of maximum entropy.
+Section \ref{notforma} introduces the notations and the formalism, as well as the statement of the problem and the iterative solution method, which consist of three consecutive steps: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior; an update of the marginal flows to avoid transfer overflow; and an update of the prior distribution by shrinking the components responsible for overflow.
+This iterative procedure clearly evokes the EM algorithm (see e.g.
+\cite{dempster1977maximum}  \cite{bavaud2009information}), with the first step corresponding to the ``expectation step" and last two steps to the ``maximisation step".
+Moreover, this solution
+
+whose principled evaluation require some methodological reflexion and experimentation.
+: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior (section \ref{maxenso}); an update of the marginal flows to avoid transfer overflow (section \ref{marginup}); and an update of the prior distribution  (section \ref{priorup}) by shrinking the components responsible for overflow.
 
@@
 The initial margins were chosen to initially match $g_{st}$, namely $\alpha_s=g_{s \bullet}$ and $\beta_t=g_{\bullet t}$ for all stops.
+However, $g_{st}$ could be chosen more carefully in case of additional data.
+As a matter of fact, they represent \emph{prior attractions} between nodes in equation $\ref{Sol}$, before margin correction given by $\phi_s$ and $\psi_t$ and subsequent algorithm steps.
+An urban planner could chose to increase or decrease some of these values in order to take into account prior knowledge on commuting habits of inhabitants.
+
+\subsubsection{Embarkment, disembarkment constraints and hyperparameter $\theta$}
+\label{constraints}
+After a single iterative fitting step, the resulting transportation flow $\mathbf{N}=(n_{st})$ and transfer flow $\mathbf{Z}=(z_{ij})$ have little chance to fulfill constraints (\ref{entrer}) and (\ref{sortir}) defined by $\mathbf{a}$ and $\mathbf{b}$, and prior distributions $g_{st}$, $\alpha_s$ and $\beta_t$ must be corrected accordingly.
+
+When $z_{\bullet i} > a_i$, or $z_{i \bullet} > b_i$, the found solution typically predict that there are more passengers entering, respectively exiting, at a stop $i$ than the actual measured quantity.
+One could be tempted to correct prior distributions in order to have $z_{\bullet i} = a_i$, or $z_{i \bullet} = b_i$, on every problematic nodes $i$, but the latter solution would mean that all passengers entering (resp.
+exiting) the line at this stop are transiting, which seems unrealistic in most real life scenarios.
+
+We define here the hyperparameter $ \theta\in [0, 1)$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the \textbf{network} at each stop} (in other words, not transferring), that is
+\begin{equation}
+n_{s\bullet}\ge \theta a_s \qquad\qquad \qquad n_{\bullet t}\ge \theta b_t
+\end{equation}
+or 
+\begin{equation}
+	z_{\bullet s} \le (1 - \theta) a_s\qquad\qquad \qquad z_{t \bullet} \le  (1 - \theta) b_t
+\end{equation}
+and updates of prior distributions will be made accordingly.
+
+Note that, if additional data would be known, we could set a particular value of $\theta$ for every node, and differing for embarkments and disembarkments.
+However, without addition information, we will restrain to this simpler case.
+
+With
 
@@
 \label{marginup}
+
+By limiting the transfer flow
+
 Let us define the hyperparameter $ \theta\in [0, 1)$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the network at each stop}, that is $n_{s\bullet}\ge \theta a_s$ and $n_{\bullet t}\ge \theta b_t$.
@@
 
-		\State $\alpha_s \leftarrow a_s / a_\bullet, \; \forall s$ \Comment{{\footnotesize Initialize the network ingoing distribution}}
+		\State $\alpha_s \leftarrow g_{s\bullet}, \; \forall s$ \Comment{{\footnotesize Initialize the network ingoing distribution}}
 
-		\State $\beta_t \leftarrow b_t / b_\bullet, \; \forall t$ \Comment{{\footnotesize Initialize the network outgoing distribution}}
+		\State $\beta_t \leftarrow g_{\bullet t}, \; \forall t$ \Comment{{\footnotesize Initialize the network outgoing distribution}}
 
@@
 \\
-As for the hyperparameter study, found in Figure \ref{MTE_vs_theta}, we see that the optimal parameter seems to decreases as network size increases. However,
+As for the hyperparameter study, found in Figure \ref{MTE_vs_theta}, we see that the optimal parameter seems to decreases as network size increases, reaching value close to $0$ for $8$ line tours.
+However, this is not the only factor which helps to decide this parameter.
+As a matter of fact, it seems unrealistic, in some real life scenarios, that all passengers embarking or disembarking at junctions are all transiting, and this value will be kept at $0.1$ for our real life study.
 
@@
 	\includegraphics[width=0.98\textwidth]{fig/MTE_passengers_theta01.pdf}
-	\caption{\emph{Mean transportation error} (MTE) vs the number of randomly drawn passengers, in toy examples with number of tours $p$ varying from 2 to 8.
-The hyperparameter was set to $\theta=0.001$ and $10$ different drawn where performed for each data point.}
+	\caption{Mean transportation error (MTE) vs the number of randomly drawn passengers, in toy examples with numbers of tours $p$ varying from 2 to 8.
+The hyperparameter is set to $\theta=0.001$ and $10$ different draws are performed for each data point.}
 	\label{MTE_vs_passengers}
@@
 	\includegraphics[width=0.98\textwidth]{fig/MTE_tours.pdf}
-	\caption{\emph{Mean transportation error} (MTE), according to the number of tours $p \in \{2,3,4,5,6,7,8\}$ and resp.
-to the number of passengers into the network per line tour and the minimum proportion $\theta =0.1$ of passengers entering/leaving the network at each stop with a proportional number of passengers to the shortest paths of each line tour.}
+	\caption{Mean transportation error (MTE) vs hyperparameter $\theta$, in toy examples with numbers of tours $p$ varying from 2 to 8.
+The dot on each curve denotes the minimum.
+The number of drawn passengers is different for every network size and is set to $20 \cdot \vert P \vert$, in order to keep the randomness level constant.
+$10$ different draws are performed for each data point.}
 	\label{MTE_vs_theta}

### 2023-02-24 latex_files/applied_network_science_article/article_v0.tex
write article

@@
 In this article, we take a different approach, which is to give the best estimation of transportation trajectories without additional data, using the principle of maximum entropy.
-Section \ref{notforma} introduces the notations and the formalism, as well as the statement of the problem and the iterative solution method, which consist of three consecutive steps: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior; an update of the marginal flows to avoid transfer overflow; and an update of the prior distribution by shrinking the components responsible for overflow.
+Section \ref{notforma} introduces the notations and the formalism, as well as the statement of the problem and the iterative solution method, which consist of three consecutive steps: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior;  an update of the prior distribution by shrinking the components responsible for overflow; and an update of the marginal flows to avoid transfer overflow.
 This iterative procedure clearly evokes the EM algorithm (see e.g.
@@
 \end{equation}
-Also, $\mathbf{a}$ and $\mathbf{b}$ must be consistent, in the sense that $A_i\ge B_i$, where $A_i$ (respectively $B_i$) is the \emph{cumulated number of embarked 
+Also, $\mathbf{a}$ and $\mathbf{b}$ must be consistent, in the sense that $A_{B(i)}\ge B_i$, where $A_i$ (respectively $B_i$) is the \emph{cumulated number of embarked 
 (resp.
@@
 \begin{itemize}
-\item[(1)] we shall first suppose that the empirical margins $\alpha_s=f_{s\bullet}$ and $\beta_t=f_{\bullet t}$ are known.  
+\item[(1)] we first suppose that the empirical margins $\alpha_s=f_{s\bullet}$ and $\beta_t=f_{\bullet t}$ are known.  
 Then $f_{st}$ can be determined as the maximum entropy solution (section \ref{maxenso}), i.e.
 as the distribution closest to $g_{st}$ in the Kullback-Leibler divergence sense under the margin constraints, to be calibrated by an iterative fitting inner loop
-\item[(2)] then (section \ref{marginup}), the margins will be updated to $\tilde{\alpha}_s$ and $\tilde{\beta}_t$   by requiring a \emph{minimum proportion} $\theta\in [0,1)$ of passengers entering/leaving the network at each stop, as well as avoiding transfer overflow exceeding the embarking and disembarking counts at each stop
- \item[(3)] finally (section \ref{priorup}), the prior will be updated to $\tilde{g}_{st}$ by shrinking, if necessary, the priors $g_{st}$ associated to overflows. 
+\item[(2)] then (section \ref{priorup}), the prior is updated to $\tilde{g}_{st}$ by shrinking, if necessary, the priors $g_{st}$, thus avoiding transfer overflow exceeding the embarking and disembarking counts at each stop.
+Moreover, an hyperparameter $\theta \in [0, 1]$ is used at this stage in order to control the minimum proportion of passengers entering/leaving the network at each stop.
+\item[(3)] finally (section \ref{marginup}), the margins are updated to $\tilde{\alpha}_s$ and $\tilde{\beta}_t$.
 \end{itemize}
@@
 As a matter of fact, the first  step  (maximum entropy) exactly correspond to the ``expectation step" of the EM algorithm (see e.g.
-\cite{dempster1977maximum}  \cite{bavaud2009information}), but steps two and three, aiming at calibrating parameters $\alpha_s$, $\beta_t$ and $g_{st}$, do not follow the maximum likelihood rationale of the ``maximisation step".
+\cite{dempster1977maximum}  \cite{bavaud2009information}), but steps two and three, aiming at calibrating parameters $g_{st}$, $\alpha_s$ and $\beta_t$, do not follow the maximum likelihood rationale of the ``maximisation step".
 Pseudocode of the algorithm is shown with Algorithm \ref{algo1}.
@@
 whose plugging into (\ref{equationGG}) yields the intra-line edge flows $\mathbf{Y}=(y_{ij})$ and the transfer edge flows $\mathbf{Z}=(z_{ij})$.
-
+Only the latter is required for subsequent algorithm steps and can be computed with
+\begin{equation}
+	z_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} n_{st}
+\end{equation}
+where $I(.)$ denotes the 0/1 indicator function.
 \subsubsection{Initialization of the prior and the margins}
@@
 
-We define here the hyperparameter $ \theta\in [0, 1)$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the \textbf{network} at each stop} (in other words, not transferring), that is
+We define here the hyperparameter $ \theta\in [0, 1]$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the \textbf{network} at each stop} (in other words, not transferring), that is
 \begin{equation}
@@
 \end{equation}
-and updates of prior distributions will be made accordingly.
+and updates of distributions will be made accordingly.
 
@@
 However, without addition information, we will restrain to this simpler case.
-
-With
-
-\subsubsection{Updating the margin distributions}
-\label{marginup}
-
-By limiting the transfer flow
-
-Let us define the hyperparameter $ \theta\in [0, 1)$ as the \emph{minimum proportion of passengers (among $a_i$ and $b_i$) entering/leaving the network at each stop}, that is $n_{s\bullet}\ge \theta a_s$ and $n_{\bullet t}\ge \theta b_t$.
-Note that we could set a different hyperparameter for each node, and differing for embarkments and disembarkments, but without addition information, we will restrain to this simpler case.
-Identities (\ref{entrer}) and  (\ref{sortir}) then imply the inequalities
-\begin{displaymath}
-z_{\bullet s} \le (1 - \theta) a_s\qquad\qquad \qquad z_{t \bullet} \le  (1 - \theta) b_t
-\end{displaymath}
-the violation of which constitutes transfer overflow. Hence requiring a maximal transfer yet avoiding overflow can be granted with the following updating of margins
-\begin{equation}
-\label{alpha_beta_update}
-\widetilde{\alpha}_s = \frac{\min(\theta a_s, a_s - z_{\bullet s})}{\sum_{s'} \min(\theta a_{s'}, a_{s'} - z_{\bullet {s'}})}   \qquad \qquad \qquad
-	\widetilde{\beta}_t = \frac{\min(\theta b_t, b_t - z_{t \bullet})}{\sum_{t'} \min(\theta b_{t'}, b_{t'} - z_{{t'} \bullet})}  \enspace. 
-\end{equation}
 
@@
 where $\phi_s$ and $\psi_t$ are the values (\ref{Sol_LagMult}) obtained in the previous maximum entropy step.
+
+\subsubsection{Updating the margin distributions}
+\label{marginup}
+
+By construction, the corrected flow $\widetilde{n}_{st}$ found in (\ref{update_flow}) now respect embarkment and disembarkment constraints. We can compute the new transfer flow on edge with
+\begin{equation}
+\widetilde{z}_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} \widetilde{n}_{st}
+\end{equation}
+and, with (\ref{entrer}) and (\ref{sortir}), updating margin distributions is straightforward 
+\begin{equation}
+\label{alpha_beta_update}
+\widetilde{\alpha}_s = \frac{a_s - \widetilde{z}_{\bullet s}}{\sum_{s'} (a_{s'} - z_{\bullet {s'}})}   \qquad \qquad \qquad
+	\widetilde{\beta}_t = \frac{b_t - \widetilde{z}_{t \bullet}}{\sum_{t'} (b_{t'} - z_{{t'} \bullet})}  \enspace. 
+\end{equation}
 
@@
 
-		\State $\alpha_s \leftarrow \frac{\min(\theta a_s, a_s - z_{\bullet s})}{\sum_{s'} \min(\theta a_{s'}, a_{s'} - z_{\bullet {s'}})}, \; \forall s$ \Comment{{\footnotesize Update the network ingoing distribution}}
-
-		\State $\beta_t \leftarrow \frac{\min(\theta b_t, b_t - z_{t \bullet})}{\sum_{t'} \min(\theta b_{t'}, b_{t'} - z_{{t'} \bullet})}, \; \forall t$ \Comment{{\footnotesize Update the network outgoing distribution}}
-
 		\State $r_{ij} \leftarrow \max \left(1, \frac{z_{i \bullet}}{(1 - \theta)b_i}, \frac{z_{\bullet j}}{(1 - \theta)a_j} \right), \; \forall i,j$
@@
 
-		\State $\widetilde{g}_{st} \leftarrow \frac{\left( \frac{n_{st}}{\phi_s \psi_t \bar{r}_{st} + \epsilon} \right)}{\sum_{s',t'} \left( \frac{n_{s't'}}{\phi_{s'} \psi_{t'}\bar{r}_{s't'} + \epsilon} \right) + \epsilon}, \; \forall s,t$ \Comment{{\footnotesize Update the prior distribution}}
+		\State $\widetilde{n}_{st} \leftarrow \frac{n_{st}}{\bar{r}_{st}}, \; \forall s,t$
+
+		\State $\widetilde{g}_{st} \leftarrow \frac{\left( \frac{\widetilde{n}_{st}}{\phi_s \psi_t + \epsilon} \right)}{\sum_{s',t'} \left( \frac{\widetilde{n}_{s't'}}{\phi_{s'} \psi_{t'} + \epsilon} \right) + \epsilon}, \; \forall s,t$ \Comment{{\footnotesize Update the prior distribution}}
+
+		\State $\widetilde{z}_{ij} \leftarrow I((i,j) \in T)\sum_{st} \chi_{ij}^{st} \widetilde{n}_{st}, \; \forall i,j$
+
+		\State $\alpha_s \leftarrow \frac{a_s - \widetilde{z}_{\bullet s}}{\sum_{s'}(a_{s'} - \widetilde{z}_{\bullet {s'}})}, \; \forall s$ \Comment{{\footnotesize Update the network ingoing distribution}}
+
+		\State $\beta_t \leftarrow \frac{b_t - \widetilde{z}_{t \bullet}}{\sum_{t'} (b_{t'} - \widetilde{z}_{{t'} \bullet})}, \; \forall t$ \Comment{{\footnotesize Update the network outgoing distribution}}
 
@@
 $F(i)=i+1$.
-The initial prior is simply $g_{st}=c\; I(s<t)$ and captures solely the unidirectional nature of trips,  where $I(.)$ denotes the 0/1  indicator function and $c=\frac{1}{(l-1)(l-2)}$.
+The initial prior is simply $g_{st}=c\; I(s<t)$ and captures solely the unidirectional nature of trips, where $c=\frac{1}{(l-1)(l-2)}$.
 The margins of the empirical distribution $f_{st}$, as well as the total flow, are here known :
@@
 
-First, we shows some algorithm iterations on a toy example with $p=2$, where $100$ passengers where drawn uniformly across the $\vert P \vert = 20$ possible $st$-trips.
-Some iterations of the algorithm with $\theta=0.1$, along with MTE and MME errors, are shown in Figure \ref{iteration_plots}.
+First, we shows some algorithm iterations on a toy example with $p=2$, where $50$ passengers where drawn uniformly across the $\vert P \vert = 20$ possible $st$-trips.
+Some iterations of the algorithm with $\theta=0.001$, along with MTE and MME errors, are shown in Figure \ref{iteration_plots}.
 \begin{figure}[h]
 	\includegraphics[width=0.98\textwidth]{fig/iterations.pdf}
-	\caption{Real transportation flow (green arrows) obtained by randomly drawing $100$ passengers on the graph toy example with $p=2$ line tours, along with iterations $1, 2, 4, 7, 10$ of the algorithm with $\theta=0.1$.
+	\caption{Real transportation flow (green arrows) obtained by randomly drawing $50$ passengers on the graph toy example with $p=2$ line tours, along with iterations $1, 2, 4, 7, 15$ of the algorithm with $\theta=0.001$.
 MTE and MME errors are computed, and embarkment and disembarkment flows are represented respectively by the red and blue colors on nodes.}
@@
 \end{figure}
-On this small example, we can see that the algorithm quickly find an estimation giving small MME error, but still give an MTE of 0.252.
+On this small example, we can see that the algorithm quickly find an estimation giving small MME error, but still give an MTE of 0.256.
 This result is due to the fact that only $50$ passengers are drawn, giving a large deviation compared to the optimally found solution which maximize the entropy.
-Interestingly, we see that a better result is found on iteration , but margins constraints at this point are not perfectly respected yet.
 
@@
 \begin{figure}
-	\includegraphics[width=0.98\textwidth]{fig/MTE_passengers_theta01.pdf}
+	\includegraphics[width=0.98\textwidth]{fig/MTE_passengers.pdf}
 	\caption{Mean transportation error (MTE) vs the number of randomly drawn passengers, in toy examples with numbers of tours $p$ varying from 2 to 8.
@@
 
-after some preliminary, undocumented corrections (i.e.
-the components of $\mathbf{a}$ and $\mathbf{b}$ can be non-integer).
-It may also happen that, on some lines $\ell$,  raw data do not obey the necessary consistency condition  $a_\bullet^\ell=b_\bullet^\ell$ (where the latter quantities denote the total embarkments and disembarkments on line $\ell$), in which case we did rescale the embarking and disembarking line counts as
-\begin{displaymath}
-	\hat{a}_i=(1-\frac{a_\bullet^\ell-b_\bullet^\ell}{a_\bullet^\ell+b_\bullet^\ell})\, a_i 
-	\qquad\qquad\qquad
-	\hat{b}_i=(1+\frac{a_\bullet^\ell-b_\bullet^\ell}{a_\bullet^\ell+b_\bullet^\ell})\,  b_i 
-\end{displaymath}
-ensuring $\hat{a}_\bullet^\ell=\hat{b}_\bullet^\ell=2a_\bullet^\ell b_\bullet^\ell/(a_\bullet^\ell+b_\bullet^\ell)$.
-However, strongly unbalanced lines such that  $|a_\bullet^\ell-b_\bullet|^\ell/a_\bullet^\ell > 0.3$ or $|a_\bullet^\ell-b_\bullet^\ell|/b_\bullet^\ell > 0.3$  (which always turned out to be temporary lines with small  counts) were simply disregarded and line $\ell$ removed from the network.
-
-Also, the geometry of the network permits to derive the edge-trip incidence matrix $\bm{\chi}$ defined in (\ref{edgetrip}).
-
-\subsubsection{Dataset construction}
-This research presents a dataset derived from the public transportation agency in Lausanne (tl), Switzerland.
+\subsubsection{The dataset}
+The dataset derived from the public transportation agency in the city of Lausanne (tl), in Switzerland.
 Automatic passenger counters were used to collect data on the number of passengers entering and leaving buses and metros at each stop.
@@
 Lines with the most errors were removed, resulting in a final dataset of 35 transportation lines and 1216 stops.
-The threshold for invalid lines was set at a difference of 15\% between boarded and disembarked passenger counters.
+The threshold for invalid lines was set when a difference of between total embarkment and disembarked differs more than 15\% of their means.
+For the rest all other line, the correction found in appendix \ref{}
 
@@
 \section*{Appendix}
-Text for this section\ldots
+
+\subsection*{Embarkment and disembarkment corrections for lines}
+It may happen that, on some lines $\ell$, with stop indexed in order as $1, \ldots, l$, raw data do not obey the following necessary consistency conditions 
+\begin{align*}
+	a_l &= 0 \\
+	b_1 &= 0 \\
+	A_{i-1} &\geq B_i \qquad \forall i \in \{1, \ldots, l-1\} \\
+	A_{l} &= B_{l} \qquad \text{for terminal stop $l$}
+\end{align*}
+where $A_i$ (respectively $B_i$) is the cumulated number of embarked (resp.
+disembarked) passengers on the line.
+The first two conditions are easy to correct, and we will assume that they are valid, but the last two require a recursive procedure.
+
+We first identify all stops $i$ where $A_{i - 1} < B_i$ and store them, along with stop $1$ and $l$, in a order set $\{1, s_1, \ldots, s_r, l\}$.
+
+$a_\bullet^\ell=b_\bullet^\ell$ (where the latter quantities denote the total embarkments and disembarkments on line $\ell$), in which case we did rescale the embarking and disembarking line counts as
+\begin{displaymath}
+	\hat{a}_i=(1-\frac{a_\bullet^\ell-b_\bullet^\ell}{a_\bullet^\ell+b_\bullet^\ell})\, a_i 
+	\qquad\qquad\qquad
+	\hat{b}_i=(1+\frac{a_\bullet^\ell-b_\bullet^\ell}{a_\bullet^\ell+b_\bullet^\ell})\,  b_i 
+\end{displaymath}
+ensuring $\hat{a}_\bullet^\ell=\hat{b}_\bullet^\ell=2a_\bullet^\ell b_\bullet^\ell/(a_\bullet^\ell+b_\bullet^\ell)$. This process is applied for every
 

### 2023-02-24 2_compute_OD_matrix.R

@@ -20,24 +20,24 @@ source("local_functions.R")
 
 # --- Test 6789
 
-# # The folder containing pre-processed data
-# data_folder = "multilines_data/preprocessed_data/test_mc_6789"
-# # Output folder for results
-# out_folder = "results/test_mc_6789"
+# The folder containing pre-processed data
+data_folder = "multilines_data/preprocessed_data/test_mc_6789"
+# Output folder for results
+out_folder = "results/test_mc_6789"
 
 # --- all lines
 
-# The folder containing pre-processed data
-data_folder = "multilines_data/preprocessed_data/all_lines"
-# Output folder for results
-out_folder = "results/all_lines"
+# # The folder containing pre-processed data
+# data_folder = "multilines_data/preprocessed_data/all_lines"
+# # Output folder for results
+# out_folder = "results/all_lines"
 
 # Conv threshold for iterative fitting
 conv_thres_if = 1e-5
 # Conv threshold
 conv_thres_algo = 1e-7
 # proportional limit 
-min_p_ntwk = 0.1
+min_p_ntwk = 0
 # epsilon
 epsilon = 1e-40
 # max iterations

### 2023-02-24 4.2_plot_toy_iterations.R

@@ -34,19 +34,19 @@ conv_thres_if = 1e-6
 # Conv threshold
 conv_thres_algo = 1e-10
 # proportional limit 
-min_p_ntwk = 0.01
+min_p_ntwk = 0.001
 # epsilon
 epsilon = 1e-40
 # max iterations
-max_it = 50
+max_it = 15
 # max iterations for iterative fitting
 max_it_if = 1000
 
 # Seed
-set.seed(3)
+set.seed(30)
 
 # It list
-it_list = c(1, 2, 4, 7, 50)
+it_list = c(1, 2, 4, 7, 15)
 
 #--------------------------------
 # Code
@@ -128,7 +128,7 @@ e_flow_size = 3
 v_size = 20
 v_label_size = 0.00001
 
-pdf(paste0(result_folder, "/iterations_test.pdf"))
+pdf(paste0(result_folder, "/iterations.pdf"))
 par(mfrow=c(2,3))
 plot_flow_graph(adj, n_real, layout, v_names=stop_names, main="Real", 
                 e_line_size=e_line_size, e_flow_size=e_flow_size, 

### 2023-02-25 latex_files/applied_network_science_article/article_v0.tex
annexe written

@@
 This iterative procedure clearly evokes the EM algorithm (see e.g.
-\cite{dempster1977maximum}  \cite{bavaud2009information}), with the first step corresponding to the ``expectation step" and last two steps to the ``maximisation step".
-Moreover, this solution
+\cite{dempster1977maximum}  \cite{bavaud2009information}), with the first step corresponding to the ``expectation step" and last two steps to the ``maximization step".
+The fist step only is required for solving the single line case (section \ref{Single line}), naturally much simpler but  yet  not trivial, and exhibiting a disembarking probability independent of the embarking stop (Markov property).
+Moreover, this solution offer some flexibility, as it is possible to set a prior distribution taking into account attraction or repulsion poles among stops, and to fix hyperparameters limiting the number of transfers between lines.
 
-whose principled evaluation require some methodological reflexion and experimentation.
-: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior (section \ref{maxenso}); an update of the marginal flows to avoid transfer overflow (section \ref{marginup}); and an update of the prior distribution  (section \ref{priorup}) by shrinking the components responsible for overflow.
-
-The fist step only is required for solving the single line case (section \ref{Single line}), naturally much simpler but  yet  not trivial, and exhibiting a disembarking probability independent of the embarking stop (Markov property).
-
-Cases studies are presented in section \ref{casestudies} *** à poursuivre ***
+In section \ref{casestudies}, we test the propose method on toy examples and the data of the public transportation network of the city of Lausanne.
+Toy examples offer some kind of validation, as a transportation flow can be set on toy networks and compared to the solution given by our algorithm.
+The real case scenario with the data from the public transportation network of the city of Lausanne shows that this method is applicable on a large transportation network and can give pertinent insights about travelers habits.
 
@@
 \subsection*{Embarkment and disembarkment corrections for lines}
-It may happen that, on some lines $\ell$, with stop indexed in order as $1, \ldots, l$, raw data do not obey the following necessary consistency conditions 
+It may happen that, on some lines $\ell$, with stops indexed in order as $1, \ldots, l$, raw data do not obey the following necessary consistency conditions 
 \begin{align*}
-	a_l &= 0 \\
-	b_1 &= 0 \\
-	A_{i-1} &\geq B_i \qquad \forall i \in \{1, \ldots, l-1\} \\
+	a_l = 0&, \; b_1 = 0 \\
+	A_{(i-1)} &\geq B_i \qquad \forall i \in \{1, \ldots, l-1\} \\
 	A_{l} &= B_{l} \qquad \text{for terminal stop $l$}
@@
 where $A_i$ (respectively $B_i$) is the cumulated number of embarked (resp.
-disembarked) passengers on the line.
-The first two conditions are easy to correct, and we will assume that they are valid, but the last two require a recursive procedure.
+disembarked) passengers on the line at stop $i$.
+The first condition is easy to correct (by setting both quantities to $0$), and we will assume that they are valid.
+The last two require an iterative procedure, explained here.
 
-We first identify all stops $i$ where $A_{i - 1} < B_i$ and store them, along with stop $1$ and $l$, in a order set $\{1, s_1, \ldots, s_r, l\}$.
-
-$a_\bullet^\ell=b_\bullet^\ell$ (where the latter quantities denote the total embarkments and disembarkments on line $\ell$), in which case we did rescale the embarking and disembarking line counts as
-\begin{displaymath}
-	\hat{a}_i=(1-\frac{a_\bullet^\ell-b_\bullet^\ell}{a_\bullet^\ell+b_\bullet^\ell})\, a_i 
-	\qquad\qquad\qquad
-	\hat{b}_i=(1+\frac{a_\bullet^\ell-b_\bullet^\ell}{a_\bullet^\ell+b_\bullet^\ell})\,  b_i 
-\end{displaymath}
-ensuring $\hat{a}_\bullet^\ell=\hat{b}_\bullet^\ell=2a_\bullet^\ell b_\bullet^\ell/(a_\bullet^\ell+b_\bullet^\ell)$. This process is applied for every
+We first identify all stops $i$ where $A_{i - 1} < B_i$ and store them, along with stop $1$ and $l$, in a order set $S$.
+Let $\text{Prev}(i)$ be the item right before $i$ in the set $S$.
+For all $i \in S \setminus \{1\}$, we do :
+\begin{align*}
+		\hat{a}_j&=\left( 1-\frac{(A_{(i-1)} - A_{(\text{Prev}(i)-1)}) - (B_{i} -  B_{\text{Prev}(i)})}{(A_{(i-1)} - A_{(\text{Prev}(i)-1)}) + (B_{i} -  B_{\text{Prev}(i)})}\right)\, a_j  \qquad \forall j \in \{ \text{Prev}(i), \ldots, i-1 \} \\
+	\hat{b}_j&=\left( 1+\frac{(A_{(i-1)} - A_{(\text{Prev}(i)-1)}) - (B_{i} -  B_{\text{Prev}(i)})}{(A_{(i-1)} - A_{(\text{Prev}(i)-1)}) + (B_{i} -  B_{\text{Prev}(i)})}\right)\, b_j  \qquad \forall j \in \{ \text{Prev}(i) + 1, \ldots, i \}
+\end{align*}
+Before the last step, all conditions $A_{(i-1)} \geq  B_{i}$, should be respected except for node $l$.
+The last step ensures that $A_{l} = B_{l} $, however, as it can sometimes lowers the embarkment count and increases the disembarkment count on previous nodes, some new nodes can now violate $A_{(i-1)} \geq B_i$.
+This is why this procedure should be iterated until all stops on the line verify consistency conditions.
 

### 2023-02-25 latex_files/applied_network_science_article/article_v0.tex
write article, transfer computations

@@
 \end{align} 
-where $z_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} n_{st}$ is the flow on transfer edges $T$. Both errors can be interpreted as a weighted mean deviation percentage.
+where $z_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} n_{st}$ is the flow on transfer edges $T$. Both errors can be interpreted as a weighted mean of a percentage of error.
 
 Note that, by construction, the MME should be null when the algorithm converges.
-However, it can be informative to track down this error along iterations and, in some practical cases where the network is large, the algorithm convergence criterion is reached without having margin constraints perfectly respected.
+However, it can be informative to track down this error along iterations and, in some practical cases where margin constraints are impossible to fulfill, the algorithm convergence criterion is reached with MME $> 0$.
 
@@
 In fact, all MTE should converge to $0$ eventually, however this rate of convergence seems to be low.
-\\
-As for the hyperparameter study, found in Figure \ref{MTE_vs_theta}, we see that the optimal parameter seems to decreases as network size increases, reaching value close to $0$ for $8$ line tours.
-However, this is not the only factor which helps to decide this parameter.
-As a matter of fact, it seems unrealistic, in some real life scenarios, that all passengers embarking or disembarking at junctions are all transiting, and this value will be kept at $0.1$ for our real life study.
-
-\begin{figure}
+\begin{figure}[h]
 	\includegraphics[width=0.98\textwidth]{fig/MTE_passengers.pdf}
@@
 \end{figure}
-
-\begin{figure}
+As for the hyperparameter study, found in Figure \ref{MTE_vs_theta}, we see that the optimal parameter seems to decreases as network size increases, reaching value close to $0$ for $8$ line tours.
+However, this is not the only factor which helps to decide this parameter.
+As a matter of fact, it seems unrealistic, in some real life scenarios, that all passengers embarking or disembarking at junctions are all transiting, and this value will be kept at $0.1$ for our real life study.
+\begin{figure}[h]
 	\includegraphics[width=0.98\textwidth]{fig/MTE_tours.pdf}
@@
 
-\subsubsection{The dataset}
+\subsubsection{The dataset and algorithm setup}
+\label{data_algo}
 The dataset derived from the public transportation agency in the city of Lausanne (tl), in Switzerland.
@@
 The threshold for invalid lines was set when a difference of between total embarkment and disembarked differs more than 15\% of their means.
-For the rest all other line, the correction found in appendix \ref{}
+For the rest all other lines, the correction found in appendix was applied to ensure consistency conditions.
+
+The matrix $P$, containing valid $st$-trips, was built using same condition found in toy example with an additional one.
+If the estimated pedestrian time (using Open Trip Planner (REF)) between two stops $s$ and $t$ was shorter than the estimated time by taking the public transportation network, this $st$-trip was also removed from $P$.
+The graph structure, consisting in matrices $\bm{\chi}$ and $P$, was constructed beforehand testing the method.
+
+The algorithm was run with hyperparameter $\theta = 0.1$.
+Tracking the error on margins (MME) showed a rapid decreases, as seen in Figure \ref{MME_vs_iteration}, but the (conservative) convergence criterion (chosen as when $\sum_{st} \vert f^{prev}_{st} - f_{st} \vert< 10^{-6} $) was reach after 316 iterations, which corresponds to approximately one hour of computing time with a single thread.
+
+\begin{figure}[h]
+	\includegraphics[width=0.98\textwidth]{fig/MME_vs_iteration.pdf}
+	\caption{Mean margin error (MME) vs iteration for the algorithm run with the public transportation network data of city of Lausanne (35 lines, 1216 stops).}
+	\label{MME_vs_iteration}
+\end{figure}
+
+\subsubsection{Results}
+\label{lausanne_results}
 
@@
 \subsection*{Embarkment and disembarkment corrections for lines}
+\label{flow_correction}
 It may happen that, on some lines $\ell$, with stops indexed in order as $1, \ldots, l$, raw data do not obey the following necessary consistency conditions 
@@
 \end{align*}
-Before the last step, all conditions $A_{(i-1)} \geq  B_{i}$, should be respected except for node $l$.
-The last step ensures that $A_{l} = B_{l} $, however, as it can sometimes lowers the embarkment count and increases the disembarkment count on previous nodes, some new nodes can now violate $A_{(i-1)} \geq B_i$.
-This is why this procedure should be iterated until all stops on the line verify consistency conditions.
+Before the last step, all conditions $A_{(i-1)} \geq  B_{i}$ should be respected except for node $l$.
+The last step ensures that $A_{l} = B_{l}$, however, as it can sometimes lower the embarkment count and increase the disembarkment count on previous nodes, some new nodes can now violate $A_{(i-1)} \geq B_i$.
+This is why this procedure must be iterated until all stops on the line verify consistency conditions.
 
@@
 
-\section*{Funding}%% if any
-Text for this section\ldots
-
-\section*{Abbreviations}%% if any
-Text for this section\ldots
-
-\section*{Availability of data and materials}%% if any
-Text for this section\ldots
-
-\section*{Ethics approval and consent to participate}%% if any
-Text for this section\ldots
-
 \section*{Competing interests}
 The authors declare that they have no competing interests.
-
-\section*{Consent for publication}%% if any
-Text for this section\ldots
-
-\section*{Authors' contributions}
-Text for this section \ldots
-
-\section*{Authors' information}%% if any
-Text for this section\ldots
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  Bmc_mathpys.bst  will be used to                       %%
-%%  create a .BBL file for submission.                     %%
-%%  After submission of the .TEX file,                     %%
-%%  you will be prompted to submit your .BBL file.         %%
-%%                                                         %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \bibliography{article_tl.bib}      % Bibliography file (usually '*.bib' )
-% for author-year bibliography (bmc-mathphys or spbasic)
-% a) write to bib file (bmc-mathphys only)
-% @settings{label, options="nameyear"}
-% b) uncomment next line
-%\nocite{label}
-
-% or include bibliography directly:
-% \begin{thebibliography}
-% \bibitem{b1}
-% \end{thebibliography}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Figures                       %%
-%%                               %%
-%% NB: this is for captions and  %%
-%% Titles. All graphics must be  %%
-%% submitted separately and NOT  %%
-%% included in the Tex document  %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%
-%% Do not use \listoffigures as most will included as separate files
-
-\section*{Figures}
-  \begin{figure}[h!]
-  \caption{Sample figure title}
-\end{figure}
-
-\begin{figure}[h!]
-  \caption{Sample figure title}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Tables                        %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Use of \listoftables is discouraged.
-%%
-\section*{Tables}
-\begin{table}[h!]
-\caption{Sample table title. This is where the description of the table should go}
-  \begin{tabular}{cccc}
-    \hline
-    & B1  &B2   & B3\\ \hline
-    A1 & 0.1 & 0.2 & 0.3\\
-    A2 & ... & ..  & .\\
-    A3 & ..  & .   & .\\ \hline
-  \end{tabular}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Additional Files              %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Additional Files}
-  \subsection*{Additional file 1 --- Sample additional file title}
-    Additional file descriptions text (including details of how to
-    view the file, if it is in a non-standard format or the file extension).  This might
-    refer to a multi-page table or a figure.
-
-  \subsection*{Additional file 2 --- Sample additional file title}
-    Additional file descriptions text.
 

### 2023-02-25 3.1_iteration_graph.R

@@ -1,0 +1,37 @@
+#-------------------------------------------------------------------------------
+#
+# Plot the graphic for MME error vs iterations
+#
+#-------------------------------------------------------------------------------
+
+#--------------------------------
+# Head
+#--------------------------------
+
+# Load packages 
+library("tidyverse")
+
+#--------------------------------
+# Parameters
+#--------------------------------
+
+# The folder containing results
+results_folder = "results/all_lines2"
+
+#--------------------------------
+# Process
+#--------------------------------
+
+# Load files 
+iterations_df = read_csv(paste0(results_folder, "/iterations_df.csv"))
+iterations_df = iterations_df %>%
+  mutate(mean_error = (error_in + error_out)/2) %>%
+  filter(it_algo %in% 2:15)
+  
+iterations_df %>%
+  ggplot() +
+  geom_line(aes(x=it_algo, y=mean_error), color="blue") +
+  xlab("Iteration") +
+  ylab("MME") 
+
+ggsave(paste0(results_folder, "/MME_vs_iteration.pdf"), width = 15, height = 5)

### 2023-02-25 3.3_map_results_6789.R

@@ -35,6 +35,10 @@ results_folder = "results/test_mc_6789"
 
 # --- Reading files
 
+# Reading graph data
+x_res = compute_x_from_n(n_mat, edge_ref, sp_ref, sp_edge_link)
+node_in_btw = colSums(x_res$x_btw)
+node_out_btw = rowSums(x_res$x_btw)
 # Reading results data 
 line_res = read.csv(paste0(results_folder, "/line_res.csv")) 
 n_mat = as.matrix(read.csv(paste0(results_folder, "/n_mat.csv"), header=F))

### 2023-02-27 5.5_visu_AFC.R
up down

@@ -23,10 +23,10 @@ library(RColorBrewer)
 library(leaflet)
 
 # Load functions
-source("local_functions.R")
+# source("local_functions.R")
 
 # The folder containing pre-processed data
-data_folder = "multilines_data/preprocessed_data/all_lines"
+# data_folder = "multilines_data/preprocessed_data/all_lines"
 
 # The folder containing results
 results_folder = "results/all_lines2"
@@ -34,6 +34,9 @@ results_folder = "results/all_lines2"
 #--------------------------------
 # Process
 #--------------------------------
+
+# up or down ?
+up_down = "up"
 
 # --- Reading files
 
@@ -60,7 +63,13 @@ rownames(n_mat) = stop_names
 
 res.ca <- CA(n_mat, graph = FALSE)
 
-CA_dim = res.ca$row$coord[,1:2]
+if (up_down == "up") {
+  CA_dim = res.ca$row$coord[,1:2]
+}
+if (up_down == "down") {
+  CA_dim = res.ca$col$coord[,1:2]
+}
+
 colnames(CA_dim) <- c("dim1","dim2")
 CA_dim = as.data.frame(CA_dim)
 CA_dim$stop_names = rownames(CA_dim)
@@ -116,6 +125,14 @@ CA_shp = right_join(stops_shp, CA_dim, by = "stop_names")
 lines_shp_sample = right_join(lines_shp, CA_dim, by = "stop_names")
 
 
+if (up_down == "up") {
+  CA_shp$CA_radius = CA_shp$montees
+}
+
+if (up_down == "down") {
+  CA_shp$CA_radius = CA_shp$descentes
+}
+
 
 # Mapping results
 
@@ -130,7 +147,8 @@ CA_map2 = leaflet(CA_shp) %>%
                color = "#eeeeee") %>%
   addCircles(
     group = "dim1",
-    radius = ((CA_shp$montees+CA_shp$descentes)/max(CA_shp$montees+CA_shp$descentes)*100000)^0.45, # coefficient de correction avec un minimum de largeur de 1
+    # radius = ((CA_shp$montees)/max(CA_shp$montees)*100000)^0.45, # coefficient de correction avec un minimum de largeur de 1
+    radius = ((CA_shp$CA_radius)/max(CA_shp$CA_radius)*100000)^0.45, # coefficient de correction avec un minimum de largeur de 1
     fillColor = "transparent",
     opacity = 1,
     color = ~colorPal1(dim1),
[…]

### 2023-02-27 latex_files/applied_network_science_article/article_v0.tex
new version

@@
 \usepackage{amssymb}
+\usepackage{hyperref}
 
@@
 \end{equation}
-until convergence to $\phi_s$ and $\psi_t$ obeying (\ref{Sol_LagMult}).
+until convergence to $\phi_s$ and $\psi_t$ obeying (\ref{Sol_LagMult}).\footnote{with possibly a small quantity $\epsilon > 0$ added on null components of $g_{st}$.}
 
@@
 	\includegraphics[width=0.98\textwidth]{fig/MTE_passengers.pdf}
-	\caption{Mean transportation error (MTE) vs the number of randomly drawn passengers, in toy examples with numbers of tours $p$ varying from 2 to 8.
+	\caption{Mean Transportation Error (MTE) vs the number of randomly drawn passengers, in toy examples with numbers of tours $p$ varying from 2 to 8.
 The hyperparameter is set to $\theta=0.001$ and $10$ different draws are performed for each data point.}
@@
 	\includegraphics[width=0.98\textwidth]{fig/MTE_tours.pdf}
-	\caption{Mean transportation error (MTE) vs hyperparameter $\theta$, in toy examples with numbers of tours $p$ varying from 2 to 8.
+	\caption{Mean Transportation Error (MTE) vs hyperparameter $\theta$, in toy examples with numbers of tours $p$ varying from 2 to 8.
 The dot on each curve denotes the minimum.
@@
 The matrix $P$, containing valid $st$-trips, was built using same condition found in toy example with an additional one.
-If the estimated pedestrian time (using Open Trip Planner (REF)) between two stops $s$ and $t$ was shorter than the estimated time by taking the public transportation network, this $st$-trip was also removed from $P$.
+If the estimated pedestrian time (using Open Trip Planner\footnote{\url{https://docs.opentripplanner.org/en/v2.2.0/}} \cite{MalcolmMorgan2019}) between two stops $s$ and $t$ was shorter than the estimated time by taking the public transportation network, this $st$-trip was also removed from $P$.
 The graph structure, consisting in matrices $\bm{\chi}$ and $P$, was constructed beforehand testing the method.
@@
 The algorithm was run with hyperparameter $\theta = 0.1$.
-Tracking the error on margins (MME) showed a rapid decreases, as seen in Figure \ref{MME_vs_iteration}, but the (conservative) convergence criterion (chosen as when $\sum_{st} \vert f^{prev}_{st} - f_{st} \vert< 10^{-6} $) was reach after 316 iterations, which corresponds to approximately one hour of computing time with a single thread.
+Tracking the Mean Margin Error (MME) showed a rapid decreases but the (conservative) convergence criterion (chosen as when $\sum_{st} \vert f^{prev}_{st} - f_{st} \vert< 10^{-6} $) was reach after 316 iterations, which corresponds to approximately one hour of computing time with a single thread.
+
+\subsubsection{Results}
+\label{real_results}
+
+As the result of our algorithm consists in the $(1216 \times 1216)$ matrix $\mathbf{N} = (n_{st})$, it is difficult to present them succinctly.
+We chose here to focus on two aspects: (1) displaying the localization of the predicted transfers; and (2) mapping the origin and destination profiles of stops with the help of Correspondence Analysis (CA) \cite{benzecri1977histoire}.
+
+The first and second factorial dimensions of the destination and origin profiles of stops, obtained with CA applied on $\mathbf{N}$ , can be found on Figure \ref{afc}.
+We can see that origin and destination of each stop are quite similar when displayed on the first two factorial axis.
 
 \begin{figure}[h]
-	\includegraphics[width=0.98\textwidth]{fig/MME_vs_iteration.pdf}
-	\caption{Mean margin error (MME) vs iteration for the algorithm run with the public transportation network data of city of Lausanne (35 lines, 1216 stops).}
-	\label{MME_vs_iteration}
+	\includegraphics[width=0.46\textwidth]{fig/afc_in_dim1.png}
+	\includegraphics[width=0.46\textwidth]{fig/afc_in_dim2.png} \\
+	\includegraphics[width=0.46\textwidth]{fig/afc_out_dim1.png}
+	\includegraphics[width=0.46\textwidth]{fig/afc_out_dim2.png}
+	\caption{Color representations (red positive value, blue negative values) of factorial coordinates of stops, after Correspondence analysis (CA) performed on matrix $\mathbf{N}$.
+Top figures map stops as ``origin profiles" (``where passengers are going"), bottom figures map stops as ``destination profile" (``where passengers are coming from").
+Left column is the first factorial dimension, right column is the second factorial dimension.
+Ring sizes correspond to the number of embarkments (top) and disembarkment (bottom).}
+	\label{afc}
 \end{figure}

### 2023-02-27 5.6_visu_nodes.R
Visu transfer nodes

@@ -1,6 +1,6 @@
 #-------------------------------------------------------------------------------
 #-------------------------------------------------------------------------------
-# Transfert visualisation
+# Transfer visualization
 #-------------------------------------------------------------------------------
 #-------------------------------------------------------------------------------
 # - Map important nodes
@@ -12,20 +12,9 @@
 
 # Libraries
 #--------------------------------
-# library(dplyr)
-# library(ggplot2)
-# library(gridExtra)
-# library(reshape2)
-# library(FactoMineR)
-# library(sf)
-# library(RColorBrewer)
-# library(leaflet)
-
-# # Load functions
-# source("local_functions.R")
-
-# # The folder containing pre-processed data
-# data_folder = "multilines_data/preprocessed_data/all_lines"
+library(dplyr)
+library(leaflet)
+library(BAMMtools) # Jenks library
 
 # The folder containing results
 results_folder = "results/all_lines2"
@@ -35,9 +24,71 @@ results_folder = "results/all_lines2"
 #--------------------------------
 
 # --- Reading files
+transfer_df = read.csv(paste0(results_folder, "/transfer_df.csv"))
+coord = read.csv2("utilities/data_processing/stop_frequentation2.csv")
 
-# Reading results data 
-line_res = read.csv(paste0(results_folder, "/line_res.csv")) 
-n_mat = as.matrix(read.csv(paste0(results_folder, "/n_mat.csv"), header=F))
-ped_time = read.csv(as.matrix("multilines_data/formatted_data/all_lines/ped_time.csv"), header=F)
-stop_names = line_res$stop_names
+# --- Adaptation of coord
+coord$id = paste0("S", coord$code_ligne_theo,"_", coord$direction_voy_theo, "_",
+                  coord$code_arret_theo)
+coord = coord[,c(11, 9, 10)]
+
+# Join i
+transfer_df <- left_join(
+  transfer_df,
+  coord,
+  by = c("tag_i" = "id")
+  )
+colnames(transfer_df)[12] = "lon_i"
+colnames(transfer_df)[13] = "lat_i"
+
+# Join j
+transfer_df <- left_join(
+  transfer_df,
+  coord,
+  by = c("tag_j" = "id")
+)
+colnames(transfer_df)[14] = "lon_j"
+colnames(transfer_df)[15] = "lat_j"
+
[…]

### 2023-02-27 5.6_visu_nodes.R
Visu transfer nodes

@@ -1,6 +1,6 @@
 #-------------------------------------------------------------------------------
 #-------------------------------------------------------------------------------
-# Transfer visualization
+# Transfer visualisation
 #-------------------------------------------------------------------------------
 #-------------------------------------------------------------------------------
 # - Map important nodes

### 2023-02-27 3.2_extract_transfers.R
compute aggregate tansfer counts

@@ -62,7 +62,14 @@ while(sum(x_to_empty) != 0){
   x_to_empty[indices_stop[1], indices_stop[2]] = 0
 }
 
+aggregate_transfer_df = transfer_df %>% 
+  group_by(stop_i) %>% 
+  summarise(sum(number_of_transfers))
+
 write.table(transfer_df, paste0(results_folder, "/transfer_df.csv"), sep=",",
+            row.names=F)
+write.table(aggregate_transfer_df, 
+            paste0(results_folder, "/aggregate_transfers_df.csv"), sep=",",
             row.names=F)
 
 

### 2023-02-27 latex_files/applied_network_science_article/article_v0.tex
write article

@@
 
-In section \ref{casestudies}, we test the propose method on toy examples and the data of the public transportation network of the city of Lausanne.
+In section \ref{casestudies}, we test the propose method on toy examples and with the data of the public transportation network of the city of Lausanne.
 Toy examples offer some kind of validation, as a transportation flow can be set on toy networks and compared to the solution given by our algorithm.
@@
 As explained in section \ref{Lines and junctions}, transfers can only occur at junctions, that is $z_{ij}>0$ implies $(i, j) \in T$.
-In particular,  $z_{ii}=0$ : no traveller is supposed to disembark and re-embark later at the same stop.
+In particular,  $z_{ii}=0$ : no traveler is supposed to disembark and re-embark later at the same stop.
 
@@
 \emph{Automatic passenger counters} measure the number of passengers entering and leaving lines at each stop [Boyle, 1998], that is $\mathbf{a}$ and $\mathbf{b}$, which provide the basic raw data of the present study, kindly provided by the Lausanne Transportation Agency (tl) for the case study in section \ref{real_data}.
-We will suppose here that this data obeys the necessary consistency condition $a_\bullet^\ell=b_\bullet^\ell$ (where the latter quantities denote the total embarkments and disembarkments on line $\ell$), even if, in real case studies, a rescaling must usually be performed to balance in and out-flows on each lines.
+We will suppose here that this data obeys the necessary consistency conditions for embarkment and disembarkment counts which can be found in the Appendix, even if, in real case studies, a rescaling must usually be performed to balance in and out-flows on each lines (also given in the Appendix).
 
@@
 As the result of our algorithm consists in the $(1216 \times 1216)$ matrix $\mathbf{N} = (n_{st})$, it is difficult to present them succinctly.
-We chose here to focus on two aspects: (1) displaying the localization of the predicted transfers; and (2) mapping the origin and destination profiles of stops with the help of Correspondence Analysis (CA) \cite{benzecri1977histoire}.
+We chose here to focus on two aspects: (1) mapping the origin and destination profiles of stops with the help of Correspondence Analysis (CA) \cite{benzecri1977histoire}; and (2) mapping the predicted transfers between lines.
 
 The first and second factorial dimensions of the destination and origin profiles of stops, obtained with CA applied on $\mathbf{N}$ , can be found on Figure \ref{afc}.
-We can see that origin and destination of each stop are quite similar when displayed on the first two factorial axis.
+We can see that origin and destination of each stop are quite similar when displayed on the first two factorial axes.
+This result can be explained if we understand that, in fact, origin and destination profiles are in fact kinda symmetrical: at the beginning of a line, stops have several choices of destinations but very few origins ``point" to them, and this is the opposite for stops at the end of the line.
+Intermediary stops are generally located at the center of the city, and have a similar and rich origin and destination profiles.
+Concerning the dimensions, we can see that the first one divided the city stops between west and east, a known dichotomy in the city of Lausanne in public transportation usage: the east of the city is less well deserved and often use cars, while the west intensively use the public transportation network.
+The second dimension is harder to interpret, but seems to highlight the radial structure of the transportation network in Lausanne.
+
+Figure \ref{transfers} maps the geographically aggregate transfer flows $z_{ij}$ between lines.
+As expected, most transfers occur in the city center, with top 5 largest hubs located at St-François (2,132,597 transfers), Lausanne-Gare (2,035,543 transfers), Bel-Air (1,908,168 transfers), Ours (1,057,282 transfers), and Chauderon (969,789 transfers).
+When taken individually, four of the top 5 transfer counts is predicted between line $1$ and line $72$ occurring at the train station (Lausanne-Gare).
+Line 72 is actually a subway line connecting north and south region of the city, and is by far the most used line in the network.
+Line $1$ is also a very frequented line and operated toward the western region of the city.
+It is not surprising to see most transfer occurring between these two lines, as they permit to connect the region with an high density of public transportation network users (the western part of Lausanne), to the main artery of the network (line 72).
 
@@
 	\includegraphics[width=0.46\textwidth]{fig/afc_out_dim2.png}
-	\caption{Color representations (red positive value, blue negative values) of factorial coordinates of stops, after Correspondence analysis (CA) performed on matrix $\mathbf{N}$.
-Top figures map stops as ``origin profiles" (``where passengers are going"), bottom figures map stops as ``destination profile" (``where passengers are coming from").
+	\caption{Color representations (positive values in red, negative values in blue) of the first two factorial coordinates of stops, after Correspondence analysis (CA) performed on transportation matrix $\mathbf{N}$.
+Top row maps stops as ``origin profiles" (``where passengers are going"), bottom row maps stops as ``destination profile" (``where passengers are coming from").
 Left column is the first factorial dimension, right column is the second factorial dimension.
@@
 	\label{afc}
+\end{figure}
+
+\begin{figure}[h]
+	\includegraphics[width=0.97\textwidth]{fig/transfers.png}
+	\caption{Location and size (ring size and color) of estimated transfer flows $z_{ij}$ between lines.}
+	\label{transfers}
 \end{figure}

### 2023-02-27 5.4_visu_tl_nodes.R
visu AC stops aggregation

@@ -31,6 +31,7 @@ coord = read.csv2("utilities/data_processing/stop_frequentation2.csv")
 coord$id = paste0("S", coord$code_ligne_theo,"_", coord$direction_voy_theo, "_",
                   coord$code_arret_theo)
 coord = coord[,c(11, 9, 10)]
+coord$id <- gsub("\\s+", "", coord$id)
 
 # Join i
 transfer_df <- left_join(

### 2023-02-27 5.5_visu_tl_agg_nodes.R

@@ -1,0 +1,64 @@
+setwd("/Users/rloup/Documents/r_projects/tl_study")
+#-------------------------------------------------------------------------------
+#-------------------------------------------------------------------------------
+# Transfer visualisation
+#-------------------------------------------------------------------------------
+#-------------------------------------------------------------------------------
+# - Map important nodes
+
+#--------------------------------
+# Head
+#--------------------------------
+
+
+# Libraries
+#--------------------------------
+library(dplyr)
+library(leaflet)
+library(BAMMtools) # Jenks library
+
+# The folder containing results
+results_folder = "results/all_lines2"
+
+#--------------------------------
+# Process
+#--------------------------------
+
+# --- Reading files
+aggregate_transfers_df = read.csv("results/all_lines2/aggregate_transfers_df.csv")
+coord = read.csv2("utilities/data_processing/stop_frequentation2.csv")
+
+# --- Adaptation of coord
+coord$id = paste0("S", coord$code_ligne_theo,"_", coord$direction_voy_theo, "_",
+                  coord$code_arret_theo)
+coord$id <- gsub("\\s+", "", coord$id)
+coord = coord[,c(11, 4, 9, 10)]
+colnames(coord)[2] = "name"
+
+# Mean of coord
+coord_mean <- aggregate(cbind(lon, lat) ~ name, data = coord, FUN = mean)
+
+# Join data
+aggregate_transfers_df <- left_join(
+  aggregate_transfers_df,
+  coord_mean,
+  by = c("stop_i" = "name")
+)
+
+
+# Color palette from white to red
+colorPal = colorBin("Reds", aggregate_transfers_df$sum.number_of_transfers.,
+                    bins = getJenksBreaks(
+                      aggregate_transfers_df$sum.number_of_transfers., 
+                      9, subset = NULL))
+# Map results
+nodes_map = leaflet() %>%
+  addProviderTiles(providers$CartoDB.DarkMatter) %>%
+  setView(lat=46.54033, lng=6.63575 , zoom=13) %>%
+  addCircles(lng = aggregate_transfers_df$lon, lat = aggregate_transfers_df$lat,
+             radius = aggregate_transfers_df$sum.number_of_transfers./15000,
+             color = colorPal(aggregate_transfers_df$sum.number_of_transfers.),
+             fillOpacity = 1,
+             opacity = 1
+  )
+nodes_map

### 2023-02-27 5.5_visu_tl_agg_nodes.R
update

@@ -1,4 +1,3 @@
-setwd("/Users/rloup/Documents/r_projects/tl_study")
 #-------------------------------------------------------------------------------
 #-------------------------------------------------------------------------------
 # Transfer visualisation

### 2023-02-27 latex_files/applied_network_science_article/article_v0.tex
write article

@@
 Yet, if automatic counting makes the embarkment (boarding) and disembarkment (alighting) counts in vehicles known, it often happens that pedestrian transfers between lines are harder to track, and require costly and invasive devices (e.g., facial recognition system) to be estimated.
-In this contribution, we propose a method, based on maximum entropy and involving an iterative fitting procedure, which estimates the passenger flow between origins and destinations solely based on embarkment and disembarkment flows.
+In this contribution, we propose a method, based on maximum entropy and involving an iterative fitting procedure, which estimates the passenger flow between origins and destinations solely based on embarkment and disembarkment counts.
 Moreover, this method is flexible enough to provide an adaptable framework in case additional data is known, such as attraction poles between certain points in the network, or percentages of transferring passengers between some lines.
-This method is tested on toy examples, as well as with the data of the public transportation network of the city of Lausanne provided by its Transportation Agency (tl), and gives convincing estimations of the transportation flow.
+This method is tested on toy examples, as well as with the data of the public transportation network of the city of Lausanne provided by its Transportation Agency (tl), and gives arguably convincing estimations of the transportation flow.
 \end{abstract}
@@
 
-Pedestrian transfers of travelers between different lines here constitute the missing information, which planners traditionally try to estimate using census, or more recently with costly and invasive devices located at station, such as mobile phone tracking or facial recognition systems.
-In this article, we take a different approach, which is to give the best estimation of transportation trajectories without additional data, using the principle of maximum entropy.
-Section \ref{notforma} introduces the notations and the formalism, as well as the statement of the problem and the iterative solution method, which consist of three consecutive steps: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior;  an update of the prior distribution by shrinking the components responsible for overflow; and an update of the marginal flows to avoid transfer overflow.
+Pedestrian transfers of travelers between different lines here constitute the missing information, which planners traditionally attempt to estimate using census, or more recently with costly and invasive devices located at station, such as mobile phone tracking or facial recognition systems.
+In this article, we take a different approach, which is to produce the optimal estimation of transportation trajectories without additional data, using the principle of maximum entropy.
+The proposed solution consists of three consecutive steps: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior;  an update of the prior distribution by shrinking the components responsible for overflow; and an update of the marginal flows to avoid transfer overflow.
 This iterative procedure clearly evokes the EM algorithm (see e.g.
 \cite{dempster1977maximum}  \cite{bavaud2009information}), with the first step corresponding to the ``expectation step" and last two steps to the ``maximization step".
-The fist step only is required for solving the single line case (section \ref{Single line}), naturally much simpler but  yet  not trivial, and exhibiting a disembarking probability independent of the embarking stop (Markov property).
-Moreover, this solution offer some flexibility, as it is possible to set a prior distribution taking into account attraction or repulsion poles among stops, and to fix hyperparameters limiting the number of transfers between lines.
+The fist step only is required for solving the single line case, naturally much simpler but  yet  not trivial, and exhibiting a disembarking probability independent of the embarking stop (Markov property).
+Moreover, this solution offers some flexibility, as it is possible to set a prior distribution taking into account attraction or repulsion poles among stops, and to fix hyperparameters limiting the number of transfers between lines.
 
-In section \ref{casestudies}, we test the propose method on toy examples and with the data of the public transportation network of the city of Lausanne.
-Toy examples offer some kind of validation, as a transportation flow can be set on toy networks and compared to the solution given by our algorithm.
-The real case scenario with the data from the public transportation network of the city of Lausanne shows that this method is applicable on a large transportation network and can give pertinent insights about travelers habits.
+As case studies, we test the proposed method on toy examples as well as with the data of the public transportation network of the city of Lausanne, in Switzerland.
+Toy examples offer some kind of validation, as a transportation flow can be entirely set on toy networks and compared to the solution given by our algorithm.
+The real case scenario with the data from the Lausanne public transportation network demonstrates that this method is applicable on a large transportation network and can yield pertinent insights about traveler habits.
+
+Section \ref{notforma} introduces the notations and the formalism, as well as the statement of the problem and the proposed solution.
+Section \ref{casestudies} contains case studies of the method on toy examples and on the data from the Lausanne public transportation network.
+Section \ref{conclusion} concludes the article.
 
@@
 Note that we can also forbid some aberrant trips across the network, for example, trips $[s, t]$ where $(s, t)$ is a transfer edge (making this trip do not actually use the line network).
-The \emph{set of permitted trips} across the network is denoted by $P$, and can be defined regarding some conditions.
+The \emph{set of permitted trips} across the network is denoted by $P$, and can be defined following some conditions (see, e.g., section \ref{example_construction}).
 
@@
 \emph{Automatic passenger counters} measure the number of passengers entering and leaving lines at each stop [Boyle, 1998], that is $\mathbf{a}$ and $\mathbf{b}$, which provide the basic raw data of the present study, kindly provided by the Lausanne Transportation Agency (tl) for the case study in section \ref{real_data}.
-We will suppose here that this data obeys the necessary consistency conditions for embarkment and disembarkment counts which can be found in the Appendix, even if, in real case studies, a rescaling must usually be performed to balance in and out-flows on each lines (also given in the Appendix).
+We will suppose here that this data obeys the necessary consistency conditions for embarkment and disembarkment counts which can be found in appendix, even if, in real case studies, a rescaling must usually be performed to balance in and out-flows on each lines (also given in appendix).
 
@@
 Many consistent solutions coexist in general, even for a single line with no transferts (section \ref{Single line}).
-This issue of incompletely observed data can be tackled by the maximum entropy formalism \cite{jaynes1957information}, which has often been the case in transportation modelling researches \cite{wilson1967statistical}  \cite{erlander1990gravity}.
+This issue of incompletely observed data can be tackled by the maximum entropy formalism \cite{jaynes1957information}, which has often been invoked in transportation modeling researches \cite{wilson1967statistical}  \cite{erlander1990gravity}.
 
@@
 \end{itemize}
-With the new prior distribution $\widetilde{g}_{st}$ and the new margin distributions $\widetilde{\alpha}_s$, $\widetilde{\beta}_t$, we can iterate the 
- the above steps, until convergence. The only free parameter is $\theta$, whose effect is studied on toy examples in section \ref{toy_examples}.
+With the new prior distribution $\widetilde{g}_{st}$ and the new margin distributions $\widetilde{\alpha}_s$, $\widetilde{\beta}_t$, we can iterate the above steps, until convergence.
+The only free parameter is $\theta$, whose effect is studied on toy examples in section \ref{toy_examples}.
 
@@
 However, $g_{st}$ could be chosen more carefully in case of additional data.
-As a matter of fact, they represent \emph{prior attractions} between nodes in equation $\ref{Sol}$, before margin correction given by $\phi_s$ and $\psi_t$ and subsequent algorithm steps.
+As a matter of fact, $g_{st}$ represent \emph{prior attractions} between nodes in equation (\ref{Sol}), before margin correction given by $\phi_s$ and $\psi_t$ and subsequent algorithm steps.
 An urban planner could chose to increase or decrease some of these values in order to take into account prior knowledge on commuting habits of inhabitants.
@@
 
-When $z_{\bullet i} > a_i$, or $z_{i \bullet} > b_i$, the found solution typically predict that there are more passengers entering, respectively exiting, at a stop $i$ than the actual measured quantity.
+When $z_{\bullet i} > a_i$, or $z_{i \bullet} > b_i$, the found solution typically predicts that there are more passengers entering, respectively exiting, at a stop $i$ than the actual measured quantity.
 One could be tempted to correct prior distributions in order to have $z_{\bullet i} = a_i$, or $z_{i \bullet} = b_i$, on every problematic nodes $i$, but the latter solution would mean that all passengers entering (resp.
@@
 \end{equation}
-or 
+or equivalently 
 \begin{equation}
@@
 
-Note that, if additional data would be known, we could set a particular value of $\theta$ for every node, and differing for embarkments and disembarkments.
-However, without addition information, we will restrain to this simpler case.
+Note that, if additional data were known, we could set a particular value of $\theta$ for every node, and differing for embarkments and disembarkments.
+However, without additional information, we will restrain to this simpler case.
 
@@
 \begin{algorithm}[h]
-	\caption{Compute the transportation flow matrix $\mathbf{N} = (n_{st})$ knowing the edge-trip incidence matrix $\bm{\chi} = (\chi_{ij}^{st})$, the set of transfer edges $T$, the set of permitted trips $P$, the embarking flow $\mathbf{a}$, the disambarking flow $\mathbf{b}$, the index of an isolated source node $\tilde{s}$, and the minimum proportion of passengers entering/leaving the network $\theta$.}
+	\caption{Computes the transportation flow matrix $\mathbf{N} = (n_{st})$, knowing the edge-trip incidence matrix $\bm{\chi} = (\chi_{ij}^{st})$, the set of transfer edges $T$, the set of permitted trips $P$, the embarking flow $\mathbf{a}$, the disambarking flow $\mathbf{b}$, the index of an isolated source node $\tilde{s}$, and the minimum proportion of passengers entering/leaving the network $\theta$.}
 	\label{algo1}
@@
 
-Let  $i=1,\ldots, l$ enumerate the bus stops in increasing order,  i.e.
-$F(i)=i+1$.
+Let  $i=1,\ldots, l$ enumerate the bus stops in increasing order, that is $F(i)=i+1$.
 The initial prior is simply $g_{st}=c\; I(s<t)$ and captures solely the unidirectional nature of trips, where $c=\frac{1}{(l-1)(l-2)}$.
-The margins of the empirical distribution $f_{st}$, as well as the total flow, are here known :
-\begin{displaymath}
+The margins of the empirical distribution $f_{st}$, as well as the total flow, are here known
+\begin{equation}
 \alpha_s=\frac{a_s}{a_\bullet}\qquad\qquad\qquad \beta_t=\frac{b_t}{b_\bullet}\qquad\qquad\qquad n_{\bullet\bullet}=a_{\bullet}=b_\bullet\enspace. 
-\end{displaymath}
+\end{equation}
 Following  (\ref{Sol}) maximum entropy flows are of the form
@@
 Case studies are divided in two sections.
-In the first section, we test the algorithm on toy examples, which are artificial networks where the transportation flow $n_{st}$ is randomly drawn.
-These examples enable some kind of validation of the algorithm, as the "real" transportation flow is known and can be compared to solutions given by our method.
+In the first section, we test the algorithm on toy examples, which are artificial networks where the transportation flow $\mathbf{N}_\text{ref} = (n^\text{ref}_{st})$ is randomly drawn.
+These examples enable some kind of validation of the algorithm, as the "real" transportation flow is known and can be compared to the solution $\mathbf{N} = (n_{st})$ given by our method.
 This setup differs from the second section, which is dedicated to applying the algorithm to the real case of the public transportation network of the city of Lausanne (tl), where embankment and disembarkment flows are measured but the real transportation flow is unknown.
@@
 There are two types of dissimilarity measures between the data and the solution proposed by the algorithm: (1) if we have access to $\mathbf{N}_\text{ref}$, how much $\mathbf{N}$ differs from it, and (2) how well constraints defined by $\mathbf{a}_\text{ref}$ and $\mathbf{b}_\text{ref}$ are respected.
-The first dissimilarity is measured through the \emph{mean transportation error}, denoted by $\text{MTE}(\mathbf{N})$, and computed with
+The first dissimilarity is measured through the \emph{mean transportation error}, denoted by $\text{MTE}(\mathbf{N})$, and computed as
 \begin{equation}
@@
 \end{equation} 
-and the second one with the \emph{mean margin error}, noted $\text{MME}(\mathbf{N})$, obtained with
+and the second one with the \emph{mean margin error}, noted $\text{MME}(\mathbf{N})$, defined as
 \begin{align}
@@
 
-Note that, by construction, the MME should be null when the algorithm converges.
+Note that, by construction, the MME should be zero when the algorithm converges.
 However, it can be informative to track down this error along iterations and, in some practical cases where margin constraints are impossible to fulfill, the algorithm convergence criterion is reached with MME $> 0$.
@@
 All constructed toy examples are built following the same approach, which aims at being simple but somewhat realistic.
-We fix a number of \emph{line tours} $p \geq 2$, each of which is constituted of a forward line and a backward line, for a total of $q = 2p$ lines.
-Every line has a starting and ending node, which are isolated node, and posses $p-1$ intermediary nodes which allows transfers to the other tour lines, giving a total of $n = 2p(p + 1)$ nodes in the network.
+We fix a number of \emph{round trips} $p \geq 2$, each of which is constituted of a forward line and a backward line, for a total of $q = 2p$ lines.
+Every line has a starting and ending node, which are isolated nodes, and possesses $p-1$ intermediary nodes which allow transfers to the other round trips, yielding a total of $n = 2p(p + 1)$ nodes in the network.
 Examples of these toy networks can be found in Figure \ref{toy_example_plots}.
@@
 	\includegraphics[width=0.3\textwidth]{fig/toy_4_display.pdf}
-	\caption{3 toy examples, with the number of tours $p \in \{2,3,4\}$.
-Tours are displayed in the same color, line have a unique letter, and each stop a unique combination of a letter and a number.
-Clusters of nodes represent positions where transfer between tours are possible.}
+	\caption{3 toy examples, with the number of round trips $p \in \{2,3,4\}$.
+Trips are displayed in the same color, lines possess a unique letter, and each stop correspond to a unique combination of a letter and a number.
+Clusters of nodes represent positions where transfers between round trips are possible.}
 	\label{toy_example_plots}
@@
 
-In order to be realistic, permitted $st$-trips set $P$ is constructed considering all shortest-path between pair of nodes, excluding :
+In order to be realistic, the permitted $st$-trips set $P$ is constructed by considering all shortest-path between pair of nodes, excluding :
 \begin{itemize}
-\item $s$ and $t$ that are on the same line but with $t$ preceding $s$ in the line order.
-\item $s$ and $t$ that are on the same tour but opposite line.
+\item $s$ and $t$ that are on the same line but with $t$ preceding (or equal to) $s$ in the line order.
+\item $s$ and $t$ that are on the same round trip but on opposite lines.
 \item $s$ and $t$ whose shortest-path starts with a transfer edge, ends with a transfer edge, or possesses two (or more) consecutive transfer edges.
@@
 
-First, we shows some algorithm iterations on a toy example with $p=2$, where $50$ passengers where drawn uniformly across the $\vert P \vert = 20$ possible $st$-trips.
+First, we exhibit some algorithm iterations on a toy example with $p=2$, where $50$ passengers where drawn uniformly across the $\vert P \vert = 20$ possible $st$-trips.
 Some iterations of the algorithm with $\theta=0.001$, along with MTE and MME errors, are shown in Figure \ref{iteration_plots}.
@@
 	\includegraphics[width=0.98\textwidth]{fig/iterations.pdf}
-	\caption{Real transportation flow (green arrows) obtained by randomly drawing $50$ passengers on the graph toy example with $p=2$ line tours, along with iterations $1, 2, 4, 7, 15$ of the algorithm with $\theta=0.001$.
+	\caption{Real transportation flow (green arrows) obtained by randomly drawing $50$ passengers on the graph toy example with $p=2$ round trips, along with iterations $1, 2, 4, 7, 15$ of the algorithm with $\theta=0.001$.
 MTE and MME errors are computed, and embarkment and disembarkment flows are represented respectively by the red and blue colors on nodes.}
@@
 On this small example, we can see that the algorithm quickly find an estimation giving small MME error, but still give an MTE of 0.256.
-This result is due to the fact that only $50$ passengers are drawn, giving a large deviation compared to the optimally found solution which maximize the entropy.
+This result is due to the fact that only $50$ passengers are drawn, giving a large deviation compared to the optimally found solution which maximizes the entropy.
 
@@
 
-The main goal of toy examples, since we have access to the real transportation flow, is to study how the resulting $MTE$ behave regarding passenger $st$-trips distribution and hyperparameter $\theta$, on different network sizes.
+The main goal of toy examples, since we have access to the real transportation flow, is to study how the resulting MTE behaves regarding passenger $st$-trips distribution and hyperparameter $\theta$, on different network sizes.
 
 The randomness of the $st$-trips distribution is controlled by the number of drawn passengers and we can see that the algorithm performs better if this number increases, as seen in Figure \ref{MTE_vs_passengers}.
-This behavior can be expected as the method is constructed to find the solution maximizing the entropy while respecting embarkment and disembarkment constraint.
+This behavior can be expected as the method is constructed to find the solution maximizing the entropy while respecting embarkment and disembarkment constraints.
 In fact, all MTE should converge to $0$ eventually, however this rate of convergence seems to be low.
@@
 	\includegraphics[width=0.98\textwidth]{fig/MTE_passengers.pdf}
-	\caption{Mean Transportation Error (MTE) vs the number of randomly drawn passengers, in toy examples with numbers of tours $p$ varying from 2 to 8.
+	\caption{Mean Transportation Error (MTE) vs the number of randomly drawn passengers, in toy examples with numbers of round trips $p$ varying from 2 to 8.
 The hyperparameter is set to $\theta=0.001$ and $10$ different draws are performed for each data point.}
@@
 \end{figure}
-As for the hyperparameter study, found in Figure \ref{MTE_vs_theta}, we see that the optimal parameter seems to decreases as network size increases, reaching value close to $0$ for $8$ line tours.
-However, this is not the only factor which helps to decide this parameter.
+As for the hyperparameter study, found in Figure \ref{MTE_vs_theta}, we see that the optimal parameter seems to decrease as the network size increases, reaching value close to $0$ for $8$ round trips.
+However, this is not the only factor which helps to calibrate this parameter.
 As a matter of fact, it seems unrealistic, in some real life scenarios, that all passengers embarking or disembarking at junctions are all transiting, and this value will be kept at $0.1$ for our real life study.
@@
 	\includegraphics[width=0.98\textwidth]{fig/MTE_tours.pdf}
-	\caption{Mean Transportation Error (MTE) vs hyperparameter $\theta$, in toy examples with numbers of tours $p$ varying from 2 to 8.
+	\caption{Mean Transportation Error (MTE) vs hyperparameter $\theta$, in toy examples with numbers of round trips $p$ varying from 2 to 8.
 The dot on each curve denotes the minimum.
-The number of drawn passengers is different for every network size and is set to $20 \cdot \vert P \vert$, in order to keep the randomness level constant.
+The number of drawn passengers is different for every network size and is set to $5 \cdot \vert P \vert$, in order to keep the randomness level constant.
 $10$ different draws are performed for each data point.}
@@
 The dataset derived from the public transportation agency in the city of Lausanne (tl), in Switzerland.
-Automatic passenger counters were used to collect data on the number of passengers entering and leaving buses and metros at each stop.
-The dataset comprises 44 lines of metro and bus transportation, 1361 stops, and over 115 million passengers per year.
+Automatic passenger counters were used to collect data on the number of passengers entering and leaving buses and subways at each stop.
+The dataset comprises 44 lines, 1361 stops, and over 115 million passengers per year.
 Each round trip forms two separate lines, which often have stops that correspond to both the inbound and outbound directions, but this is not always the case.
@@
 Lines with the most errors were removed, resulting in a final dataset of 35 transportation lines and 1216 stops.
-The threshold for invalid lines was set when a difference of between total embarkment and disembarked differs more than 15\% of their means.
+The threshold for invalid lines was set when a difference between total embarkment and disembarked differs more than 15\% of their means.
 For the rest all other lines, the correction found in appendix was applied to ensure consistency conditions.
 
-The matrix $P$, containing valid $st$-trips, was built using same condition found in toy example with an additional one.
+The matrix $P$, containing valid $st$-trips, was built using same condition found in toy example (section \ref{example_construction}) with an additional one.
 If the estimated pedestrian time (using Open Trip Planner\footnote{\url{https://docs.opentripplanner.org/en/v2.2.0/}} \cite{MalcolmMorgan2019}) between two stops $s$ and $t$ was shorter than the estimated time by taking the public transportation network, this $st$-trip was also removed from $P$.
-The graph structure, consisting in matrices $\bm{\chi}$ and $P$, was constructed beforehand testing the method.
+The graph structure, consisting in matrices $\bm{\chi}$, $P$ and $T$, was constructed beforehand testing the method.
 
 The algorithm was run with hyperparameter $\theta = 0.1$.
-Tracking the Mean Margin Error (MME) showed a rapid decreases but the (conservative) convergence criterion (chosen as when $\sum_{st} \vert f^{prev}_{st} - f_{st} \vert< 10^{-6} $) was reach after 316 iterations, which corresponds to approximately one hour of computing time with a single thread.
+Tracking the Mean Margin Error (MME) showed a rapid decrease but the (conservative) convergence criterion (chosen as when $\sum_{st} \vert f^{prev}_{st} - f_{st} \vert< 10^{-6} $) was reach after 316 iterations, which corresponds to approximately one hour of computing time with a single thread on a AMD Epyc2 7402 with 32GB of RAM.
 
@@
 As the result of our algorithm consists in the $(1216 \times 1216)$ matrix $\mathbf{N} = (n_{st})$, it is difficult to present them succinctly.
-We chose here to focus on two aspects: (1) mapping the origin and destination profiles of stops with the help of Correspondence Analysis (CA) \cite{benzecri1977histoire}; and (2) mapping the predicted transfers between lines.
+We chose here to focus on two aspects: (1) mapping the origin and destination profiles of stops with the help of Correspondence Analysis (CA) \cite{benzecri1977histoire}; and (2) mapping the aggregated transfers between lines.
 
-The first and second factorial dimensions of the destination and origin profiles of stops, obtained with CA applied on $\mathbf{N}$ , can be found on Figure \ref{afc}.
-We can see that origin and destination of each stop are quite similar when displayed on the first two factorial axes.
-This result can be explained if we understand that, in fact, origin and destination profiles are in fact kinda symmetrical: at the beginning of a line, stops have several choices of destinations but very few origins ``point" to them, and this is the opposite for stops at the end of the line.
-Intermediary stops are generally located at the center of the city, and have a similar and rich origin and destination profiles.
-Concerning the dimensions, we can see that the first one divided the city stops between west and east, a known dichotomy in the city of Lausanne in public transportation usage: the east of the city is less well deserved and often use cars, while the west intensively use the public transportation network.
+The first and second factorial dimensions of the destination and origin profiles of stops, obtained with CA applied on $\mathbf{N}$, can be found on Figure \ref{afc}.
+We can see that origin and destination coordinates of each stop are quite similar when displayed on the first two factorial axes.
+This result can be explained if we understand that, in fact, origin and destination profiles are in fact kinda symmetrical: at the beginning of a line, stops have several choices of destinations, but very few origins ``point" at them, which is exactly reversed for stops at the end of a line.
+Intermediary stops are generally located near city center, and have a similarly rich origin and destination profiles.
+Concerning the dimensions, we can see that the first one divided the city stops between west and east, a known dichotomy in the public transportation usage in Lausanne: the east of the city is less well served by public transport and inhabitants often use cars, while the west intensively use public transports.
 The second dimension is harder to interpret, but seems to highlight the radial structure of the transportation network in Lausanne.
 
-Figure \ref{transfers} maps the geographically aggregate transfer flows $z_{ij}$ between lines.
+Figure \ref{transfers} maps predicted transfer hubs, obtained by aggregating every transfer counts $z_{ij}$ between lines when they occurs in a spatial cluster of stops.
 As expected, most transfers occur in the city center, with top 5 largest hubs located at St-François (2,132,597 transfers), Lausanne-Gare (2,035,543 transfers), Bel-Air (1,908,168 transfers), Ours (1,057,282 transfers), and Chauderon (969,789 transfers).
 When taken individually, four of the top 5 transfer counts is predicted between line $1$ and line $72$ occurring at the train station (Lausanne-Gare).
-Line 72 is actually a subway line connecting north and south region of the city, and is by far the most used line in the network.
-Line $1$ is also a very frequented line and operated toward the western region of the city.
+Line 72 is actually a subway line connecting the north and south regions of the city, and is by far the most used line in the network.
+Line $1$ is also a highly frequented line and operated toward the western region of the city.
 It is not surprising to see most transfer occurring between these two lines, as they permit to connect the region with an high density of public transportation network users (the western part of Lausanne), to the main artery of the network (line 72).
@@
 	\includegraphics[width=0.97\textwidth]{fig/transfers.png}
-	\caption{Location and size (ring size and color) of estimated transfer flows $z_{ij}$ between lines.}
+	\caption{Location and size of transfer hubs, obtained by aggregating estimated transfer flows between lines $z_{ij}$ in geographical clusters.}
 	\label{transfers}
 \end{figure}
-
-\subsubsection{Results}
-\label{lausanne_results}
 
@@
 \section{Conclusion}
+\label{conclusion}
 
@@
 \label{flow_correction}
-It may happen that, on some lines $\ell$, with stops indexed in order as $1, \ldots, l$, raw data do not obey the following necessary consistency conditions 
+It may happen that, on some lines $\ell$ with stops indexed in order as $1, \ldots, l$, raw data $\mathbf{a}=(a_i)$ and $\mathbf{b}=(b_i)$ do not obey the following necessary consistency conditions 
 \begin{align*}
@@
 where $A_i$ (respectively $B_i$) is the cumulated number of embarked (resp.
-disembarked) passengers on the line at stop $i$.
+disembarked) passengers on the line at stop $i$, i.e.
+$A_i = \sum_{j=1}^{i} a_j$ and $B_i = \sum_{j=1}^{i} b_j$.
 The first condition is easy to correct (by setting both quantities to $0$), and we will assume that they are valid.
@@
 Before the last step, all conditions $A_{(i-1)} \geq  B_{i}$ should be respected except for node $l$.
-The last step ensures that $A_{l} = B_{l}$, however, as it can sometimes lower the embarkment count and increase the disembarkment count on previous nodes, some new nodes can now violate $A_{(i-1)} \geq B_i$.
+The last step ensures that $A_{l} = B_{l}$.
+However, as this last step can sometimes lower the embarkment count and increase the disembarkment count on previous nodes, some new nodes can now violate $A_{(i-1)} \geq B_i$.
 This is why this procedure must be iterated until all stops on the line verify consistency conditions.
@@
 \section*{Acknowledgements}%% if any
-Text for this section\ldots
+This paper would not have been possible without the dataset given by the transportation agency of the city of Lausanne (tl).
+We wish to thank the agency for its kindness and hope this research will help them for future developments.
 

### 2023-02-27 latex_files/applied_network_science_article/article_v0.tex
finish v0 of article

@@
 
-As case studies, we test the proposed method on toy examples as well as with the data of the public transportation network of the city of Lausanne, in Switzerland.
+As case studies, we test the proposed method on toy examples as well as with the data of the public transportation network\footnote{\url{https://www.t-l.ch/}} of the city of Lausanne, in Switzerland.
 Toy examples offer some kind of validation, as a transportation flow can be entirely set on toy networks and compared to the solution given by our algorithm.
@@
 Section \ref{conclusion} concludes the article.
+Data, codes and extensive results can be found in the GitHub repository of the article\footnote{\url{https://github.com/sliunil/tl_study}}.
 
@@
 We can see that origin and destination coordinates of each stop are quite similar when displayed on the first two factorial axes.
-This result can be explained if we understand that, in fact, origin and destination profiles are in fact kinda symmetrical: at the beginning of a line, stops have several choices of destinations, but very few origins ``point" at them, which is exactly reversed for stops at the end of a line.
-Intermediary stops are generally located near city center, and have a similarly rich origin and destination profiles.
-Concerning the dimensions, we can see that the first one divided the city stops between west and east, a known dichotomy in the public transportation usage in Lausanne: the east of the city is less well served by public transport and inhabitants often use cars, while the west intensively use public transports.
-The second dimension is harder to interpret, but seems to highlight the radial structure of the transportation network in Lausanne.
+This result can be explained if we understand that, in fact, origin and destination profiles are in fact, to some extent, symmetrical: at the beginning of a line, stops have several choices of destinations, but very few origins ``point" at them, which is exactly the reverse for stops at the end of a line.
+Intermediary stops are generally located near the city center, and have a similarly rich origin and destination profiles.
+Concerning the dimensions, we can see that the first component divides the city stops between west and east, a known dichotomy in the public transportation usage in Lausanne: the east of the city is less well served by public transport and inhabitants often use cars, while the west intensively uses public transports.
+The second component is harder to interpret, but seems to highlight the radial structure of the transportation network in Lausanne.
 
-Figure \ref{transfers} maps predicted transfer hubs, obtained by aggregating every transfer counts $z_{ij}$ between lines when they occurs in a spatial cluster of stops.
-As expected, most transfers occur in the city center, with top 5 largest hubs located at St-François (2,132,597 transfers), Lausanne-Gare (2,035,543 transfers), Bel-Air (1,908,168 transfers), Ours (1,057,282 transfers), and Chauderon (969,789 transfers).
-When taken individually, four of the top 5 transfer counts is predicted between line $1$ and line $72$ occurring at the train station (Lausanne-Gare).
+Figure \ref{transfers} maps predicted transfer hubs, obtained by aggregating every transfer counts $z_{ij}$ between lines when they occurs between stops located in the same spatial cluster (same ``superstop" name in the dataset).
+As expected, most transfers occur in the city center, with the top 5 largest hubs located at St-François (2,132,597 transfers), Lausanne-Gare (2,035,543 transfers), Bel-Air (1,908,168 transfers), Ours (1,057,282 transfers), and Chauderon (969,789 transfers).
+When taken individually, four of the top 5 transfer counts are predicted between line $1$ and line $72$ occurring at the train station (Lausanne-Gare).
 Line 72 is actually a subway line connecting the north and south regions of the city, and is by far the most used line in the network.
@@
 	\includegraphics[width=0.46\textwidth]{fig/afc_out_dim2.png}
-	\caption{Color representations (positive values in red, negative values in blue) of the first two factorial coordinates of stops, after Correspondence analysis (CA) performed on transportation matrix $\mathbf{N}$.
+	\caption{Color representations (positive values in red, negative values in blue) of the first two factorial coordinates of stops, obtained from Correspondence analysis (CA) performed on the transportation matrix $\mathbf{N}$.
 Top row maps stops as ``origin profiles" (``where passengers are going"), bottom row maps stops as ``destination profile" (``where passengers are coming from").
-Left column is the first factorial dimension, right column is the second factorial dimension.
-Ring sizes correspond to the number of embarkments (top) and disembarkment (bottom).}
+Left column depicts the first factorial dimension, right column the second factorial dimension.
+Ring sizes correspond to the number of embarkments (top) and disembarkments (bottom).}
 	\label{afc}
@@
 \label{conclusion}
-
-\newpage
+At first glance, the task of estimating the origin-destination trip counts of public transportation networks on the sole basis of embarkments and disembarkments stop counts might appear as too ambitious.
+The project is indeed challenging, but we hope to have convinced the reader that meaningful estimates can be obtained by applying a succession of carefully chosen, principled yet flexible steps.
+Some of the ingredients (maximum entropy, iterative fitting) are familiar in transportation theory, and some others (shrinkage of priors, minimal amount of entering and leaving the network) seem original.
+Possible theoretical and algorithmic improvements are still under investigation.
+Due to length constraints, the exploitation and interpretation of the estimated flow has been kept to a minimum.
+Yet, many issues, such as disaggregating data and regimes (week days, week-ends or holidays; time in the day), assessing the centrality of stops and junctions, or characterizing the imbalance between uphill and downhill flows (Lausanne is known for its slopes) are of primary interest for urban planning and transportation geography, and will be developed in a forthcoming work.
+Also, visualizing the numerous and various quantities of interest, on a spatial map or else, constitutes a challenge in itself, requiring a particular blend of creativity and rigor.
 

### 2023-02-27 5.3_visu_tl_CA.R
update visu of CA

@@ -64,10 +64,10 @@ rownames(n_mat) = stop_names
 res.ca <- CA(n_mat, graph = FALSE)
 
 if (up_down == "up") {
-  CA_dim = res.ca$row$coord[,1:2]
+  CA_dim = res.ca$row$coord[, 1:2]
 }
 if (up_down == "down") {
-  CA_dim = res.ca$col$coord[,1:2]
+  CA_dim = res.ca$col$coord[, 1:2]
 }
 
 colnames(CA_dim) <- c("dim1","dim2")
@@ -150,7 +150,7 @@ CA_map2 = leaflet(CA_shp) %>%
     # radius = ((CA_shp$montees)/max(CA_shp$montees)*100000)^0.45, # coefficient de correction avec un minimum de largeur de 1
     radius = ((CA_shp$CA_radius)/max(CA_shp$CA_radius)*100000)^0.45, # coefficient de correction avec un minimum de largeur de 1
     fillColor = "transparent",
-    opacity = 1,
+    opacity = ((CA_shp$CA_radius)/max(CA_shp$CA_radius))^0.05,
     color = ~colorPal1(dim1),
     highlight = highlightOptions(
       color = "white",
@@ -162,7 +162,7 @@ CA_map2 = leaflet(CA_shp) %>%
     # radius = ((CA_shp$montees)/max(CA_shp$montees)*100000)^0.45, # coefficient de correction avec un minimum de largeur de 1
     radius = ((CA_shp$CA_radius)/max(CA_shp$CA_radius)*100000)^0.45, # coefficient de correction avec un minimum de largeur de 1
     fillColor = "transparent",
-    opacity = 1,
+    opacity = ((CA_shp$CA_radius)/max(CA_shp$CA_radius))^0.05,
     color = ~colorPal2(dim2),
     highlight = highlightOptions(
       color = "white",

### 2023-02-27 latex_files/applied_network_science_article/article_v0.tex
update visu of CA
write article

@@
 In this article, we take a different approach, which is to produce the optimal estimation of transportation trajectories without additional data, using the principle of maximum entropy.
-The proposed solution consists of three consecutive steps: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior;  an update of the prior distribution by shrinking the components responsible for overflow; and an update of the marginal flows to avoid transfer overflow.
+The proposed solution consists of three consecutive steps: a maximum-entropy computation of the trip distributions, obeying marginal constraints and with a given prior;  an update of the prior distribution by shrinking the components responsible for transfer overflow; and an update of marginal distributions.
 This iterative procedure clearly evokes the EM algorithm (see e.g.
@@
 
-The matrix $P$, containing valid $st$-trips, was built using same condition found in toy example (section \ref{example_construction}) with an additional one.
-If the estimated pedestrian time (using Open Trip Planner\footnote{\url{https://docs.opentripplanner.org/en/v2.2.0/}} \cite{MalcolmMorgan2019}) between two stops $s$ and $t$ was shorter than the estimated time by taking the public transportation network, this $st$-trip was also removed from $P$.
-The graph structure, consisting in matrices $\bm{\chi}$, $P$ and $T$, was constructed beforehand testing the method.
+Line edges were built from the given data, and to obtain the full network structure, we added transfer edges between different lines if the estimated walking time (using Open Trip Planner\footnote{\url{https://docs.opentripplanner.org/en/v2.2.0/}} \cite{MalcolmMorgan2019}) was less than $120$ seconds.
+Shortest-paths between all pairs of stops were computed using the breath-first search algorithm \cite{west2001introduction} implemented in the \emph{igraph}\footnote{\url{https://igraph.org/}} R library \cite{Csardi2006}, thus giving $\bm{\chi}$.
+The matrix $P$, containing valid $st$-trips, was built using the same conditions found in the toy example experiments (section \ref{example_construction}) with an additional one.
+If the estimated pedestrian time between two stops $s$ and $t$ was shorter than the estimated time by taking the public transportation network, this $st$-trip was also removed from $P$.
+The graph structure, consisting in the edge-trip incidence matrix $\bm{\chi}$, the valid $st$-trips set $P$, and the transfer edges set $T$, was constructed beforehand testing the method.
 
 The algorithm was run with hyperparameter $\theta = 0.1$.
-Tracking the Mean Margin Error (MME) showed a rapid decrease but the (conservative) convergence criterion (chosen as when $\sum_{st} \vert f^{prev}_{st} - f_{st} \vert< 10^{-6} $) was reach after 316 iterations, which corresponds to approximately one hour of computing time with a single thread on a AMD Epyc2 7402 with 32GB of RAM.
+Tracking the Mean Margin Error (MME) showed a rapid decrease ($<0.001$ after 13 iterations) but the (conservative) convergence criterion (chosen as when $\sum_{st} \vert f^\text{prev}_{st} - f_{st} \vert< 10^{-6} $) was reach after 316 iterations, which corresponds to approximately one hour of computing time with a single thread on a AMD Epyc2 7402 with 32GB of RAM.
 
@@
 	\includegraphics[width=0.97\textwidth]{fig/transfers.png}
-	\caption{Location and size of transfer hubs, obtained by aggregating estimated transfer flows between lines $z_{ij}$ in geographical clusters.}
+	\caption{Location and size of transfer hubs, obtained by aggregating estimated transfer flows between lines $z_{ij}$ by ``superstop" names.}
 	\label{transfers}

### 2023-02-28 5.2_visu_toy_lines.R
visu theta v2

@@ -60,7 +60,6 @@ n_test = 10
 hyper_par = c(seq(0.001,0.009, 0.001), seq(0.01,0.1, 0.01), seq(0.12,0.2, 0.02),
               0.25, 0.3, 0.35, seq(0.4,1, 0.1))
 
-
 # number of passengers
 # n_passengers = 2000
 nb_pass_stop = 5
@@ -77,11 +76,12 @@ seq_lines = 2:8
 res_mat_line = c()
 
 for (j in 1:length(hyper_par)) {
+  print(paste("hyperpar", j))
   for (i in seq_lines) {
     nb_lines = i
     nb_stops = i + 1
     # Network properties output
-    network_prop_res = network_prop(nb_lines, nb_stops)
+    network_prop_res = network_prop(nb_lines, nb_stops, mc.cores )
     # Unlist and save variables
     list2env(network_prop_res, .GlobalEnv)
     
@@ -103,12 +103,14 @@ for (j in 1:length(hyper_par)) {
   mean_line = cbind(seq_lines, mean_line, sd_line)
   mean_line$param = row.names(mean_line)
   colnames(mean_line) <- c("Lines", "mean_error", "sd_error","param")
-  mean_line$param = substr(mean_line$param, 4, 7)
+  mean_line$param = substring(mean_line$param, 4)
 }
-write.csv(res_mat_line, "results/5_toy_ex_outputs/res_line_param.csv", row.names = F)
-write.csv(mean_line, "results/5_toy_ex_outputs/mean_line_param.csv", row.names = F)
+# write.csv(res_mat_line, "results/5_toy_ex_outputs/res_line_param.csv", row.names = F)
+# write.csv(mean_line, "results/5_toy_ex_outputs/mean_line_param.csv", row.names = F)
 # Read saved data
-# mean_line = read.csv("results/5_toy_ex_outputs/mean_line_param.csv")
+mean_line = read.csv("results/5_toy_ex_outputs/mean_line_param.csv")
+
+mean_line$param = as.numeric(mean_line$param)
 
 # Extract min/max
 min_max = as.data.frame(mean_line %>% group_by(Lines) %>% top_n(-1, mean_error))

### 2023-02-28 latex_files/applied_network_science_article/article_v0.tex
do last corrections

@@
 In this contribution, we propose a method, based on maximum entropy and involving an iterative fitting procedure, which estimates the passenger flow between origins and destinations solely based on embarkment and disembarkment counts.
-Moreover, this method is flexible enough to provide an adaptable framework in case additional data is known, such as attraction poles between certain points in the network, or percentages of transferring passengers between some lines.
+Moreover, this method is flexible enough to provide an adaptable framework in case additional data is known, such as attraction poles between certain nodes in the network, or percentages of transferring passengers between some lines.
 This method is tested on toy examples, as well as with the data of the public transportation network of the city of Lausanne provided by its Transportation Agency (tl), and gives arguably convincing estimations of the transportation flow.
@@
 \kwd{boarding and alighting counts}
-\kwd{transit flows}
 \kwd{maximum entropy estimation}
@@
 \cite{dempster1977maximum}  \cite{bavaud2009information}), with the first step corresponding to the ``expectation step" and last two steps to the ``maximization step".
-The fist step only is required for solving the single line case, naturally much simpler but  yet  not trivial, and exhibiting a disembarking probability independent of the embarking stop (Markov property).
-Moreover, this solution offers some flexibility, as it is possible to set a prior distribution taking into account attraction or repulsion poles among stops, and to fix hyperparameters limiting the number of transfers between lines.
+The first step only is required for solving the single line case, naturally much simpler but  yet  not trivial, and exhibiting a disembarking probability independent of the embarking stop (Markov property).
+This method also offers some flexibility, as it is possible to set a prior distribution taking into account attraction or repulsion poles among stops, and to fix hyperparameters limiting the number of transfers between lines.
 
@@
 Toy examples offer some kind of validation, as a transportation flow can be entirely set on toy networks and compared to the solution given by our algorithm.
-The real case scenario with the data from the Lausanne public transportation network demonstrates that this method is applicable on a large transportation network and can yield pertinent insights about traveler habits.
+The real case scenario with the data from the Lausanne public transportation network demonstrates that this method is applicable on a real transportation network and can yield pertinent insights about traveler habits.
 
 Section \ref{notforma} introduces the notations and the formalism, as well as the statement of the problem and the proposed solution.
-Section \ref{casestudies} contains case studies of the method on toy examples and on the data from the Lausanne public transportation network.
+Section \ref{casestudies} contains case studies with toy examples and the Lausanne public transportation network.
 Section \ref{conclusion} concludes the article.
@@
 A \emph{$st$-trip}, noted $[s,t]$, consists of entering into the network at stop $s$, and leaving the network at $t$, by following the shortest-path (i.e.
-achieving the minimum distance,  minimum time, or  minimum cost), supposed unique, leading to $s$ from $t$.
+achieving the minimum distance,  minimum time, or  minimum cost), supposed unique, leading from $s$ to $t$.
 
@@
 
-Note that we can also forbid some aberrant trips across the network, for example, trips $[s, t]$ where $(s, t)$ is a transfer edge (making this trip do not actually use the line network).
+Note that we can also forbid some aberrant trips across the network, for example, trips $[s, t]$ where $(s, t)$ is a transfer edge (travelers making this trip do not actually do not use the transportation network).
 The \emph{set of permitted trips} across the network is denoted by $P$, and can be defined following some conditions (see, e.g., section \ref{example_construction}).
@@
 \emph{Automatic passenger counters} measure the number of passengers entering and leaving lines at each stop [Boyle, 1998], that is $\mathbf{a}$ and $\mathbf{b}$, which provide the basic raw data of the present study, kindly provided by the Lausanne Transportation Agency (tl) for the case study in section \ref{real_data}.
-We will suppose here that this data obeys the necessary consistency conditions for embarkment and disembarkment counts which can be found in appendix, even if, in real case studies, a rescaling must usually be performed to balance in and out-flows on each lines (also given in appendix).
+We will suppose here that this data obeys the necessary consistency conditions for embarkment and disembarkment counts, even if, in real case studies, a rescaling must usually be performed to balance in and out-flows on each lines (given in the appendix).
 
@@
 Many consistent solutions coexist in general, even for a single line with no transferts (section \ref{Single line}).
-This issue of incompletely observed data can be tackled by the maximum entropy formalism \cite{jaynes1957information}, which has often been invoked in transportation modeling researches \cite{wilson1967statistical}  \cite{erlander1990gravity}.
+This issue of incompletely observed data can be tackled by the maximum entropy formalism \cite{jaynes1957information}, which has often been invoked in transportation modeling research \cite{wilson1967statistical}  \cite{erlander1990gravity}.
 
@@
 \cite{dempster1977maximum}  \cite{bavaud2009information}), but steps two and three, aiming at calibrating parameters $g_{st}$, $\alpha_s$ and $\beta_t$, do not follow the maximum likelihood rationale of the ``maximisation step".
-Pseudocode of the algorithm is shown with Algorithm \ref{algo1}.
+Pseudocode of the algorithm is shown in Algorithm \ref{algo1}.
 
@@
 As a matter of fact, $g_{st}$ represent \emph{prior attractions} between nodes in equation (\ref{Sol}), before margin correction given by $\phi_s$ and $\psi_t$ and subsequent algorithm steps.
-An urban planner could chose to increase or decrease some of these values in order to take into account prior knowledge on commuting habits of inhabitants.
+An urban planner could choose to increase or decrease some of these values in order to take into account prior knowledge on commuting habits of inhabitants.
 
@@
 
-By construction, the corrected flow $\widetilde{n}_{st}$ found in (\ref{update_flow}) now respect embarkment and disembarkment constraints. We can compute the new transfer flow on edge with
+By construction, the corrected flow $\widetilde{n}_{st}$ found in (\ref{update_flow}) now respects embarkment and disembarkment constraints. We can compute the new transfer flow on edges with
 \begin{equation}
@@
 
-Let  $i=1,\ldots, l$ enumerate the bus stops in increasing order, that is $F(i)=i+1$.
+Let  $i=1,\ldots, l$ enumerate the stops in increasing order, that is $F(i)=i+1$.
 The initial prior is simply $g_{st}=c\; I(s<t)$ and captures solely the unidirectional nature of trips, where $c=\frac{1}{(l-1)(l-2)}$.
@@
 
-This constancy entails the following Markovian behaviour for flows: let $m_{st}$ be the number of travelers embarking at stop $s$ and still inside the bus at stop $t>s$, and let $\rho_{st}$ the probability that travelers embarking at $s$ will disembark at $t$.
+This constancy entails the following Markovian behaviour for flows: let $m_{st}$ be the number of travelers embarking at stop $s$ and still inside the line at stop $t>s$, and let $\rho_{st}$ the probability that travelers embarking at $s$ will disembark at $t$.
 By (\ref{nosignle}),
@@
 which depends on $t$ only: it appears that the disembarkment probability $\rho_t=\frac{\psi_t}{\psi_t+\Psi_t}$ at $t$ is {\em independent} of the embarkment stop $s$.
-Said otherwise, a traveler embarking at any stop $s$ (and thus necessarily in the bus at $F(s)=s+1$) experiences the {\em same disembarkment probability} at each further stop $t>s$.
+Said otherwise, a traveler embarking at any stop $s$ (and thus necessarily in the line at $F(s)=s+1$) experiences the {\em same disembarkment probability} at each further stop $t>s$.
 
@@
 This setup differs from the second section, which is dedicated to applying the algorithm to the real case of the public transportation network of the city of Lausanne (tl), where embankment and disembarkment flows are measured but the real transportation flow is unknown.
-This second case study shows that the algorithm is applicable on large, real datasets and can gives insights about passengers probable routes in the network.
+This second case study shows that the algorithm is applicable on large, real datasets and can give insights about passengers probable routes in the network.
 
@@
 \end{align} 
-where $z_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} n_{st}$ is the flow on transfer edges $T$. Both errors can be interpreted as a weighted mean of a percentage of error.
+where $z_{ij} = I((i,j) \in T)\sum_{st} \chi_{ij}^{st} n_{st}$ is the flow on transfer edges $T$. Both errors can be interpreted as a weighted mean of error percentages.
 
@@
 We fix a number of \emph{round trips} $p \geq 2$, each of which is constituted of a forward line and a backward line, for a total of $q = 2p$ lines.
-Every line has a starting and ending node, which are isolated nodes, and possesses $p-1$ intermediary nodes which allow transfers to the other round trips, yielding a total of $n = 2p(p + 1)$ nodes in the network.
+Every line has a starting and an ending node, which are isolated nodes, and possesses $p-1$ intermediary nodes which allow transfers to the other round trips, yielding a total of $n = 2p(p + 1)$ nodes in the network.
 Examples of these toy networks can be found in Figure \ref{toy_example_plots}.
@@
 
-First, we exhibit some algorithm iterations on a toy example with $p=2$, where $50$ passengers where drawn uniformly across the $\vert P \vert = 20$ possible $st$-trips.
+First, we exhibit some algorithm iterations on a toy example with $p=2$, where $50$ passengers were drawn uniformly across the $\vert P \vert = 20$ possible $st$-trips.
 Some iterations of the algorithm with $\theta=0.001$, along with MTE and MME errors, are shown in Figure \ref{iteration_plots}.
@@
 	\caption{Real transportation flow (green arrows) obtained by randomly drawing $50$ passengers on the graph toy example with $p=2$ round trips, along with iterations $1, 2, 4, 7, 15$ of the algorithm with $\theta=0.001$.
-MTE and MME errors are computed, and embarkment and disembarkment flows are represented respectively by the red and blue colors on nodes.}
+MTE and MME errors are computed, and embarkment and disembarkment counts are represented respectively by the red and blue colors on nodes.}
 	\label{iteration_plots}
 \end{figure}
-On this small example, we can see that the algorithm quickly find an estimation giving small MME error, but still give an MTE of 0.256.
+On this small example, we can see that the algorithm quickly find an estimation giving a small MME error, but still give a MTE of 0.256.
 This result is due to the fact that only $50$ passengers are drawn, giving a large deviation compared to the optimally found solution which maximizes the entropy.
@@
 	\includegraphics[width=0.98\textwidth]{fig/MTE_passengers.pdf}
-	\caption{Mean Transportation Error (MTE) vs the number of randomly drawn passengers, in toy examples with numbers of round trips $p$ varying from 2 to 8.
+	\caption{Mean transportation error (MTE) vs the number of randomly drawn passengers, in toy examples with numbers of round trips $p$ varying from 2 to 8.
 The hyperparameter is set to $\theta=0.001$ and $10$ different draws are performed for each data point.}
@@
 	\includegraphics[width=0.98\textwidth]{fig/MTE_tours.pdf}
-	\caption{Mean Transportation Error (MTE) vs hyperparameter $\theta$, in toy examples with numbers of round trips $p$ varying from 2 to 8.
+	\caption{Mean transportation error (MTE) vs hyperparameter $\theta$, in toy examples with numbers of round trips $p$ varying from 2 to 8.
 The dot on each curve denotes the minimum.
-The number of drawn passengers is different for every network size and is set to $5 \cdot \vert P \vert$, in order to keep the randomness level constant.
+The number of drawn passengers is different for every network size and is set to $5 \cdot \vert P \vert$, in order to keep an (almost) constant variance.
 $10$ different draws are performed for each data point.}
@@
 \label{data_algo}
-The dataset derived from the public transportation agency in the city of Lausanne (tl), in Switzerland.
+The dataset was constructed from data collected by the public transportation agency in the city of Lausanne (tl), in Switzerland.
 Automatic passenger counters were used to collect data on the number of passengers entering and leaving buses and subways at each stop.
@@
 Some stops may be unique to either the inbound or outbound direction.
-The initial dataset contained 13 million data rows, which were aggregated to the year 2019, and the passenger data was filtered to include only passengers who boarded and disembarked at each stop.
+The initial dataset contained 13 million data rows, which were aggregated for the year 2019, and the passenger data was filtered to include only passengers who embarked and disembarked at each stop.
 Lines with the most errors were removed, resulting in a final dataset of 35 transportation lines and 1216 stops.
-The threshold for invalid lines was set when a difference between total embarkment and disembarked differs more than 15\% of their means.
-For the rest all other lines, the correction found in appendix was applied to ensure consistency conditions.
+The threshold for invalid lines was set when a difference between total embarkment and disembarkment counts differs more than 15\% of their means.
+For the rest all other lines, the correction found in the appendix was applied to ensure consistency conditions.
 
@@
 Shortest-paths between all pairs of stops were computed using the breath-first search algorithm \cite{west2001introduction} implemented in the \emph{igraph}\footnote{\url{https://igraph.org/}} R library \cite{Csardi2006}, thus giving $\bm{\chi}$.
-The matrix $P$, containing valid $st$-trips, was built using the same conditions found in the toy example experiments (section \ref{example_construction}) with an additional one.
+The matrix $P$, containing permitted $st$-trips, was built using the same conditions found in the toy example experiments (section \ref{example_construction}) with an additional one.
 If the estimated pedestrian time between two stops $s$ and $t$ was shorter than the estimated time by taking the public transportation network, this $st$-trip was also removed from $P$.
-The graph structure, consisting in the edge-trip incidence matrix $\bm{\chi}$, the valid $st$-trips set $P$, and the transfer edges set $T$, was constructed beforehand testing the method.
+The graph structure, consisting in the edge-trip incidence matrix $\bm{\chi}$, the permitted $st$-trips set $P$, and the transfer edges set $T$, was constructed beforehand testing the method.
 
 The algorithm was run with hyperparameter $\theta = 0.1$.
-Tracking the Mean Margin Error (MME) showed a rapid decrease ($<0.001$ after 13 iterations) but the (conservative) convergence criterion (chosen as when $\sum_{st} \vert f^\text{prev}_{st} - f_{st} \vert< 10^{-6} $) was reach after 316 iterations, which corresponds to approximately one hour of computing time with a single thread on a AMD Epyc2 7402 with 32GB of RAM.
+Tracking the mean margin error (MME) showed a rapid decrease ($<0.001$ after 13 iterations) but the (conservative) convergence criterion (chosen as when $\sum_{st} \vert f^\text{prev}_{st} - f_{st} \vert< 10^{-6} $) was reach after 316 iterations, which corresponds to approximately one hour of computing time with a single thread on a AMD Epyc2 7402 with 32GB of RAM.
 
@@
 We can see that origin and destination coordinates of each stop are quite similar when displayed on the first two factorial axes.
-This result can be explained if we understand that, in fact, origin and destination profiles are in fact, to some extent, symmetrical: at the beginning of a line, stops have several choices of destinations, but very few origins ``point" at them, which is exactly the reverse for stops at the end of a line.
+This result can be explained if we understand that, in fact, origin and destination profiles are, to some extent, symmetrical: at the beginning of a line, stops have several choices of destinations, but very few origins ``point" at them, which is exactly the reverse for stops at the end of a line.
 Intermediary stops are generally located near the city center, and have a similarly rich origin and destination profiles.
@@
 When taken individually, four of the top 5 transfer counts are predicted between line $1$ and line $72$ occurring at the train station (Lausanne-Gare).
-Line 72 is actually a subway line connecting the north and south regions of the city, and is by far the most used line in the network.
+Line 72 is actually a subway line connecting the northern and southern regions of the city, and is by far the most used line in the network.
 Line $1$ is also a highly frequented line and operated toward the western region of the city.
-It is not surprising to see most transfer occurring between these two lines, as they permit to connect the region with an high density of public transportation network users (the western part of Lausanne), to the main artery of the network (line 72).
+It is not surprising to see most transfers occurring between these two lines, as they permit to connect the region with a high density of public transportation network users (the western part of Lausanne), to the main artery of the network (line 72).
 
@@
 	\includegraphics[width=0.97\textwidth]{fig/transfers.png}
-	\caption{Location and size of transfer hubs, obtained by aggregating estimated transfer flows between lines $z_{ij}$ by ``superstop" names.}
+	\caption{Localization and size of transfer hubs, obtained by aggregating estimated transfer flows between lines $z_{ij}$ in ``superstops".}
 	\label{transfers}
@@
 
-\subsection*{Embarkment and disembarkment corrections for lines}
+\subsection*{Correction of the embarkment and disembarkment counts in a single line}
 \label{flow_correction}
-It may happen that, on some lines $\ell$ with stops indexed in order as $1, \ldots, l$, raw data $\mathbf{a}=(a_i)$ and $\mathbf{b}=(b_i)$ do not obey the following necessary consistency conditions 
+It may happen that, in a line $\ell$ with stops indexed in order as $1, \ldots, l$, raw data $\mathbf{a}=(a_i)$ and $\mathbf{b}=(b_i)$ do not obey the following necessary consistency conditions 
 \begin{align*}
@@
 The last step ensures that $A_{l} = B_{l}$.
-However, as this last step can sometimes lower the embarkment count and increase the disembarkment count on previous nodes, some new nodes can now violate $A_{(i-1)} \geq B_i$.
+However, as this last step can sometimes lower the embarkment count and increase the disembarkment count on previous nodes (if $A_l > B_l$ before correction), some new nodes can now violate $A_{(i-1)} \geq B_i$.
 This is why this procedure must be iterated until all stops on the line verify consistency conditions.

### 2023-02-28 5.1_visu_toy_passengers.R
clean 5.x files

@@ -4,15 +4,12 @@
 # Visualisation output
 #-------------------------------------------------------------------------------
 #-------------------------------------------------------------------------------
-# - Histogram of mean error, according to the number of passengers travelling
-#   into the network (only 1 hyper parameter)
+# - Line plot of the mean error, according to the number of passengers
+#   travelling into the network (fixed theta hyper parameter)
 
 #--------------------------------
 # Head
 #--------------------------------
-
-# Set working directory path
-# setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
 # Libraries
 #--------------------------------
@@ -24,19 +21,16 @@ library(reshape2)
 # Load functions
 source("local_functions.R")
 
-# The folder containing pre-processed data
-data_folder = "multilines_data/preprocessed_data/loop_toy_example"
-
-# Output folder for results
-out_folder = "results/loop_toy_example"
-
 #--------------------------------
 # Network parameters
 #--------------------------------
-# Choose number of lines
+# Choose tours
 nb_lines = 2:8
-# Choose number of stops
+# Nb stop depends of tours
 nb_stops = nb_lines + 1
+# Passengers into the network
+seq_passengers = c(100,500,1000,1500,2000,2500,3000,3500,4000,5000,6000,7500,
+                   10000,12500,15000,20000,30000,50000)
 
 #--------------------------------
 # Algorithm parameters
@@ -63,16 +57,9 @@ hyper_par = 0.001
 
 # Network properties output
 #--------------------------
-# network_prop_list = network_prop(nb_lines, nb_stops)
-# Unlist and save variables
-# list2env(network_prop_list, .GlobalEnv)
 
 # Create a data frame with all different number of passengers into the network
 res_mean = c()
-seq_passengers = c(100,500,1000,1500,2000,2500,3000,3500,4000,5000,6000,7500,
-                   10000,12500,15000,20000,30000,50000)
-
-# rbind(seq_passengers,seq_passengers2)
 
 for (j in 1:length(nb_lines)) {
   
@@ -103,9 +90,21 @@ for (j in 1:length(nb_lines)) {
   res_mean = rbind(res_mean, mean_pass)
 }
 
-write.csv(res_mean, "results/5_toy_ex_outputs/res_mean_passengers.csv", row.names = F)
+# Save results
[…]

### 2023-02-28 5.2_visu_toy_theta.R

@@ -4,15 +4,12 @@
 # Visualisation output
 #-------------------------------------------------------------------------------
 #-------------------------------------------------------------------------------
-# - Histogram of mean error, according to the number of lines into the network
-#   (possible to have more than 1 hyper parameter)
+# - Line plot of the mean error, according to the theta hyper parameter and the
+#   number of passengers depends on number of line tours
 
 #--------------------------------
 # Head
 #--------------------------------
-
-# Set working directory path
-# setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
 # Libraries
 #--------------------------------
@@ -24,18 +21,16 @@ library(reshape2)
 # Load functions
 source("local_functions.R")
 
-
-# The folder containing pre-processed data
-data_folder = "multilines_data/preprocessed_data/loop_toy_example"
-
-# Output folder for results
-out_folder = "results/loop_toy_example"
-
 #--------------------------------
 # Network parameters
 #--------------------------------
-# Choose number of lines
-# below
+# Choose tours
+seq_lines = 2:8
+# Choose prop limit theta
+hyper_par = c(seq(0.001,0.009, 0.001), seq(0.01,0.1, 0.01), seq(0.12,0.2, 0.02),
+              0.25, 0.3, 0.35, seq(0.4,1, 0.1))
+# number of passengers per node (n_passengers = sum(paths)*nb_pass_stop)
+nb_pass_stop = 5
 
 #--------------------------------
 # Algorithm parameters
@@ -53,20 +48,8 @@ max_it = 1000
 display_it = F
 # number of iterations
 n_test = 10
-# prop limit
-# hyper_par = c(0, 0.01, 0.02, 0.03, 0.04, 0.05, 0.06, 0.07, 0.08, 0.09, 0.1,
-#               0.12, 0.14, 0.16, 0.18, 0.2,
-#               0.25, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1)
-hyper_par = c(seq(0.001,0.009, 0.001), seq(0.01,0.1, 0.01), seq(0.12,0.2, 0.02),
-              0.25, 0.3, 0.35, seq(0.4,1, 0.1))
-
-# number of passengers
-# n_passengers = 2000
-nb_pass_stop = 5
-# number of cores
+# multicors
[…]

### 2023-02-28 5.3_visu_tl_CA.R

@@ -1,15 +1,14 @@
 #-------------------------------------------------------------------------------
 #-------------------------------------------------------------------------------
-# AFC
+# Correspondence Analysis (CA)
 # Visualisation output
 #-------------------------------------------------------------------------------
 #-------------------------------------------------------------------------------
-# - Map AFC
+# - Map CA
 
 #--------------------------------
 # Head
 #--------------------------------
-
 
 # Libraries
 #--------------------------------
@@ -22,12 +21,6 @@ library(sf)
 library(RColorBrewer)
 library(leaflet)
 
-# Load functions
-# source("local_functions.R")
-
-# The folder containing pre-processed data
-# data_folder = "multilines_data/preprocessed_data/all_lines"
-
 # The folder containing results
 results_folder = "results/all_lines2"
 
@@ -35,7 +28,7 @@ results_folder = "results/all_lines2"
 # Process
 #--------------------------------
 
-# up or down ?
+# Choose "up" or "down" to set visualisation
 up_down = "up"
 
 # --- Reading files
@@ -61,15 +54,16 @@ rownames(n_mat) = stop_names
 # ped_acc <- apply(ped_time, MARGIN = c(1, 2), FUN = access)
 # ANA = ped_acc %*% n_mat %*% ped_acc
 
+# Compute CA
 res.ca <- CA(n_mat, graph = FALSE)
 
+# Dim 1 and 2 according to up_down
 if (up_down == "up") {
   CA_dim = res.ca$row$coord[, 1:2]
 }
 if (up_down == "down") {
   CA_dim = res.ca$col$coord[, 1:2]
 }
-
 colnames(CA_dim) <- c("dim1","dim2")
 CA_dim = as.data.frame(CA_dim)
 CA_dim$stop_names = rownames(CA_dim)
@@ -124,7 +118,7 @@ CA_shp = right_join(stops_shp, CA_dim, by = "stop_names")
 
 lines_shp_sample = right_join(lines_shp, CA_dim, by = "stop_names")
 
-
+# Radius of the proportional circles
 if (up_down == "up") {
   CA_shp$CA_radius = CA_shp$montees
 }
@@ -134,8 +128,9 @@ if (up_down == "down") {
 }
 
 
-# Mapping results
+# --- Mapping results
 
+# Color palette for 2 dimensions
[…]

### 2023-02-28 5.4_visu_tl_nodes.R

@@ -9,7 +9,6 @@
 # Head
 #--------------------------------
 
-
 # Libraries
 #--------------------------------
 library(dplyr)
@@ -74,10 +73,9 @@ nodes_map = leaflet() %>%
              fillColor = "transparent",
              opacity = 1
   )
-
 nodes_map
 
-# # Flow map
+# Flow map (not used)
 # library(leaflet.minicharts)
 # 
 # leaflet() %>%

### 2023-02-28 5.5_visu_tl_agg_nodes.R

@@ -3,12 +3,11 @@
 # Transfer visualisation
 #-------------------------------------------------------------------------------
 #-------------------------------------------------------------------------------
-# - Map important nodes
+# - Map important nodes, nodes aggregated by stop names
 
 #--------------------------------
 # Head
 #--------------------------------
-
 
 # Libraries
 #--------------------------------

### 2023-03-02 latex_files/applied_network_science_article/article_v1.tex
format article
add declarations

--- 2023-02-28 latex_files/applied_network_science_article/article_v0.tex
@@
 \usepackage{hyperref}
-
-\usepackage{caption}
 
@@
   addressref={aff1},                                       
-  corref={aff1},                       
   email={gguex@unil.ch}
@@
 
+\textbf{\large Declarations}
+
+\section*{Ethical Approval}
+Not applicable.
+
+\section*{Competing interests}
+The authors declare that they have no competing interests.
+
+\section*{Authors' contributions}
+G.G created the algorithm and participated to the case studies and the manuscript redaction.
+R.L.
+participated to the case studies and manuscript redaction, F.B.
+participated to the manuscript redaction.
+
+\section*{Funding}
+No funding was received for this research.
+
+\section*{Availability of data and materials}
+All data, codes, and results can be found on the GitHub repository of the project : \url{https://github.com/sliunil/tl_study}.
+
 \section*{Acknowledgements}%% if any
@@
 We wish to thank the agency for its kindness and hope this research will help them for future developments.
-
-\section*{Competing interests}
-The authors declare that they have no competing interests.
 
